# Data preparation


## Searching for the MAGs
1) Choose the MAGs   
In this case, we will be working with _Parabacteroides distasonis_ and _Phocaeicola vulgatus_. In the EHI database, select the MAGs with > 90% completeness and < 2.5 contamination. Do the same in the GTDB, but also filter for isolation source CONTAINS feces.

## Downloading the MAGs
2) Download genome indices and metadata    
Download the EHI_MAG index for each species (in /data/mags_metadata folder) and the curl file and search metadata tsv from the GTDB and place in each species directory in Mjolnir.

2.5) Extract genome metadata   
- Use the GTDB search tsv to run this script to obtain more metadata. (modify script a bit to make it scalable).
```{bash, eval = FALSE}
 python scripts/extract_gtdb_metadata.py
```


3) Create the master index   
Run the create_index.R script to create an index of all the MAGs and the download paths (needs a list of species as input  and right now it is hardcoded in the script)
```{bash, eval = FALSE}
conda activate mags_r_env
Rscript scripts/create_index.R 
```

4) Run the downloading_mags.smk to download all the genomes
```{bash, eval = FALSE}
snakemake -s snakefiles/downloading_mags.smk \
    --executor slurm \
    --jobs 50 \
    --rerun-incomplete \
    --keep-going
    
    
#testing with the HUMAN_EHI
snakemake -s snakefiles/downloading_and_unzipping.smk \
    --executor slurm \
    --jobs 50 \
    --rerun-incomplete \
    --keep-going
```


5) Unzip all the MAGs and make sure they are in the same folder: run    
(maybe add this to the downloading_mags.smk)
```{bash, eval = FALSE}
bash scripts/unzip_mags.sh parabacteroides_distasonis  
bash scripts/unzip_mags.sh phocaeicola_vulgatus
```

6) Make a screen session for each species.
```{bash, eval = FALSE}
screen -S parabacteroides_distasonis
```


7) Run drakkar annotating_function.smk to re-annotate all the MAGs:
```{bash, eval = FALSE}
drakkar annotating -b /maps/projects/alberdilab/people/pjq449/comparative_metagenomics/data/parabacteroides_distasonis/mags/unzipped -o /maps/projects/alberdilab/people/pjq449/comparative_metagenomics --env_path /projects/alberdilab/data/environments/drakkar --annotation-type function 
```


8) Annotate with prokka and make phylogenetic tree with getphylo
```{bash, eval = FALSE}
snakemake -s snakefiles/prokka_and_pangenome.smk --profile snakefiles/slurm_profile

```

9) Run genome_mapping.sh to map the mag ids to the corresponding names 


## R Studio- Data Analysis
## Load data

### EHI MAGs
```{r load ehi mags first}
ehi_mags <- read_csv("data/ehi_mags.csv")
```
### Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
ehi_mags_p <- ehi_mags %>%
  mutate(phylum=str_remove_all(phylum, "p__")) 

phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate(phylum=str_remove_all(phylum, "p__")) %>%
    right_join(ehi_mags_p, by=join_by(phylum == phylum)) %>%
    dplyr::select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

```


```{r colors}
#source_colors <- c(EHI = "#8BC63F", GTDB = "#2D522D")
source_colors <- c(EHI = "#8BC63F", GTDB = "#A90D00")
```

### Sample metadata
```{r load gtdb search metadata}
gtdb_search_metadata <- read_tsv("data/p_dist_gtdb_metadata.tsv")
```

```{r load gtdb accession metadata}
gtdb_accession_metadata<- read_tsv("data/parabacteroides_distasonis_GTDB_accession_metadata.tsv") 
```


```{r load EHI metadata}
ehi_metadata <- read_tsv("data/mags_metadata/parabacteroides_distasonis_metadata.tsv") %>%
  mutate(GC = as.numeric(str_remove(GC, "%")))
```
### Load genome mapping
```{r load genome mapping}
genome_mapping <- read_tsv("data/genome_mapping.tsv") 

gtdb_accession_metadata_2 <- left_join(gtdb_accession_metadata, genome_mapping, by = c("search_accession" = "genome_id")
  )
```

```{r join the metadatas}
# Prepare EHI metadata with standardized column names
ehi_clean <- ehi_metadata %>%
  select(
    ID,
    species = species,  
    completeness,
    contamination,
    genome_size = size,
    GC,
    N50,
    contigs,
    host_species,
    host_order,
    host_class,
    assembly_type,
    isolation_source= sample_type,
    mag_name, 
    eha_number
  ) %>%
  mutate(source = "EHI")

# Prepare GTDB metadata
# Join search and accession metadata
gtdb_metadata <- full_join(gtdb_accession_metadata_2, gtdb_search_metadata,  by = c("search_accession" = "Accession")
  )



# Prepare GTDB metadata
gtdb_clean <- gtdb_metadata %>%
  select(
    ID = search_accession,
    species = "Organism Name",
    completeness = "CheckM Completeness",
    contamination = "CheckM2 Contamination",
    gtdb_taxonomy = "GTDB Taxonomy",
    isolation_source = "Isolation Source",
    genome_size,
    GC = gc_percentage,
    N50 = n50_contigs,
    contigs = contig_count,
    ncbi_country,
    mimag_high_quality,
    mimag_medium_quality,
    gtdb_representative,
    mag_name = genome_file
  )%>%
  mutate(source = "GTDB")

#Join
genome_metadata <- bind_rows(ehi_clean, gtdb_clean)

```




```{r load master index}
# Read master index
master_index <- read_tsv("data/master_mag_index.tsv") %>%
  filter(species == "parabacteroides_distasonis") %>%
  mutate(
    # Extract the actual genome identifier from out_path
    genome_identifier = case_when(
      # For EHI: extract EHA00531_bin.1 from the filename
      str_detect(out_path, "EHA") ~ str_extract(out_path, "EHA[0-9]+_bin\\.[0-9]+"),
      # For NCBI: use the ID as is (GCA/GCF number)
      TRUE ~ ID
    )
  )
```
```{r mapping the ids to the genome identifiers}
# Check the mapping
print("Master index mapping:")
master_index %>% 
  select(ID, genome_identifier, out_path) %>%
  head(10) %>%
  print()

# Read contig-to-genome mapping
contig_to_genome <- read_tsv("data/p_dist_contig_to_genome.tsv",
                              col_names = c("contig", "genome_filename"))
```


### Genome annotations

```{r load genome annotations with ids}
# Read annotations and add IDs
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz") %>%
  mutate(contig = sub("_[^_]*$", "", gene)) %>%
  left_join(contig_to_genome, by = "contig") %>%
  # Extract accession from genome filename
  mutate(
    accession = case_when(
      # For NCBI genomes: extract GCA/GCF number
      str_detect(genome_filename, "^GC[AF]_") ~ str_extract(genome_filename, "^GC[AF]_[0-9]+\\.[0-9]+"),
      # For EHI genomes: the filename IS the identifier
      TRUE ~ genome_filename
    )
  ) %>%
  # Join with master index using genome_identifier
  left_join(
    master_index %>% select(mag_id = ID, genome_identifier),
    by = c("accession" = "genome_identifier")
  ) %>%
  select(gene, mag_id, genome = genome_filename, contig, accession, everything()) %>%
  filter(!is.na(mag_id))

```

### Distill annotations into GIFTs 

```{r distill_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol= 2, annotcol=c(9,10,11,12), verbosity = T)
```



### Load trees
```{r load trees}
genome_metadata$mag_name <- sub("\\.fa$", "", genome_metadata$mag_name)#remove .fa from the mag names so that they match the tree ids


getphylo_tree <- read_tree("data/combined_alignment.tree")
getphylo_tree$tip.label <- gsub("'", "", getphylo_tree$tip.label)
tip_df <- data.frame(label = getphylo_tree$tip.label) %>%
  left_join(genome_metadata, by = c("label" = "mag_name"))

#replace long mag names with IDs
getphylo_tree$tip.label <- tip_df$ID

```


## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(ehi_mags,
     phylum_colors,
     genome_annotations,
     genome_gifts,
     contig_to_genome,
     gtdb_metadata,
     ehi_metadata,
     master_index,
     genome_metadata,
     source_colors,
     getphylo_tree,
     file = "data/data.Rdata")
```

