# NCBI MAGs exploration

## NCBI MAGs metadata
```{r Gut microbiome keywords}
gut_keywords <- c("ceaca", "caeca", "excrement", "excrements", "rectal", "feces", 
                  "stool", "stools", "gut", "cloaca", "cloacal", "fecal", "intestines",
                  "intestine", "intestinal", "cecum", "ileum", "jejunum", "colon", "rectum", "rumen",
                  "anal", "anus", "manure", "faeces")

# Wrap each keyword in word boundaries: \bkeyword\b --otherwise artisanal passes the filter 
pattern <- paste0("\\b(", paste(gut_keywords, collapse = "|"), ")\\b")


# Filter the data
#gut_samples <- metadata_wide %>%
#  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))
```


Download metadata for each taxa
```{bash, eval = FALSE}
# Trying to download ncbi metadata for mags

conda activate drakkar_env
conda install -c conda-forge ncbi-datasets-cli



## CITROBACTER BRAAKII
datasets summary genome taxon 57706 \
  --as-json-lines > genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' genomes.jsonl > high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
  .accession as $asm
  | .checkm_info.completeness as $comp
  | .checkm_info.contamination as $cont
  | .assembly_stats.total_sequence_length as $genome_size
  | .assembly_stats.gc_percent as $gc
  | .assembly_info.biosample.attributes[]?
  | [$asm, $comp, $cont, $genome_size, $gc, .name, .value]
  | @tsv
' high_quality_genomes.jsonl > citrobacter_biosample_attributes.tsv


```




### Citrobacter braakii
```{r citrobacter braakii, eval = FALSE}
c_braki_metadata <- read_tsv("data/mags_metadata/citrobacter_braakii_biosample_attributes.tsv",
  col_names = c(
    "accession",
    "completeness",
    "contamination",
    "genome_size",
    "gc_percent",
    "name",
    "value"
  )
)

wide <- c_braki_metadata %>%
  pivot_wider(
    id_cols = c(
      accession,
      completeness,
      contamination,
      genome_size,
      gc_percent
    ),
    names_from  = name,
    values_from = value
  )


source_counts <- wide %>%
  filter(!is.na(isolation_source)) %>%
  dplyr::count(isolation_source, sort = TRUE)

source_counts


host_counts <- wide %>%
  filter(!is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

host_counts

wide %>%
  filter(host == "Homo sapiens")

citrobacter_gut <- wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

citrobacter_gut %>%tt()

## HUMAN
summary_counts <- citrobacter_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- citrobacter_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Lactococcus lactis
```{bash, eval = FALSE}
## LACTOCOCCUS LACTIS
datasets summary genome taxon 1358 \
  --as-json-lines > lactococcus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' lactococcus_genomes.jsonl > lactococcus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
  .accession as $asm
  | .checkm_info.completeness as $comp
  | .checkm_info.contamination as $cont
  | .assembly_stats.total_sequence_length as $genome_size
  | .assembly_stats.gc_percent as $gc
  | .assembly_info.biosample.attributes[]?
  | [$asm, $comp, $cont, $genome_size, $gc, .name, .value]
  | @tsv
' lactococcus_high_quality_genomes.jsonl > lactococcus_biosample_attributes.tsv



```


```{r lactococcus, eval = FALSE}
lactococcus_metadata <- read_tsv("data/mags_metadata/lactococcus_lactis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- lactococcus_metadata %>%
  filter(attribute != "assembly") %>%  # assembly is in some attribute column, causing issues
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = list(value = ~ paste(.x, collapse = "; "))
  )


lactococcus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- lactococcus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- lactococcus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Hafnia paralvei
```{bash, eval = FALSE}
## HAFNIA PARALVEI
datasets summary genome taxon 546367 \
  --as-json-lines > hafnia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' hafnia_genomes.jsonl > hafnia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' hafnia_high_quality_genomes.jsonl > hafnia_biosample_attributes.tsv


```


```{r hafnia, eval = FALSE}
hafnia_metadata <- read_tsv("data/ncbi/hafnia_paralvei_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- hafnia_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )

hafnia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- hafnia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- hafnia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Enterococcus faecalis
```{bash, eval = FALSE}
## ENTEROCOCCUS FAECALIS
datasets summary genome taxon 1351 \
  --as-json-lines > enterococcus_faecalis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' enterococcus_faecalis_genomes.jsonl > enterococcus_faecalis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' enterococcus_faecalis_high_quality_genomes.jsonl > enterococcus_faecalis_biosample_attributes.tsv


```


```{r enterococcus_faecalis, eval = FALSE}
enterococcus_faecalis_metadata <- read_tsv("data/ncbi/enterococcus_faecalis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- enterococcus_faecalis_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


enterococcus_faecalis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- enterococcus_faecalis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- enterococcus_faecalis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



### Enterococcus hirae
```{bash, eval = FALSE}
## ENTEROCOCCUS HIRAE
datasets summary genome taxon 1354 \
  --as-json-lines > enterococcus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' enterococcus_genomes.jsonl > enterococcus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' enterococcus_high_quality_genomes.jsonl > enterococcus_biosample_attributes.tsv


```


```{r enterococcus_hirae, eval = FALSE}
enterococcus_metadata <- read_tsv("data/ncbi/enterococcus_hirae_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- enterococcus_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


enterococcus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- enterococcus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- enterococcus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Bacteroides uniformis
```{bash, eval = FALSE}
## BACTEROIDES UNIFORMIS
datasets summary genome taxon 820 \
  --as-json-lines > b_uniformis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' b_uniformis_genomes.jsonl > b_uniformis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' b_uniformis_high_quality_genomes.jsonl > b_uniformis_biosample_attributes.tsv


```


```{r bacteroides_uniformis, eval = FALSE}
b_uniformis_metadata <- read_tsv("data/ncbi/bacteroides_uniformis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- b_uniformis_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


b_uniformis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- b_uniformis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- b_uniformis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



### Phocaeicola vulgatus
```{bash, eval = FALSE}
## PHOCAEICOLA VULGATUS
datasets summary genome taxon 821 \
  --as-json-lines > p_vulgatus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' p_vulgatus_genomes.jsonl > p_vulgatus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' p_vulgatus_high_quality_genomes.jsonl > p_vulgatus_biosample_attributes.tsv


```


```{r phocaeicola_vulgatus, eval = FALSE}
p_vulgatus_metadata <- read_tsv("data/ncbi/p_vulgatus_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- p_vulgatus_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


p_vulgatus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- p_vulgatus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- p_vulgatus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Parabacteroides goldsteinii
```{bash, eval = FALSE}
## PARABACTEROIDES GOLDSTEINII
datasets summary genome taxon 328812 \
  --as-json-lines > p_goldsteinii_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' p_goldsteinii_genomes.jsonl > p_goldsteinii_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' p_goldsteinii_high_quality_genomes.jsonl > p_goldsteinii_biosample_attributes.tsv


```


```{r parabacteroides_goldsteinii, eval = FALSE}
p_goldsteinii_metadata <- read_tsv("data/ncbi/parabacteroides_goldsteinii_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- p_goldsteinii_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


p_goldsteinii_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- p_goldsteinii_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- p_goldsteinii_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```
### Akkermansia municiphila
```{bash, eval = FALSE}
## AKKERMANSIA MUNICIPHILA
datasets summary genome taxon 239935 \
  --as-json-lines > akkermansia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' akkermansia_genomes.jsonl > akkermansia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' akkermansia_high_quality_genomes.jsonl > akkermansia_biosample_attributes.tsv


```


```{r akkermansia, eval = FALSE}
akkermansia_metadata <- read_tsv("data/ncbi/akkermansia_muciniphila_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- akkermansia_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


akkermansia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- akkermansia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- akkermansia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```
### Bacteroides fragilis
```{bash, eval = FALSE}
## BACTEROIDES FRAGILIS
datasets summary genome taxon 817 \
  --as-json-lines > b_fragilis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' b_fragilis_genomes.jsonl > b_fragilis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' b_fragilis_high_quality_genomes.jsonl > b_fragilis_biosample_attributes.tsv


```


```{r bacteroides_fragilis, eval = FALSE}
b_fragilis_metadata <- read_tsv("data/ncbi/bacteroides_fragilis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- b_fragilis_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


b_fragilis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- b_fragilis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- b_fragilis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Providencia rettgeri
```{bash, eval = FALSE}
## PROVIDENCIA RETTGERI
datasets summary genome taxon 587 \
  --as-json-lines > providencia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' providencia_genomes.jsonl > providencia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' providencia_high_quality_genomes.jsonl > providencia_biosample_attributes.tsv


```


```{r providencia rettgeri, eval = FALSE}
providencia_metadata <- read_tsv("data/ncbi/providencia_rettgeri_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- providencia_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


providencia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- providencia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- providencia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



```{r final selection, eval = FALSE}
species <- c(
  "lactococcus_lactis",
  "hafnia_paralvei",
  "enterococcus_faecalis",
  "enterococcus_hirae",
  "bacteroides_uniformis",
  "phocaeicola_vulgatus",
  "parabacteroides_goldsteinii",
  "akkermansia_muciniphila",
  "bacteroides_fragilis",
  "providencia_rettgeri",
  "citrobacter_braakii"
)

for (sp in species) {

  message("Processing ", sp)

  infile <- file.path(
    "data/mags_metadata",
    paste0(sp, "_biosample_attributes.tsv")
  )

  metadata_long <- read_tsv(
    infile,
    col_names = c(
      "accession",
      "completeness",
      "contamination",
      "genome_size",
      "gc_percent",
      "name",
      "value"
    ),
    show_col_types = FALSE
  )

  metadata_wide <- metadata_long %>%
    pivot_wider(
      id_cols = c(
        accession,
        completeness,
        contamination,
        genome_size,
        gc_percent
      ),
      names_from  = name,
      values_from = value,
      values_fn   = ~ dplyr::first(na.omit(.x)),
      names_repair = "unique"
    )

  ## filter gut-associated genomes
  if ("isolation_source" %in% colnames(metadata_wide)) {

    gut_metadata <- metadata_wide %>%
      filter(
        str_detect(
          isolation_source,
          regex(pattern, ignore_case = TRUE)
        )
      )

  } else {
    warning(sp, " has no isolation_source column")
    gut_metadata <- metadata_wide[0, ]
  }

  ## save ONLY the gut-filtered metadata
  saveRDS(
    gut_metadata,
    file.path(
      "data/mags_metadata",
      paste0(sp, "_ncbi_metadata.rds")
    )
  )

  selected_accessions <- gut_metadata %>%
    pull(accession) %>%
    unique()

  selected_accessions <- selected_accessions[!is.na(selected_accessions)]

  writeLines(
    selected_accessions,
    file.path(
      "data/mags_metadata",
      paste0(sp, "_ncbi_selected_accessions.txt")
    )
  )
}

```
## Downloading the MAGs after filtering

Download the MAGs when you finish selecting them
```{bash, eval = FALSE}
### DOWNLOAD THE MAGS

datasets download genome accession --inputfile ../bacteroides_unifmormis_ncbi_selected_accessions.txt --include genome --filename hafnia_paralvei_selected_genomes.zip

```


```{r checking if mags in gtdb are same as ncbi, eval = FALSE}
mags_gtdb <- read_tsv("data/mags_metadata/master_mag_index_lactococcus_lactis.tsv") %>% pull(ID)

intersect(selected, mags_gtdb)


```

```{r in ncbi but not gtdb, eval = FALSE}
# in selected but NOT in mags_gtdb
setdiff(selected, mags_gtdb)

```

```{r in gtdb not in ncbi, eval = FALSE}

# in mags_gtdb but NOT in selected
setdiff(mags_gtdb, selected)
```

