# NCBI MAGs exploration

## NCBI MAGs metadata


```{r Gut microbiome keywords}
gut_keywords <- c("ceaca", "caeca", "excrement", "excrements", "rectal", "feces", 
                  "stool", "stools", "gut", "cloaca", "cloacal", "fecal", "intestines",
                  "intestine", "intestinal", "cecum", "ileum", "jejunum", "colon", "rectum", "rumen",
                  "anal", "anus", "manure", "faeces")

# Wrap each keyword in word boundaries: \bkeyword\b --otherwise artisanal passes the filter 
pattern <- paste0("\\b(", paste(gut_keywords, collapse = "|"), ")\\b")


# Filter the data
#gut_samples <- metadata_wide %>%
#  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))
```


Download metadata for each taxa
```{bash, eval = FALSE}
# Trying to download ncbi metadata for mags

conda activate drakkar_env
conda install -c conda-forge ncbi-datasets-cli



## CITROBACTER BRAAKII
datasets summary genome taxon 57706 \
  --as-json-lines > genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' genomes.jsonl > high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' high_quality_genomes.jsonl > biosample_attributes.tsv


```

### Citrobacter braakii
```{r citrobacter braakii}
c_braki_metadata <- read_tsv("data/ncbi/c_braakii_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- c_braki_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


source_counts <- metadata_wide %>%
  filter(!is.na(isolation_source)) %>%
  dplyr::count(isolation_source, sort = TRUE)

source_counts


host_counts <- metadata_wide %>%
  filter(!is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

host_counts

metadata_wide %>%
  filter(host == "Homo sapiens")

citrobacter_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

citrobacter_gut %>%tt()

## HUMAN
summary_counts <- citrobacter_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- citrobacter_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Lactococcus lactis
```{bash, eval = FALSE}
## LACTOCOCCUS LACTIS
datasets summary genome taxon 1358 \
  --as-json-lines > lactococcus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' lactococcus_genomes.jsonl > lactococcus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' lactococcus_high_quality_genomes.jsonl > lactococcus_biosample_attributes.tsv


```


```{r lactococcus}
lactococcus_metadata <- read_tsv("data/ncbi/lactococcus_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- lactococcus_metadata %>%
  filter(attribute != "assembly") %>%  # assembly is in some attribute column, causing issues
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = list(value = ~ paste(.x, collapse = "; "))
  )


lactococcus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- lactococcus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- lactococcus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Hafnia paralvei
```{bash, eval = FALSE}
## HAFNIA PARALVEI
datasets summary genome taxon 546367 \
  --as-json-lines > hafnia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' hafnia_genomes.jsonl > hafnia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' hafnia_high_quality_genomes.jsonl > hafnia_biosample_attributes.tsv


```


```{r hafnia}
hafnia_metadata <- read_tsv("data/ncbi/hafnia_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- hafnia_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )

hafnia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- hafnia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- hafnia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Enterococcus faecalis
```{bash, eval = FALSE}
## ENTEROCOCCUS FAECALIS
datasets summary genome taxon 1351 \
  --as-json-lines > enterococcus_faecalis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' enterococcus_faecalis_genomes.jsonl > enterococcus_faecalis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' enterococcus_faecalis_high_quality_genomes.jsonl > enterococcus_faecalis_biosample_attributes.tsv


```


```{r enterococcus_faecalis}
enterococcus_faecalis_metadata <- read_tsv("data/ncbi/enterococcus_faecalis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- enterococcus_faecalis_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


enterococcus_faecalis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- enterococcus_faecalis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- enterococcus_faecalis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



### Enterococcus hirae
```{bash, eval = FALSE}
## ENTEROCOCCUS HIRAE
datasets summary genome taxon 1354 \
  --as-json-lines > enterococcus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' enterococcus_genomes.jsonl > enterococcus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' enterococcus_high_quality_genomes.jsonl > enterococcus_biosample_attributes.tsv


```


```{r enterococcus_hirae}
enterococcus_metadata <- read_tsv("data/ncbi/enterococcus_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- enterococcus_metadata %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


enterococcus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- enterococcus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- enterococcus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Bacteroides uniformis
```{bash, eval = FALSE}
## BACTEROIDES UNIFORMIS
datasets summary genome taxon 820 \
  --as-json-lines > b_uniformis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' b_uniformis_genomes.jsonl > b_uniformis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' b_uniformis_high_quality_genomes.jsonl > b_uniformis_biosample_attributes.tsv


```


```{r bacteroides_uniformis}
b_uniformis_metadata <- read_tsv("data/ncbi/b_uniformis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- b_uniformis_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


b_uniformis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- b_uniformis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- b_uniformis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



### Phocaeicola vulgatus
```{bash, eval = FALSE}
## PHOCAEICOLA VULGATUS
datasets summary genome taxon 821 \
  --as-json-lines > p_vulgatus_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' p_vulgatus_genomes.jsonl > p_vulgatus_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' p_vulgatus_high_quality_genomes.jsonl > p_vulgatus_biosample_attributes.tsv


```


```{r phocaeicola_vulgatus}
p_vulgatus_metadata <- read_tsv("data/ncbi/p_vulgatus_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- p_vulgatus_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


p_vulgatus_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- p_vulgatus_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- p_vulgatus_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Parabacteroides goldsteinii
```{bash, eval = FALSE}
## PARABACTEROIDES GOLDSTEINII
datasets summary genome taxon 328812 \
  --as-json-lines > p_goldsteinii_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' p_goldsteinii_genomes.jsonl > p_goldsteinii_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' p_goldsteinii_high_quality_genomes.jsonl > p_goldsteinii_biosample_attributes.tsv


```


```{r parabacteroides_goldsteinii}
p_goldsteinii_metadata <- read_tsv("data/ncbi/p_goldsteinii_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- p_goldsteinii_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


p_goldsteinii_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- p_goldsteinii_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- p_goldsteinii_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```
### Akkermansia municiphila
```{bash, eval = FALSE}
## AKKERMANSIA MUNICIPHILA
datasets summary genome taxon 239935 \
  --as-json-lines > akkermansia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' akkermansia_genomes.jsonl > akkermansia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' akkermansia_high_quality_genomes.jsonl > akkermansia_biosample_attributes.tsv


```


```{r akkermansia}
akkermansia_metadata <- read_tsv("data/ncbi/akkermansia_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- akkermansia_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


akkermansia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- akkermansia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- akkermansia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```
### Bacteroides fragilis
```{bash, eval = FALSE}
## BACTEROIDES FRAGILIS
datasets summary genome taxon 817 \
  --as-json-lines > b_fragilis_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' b_fragilis_genomes.jsonl > b_fragilis_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' b_fragilis_high_quality_genomes.jsonl > b_fragilis_biosample_attributes.tsv


```


```{r bacteroides_fragilis}
b_fragilis_metadata <- read_tsv("data/ncbi/b_fragilis_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- b_fragilis_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


b_fragilis_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- b_fragilis_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- b_fragilis_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```


### Providencia rettgeri
```{bash, eval = FALSE}
## PROVIDENCIA RETTGERI
datasets summary genome taxon 587 \
  --as-json-lines > providencia_genomes.jsonl


#filtering by contamination and completeness
jq -c '
select(
  .checkm_info.completeness >= 90 and
  .checkm_info.contamination <= 2.5
)
' providencia_genomes.jsonl > providencia_high_quality_genomes.jsonl



#put metadata in tsv
jq -r '
.accession as $asm |
.assembly_info.biosample.attributes[]? |
[$asm, .name, .value] | @tsv
' providencia_high_quality_genomes.jsonl > providencia_biosample_attributes.tsv


```


```{r providencia rettgeri}
providencia_metadata <- read_tsv("data/ncbi/providencia_biosample_attributes.tsv",  col_names = c("assembly", "attribute", "value"))


metadata_wide <- providencia_metadata %>%
  filter(attribute != "assembly") %>%
  pivot_wider(
    id_cols = assembly,
    names_from = attribute,
    values_from = value,
    values_fn = function(x) x[1]
  )


providencia_gut <- metadata_wide %>%
  filter(str_detect(isolation_source, regex(pattern, ignore_case = TRUE)))

## HUMAN
summary_counts <- providencia_gut %>%
  dplyr::summarize(
    total_rows = n(),
    human_samples = sum(
      host == "Homo sapiens" | 
      stringr::str_detect(isolation_source, stringr::regex("human", ignore_case = TRUE)), 
      na.rm = TRUE
    )
  )

print(summary_counts)

## OTHER HOSTS
other_hosts <- providencia_gut %>%
  dplyr::filter(host != "Homo sapiens" | is.na(host)) %>%
  dplyr::count(host, sort = TRUE)

print(other_hosts)
```



```{r final selection, eval = FALSE}
selected <- hafnia_gut %>%
  pull(assembly) %>%
  unique()

#check
length(selected)
head(selected)

#save the accessions to download the mags in mjolnir
writeLines(selected, "data/mags_metadata/hafnia_paralvei_ncbi_selected_accessions.txt")

```
## Downloading the MAGs after filtering

Download the MAGs when you finish selecting them
```{bash, eval = FALSE}
### DOWNLOAD THE MAGS

datasets download genome accession --inputfile ../hafnia_paralvei_ncbi_selected_accessions.txt --include genome --filename hafnia_paralvei_selected_genomes.zip

```


```{r checking if mags in gtdb are same as ncbi}
mags_gtdb <- read_tsv("data/mags_metadata/master_mag_index_lactococcus_lactis.tsv") %>% pull(ID)

intersect(selected, mags_gtdb)


```

```{r in ncbi but not gtdb}
# in selected but NOT in mags_gtdb
setdiff(selected, mags_gtdb)

```

```{r in gtdb not in ncbi}

# in mags_gtdb but NOT in selected
setdiff(mags_gtdb, selected)
```

