
# Functions and helpers

## Colour schemes
```{r color schemes}
source_colors <- c(EHI = "#8BC63F", GTDB = "#2D522D", NCBI= "#20558A")

host_type_colors <- c(human = "#D81B60" , animal = "#004D40") 

host_type_2_colors <- c(human = "#db2b39" , domestic = "#29335c", wild= "#f3a712") 

host_type_3_colors <- c(human_adj = "#D81B60" , wild = "#004D40") 


host_order_colors <- c(
  # ---- Mammalia ----
  "Rodentia"        = "#FDA991",  
  "Primates"        = "#CC0000",  
  "Lagomorpha"      = "#C87F46",  
  "Proboscidea"     = "#8E5B3A",  
  "Artiodactyla"    = "#A86A3E",  
  "Carnivora"       = "#633C2C",  
  "Perissodactyla"  = "#7A4B39", 
  "Chiroptera"      = "#1B2021", 
  "Dasyuromorphia"  = "#A34F3E",  
  "Diprotodontia"   = "#B56952", 
  
  # --- Amphibian ----
  "Urodela"         = "#1C5551", 

  # ---- Aves (birds) ----
  
  "Accipitriformes" = "#6A1B9A",   
  "Galliformes"     = "#D9A6FF",   
  "Sphenisciformes" = "#8E24AA",   
  "Columbiformes"   = "#9C27B0",   
  "Strigiformes"    = "#512DA8",  
  "Charadriiformes" = "#AB47BC",   
  "Passeriformes"   = "#BA68C8",   
  "Ciconiiformes"   = "#4A148C",   
  "Anseriformes"    = "#6F2DA8",   
  "Gruiformes"      = "#9575CD",   
  "Cathartiformes"  = "#5E35B1",   


  # ---- Reptilia ----
  "Squamata"        = "#1D921B",  
  "Testudines"      = "#0B6E32",  

  # ---- Fish (Actinopterygii) ----
  "Salmoniformes"   = "#00BCD4",  
  "Cichliformes"    = "#0097A7", 

  # ---- Unclassified / NA ----
  "NA"              = "#BFBFBF"   
)


host_class_colors <- c(
  "Mammalia"  = "#A90D00",  
  "Reptilia"  = "#1D921B", 
  "Aves"      = "#6A1B9A",
  "Actinopterygii"      = "#00BCD4",
  "Amphibia"  = "#1C5551"
)


country_palette <- c(
  #  Southern Europe (oranges) 
  "Spain"      = "#E66101",
  "Italy"      = "#F17C0E",
  "Greece"     = "#FC8D59",
  "Portugal"   = "#FDB863",
  "Malta"      = "#FECC8F",
  "Turkey"     = "#E34A33",
  "Bosnia And Herzegovina" = "#F46D43",
  "Slovenia"   = "#FEE08B",
  "Croatia"    = "#FDAE61",  
  "Serbia"     = "#FEC44F",  

  #  Northern & Central Europe (blues)
  "Denmark"      = "#1F78B4",
  "Norway"       = "#3182BD",
  "Sweden"       = "#6BAED6",
  "Finland"      = "#9ECAE1",
  "Germany"      = "#08519C",
  "Netherlands"  = "#2171B5",
  "Belgium"      = "#4292C6",
  "Ireland"      = "#6BAED6",
  "United Kingdom" = "#9ECAE1",
  "Estonia"      = "#08306B",
  "Poland"       = "#4A90E2",
  "Slovakia"     = "#5DA5DA",
  "Czechia"      = "#7FA9DF",
  "Austria"      = "#4F81BD",
  "Hungary"      = "#6C9AC3",
  "Bulgaria"     = "#3B74A5",

  #  East Asia (reds) 
  "China"       = "#B2182B",
  "Japan"       = "#D6604D",
  "South Korea" = "#F4A582",
  "Taiwan"      = "#67000D",
  "Hong Kong"   = "#A50F15",
  "Cambodia"    = "#FB6A4A",
  "Malaysia"    = "#EF3B2C",

  #  South Asia (magenta)
  "India"       = "#D13C8C",

  # Middle East (browns)
  "Israel"        = "#8C510A",
  "Saudi Arabia"  = "#BF812D",
  "Turkey"        = "#E08214",

  #Africa (yellows ) 
  "Ghana"           = "#FEE391",
  "Nigeria"         = "#FEC44F",
  "Cameroon"        = "#FEE0B6",
  "Tunisia"         = "#FDD49E",
  "Tanzania"        = "#F6E8C3",
  "Equatorial Guinea" = "#F2C969",
  "South Africa"    = "#F4D03F",

  # North America (purples)
  "United States" = "#756BB1",
  "USA"           = "#756BB1",   
  "Canada"        = "#9E9AC8",
  "Mexico"        = "#BC80BD",

  # South America (greens) 
  "Brazil"      = "#1B9E77",
  "Colombia"    = "#66C2A5",
  "Chile"       = "#41AE76",

  # Oceania 
  "Australia"    = "#FFD92F",
  "New Zealand"  = "#E5C100",

  #  Polar 
  "Greenland"   = "#A6CEE3",
  "Antarctica"  = "#BFD3E6",

  # Missing / meta categories (greys)-
  "None"           = "grey70",
  "Not Applicable" = "grey75",
  "Not Provided"   = "grey80",
  "Not Collected"  = "grey85",
  "Missing"        = "grey60",
  "Na"             = "grey65",
  "NA"               = "grey50"
)


continent_palette <-  c(
  "Europe"      = "#08306B",
  "Asia"       = "#B2182B",
  "Africa"         = "orange",
  "North America"        = "#BC80BD",
  "South America"      = "#1B9E77",
  "Oceania"    = "#FFD92F",
  "Greenland"   = "#A6CEE3",
  "Antarctica"  = "#BFD3E6",
  # Missing 
  "None"           = "grey70",
  "Not Applicable" = "grey75",
  "Not Provided"   = "grey80",
  "Not Collected"  = "grey85",
  "Missing"        = "grey60",
  "Na"             = "grey65",
  "NA"               = "grey50"
)

```

## DATA PREPARATION
### Load all the files

We have the following files per species:
1) metadata from the GTDB, NCBI and EHI MAGs (some species don't have GTDB MAGs or NCBI MAGs)
2) Ndb.csv files - ANI matrix from drep 
3) gene_annotations.tsv.xz - the gene-level annotations - includes contig names but not genome (MAGs) names
4) contig_to_mag.tsv -  maps contig names from gene_annotations to MAG names


```{r load_species_files}
## Load all the files 
load_species_files <- function(species, base_dir = "data/mags_metadata") {

  paths <- list(
    ehi_metadata       = file.path(base_dir, paste0(species, "_metadata.tsv")),
    gtdb_final         = file.path(base_dir, paste0(species, "_gtdb_final_metadata.tsv")),
    contig_map         = file.path(base_dir, paste0(species, "_contig_to_mag.tsv")),
    gene_annotations   = file.path(base_dir, paste0(species, "_gene_annotations.tsv.xz")),
    ncbi_metadata_rds  = file.path(base_dir, paste0(species, "_ncbi_metadata.rds")),
    ndb                = file.path(base_dir, paste0(species, "_Ndb.csv"))
  )

  # helper that warns when file is missing
  load_if_exists <- function(path, loader) {
    if (file.exists(path)) {
      loader(path)
    } else {
      warning(sprintf("Missing file: %s", path))
      NULL
    }
  }

  result <- list(
    ehi_metadata     = load_if_exists(paths$ehi_metadata,       \(p) readr::read_tsv(p, show_col_types = FALSE)),
    gtdb_final       = load_if_exists(paths$gtdb_final,         \(p) readr::read_tsv(p, show_col_types = FALSE)),
    contig_map       = load_if_exists(paths$contig_map,         \(p) readr::read_tsv(p, show_col_types = FALSE)),
    gene_annotations = load_if_exists(paths$gene_annotations,   \(p) readr::read_tsv(p, show_col_types = FALSE)),
    ndb              = load_if_exists(paths$ndb,                \(p) readr::read_csv(p, show_col_types = FALSE)),
    ncbi_metadata    = load_if_exists(paths$ncbi_metadata_rds,  readRDS)
  )

  return(result)
}

```

Normalize columns between the different metadata files.
```{r helpers to standardize columns}
# Find the genome size column and rename to genome_size
detect_and_rename_genome_size <- function(df) {
  if (is.null(df)) return(df)

  # If exact already present, nothing to do
  if ("genome_size" %in% names(df)) return(df)

  # Common explicit variants 
  candidates <- c(
    # EHI 
    "size", "Size", "SIZE",
    # NCBI 
    "genome_size...4", "genome_size...43",
    # other common labels
    "size_bp", "Size_bp", "genome_size_bp",
    "total_length", "assembly_length", "estimated_size"
  )

  hit <- intersect(candidates, names(df))
  if (length(hit) >= 1) {
    # If multiple hits appear, prefer the first listed above
    df <- dplyr::rename(df, genome_size = !!hit[1])
    return(df)
  }

  # Fallback: first column containing "genome_size" or "size" or "length"
  fuzzy <- grep("(genome_)?size|length", names(df), ignore.case = TRUE, value = TRUE)
  if (length(fuzzy) >= 1) {
    df <- dplyr::rename(df, genome_size = !!fuzzy[1])
    return(df)
  }

  df
}

# Normalize 'n50' (lowercase in some NCBI exports) to 'N50'
detect_and_rename_n50 <- function(df) {
  if (is.null(df)) return(df)
  if (!("N50" %in% names(df)) && ("n50" %in% names(df))) {
    df <- dplyr::rename(df, N50 = n50)
  }
  df
}

# Enforce consistent types across sources before bind_rows()
harmonize_schema <- function(df) {
  if (is.null(df)) return(NULL)


  # Column name standardization

  rename_map <- c(
    "isolation_source.x" = "isolation_source",
    "isolation_source.y" = "isolation_source",
    "isolation_source_x" = "isolation_source",
    "isolation_source_y" = "isolation_source",

    "geo_loc_name"       = "country",
    "geographic location (latitude)"  = "latitude",
    "geographic location (longitude)" = "longitude"
  )

  # Rename if present
  for (old in names(rename_map)) {
    if (old %in% names(df)) {
      names(df)[names(df) == old] <- rename_map[[old]]
    }
  }


  #  Define expected column names

  char_cols <- c(
    "ID","source","species","gtdb_taxonomy","gtdb_representative",
    "isolation_source","host","host_status","disease","diagnosis",
    "host_species","host_order","host_class","host_summary",
    "country","locality","collection_date","ncbi_biosample",
    "mag_name","eha_number","common_name","sample_name","external_id","submitter_id",
    "env_broad","env_medium","env_local","env_local_context"
  )

  num_cols <- c(
    "completeness","contamination","genome_size","GC","N50","contigs",
    "latitude","longitude"
  )

  logi_cols <- c("mimag_high_quality","mimag_medium_quality")

  #  Add missing columns

  for (cc in char_cols) if (!cc %in% names(df)) df[[cc]] <- NA_character_
  for (nc in num_cols)  if (!nc %in% names(df)) df[[nc]] <- NA_real_
  for (lc in logi_cols) if (!lc %in% names(df)) df[[lc]] <- NA


  #  Type-coerce consistently

  df <- df %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(char_cols),
        ~ as.character(.x)
      ),
      dplyr::across(
        dplyr::all_of(num_cols),
        ~ suppressWarnings(as.numeric(.x))
      ),
      dplyr::across(
        dplyr::all_of(logi_cols),
        ~ as.logical(.x)
      )
    )


  #  Return cleaned df

  df
}



#change all missing/not provided cells to NAs
clean_biosample_value <- function(v) {
  if (is.na(v)) return(NA_character_)
  v <- str_trim(v)
  m <- str_match(v, '^Attribute\\s+"(.*)"$')
  if (!is.na(m[1,2])) v <- str_trim(m[1,2])
  v_low <- tolower(v)
  if (v_low %in% c("missing","not provided","not applicable","na","n/a","none","unknown",
                   "unspecified","not collected","not determined","not available")) return(NA_character_)
  if (v == "") return(NA_character_)
  v
}

```

Add genome column to gene annotations.
```{r prepare gene annotations}
# ---- Prepare gene annotations: add the genome with contig_map ----
prep_gene_annotations_with_genome <- function(ann_raw, contig_map_raw) {
  if (is.null(ann_raw)) return(NULL)
  if (is.null(contig_map_raw)) {
    stop("contig_map is required to attach 'genome' to gene_annotations.")
  }

  ## ---- Standardize contig_map ----
  cm <- contig_map_raw %>%
    dplyr::rename(
      contig = contig_id,
      genome = mag
    ) %>%
    dplyr::mutate(
      contig = as.character(contig),
      genome = as.character(genome)
    )

  ## ---- Derive contig from gene (remove final _<gene_number>) ----
  if (!("gene" %in% names(ann_raw))) {
    stop("gene_annotations must contain column `gene`.")
  }

  an <- ann_raw %>%
    dplyr::mutate(
      contig = sub("_[0-9]+$", "", gene)
    )

  ## ---- Join contig → genome ----
  an_joined <- an %>%
    dplyr::left_join(cm, by = "contig")

  ## ---- Hard failure if something goes wrong ----
  if (all(is.na(an_joined$genome))) {
    stop(
      "No contigs matched contig_map.\n",
      "Derived contig example: ", an_joined$contig[1],
      "\nExpected contig_map example: ", cm$contig[1]
    )
  }

  an_joined
}

```


### Prepare EHI metadata
```{r prepare EHI metadata}
# Prepare EHI metadata into a standardized schema
prep_ehi_metadata <- function(ehi_raw) {
  if (is.null(ehi_raw)) return(NULL)

  ehi_raw <- detect_and_rename_genome_size(ehi_raw)
  ehi_raw <- detect_and_rename_n50(ehi_raw)

  ehi_clean <- ehi_raw %>%
    mutate(
      GC = suppressWarnings(as.numeric(str_remove(GC, "%")))
    ) %>%
    dplyr::select(
      any_of(c(
        "ID","species","completeness","contamination",
        "genome_size","GC","N50","contigs",
        "host_species","host_order","host_class",
        "sample_type","isolation_source","country","locality",
        "mag_name","eha_number"
      ))
    ) %>%
    mutate(
      isolation_source = coalesce(isolation_source, sample_type),
      source = "EHI"
    ) %>%
    dplyr::select(-sample_type)

  harmonize_schema(ehi_clean)
}
```


### Prepare GTDB metadata
```{r prepare GTDB metadata}
prep_gtdb_final <- function(gtdb_raw) {
  if (is.null(gtdb_raw)) return(NULL)

  # Standardize size & N50 labels
  gtdb_raw <- detect_and_rename_genome_size(gtdb_raw)
  gtdb_raw <- detect_and_rename_n50(gtdb_raw)

  gtdb_clean <- gtdb_raw %>%
    # Pick only columns we may need (robust to presence/absence)
    dplyr::select(dplyr::any_of(c(
      "ID","accession","gtdb_taxonomy","gtdb_representative",
      "checkm2_completeness","checkm2_contamination",
      "genome_size","gc_percentage","N50","n50_contigs",
      "contigs","contig_count",
      "ncbi_biosample","country","geo_loc_name",
      "isolation_source","isolation_source.x","isolation_source_x",
      "isolation_source.y","isolation_source_y",
      "host","host_status","disease","diagnosis","host_summary",
      # latitude/longitude may appear with ENA-style names; keep both
      "latitude","longitude","geographic location (latitude)","geographic location (longitude)",
      # environment / context variants
      "env_broad","env_broad_scale","env_medium","env_local","env_local_scale",
      "env_local_context","local environmental context",
      # names with spaces/case variants from ENA
      "common_name","common name","sample_name","external_id","External Id","submitter_id","Submitter Id",
      "collection_date"
    ))) %>%
    # Normalize N50/contigs if only alt names exist
    { if (!"N50" %in% names(.) && "n50_contigs" %in% names(.)) dplyr::rename(., N50 = n50_contigs) else . } %>%
    { if (!"contigs" %in% names(.) && "contig_count" %in% names(.)) dplyr::rename(., contigs = contig_count) else . } %>%
    # Conditional renames for ENA/GTDB variants (no-op if missing)
    { if ("env_broad_scale" %in% names(.)) dplyr::rename(., env_broad = `env_broad_scale`) else . } %>%
    { if ("env_local_scale" %in% names(.)) dplyr::rename(., env_local = `env_local_scale`) else . } %>%
    { if ("local environmental context" %in% names(.)) dplyr::rename(., env_local_context = `local environmental context`) else . } %>%
    { if ("common name" %in% names(.)) dplyr::rename(., common_name = `common name`) else . } %>%
    { if ("External Id" %in% names(.)) dplyr::rename(., external_id = `External Id`) else . } %>%
    { if ("Submitter Id" %in% names(.)) dplyr::rename(., submitter_id = `Submitter Id`) else . } %>%
    # Add core source + numeric fields
    dplyr::mutate(
      source        = "GTDB",
      completeness  = .data$checkm2_completeness,
      contamination = .data$checkm2_contamination,
      GC            = .data$gc_percentage
    )

  #  Coalesce isolation_source variants safely (handle dot/underscore suffixes)
  iso_candidates <- c("isolation_source","isolation_source.x","isolation_source_x",
                      "isolation_source.y","isolation_source_y")
  present_iso <- intersect(iso_candidates, names(gtdb_clean))
  if (length(present_iso) > 0) {
    # progressive coalesce across present variants
    iso <- gtdb_clean[[present_iso[1]]]
    if (length(present_iso) > 1) {
      for (nm in present_iso[-1]) iso <- dplyr::coalesce(iso, gtdb_clean[[nm]])
    }
    gtdb_clean[["isolation_source"]] <- iso
  }

  # Prefer GTDB 'country', else ENA 'geo_loc_name'
  if ("geo_loc_name" %in% names(gtdb_clean)) {
    gtdb_clean[["country"]] <- dplyr::coalesce(gtdb_clean[["country"]], gtdb_clean[["geo_loc_name"]])
  }

  #  Drop helper/duplicate columns we normalized away
  gtdb_clean <- gtdb_clean %>%
    dplyr::select(-dplyr::any_of(c(
      "checkm2_completeness","checkm2_contamination","gc_percentage",
      "n50_contigs","contig_count","geo_loc_name",
      "isolation_source.x","isolation_source_x","isolation_source.y","isolation_source_y"
    )))

  # Final schema/type harmonization (handles lat/long ENA names, adds missing cols)
  gtdb_clean <- harmonize_schema(gtdb_clean)

  return(gtdb_clean)
}
```


### Prepare NCBI metadata
```{r prepare ncbi metadata}
# Prepare NCBI metadata into standardized schema
prep_ncbi_metadata <- function(ncbi_raw) {
  if (is.null(ncbi_raw)) return(NULL)

  ncbi_raw <- detect_and_rename_genome_size(ncbi_raw)
  ncbi_raw <- detect_and_rename_n50(ncbi_raw)

  ncbi_clean <- ncbi_raw %>%
    dplyr::select(
      dplyr::any_of(c(
        "accession","completeness","contamination",
        "genome_size","gc_percent","isolation_source","host",
        "collection_date","geo_loc_name","N50","contigs","host_disease","host_age"
      ))
    ) %>%
    dplyr::rename(
      ID = accession,
      country = geo_loc_name,
      GC = gc_percent
    ) %>%
    dplyr::mutate(
      source = "NCBI"
    )

  harmonize_schema(ncbi_clean)
}
```



### Combine metadata sources into genome_metadata
```{r merge metadata}
# Combine EHI + GTDB(final) + NCBI into one unified table
combine_metadata_sources <- function(ehi_clean,
                                     gtdb_final_clean,
                                     ncbi_clean) {
  tables <- list(ehi_clean, gtdb_final_clean, ncbi_clean)
  tables <- tables[!vapply(tables, is.null, logical(1))]

  if (length(tables) == 0) {
    # No sources available: return an empty tibble 
    genome_metadata <- harmonize_schema(tibble::tibble())
  } else {
    # Ensure schema per table, then bind
    tables <- lapply(tables, harmonize_schema)
    genome_metadata <- dplyr::bind_rows(tables)
    # One more pass to guarantee presence of all expected columns
    genome_metadata <- harmonize_schema(genome_metadata)
  }

  # Safe numeric coercions in columns that exist
  genome_metadata <- genome_metadata %>%
  dplyr::mutate(
    completeness  = if ("completeness"   %in% names(.)) suppressWarnings(as.numeric(completeness))  else NA_real_,
    contamination = if ("contamination"  %in% names(.)) suppressWarnings(as.numeric(contamination)) else NA_real_,
    GC            = if ("GC"             %in% names(.)) suppressWarnings(as.numeric(GC))            else NA_real_,
    genome_size   = if ("genome_size"    %in% names(.)) suppressWarnings(as.numeric(genome_size))   else NA_real_,
    N50           = if ("N50"            %in% names(.)) suppressWarnings(as.numeric(N50))           else NA_real_,
    contigs       = if ("contigs"        %in% names(.)) suppressWarnings(as.numeric(contigs))       else NA_real_
  )

  # Reorder columns if present
  front_cols <- c(
    "ID","source","species","gtdb_taxonomy",
    "host_species","host_order","host_class",
    "isolation_source","host","host_summary",
    "country","locality",
    "completeness","contamination","genome_size","GC","N50","contigs",
    "collection_date","ncbi_biosample","mag_name","eha_number",
    "gtdb_representative","mimag_high_quality","mimag_medium_quality",
    "common_name","sample_name","external_id","submitter_id",
    "env_broad","env_medium","env_local","env_local_context"
  )
  front_cols <- intersect(front_cols, colnames(genome_metadata))
  genome_metadata <- genome_metadata %>%
    dplyr::relocate(dplyr::all_of(front_cols), .before = 1)

  if (nrow(genome_metadata) > 0 && all(is.na(genome_metadata$genome_size))) {
    warning("All genome_size values are NA after merging. Check size mapping.")
  }

  genome_metadata
}
```


### Infer host metadata 
```{r infer host species}
##Try to take species name and standardize to latin binomial if possible
normalize_species <- function(s) {
  s <- as.character(s)
  s[is.na(s) | s == ""] <- NA_character_

  s_low <- stringr::str_to_lower(s)

  # Canonicalize known synonyms/typos
  out <- dplyr::case_when(
    # Dog
    stringr::str_detect(s_low, "^canis\\s+familiaris$") ~ "Canis lupus familiaris",
    stringr::str_detect(s_low, "^canis\\s+lupus\\s+familiaris$") ~ "Canis lupus familiaris",
    stringr::str_detect(s_low, "^mammuthus\\s+primi") ~ "Mammuthus primigenius",
    TRUE ~ s
  )

  # Title-case genus, lower species where we have at least two tokens
  reformat <- function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    parts <- strsplit(x, "\\s+")[[1]]
    if (length(parts) < 2) return(x)
    paste0(stringr::str_to_title(parts[1]), " ", stringr::str_to_lower(parts[2]))
  }
  # Vectorize reformat over 'out'
  out[] <- vapply(out, reformat, character(1))
  out
}

#########################################################################################

# Accept only true-looking Latin binomials and reject common non-species nouns
is_latin_binomial_strict <- function(x) {
  x <- as.character(x)
  ok_shape <- grepl("^[A-Z][a-z]+\\s+[a-z]{3,}$", x)  # Genus species
  if (!any(ok_shape, na.rm = TRUE)) return(rep(FALSE, length(x)))

  bad_terms <- c(
    "human","patient","infant","child","adult",
    "feces","faeces","stool","urine","blood","saliva","swab","swabs",
    "skin","gut","intestinal","rumen","content","contents",
    "sample","metagenome","environment","environmental","tissue",
    "rectal","oral","nasal","throat","vaginal","anal", "dairy", "cabbage", "kimchi", "broiler", "subcultured"
  )

  reject <- rep(FALSE, length(x))
  ix <- which(ok_shape & !is.na(x))
  if (length(ix) > 0) {
    parts <- strsplit(x[ix], "\\s+")
    for (i in seq_along(ix)) {
      w <- tolower(parts[[i]])
      reject[ix[i]] <- any(w %in% bad_terms)
    }
  }
  ok_shape & !reject
}


#########################################################################################
## Actual inferring of host metadata function
## Searches for common names and tries to assign host species, class and order
## (made after checking which host species we have in total)

infer_host_metadata <- function(genome_metadata) {

  if (is.null(genome_metadata)) return(NULL)

  # Utility
  is_blank <- function(x) is.na(x) | x == ""
  
  tax_map <- tibble::tibble(
  host_species = c(
    # Humans / lab / livestock / pets
    "Homo sapiens",
    "Mus musculus",
    "Rattus norvegicus",
    "Sus scrofa",
    "Bos taurus",
    "Canis lupus familiaris",
    "Equus caballus",
    "Oryctolagus cuniculus",
    "Capra hircus",

    # Birds
    "Gallus gallus",
    "Cathartes aura",
    "Columba livia",
    "Spheniscus magellanicus",

    # Insects
    "Drosophila melanogaster",
    "Bombyx mori",

    # Fish
    "Danio rerio",
    "Oncorhynchus mykiss",
    "Oreochromis niloticus",

    # Marine mammals / pinnipeds
    "Balaenoptera acutorostrata",
    "Arctocephalus australis",

    # Extinct
    "Mammuthus primigenius",
    
    #More
    "Loxodonta africana",
    "Elephas maximus",
    "Felis catus",
    "Canis lupus",
    "Phascolarctos cinereus",
    
    "Grasshopper",
    "Termite",
    "Macaca fascicularis",
    "Rhinolophus capensis",
    "Fish",
    
    #Final (from total_genome_metadata)
    "Zophobas morio",
    "Mustela putorius",
    "Ovis aries",
    "Corallus caninus",
    "Meles meles",
    "Clanga pomarina",
    "Buteo buteo",
    "Accipiter gentilis",
    "Chelonia mydas",
    "Grampus griseus",
    "Sterna trudeaui",
    "Rhodnius prolixus",
    "Turdus merula",
    "Curruca curruca",
    "Columba palumbus",
    "Delichon urbicum",
    "Chloris chloris",
    "Hirundo rustica",
    "Ciconia ciconia",
    "Anas platyrhynchos",
    "Grus grus",
    "Apis mellifera",
    "Coragyps atratus",
    "Pseudoscops clamator",
    "Pandion haliaetus",
    "Sycanus bifidus",
    "Eocanthecona furcellata",
    "Arma custos",
    "Corvus frugilegus",
    "Milvus milvus",
    "Vulpes zerda"


  ),
  tax_host_order = c(
    # Humans / lab / livestock / pets
    "Primates",
    "Rodentia",
    "Rodentia",
    "Artiodactyla",
    "Artiodactyla",
    "Carnivora",
    "Perissodactyla",
    "Lagomorpha",
    "Artiodactyla",

    # Birds
    "Galliformes",
    "Accipitriformes",
    "Columbiformes",
    "Sphenisciformes",

    # Insects
    "Diptera",
    "Lepidoptera",

    # Fish
    "Cypriniformes",
    "Salmoniformes",
    "Cichliformes",

    # Marine mammals / pinnipeds
    "Artiodactyla",      # (Cetartiodactyla)
    "Carnivora",

    # Extinct
    "Proboscidea",
    
    #More
    "Proboscidea",
    "Proboscidea",
    "Carnivora",
    "Carnivora",
    "Diprotodontia",
    
    
    "Orthoptera",     # Grasshopper
    "Blattodea",      # Termite (termites now placed within Blattodea; Isoptera as infraorder)
    "Primates",       # Macaca fascicularis
    "Chiroptera",     # Rhinolophus capensis (horseshoe bat)
    NA_character_,    # Fish (no single order fits all fish)
    
    #Final (from total_genome_metadata)
    "Coleoptera",          # Zophobas morio
    "Carnivora",           # Mustela putorius
    "Artiodactyla",        # Ovis aries
    "Squamata",            # Corallus caninus
    "Carnivora",           # Meles meles
    "Accipitriformes",     # Clanga pomarina
    "Accipitriformes",     # Buteo buteo
    "Accipitriformes",     # Accipiter gentilis
    "Testudines",          # Chelonia mydas
    "Artiodactyla",        # Grampus griseus
    "Charadriiformes",     # Sterna trudeaui
    "Hemiptera",           # Rhodnius prolixus
    "Passeriformes",       # Turdus merula
    "Passeriformes",       # Curruca curruca
    "Columbiformes",       # Columba palumbus
    "Passeriformes",       # Delichon urbicum
    "Passeriformes",       # Chloris chloris
    "Passeriformes",       # Hirundo rustica
    "Ciconiiformes",       # Ciconia ciconia
    "Anseriformes",        # Anas platyrhynchos
    "Gruiformes",          # Grus grus
    "Hymenoptera",         # Apis mellifera
    "Cathartiformes",      # Coragyps atratus
    "Strigiformes",        # Pseudoscops clamator
    "Accipitriformes",     # Pandion haliaetus
    "Hemiptera",           # Sycanus bifidus
    "Hemiptera",           # Eocanthecona furcellata
    "Hemiptera",           # Arma custos
    "Passeriformes",       # Corvus frugilegus
    "Accipitriformes",     # Milvus milvus
    "Carnivora"            # Vulpes zerda


  ),
  tax_host_class = c(
    # Humans / lab / livestock / pets
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",

    # Birds
    "Aves",
    "Aves",
    "Aves",
    "Aves",

    # Insects
    "Insecta",
    "Insecta",

    # Fish
    "Actinopterygii",
    "Actinopterygii",
    "Actinopterygii",

    # Marine mammals / pinnipeds
    "Mammalia",
    "Mammalia",

    # Extinct and more
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    
        
    "Insecta",        # Grasshopper
    "Insecta",        # Termite
    "Mammalia",       # Macaca fascicularis
    "Mammalia",       # Rhinolophus capensis
    "Actinopterygii",  # Fish (bony fishes; safe default for generic "fish")

    #Final (from total_genome_metadata)
    "Insecta",        # Zophobas morio
    "Mammalia",       # Mustela putorius
    "Mammalia",       # Ovis aries
    "Reptilia",       # Corallus caninus
    "Mammalia",       # Meles meles
    "Aves",           # Clanga pomarina
    "Aves",           # Buteo buteo
    "Aves",           # Accipiter gentilis
    "Reptilia",       # Chelonia mydas
    "Mammalia",       # Grampus griseus
    "Aves",           # Sterna trudeaui
    "Insecta",        # Rhodnius prolixus
    "Aves",           # Turdus merula
    "Aves",           # Curruca curruca
    "Aves",           # Columba palumbus
    "Aves",           # Delichon urbicum
    "Aves",           # Chloris chloris
    "Aves",           # Hirundo rustica
    "Aves",           # Ciconia ciconia
    "Aves",           # Anas platyrhynchos
    "Aves",           # Grus grus
    "Insecta",        # Apis mellifera
    "Aves",           # Coragyps atratus
    "Aves",           # Pseudoscops clamator
    "Aves",           # Pandion haliaetus
    "Insecta",        # Sycanus bifidus
    "Insecta",        # Eocanthecona furcellata
    "Insecta",        # Arma custos
    "Aves",           # Corvus frugilegus
    "Aves",           # Milvus milvus
    "Mammalia"        # Vulpes zerda

  )
)
  

  # --- 1) Unified host text for pattern matching ---
  # join all columns where the host species could be mentioned for easier search
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      host_text = stringr::str_to_lower(
        stringr::str_squish(
          paste(
            dplyr::coalesce(.data$host_species, ""),
            dplyr::coalesce(.data$host, ""),
            dplyr::coalesce(.data$isolation_source, ""),
            dplyr::coalesce(.data$env_broad, ""),
            dplyr::coalesce(.data$env_local_context, ""),
            sep = " ; "
          )
        )
      )
    )

  # --- 2) Common-name / phrase heuristics (vectorized) ---
  guess_from_common <- function(txt) {
    dplyr::case_when(
      # Exact species mentions / synonyms
      stringr::str_detect(txt, "\\barctocephalus australis\\b|\\bfur seal\\b|south american fur seal") ~ "Arctocephalus australis",
      stringr::str_detect(txt, "\\bbalaenoptera acutorostrata\\b|\\bminke whale\\b") ~ "Balaenoptera acutorostrata",
      stringr::str_detect(txt, "\\bspheniscus magellanicus\\b|\\bmagellanic penguin\\b|\\bpenguin\\b") ~ "Spheniscus magellanicus",

      
 #Elephants
    stringr::str_detect(txt, "\\basian elephant\\b") ~ "Elephas maximus",
    stringr::str_detect(txt, "\\bafrican elephant\\b") ~ "Loxodonta africana",
    stringr::str_detect(txt, "\\belephants?\\b") ~ "Loxodonta africana",  # default choice
    #cats
    stringr::str_detect(txt, "\\b(cat|cats|kitten|kittens|feline|house\\s*cat)\\b") ~ "Felis catus",
    stringr::str_detect(txt, "\\bfelis\\s+catus\\b") ~ "Felis catus",

    # dogs 
    stringr::str_detect(txt, "\\b(dog|dogs|puppy|puppies|canine|pet\\s*dog)\\b") ~ "Canis lupus familiaris",
    stringr::str_detect(txt, "\\bcanis\\s+lupus\\s+familiaris\\b|\\bcanis\\s+familiaris\\b") ~ "Canis lupus familiaris",

    # gray wolf (wild Canis lupus) ---
    stringr::str_detect(txt, "\\bcanis\\s+lupus\\b|\\bgray\\s*wolf\\b|\\bgrey\\s*wolf\\b|\\bwolf\\b") ~ "Canis lupus",
 
 #grasshopper
   stringr::str_detect(txt, "\\bgrasshopper\\b|\\blocust\\b|\\blocusts\\b") ~ "Grasshopper",

 
      stringr::str_detect(txt, "\\brabbit\\b|\\bbunny\\b|\\bbunnies\\b|\\blagomorph\\b") ~ "Oryctolagus cuniculus",
      stringr::str_detect(txt, "\\bpigeon\\b|\\bdove\\b") ~ "Columba livia",
      stringr::str_detect(txt, "\\bequine\\b|\\bhorse\\b|\\bfoal\\b") ~ "Equus caballus",
      stringr::str_detect(txt, "\\bcattle\\b|\\bbovine\\b|\\bcow\\b|\\bcows\\b") ~ "Bos taurus",
      stringr::str_detect(txt, "\\bgoat\\b|\\bcaprine\\b|\\bkid\\b") ~ "Capra hircus",
      stringr::str_detect(txt, "\\bpiglet\\b|\\bpiglets\\b|\\bpig\\b|\\bswine\\b|\\bporcine\\b") ~ "Sus scrofa",
      stringr::str_detect(txt, "\\bmammoth\\b|malolyakovskiy\\s+mammoth") ~ "Mammuthus primigenius",

      # Existing helpful ones
      stringr::str_detect(txt, "homo sapiens|\\bhuman\\b|patient|adult|child|infant") ~ "Homo sapiens",
      stringr::str_detect(txt, "mus musculus|\\bmouse\\b|\\bmice\\b|murine") ~ "Mus musculus",
      stringr::str_detect(txt, "\\brattus|\\brat\\b") ~ "Rattus norvegicus",
      stringr::str_detect(txt, "sus scrofa|porcine|\\bpig\\b|swine") ~ "Sus scrofa",
      stringr::str_detect(txt, "bos taurus|bovine|\\bcow\\b|cattle") ~ "Bos taurus",
      stringr::str_detect(txt, "canis lupus familiaris|canis familiaris|\\bdog\\b|canine") ~ "Canis lupus familiaris",
     stringr::str_detect(txt, "gallus gallus|\\bchicken\\b|\\bpoultry\\b|\\bbroiler\\b") ~ "Gallus gallus",
      stringr::str_detect(txt, "bombyx mori|silkworm") ~ "Bombyx mori",
      stringr::str_detect(txt, "drosophila|fruit fly") ~ "Drosophila melanogaster",
      stringr::str_detect(txt, "danio rerio|zebrafish") ~ "Danio rerio",
      stringr::str_detect(txt, "oncorhynchus mykiss") ~ "Oncorhynchus mykiss",
      stringr::str_detect(txt, "oreochromis niloticus") ~ "Oreochromis niloticus",
 
     stringr::str_detect(txt, "\\btermite(s)?\\b") ~ "Termite",
    
        # Macaca fascicularis (Latin or common names) ---
        stringr::str_detect(txt, "\\bmacaca\\s+fascicularis\\b|\\bcynomolgus\\b|\\bcrab[-]eating\\s+macaque\\b|\\bmacaque\\b") ~ "Macaca fascicularis",
    
        # Rhinolophus capensis (Latin or common name) ---
        stringr::str_detect(txt, "\\brhinolophus\\s+capensis\\b|\\brhinolopus\\s+capensis\\b|\\bcape\\s+horseshoe\\s+bat\\b|\\bhorseshoe\\s+bat\\b") ~ "Rhinolophus capensis",
    
        # Generic fish catch-all  ---
        stringr::str_detect(txt, "\\bfish\\b|\\bfishes\\b") ~ "Fish",
 
        # Zophobas morio
        stringr::str_detect(txt, "zophobas|superworm|darkling beetle") ~ "Zophobas morio",
        
        # Mustela putorius
        stringr::str_detect(txt, "mustela putorius|polecat") ~ "Mustela putorius",
        
        # Ovis aries
        stringr::str_detect(txt, "ovis aries|\\bsheep\\b") ~ "Ovis aries",
        
        # Corallus caninus
        stringr::str_detect(txt, "corallus caninus|emerald tree boa") ~ "Corallus caninus",
        
        # Chelonia mydas
        stringr::str_detect(txt, "chelonia mydas|green sea turtle") ~ "Chelonia mydas",
        
        # Grampus griseus
        stringr::str_detect(txt, "grampus griseus|risso") ~ "Grampus griseus",
        
        # Rhodnius prolixus
        stringr::str_detect(txt, "rhodnius prolixus|kissing bug") ~ "Rhodnius prolixus",
        
        # Apis mellifera
        stringr::str_detect(txt, "apis mellifera|honey[- ]?bee") ~ "Apis mellifera",
        
        # Sycanus bifidus
        stringr::str_detect(txt, "sycanus bifidus") ~ "Sycanus bifidus",
        
        # Eocanthecona furcellata
        stringr::str_detect(txt, "eocanthecona") ~ "Eocanthecona furcellata",
        
        # Arma custos
        stringr::str_detect(txt, "arma custos") ~ "Arma custos",
        
        # Bird species
        stringr::str_detect(txt, "clanga pomarina|spotted eagle") ~ "Clanga pomarina",
        stringr::str_detect(txt, "buteo buteo|buzzard") ~ "Buteo buteo",
        stringr::str_detect(txt, "accipiter gentilis|goshawk") ~ "Accipiter gentilis",
        stringr::str_detect(txt, "trudeaui") ~ "Sterna trudeaui",
        stringr::str_detect(txt, "turdus merula|blackbird") ~ "Turdus merula",
        stringr::str_detect(txt, "curruca curruca|whitethroat") ~ "Curruca curruca",
        stringr::str_detect(txt, "columba palumbus|wood pigeon") ~ "Columba palumbus",
        stringr::str_detect(txt, "delichon urbicum|house martin") ~ "Delichon urbicum",
        stringr::str_detect(txt, "chloris chloris|greenfinch") ~ "Chloris chloris",
        stringr::str_detect(txt, "hirundo rustica|barn swallow") ~ "Hirundo rustica",
        stringr::str_detect(txt, "ciconia ciconia|white stork") ~ "Ciconia ciconia",
        stringr::str_detect(txt, "anas platyrhynchos|mallard") ~ "Anas platyrhynchos",
        stringr::str_detect(txt, "grus grus|crane") ~ "Grus grus",
        stringr::str_detect(txt, "coragyps atratus|black vulture") ~ "Coragyps atratus",
        stringr::str_detect(txt, "clamator") ~ "Pseudoscops clamator",
        stringr::str_detect(txt, "corvus frugilegus|rook") ~ "Corvus frugilegus",
        stringr::str_detect(txt, "milvus milvus|red kite") ~ "Milvus milvus",
        stringr::str_detect(txt, "pandion haliaetus|osprey") ~ "Pandion haliaetus",

      TRUE ~ NA_character_
    )
  }

  genome_metadata <- genome_metadata %>%
    dplyr::mutate(.common_guess = guess_from_common(.data$host_text))

  # --- 3) Extract explicit Latin binomial (strict) as a later fallback ---
  extract_latin_binomial_strict <- function(x) {
    x <- x[!is.na(x)]
    if (length(x) == 0) return(NA_character_)
    txt <- paste(x, collapse = " ; ")
    txt <- stringr::str_replace_all(txt, "[^A-Za-z ]+", " ")
    txt <- stringr::str_squish(txt)
    matches <- unlist(stringr::str_extract_all(
      txt, "\\b[A-Z][a-z]+\\s+[a-z]{3,}\\b"
    ))
    if (length(matches) == 0) return(NA_character_)
    valid <- matches[is_latin_binomial_strict(matches)]
    if (length(valid) == 0) return(NA_character_)
    valid[1]
  }

  genome_metadata <- genome_metadata %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      .binom_guess = extract_latin_binomial_strict(c(
        as.character(.data$host_species),
        as.character(.data$host),
        as.character(.data$isolation_source),
        as.character(.data$env_broad),
        as.character(.data$env_local_context)
      ))
    ) %>%
    dplyr::ungroup()

  # --- 4) Decide host_species (trust valid binomial, then common-name, then binomial guess) ---
  
  # bad terms -> terms that are not species latin binomials
  
  bad_terms <- "(feces|faeces|stool|urine|blood|saliva|swab|swabs|skin|gut|intestinal|rumen|content|contents|sample|metagenome|environment|environmental|patient|infant|adult|child|rectal|oral|nasal|throat|vaginal|anal)"

  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      host_species = dplyr::if_else(
        is_latin_binomial_strict(.data$host_species) &
          !stringr::str_detect(stringr::str_to_lower(.data$host_species), bad_terms),
        normalize_species(.data$host_species),
        NA_character_
      ),
      host_species = dplyr::case_when(
        !is_blank(.data$host_species) ~ .data$host_species,
        !is_blank(.data$.common_guess) ~ normalize_species(.data$.common_guess),  # PRIORITIZE common-name mapping
        !is_blank(.data$.binom_guess)  ~ normalize_species(.data$.binom_guess),  # then extracted binomial
        TRUE ~ .data$host_species
      )
    )

  # --- 5) Join to fill order/class from dictionary ---
  genome_metadata <- genome_metadata %>%
    dplyr::left_join(tax_map, by = "host_species") %>%
    dplyr::mutate(
      host_order = dplyr::coalesce(.data$host_order, .data$tax_host_order),
      host_class = dplyr::coalesce(.data$host_class, .data$tax_host_class)
    ) %>%
    dplyr::select(-dplyr::any_of(c("tax_host_order","tax_host_class")))

  # --- 6) host_type (animal or human) ---
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      host_type = dplyr::case_when(
        .data$host_species == "Homo sapiens" ~ "human",
        !is_blank(.data$host_species) ~ "animal",
        TRUE ~ NA_character_
      )
    )
  
  
  # --- 7) host_type_2 (human / domestic / wild) ---

  domestic_species <- c(
  # Pets
  "Canis lupus familiaris",  # dog
  "Felis catus",             # cat
  
  # Livestock
  "Bos taurus",              # cattle
  "Sus scrofa",              # pig
  "Ovis aries",              # sheep
  "Capra hircus",            # goat
  "Equus caballus",          # horse
  
  # Poultry
  "Gallus gallus",           # chicken
  "Anas platyrhynchos",      # duck
  
  # Lab animals
  "Mus musculus",            # mouse
  "Rattus norvegicus",       # rat
  "Oryctolagus cuniculus"    # rabbit
)
  
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      host_type_2 = dplyr::case_when(
        host_species == "Homo sapiens" ~ "human",
        host_species %in% domestic_species ~ "domestic",
        !is.na(host_species) ~ "wild",
        TRUE ~ NA_character_
      )
    )
  
   #  (putting humans and domesticated animals in one group vs wild animals)---
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      host_type_3 = dplyr::case_when(
        host_species == "Homo sapiens" ~ "human_adj",
        host_species %in% domestic_species ~ "human_adj",
        !is.na(host_species) ~ "wild",
        TRUE ~ NA_character_
      )
    )
  

  # Cleanup helpers
  genome_metadata <- genome_metadata %>%
    dplyr::select(-dplyr::any_of(c(".common_guess",".binom_guess","host_text")))

  return(genome_metadata)
}
```


### Normalize country and infer continent
```{r normalize_country}
# Normalize country names and assign continents
normalize_country_fields <- function(genome_metadata) {
  if (is.null(genome_metadata)) return(NULL)


  # 1) Simplify country strings like "Country: region"
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      country_simple = stringr::str_trim(
        stringr::str_split_fixed(dplyr::coalesce(.data$country, ""), ":", 2)[, 1]
      ),
      country_simple = dplyr::na_if(.data$country_simple, "")) 


  # 2) Normalization map for common ENA/NCBI variants (patterns in lower-case)
  country_map <- c(
    "usa" = "United States", "u\\.s\\." = "United States",
    "u\\.s\\.a\\." = "United States", "united states of america" = "United States",
    "uk" = "United Kingdom", "u\\.k\\." = "United Kingdom",
    "great britain" = "United Kingdom", "england" = "United Kingdom",
    "scotland" = "United Kingdom", "wales" = "United Kingdom",
    "northern ireland" = "United Kingdom",
    "korea,? republic of" = "South Korea", "republic of korea" = "South Korea",
    "south korea" = "South Korea", "korea,? \\(south\\)" = "South Korea",
    "pr china" = "China", "p\\.r\\. china" = "China",
    "russian federation" = "Russia", "czech republic" = "Czechia",
    "swaziland" = "Eswatini", "cote d['’]ivoire" = "Côte d’Ivoire",
    "democratic republic of the congo" = "DR Congo",
    "viet nam" = "Vietnam", "myanmar \\(burma\\)" = "Myanmar",
    "bolivia \\(plurinational state of\\)" = "Bolivia",
    "iran,? islamic republic of" = "Iran",
    "syrian arab republic" = "Syria",
    "moldova,? republic of" = "Moldova",
    "lao people'?s democratic republic" = "Laos",
    "macedonia,? the former yugoslav republic of" = "North Macedonia",
    "palestine,? state of" = "Palestine",
    "hong kong" = "Hong Kong",
    "macau|macao" = "Macau"
  )

  normalize_country <- function(x) {
    # Always return character
    y <- stringr::str_trim(tolower(dplyr::coalesce(as.character(x), "")))
    y[y == ""] <- NA_character_

    if (length(y)) {
      for (pat in names(country_map)) {
        repl <- country_map[[pat]]
        # Use ^...$ to only replace whole-string matches after lowercasing
        y <- ifelse(!is.na(y),
                    stringr::str_replace_all(y, paste0("^", pat, "$"), repl),
                    y)
      }
      #  tidy-up
      y <- ifelse(!is.na(y), stringr::str_replace_all(y, "\\s+", " "), y)
      y <- ifelse(y == "people's republic of china", "China", y)

      # Title-case anything still not mapped
      y <- ifelse(!is.na(y), stringr::str_to_title(y), y)
    }
    return(as.character(y))
  }

  # 3) Produce normalized country + continent (force character vector into countrycode)
  genome_metadata <- genome_metadata %>%
    dplyr::mutate(
      country_normalized = normalize_country(.data$country_simple),
      continent = countrycode::countrycode(
        sourcevar    = as.character(.data$country_normalized),
        origin       = "country.name",
        destination  = "continent",
        warn         = FALSE
      )
    )

  return(genome_metadata)
}
```


### Filter species that don't belong and remove duplicate MAGs
```{r filtering_deduplication}

#function to clean accessions
clean_accession <- function(x) {
  x <- basename(x)
  x <- sub("\\.(fna|fa|fasta)(\\.gz)?$", "", x, ignore.case = TRUE)
  x <- sub("^(RS_|GB_)", "", x, perl = TRUE)
  x
}


# Remove non-species MAGs and deduplicate genomes by quality,
#  merge GCA/GCF duplicates (keep GCF; merge metadata).

filter_species_MAGs <- function(genome_metadata,
                                genome_annotations,
                                mags_to_remove,
                                ndb,
                                match_on_version = TRUE) {
  if (is.null(genome_metadata) || is.null(genome_annotations)) {
    stop("Both genome_metadata and genome_annotations must be provided.")
  }

  # ---------------------------
  # Helpers
  # ---------------------------

  # Remove extensions and directories
  clean_label <- function(x) {
    x <- basename(x)
    sub("\\.(fna|fa|fasta)(\\.gz)?$", "", x, ignore.case = TRUE)
  }
  
  # clean genome IDs so RS_/GB_ prefixes are removed
  clean_accession <- function(x) {
  x <- basename(x)
  x <- sub("\\.(fna|fa|fasta)(\\.gz)?$", "", x, ignore.case = TRUE)
  x <- sub("^(RS_|GB_)", "", x, perl = TRUE)
  x
}

  # Extract (GCA|GCF) and numeric core (e.g., 015667075.1) from an ID
  # Handles optional GB_/RS_ prefixes; returns NA when not matching.
  extract_accession_fields <- function(id_vec, match_on_version = TRUE) {
    id_vec <- clean_label(id_vec)
    core_re <- if (match_on_version) "(\\d+(?:\\.\\d+)?)" else "(\\d+)"
    pat <- paste0("^(?:GB_|RS_)?(GC[AF])_", core_re, "$")

    matched <- grepl(pat, id_vec, perl = TRUE)
    acc_type <- ifelse(matched, sub(pat, "\\1", id_vec, perl = TRUE), NA_character_)
    acc_core <- ifelse(matched, sub(pat, "\\2", id_vec, perl = TRUE), NA_character_)

    data.frame(acc_type = acc_type, acc_core = acc_core, stringsAsFactors = FALSE)
  }

  # First non-NA helper (after we normalize empty "" to NA for characters)
  first_non_na <- function(x) {
    idx <- which(!is.na(x))
    if (length(idx)) x[idx[1]] else NA
  }

  # Normalize empty strings to NA for character columns
  na_if_empty_chars <- function(df) {
    df %>%
      mutate(across(where(is.character), ~ na_if(., "")))
  }

  # ---------------------------
  # Standardize removal list
  # ---------------------------
  mags_to_remove_clean <- clean_label(mags_to_remove)

  # ---------------------------
  # 1) Preprocess genome_metadata
  # ---------------------------
  has_comp    <- "completeness"  %in% names(genome_metadata)
  has_cont    <- "contamination" %in% names(genome_metadata)
  has_contigs <- "contigs"       %in% names(genome_metadata)

  gm0 <- genome_metadata %>%
    mutate(ID_clean = clean_accession(ID)) %>%
    # Make factors safe to merge
    mutate(across(where(is.factor), as.character)) %>%
    na_if_empty_chars()

  # Remove wrong-species MAGs
  gm0 <- gm0 %>%
    filter(!ID_clean %in% mags_to_remove_clean)

  # Add accession parse fields
  acc <- extract_accession_fields(gm0$ID_clean, match_on_version = match_on_version)
  gm1 <- gm0 %>%
    mutate(acc_type = acc$acc_type,
           acc_core = acc$acc_core)

  # Split into accessioned (GCA/GCF) and other IDs
  gm_acc  <- gm1 %>% filter(!is.na(acc_core))
  gm_other <- gm1 %>% filter(is.na(acc_core))

  # For ordering by quality
  gm_acc <- gm_acc %>%
    mutate(
      completeness  = if (has_comp)    suppressWarnings(as.numeric(completeness))  else NA_real_,
      contamination = if (has_cont)    suppressWarnings(as.numeric(contamination)) else NA_real_,
      contigs       = if (has_contigs) suppressWarnings(as.numeric(contigs))       else NA_real_
    )

  # Collapse GCA/GCF duplicates by acc_core:
  # - Prefer GCF over GCA if present
  # - Within type, prefer higher completeness, lower contamination, fewer contigs
  # - Per column, take first non-NA after ordering (merges metadata)
  gm_acc_collapsed <- gm_acc %>%
    group_by(acc_core) %>%
    group_modify(function(df, key) {
    
      n <- nrow(df)
    
      # Prefer GCF if present
      prefer_gcf <- any(df$acc_type == "GCF")
      pref_rank <- if (prefer_gcf) ifelse(df$acc_type == "GCF", 0L, 1L) else rep(0L, n)
    
      # Force numeric vectors of correct length
      comp <- if ("completeness" %in% names(df)) as.numeric(df$completeness) else rep(NA_real_, n)
      cont <- if ("contamination" %in% names(df)) as.numeric(df$contamination) else rep(NA_real_, n)
      ctgs <- if ("contigs" %in% names(df)) as.numeric(df$contigs) else rep(NA_real_, n)
    
      ord <- order(
        pref_rank,
        -ifelse(is.na(comp), -Inf, comp),
         ifelse(is.na(cont),  Inf, cont),
         ifelse(is.na(ctgs),  Inf, ctgs),
        na.last = TRUE
      )
    
      dfo <- df[ord, , drop = FALSE]
    
      dfo %>%
        summarise(across(everything(), first_non_na), .groups = "drop")
    }) %>%
    ungroup()

  # For non-accessioned rows: fall back to your quality-based dedup by ID_clean
  if (nrow(gm_other) > 0) {
    gm_other <- gm_other %>%
      mutate(
        completeness  = if (has_comp)    suppressWarnings(as.numeric(completeness))  else NA_real_,
        contamination = if (has_cont)    suppressWarnings(as.numeric(contamination)) else NA_real_,
        contigs       = if (has_contigs) suppressWarnings(as.numeric(contigs))       else NA_real_
      ) %>%
      arrange(
        ID_clean,
        desc(if (has_comp) completeness else 0),
        (if (has_cont) contamination else 0),
        (if (has_contigs) contigs else Inf)
      ) %>%
      distinct(ID_clean, .keep_all = TRUE)
  }

  # Combine collapsed accessioned rows and deduped others
  genome_metadata_dedup <- bind_rows(gm_acc_collapsed, gm_other) %>%
    # Drop temp fields
    dplyr::select(-acc_type, -acc_core)

  # ---------------------------
  # 2) Filter genome_annotations to match remaining genomes
  # ---------------------------

  ga0 <- genome_annotations %>%
    mutate(genome_clean = clean_accession(genome))

  # Remove wrong-species MAGs
  genome_annotations_filtered <- ga0 %>%
    filter(!genome_clean %in% mags_to_remove_clean)

  # Keep only genomes still present in metadata (after GCF/GCA merge)
  valid_genomes <- genome_metadata_dedup$ID_clean

  genome_annotations_filtered <- genome_annotations_filtered %>%
    filter(genome_clean %in% valid_genomes) %>%
    dplyr::select(-genome_clean)

  # ---------------------------
  # 3) Return cleaned objects
  # ---------------------------
  list(
    metadata    = genome_metadata_dedup,
    annotations = genome_annotations_filtered,
    ndb         = ndb
  )
}
```


### Process one species
```{r process_one_species}
process_one_species <- function(species, mags_to_remove = character(0)) {
  files <- load_species_files(species)

  ehi  <- prep_ehi_metadata(files$ehi_metadata)
  gtdb <- prep_gtdb_final(files$gtdb_final)
  ncbi <- prep_ncbi_metadata(files$ncbi_metadata)
  

  meta <- combine_metadata_sources(ehi, gtdb, ncbi) %>%
    infer_host_metadata() %>%
    normalize_country_fields() %>%
    filter(host_class != "Insecta") %>%  #remove invertebrates (in this case we only have insects)
    filter(!is.na(host_type)) # remove mags that do not have host information!

  if (is.null(files$contig_map) || is.null(files$gene_annotations)) {
    warning(sprintf(
      "[%s] Missing %s; returning metadata only.",
      species,
      if (is.null(files$contig_map) && is.null(files$gene_annotations)) "contig_map and gene_annotations"
      else if (is.null(files$contig_map)) "contig_map"
      else "gene_annotations"
    ))
    return(list(metadata = meta, annotations = NULL))
  }

  
  
  # Create cleaned genome IDs in metadata (needed here)
  meta <- meta %>%
    dplyr::mutate(
      ID_clean = clean_accession(ID)
    )

  
  ann <- prep_gene_annotations_with_genome(
    ann_raw        = files$gene_annotations,
    contig_map_raw = files$contig_map
  )
  
  ann <- ann %>%
  dplyr::mutate(
    genome = clean_accession(genome)
  )
  
  
  #make sure that we only keep mags present in the gene_annotation file (I downloaded metadata from a lot of mags, but not necessarily downloaded all the mags)
  annotated_genomes <- unique(ann$genome)
  meta <- meta %>% filter(ID_clean %in% annotated_genomes)

  
  ndb <- files$ndb

  
  cleaned <- filter_species_MAGs(
    genome_metadata    = meta,
    genome_annotations = ann,
    ndb                = ndb,
    mags_to_remove     = mags_to_remove
  )

  
    
  cleaned$annotations <- cleaned$annotations %>%
  dplyr::filter(genome %in% cleaned$metadata$ID_clean)
  
  
  cleaned
}






## for wild vs human ONLY
process_one_species_2 <- function(species, mags_to_remove = character(0)) {
  files <- load_species_files(species)

  ehi  <- prep_ehi_metadata(files$ehi_metadata)
  gtdb <- prep_gtdb_final(files$gtdb_final)
  ncbi <- prep_ncbi_metadata(files$ncbi_metadata)
  

  meta <- combine_metadata_sources(ehi, gtdb, ncbi) %>%
    infer_host_metadata() %>%
    normalize_country_fields()%>%
    filter(host_type_2 %in% c("human", "wild")) #keep only mags with human and wild metadata

  if (is.null(files$contig_map) || is.null(files$gene_annotations)) {
    warning(sprintf(
      "[%s] Missing %s; returning metadata only.",
      species,
      if (is.null(files$contig_map) && is.null(files$gene_annotations)) "contig_map and gene_annotations"
      else if (is.null(files$contig_map)) "contig_map"
      else "gene_annotations"
    ))
    return(list(metadata = meta, annotations = NULL))
  }

  ann <- prep_gene_annotations_with_genome(
    ann_raw        = files$gene_annotations,
    contig_map_raw = files$contig_map
  )
  
  ndb <- files$ndb

  cleaned <- filter_species_MAGs(
    genome_metadata    = meta,
    genome_annotations = ann,
    ndb                = ndb,
    mags_to_remove     = mags_to_remove
  )

  cleaned
}


## for domestic vs human ONLY
process_one_species_3 <- function(species, mags_to_remove = character(0)) {
  files <- load_species_files(species)

  ehi  <- prep_ehi_metadata(files$ehi_metadata)
  gtdb <- prep_gtdb_final(files$gtdb_final)
  ncbi <- prep_ncbi_metadata(files$ncbi_metadata)
  

  meta <- combine_metadata_sources(ehi, gtdb, ncbi) %>%
    infer_host_metadata() %>%
    normalize_country_fields()%>%
    filter(host_type_2 %in% c("human", "domestic")) #keep only mags with human and domestic metadata

  if (is.null(files$contig_map) || is.null(files$gene_annotations)) {
    warning(sprintf(
      "[%s] Missing %s; returning metadata only.",
      species,
      if (is.null(files$contig_map) && is.null(files$gene_annotations)) "contig_map and gene_annotations"
      else if (is.null(files$contig_map)) "contig_map"
      else "gene_annotations"
    ))
    return(list(metadata = meta, annotations = NULL))
  }

  ann <- prep_gene_annotations_with_genome(
    ann_raw        = files$gene_annotations,
    contig_map_raw = files$contig_map
  )
  
  ndb <- files$ndb

  cleaned <- filter_species_MAGs(
    genome_metadata    = meta,
    genome_annotations = ann,
    ndb                = ndb,
    mags_to_remove     = mags_to_remove
  )

  cleaned
}
  




## for domestic vs human ONLY
process_one_species_4 <- function(species, mags_to_remove = character(0)) {
  files <- load_species_files(species)

  ehi  <- prep_ehi_metadata(files$ehi_metadata)
  gtdb <- prep_gtdb_final(files$gtdb_final)
  ncbi <- prep_ncbi_metadata(files$ncbi_metadata)
  

  meta <- combine_metadata_sources(ehi, gtdb, ncbi) %>%
    infer_host_metadata() %>%
    normalize_country_fields()%>%
    filter(host_type_2 %in% c("wild", "domestic")) #keep only mags with wild and domestic metadata

  if (is.null(files$contig_map) || is.null(files$gene_annotations)) {
    warning(sprintf(
      "[%s] Missing %s; returning metadata only.",
      species,
      if (is.null(files$contig_map) && is.null(files$gene_annotations)) "contig_map and gene_annotations"
      else if (is.null(files$contig_map)) "contig_map"
      else "gene_annotations"
    ))
    return(list(metadata = meta, annotations = NULL))
  }

  ann <- prep_gene_annotations_with_genome(
    ann_raw        = files$gene_annotations,
    contig_map_raw = files$contig_map
  )
  
  ndb <- files$ndb

  cleaned <- filter_species_MAGs(
    genome_metadata    = meta,
    genome_annotations = ann,
    ndb                = ndb,
    mags_to_remove     = mags_to_remove
  )

  cleaned
}
  
```


## DREP ANALYSIS

### Recreating the drep tree
Checking for outliers 

```{r recreate_drep_tree}
## Function to recreate the drep tree from the secondary alignment (FASTANI)

recreate_drep_tree <- function(ndb){
  genomes <- unique(c(ndb$reference, ndb$querry))

# Create empty matrix
ani_matrix <- matrix(0, nrow = length(genomes), ncol = length(genomes))
rownames(ani_matrix) <- genomes
colnames(ani_matrix) <- genomes

# Fill the matrix
for(i in 1:nrow(ndb)) {
  ref <- ndb$reference[i]
  qry <- ndb$querry[i]
  ani_val <- ndb$ani[i]
  
  ani_matrix[ref, qry] <- ani_val
  ani_matrix[qry, ref] <- ani_val  # Make symmetric
}

# Set diagonal to 100
diag(ani_matrix) <- 100

# Convert to distance matrix (for tree building)
dist_matrix <- as.dist(1 - ani_matrix)

# Build tree
hc <- hclust(dist_matrix, method = "average") #drep uses hierarchical clustering
tree <- as.phylo(hc)

drep_tree <- ggtree(tree) + 
  geom_tiplab(size = 2) +
  theme_tree2()
  
print(drep_tree)
}
  
  

```


```{r drep_tree_test, fig.height = 10, fig.width = 12}
### testing with one species
ndb <- read_csv("data/mags_metadata/phocaeicola_vulgatus_Ndb.csv") 
recreate_drep_tree(ndb)
```


### Drep final tree, matrix and plots
```{r drep_tree_metadata}
# --- Helpers ------------------------------------------------------------------



# Strip file extensions so IDs match metadata
clean_label <- function(x) {
  if (is.null(x)) return(character(0))
  # Flatten factors/logicals/numerics to character, also handles tibbles columns
  x <- as.character(x)
  # basename requires character; protect against NA
  x <- ifelse(is.na(x), NA_character_, x)
  x <- basename(x)
  sub("\\.(fna|fa|fasta)(\\.gz)?$", "", x, ignore.case = TRUE)
}

# Parse accessions allowing optional external prefixes like RS_ or GB_ (possibly multiple),
# followed by GCA|GCF, digits (with leading zeros), and optional .version
# Returns: ext_prefix (may be ""), base_prefix (GCA/GCF), digits (with zeros),
#          core (numeric, no zeros), version (integer or NA)
parse_accession <- function(x) {
  x_clean <- clean_label(x)
 
  re <- "^((?:[A-Za-z0-9]+_)*)(GCA|GCF)_([0-9]+)(?:\\.([0-9]+))?$"
  m <- regexec(re, x_clean, ignore.case = FALSE)
  parts <- regmatches(x_clean, m)

  out <- lapply(parts, function(p) {
    if (length(p) == 0) {
      return(list(
        ext_prefix  = NA_character_,
        base_prefix = NA_character_,
        digits      = NA_character_,
        core        = NA_integer_,
        version     = NA_integer_
      ))
    }
    ext_prefix  <- p[2]
    base_prefix <- p[3]
    digits      <- p[4]
    version     <- if (!is.na(p[5])) as.integer(p[5]) else NA_integer_
    core        <- suppressWarnings(as.integer(sub("^0+", "", digits)))
    if (is.na(core)) core <- 0L  # for a string of zeros (rare), keep a numeric
    list(
      ext_prefix  = ext_prefix,
      base_prefix = base_prefix,
      digits      = digits,
      core        = core,
      version     = version
    )
  })

  tibble::tibble(
    ext_prefix  = vapply(out, `[[`, character(1), "ext_prefix"),
    base_prefix = vapply(out, `[[`, character(1), "base_prefix"),
    digits      = vapply(out, `[[`, character(1), "digits"),
    core        = vapply(out, `[[`, integer(1),   "core"),
    version     = vapply(out, `[[`, integer(1),   "version")
  )
}

# Canonical grouping key:
# If accession-like, group by numeric core (ignores external prefixes and leading zeros).
# Else, group by the cleaned label itself.
canonical_key <- function(x) {
  x_clean <- clean_label(x)
  acc <- parse_accession(x_clean)
  ifelse(!is.na(acc$core), paste0("ACC_CORE_", acc$core), x_clean)
}

# Build a normalized label with NO external prefix by default.

build_normalized_label <- function(base_prefix, digits, version, ext_prefix = "", keep_external_prefix = FALSE) {
  core_part <- paste0(base_prefix, "_", digits)
  if (!is.na(version) && version > 0) {
    core_part <- paste0(core_part, ".", version)
  }
  if (keep_external_prefix && nzchar(ext_prefix)) {
    paste0(ext_prefix, core_part)
  } else {
    core_part
  }
}

# Within a set of labels sharing the same canonical key, choose the representative:
# Priority:
#   1) Prefer base_prefix == "GCF"
#   2) Higher version number
#   3) Longer digits, then lexicographic for stability
# Output: normalized label (no external prefix by default)

choose_preferred_label <- function(labels, keep_external_prefix = FALSE) {
  labs <- clean_label(labels)
  acc  <- parse_accession(labs)
  is_acc <- !is.na(acc$base_prefix)

  if (any(is_acc)) {
    df <- tibble::tibble(
      lab          = labs,
      ext_prefix   = acc$ext_prefix,
      base_prefix  = acc$base_prefix,
      digits       = acc$digits,
      core         = acc$core,
      version      = acc$version
    )
    df$version[is.na(df$version)] <- 0L

    ord <- order(
      !(df$base_prefix == "GCF"),         # GCF first
      -df$version,                        # higher version
      -nchar(df$digits),                  # longer digits 
      df$lab                              # stable tie-breaker
    )
    df <- df[ord, , drop = FALSE]

    # Build normalized label for the top candidate, drop external prefix by default
    return(build_normalized_label(
      base_prefix = df$base_prefix[1],
      digits      = df$digits[1],
      version     = df$version[1],
      ext_prefix  = df$ext_prefix[1],
      keep_external_prefix = keep_external_prefix
    ))
  } else {
    # No accession-like labels, just pick a stable representative
    return(sort(unique(labs))[1])
  }
}

# Build a label rewrite map so that all labels in the same canonical group map to one preferred label
build_label_map <- function(labels, keep_external_prefix = FALSE) {
  labs <- unique(clean_label(labels))
  keys <- canonical_key(labs)
  grp  <- split(labs, keys)
  pref <- vapply(grp, choose_preferred_label, character(1), keep_external_prefix = keep_external_prefix)
  from <- unlist(grp, use.names = FALSE)
  to   <- rep(pref[names(grp)], lengths(grp))
  tibble::tibble(from = from, to = to)
}

# --- Main ---------------------------------------------------------------------

drep_tree_metadata <- function(genome_metadata = genome_metadata,
                               ndb = ndb,
                               drop_ids = character(0),
                               keep_external_prefix = FALSE) {
  metadata <- genome_metadata
  id_col   <- "ID_clean"

  # Start from ndb edges
  fastani <- ndb %>%
    dplyr::select(reference, querry, ani) %>%  # fully-qualified per your preference
    dplyr::filter(!is.na(reference), !is.na(querry), reference != querry) %>%
    dplyr::mutate(
      reference = clean_label(reference),
      querry    = clean_label(querry)
    ) %>%
    dplyr::filter(reference != querry)

  # Collect all labels we may see anywhere (pairs + metadata IDs)
  all_labels <- unique(c(
    fastani$reference,
    fastani$querry,
    clean_label(metadata[[id_col]])
  ))

  # Build rewrite map: GCA/GCF with the same core collapse to one preferred label (GCF > GCA; higher version wins).
  # External prefixes (RS_/GB_/...) are dropped by default in outputs.
  lab_map <- build_label_map(all_labels, keep_external_prefix = keep_external_prefix)

  # Helper to rewrite vectors using a named map
  rewrite_labels <- function(x) {
    x_clean <- clean_label(x)
    map_vec <- stats::setNames(lab_map$to, lab_map$from)
    m <- map_vec[x_clean]
    x_clean[!is.na(m)] <- m[!is.na(m)]
    x_clean
  }

  # Rewrite reference/querry and collapse pair duplicates by taking max ANI
  fastani <- fastani %>%
    dplyr::mutate(
      reference = rewrite_labels(reference),
      querry    = rewrite_labels(querry)
    ) %>%
    dplyr::filter(reference != querry) %>%
    dplyr::group_by(reference, querry) %>%
    dplyr::summarise(ani = max(ani), .groups = "drop")

  # Optionally drop genomes (use rewritten IDs so the drop matches what will appear in the tree)
  if (length(drop_ids) > 0) {
    drop_ids2 <- rewrite_labels(drop_ids)
    fastani <- fastani %>%
      dplyr::filter(!reference %in% drop_ids2, !querry %in% drop_ids2)
  }

  # Build a symmetric ANI matrix with diag = 1
  genomes <- sort(unique(c(fastani$reference, fastani$querry)))

  both <- dplyr::bind_rows(
    fastani %>% dplyr::transmute(reference, querry, ani),
    fastani %>% dplyr::transmute(reference = querry, querry = reference, ani)
  ) %>%
    dplyr::distinct(reference, querry, .keep_all = TRUE)

  ani_mat <- both %>%
    tidyr::complete(reference = genomes, querry = genomes) %>%
    tidyr::pivot_wider(
      names_from  = querry,
      values_from = ani,
      values_fn   = max
    ) %>%
    tibble::column_to_rownames("reference") %>%
    as.matrix()

  diag(ani_mat) <- 1

  # Build the tree (UPGMA/average as in dRep)
  dist_mat <- 1 - ani_mat
  hc <- hclust(as.dist(dist_mat), method = "average")
  tree <- ape::as.phylo(hc)

  # Deduplicate and rewrite metadata IDs with the same map so labels match the tree tips
  metadata_dedup <- metadata %>%
    dplyr::mutate(!!id_col := rewrite_labels(.data[[id_col]])) %>%
    dplyr::distinct(.data[[id_col]], .keep_all = TRUE)

  stopifnot(id_col %in% names(metadata_dedup))

  tip_df <- tibble::tibble(label = tree$tip.label) %>%
    dplyr::left_join(metadata_dedup, by = setNames(id_col, "label")) %>%
    dplyr::mutate(label_clean = label)

  return(list(
    tree = tree,
    tip_df = tip_df,
    ani_matrix = ani_mat
  ))
}

  

####-----------------------------------------------------------------------------------------

# Helper for ANI axis labels 
ani_axis <- function(p, tree, show_threshold = NULL) {
  df <- ggtree::fortify(tree)
  max_x <- max(df$x[df$isTip])
  p <- p +
    scale_x_continuous(
      labels = function(x) round(100 * (1 - (max_x - x)), 1)
    ) +
    coord_cartesian(xlim = c(0, max_x + 0.005)) +
    labs(x = "Average Nucleotide Identity (ANI, %)") +
    theme_tree2()
  if (!is.null(show_threshold)) {
    thr_delta <- 1 - (show_threshold / 100)
    p <- p + geom_vline(xintercept = max_x - thr_delta,
                        linetype = "dashed", color = "red", size = 0.8)
  }
  p
}


### --------------PLOTS ----------------------------------------
# - BASIC PLOT --
plot_tree_basic <- function(tree, tip_df, color_by = NULL, label_tips = TRUE,
                            point_size = 2.5, show_threshold = NULL) {
  # ensure completeness numeric for continuous gradients
  if ("completeness" %in% names(tip_df)) {
    tip_df <- tip_df %>% dplyr::mutate(completeness = as.numeric(completeness))
  }

  p <- ggtree(tree, size = 0.8)
  p <- p %<+% tip_df

  if (!is.null(color_by) && color_by %in% names(tip_df)) {
    p <- p + geom_tippoint(aes(color = !!rlang::sym(color_by)), size = point_size)

    # Palette/scale logic
    if (identical(color_by, "source") && exists("source_colors")) {
      p <- p + scale_color_manual(values = source_colors, name = "Source", drop = FALSE)
    } else if (identical(color_by, "host_order") && exists("host_order_colors")) {
      p <- p + scale_color_manual(values = host_order_colors, name = "Host order", drop = FALSE)
    } else if (is.numeric(tip_df[[color_by]])) {
      p <- p + scale_color_gradient(low = "white", high = "#08306B",
                                    name = paste0(color_by, " (%)"))
    }
  } else {
    p <- p + geom_tippoint(size = point_size)
  }

  if (label_tips) {
    lab_col <- if ("label_clean" %in% names(tip_df)) "label_clean" else "label"
    p <- p + geom_tiplab(size = 3, hjust = -0.1, aes(label = .data[[lab_col]]))
  }

  ani_axis(p, tree, show_threshold = show_threshold)
}

```


```{r plotting tree basic}
plot_tree_basic <- function(
  tree,
  tip_df,
  color_by = NULL,
  label_tips = TRUE,
  point_size = 2.5,
  show_threshold = NULL,
  palette = NULL   # <-- new optional palette
) {

  # make completeness numeric if needed
  if ("completeness" %in% names(tip_df)) {
    tip_df <- tip_df %>% dplyr::mutate(completeness = as.numeric(completeness))
  }

  p <- ggtree(tree, size = 0.8)
  p <- p %<+% tip_df

  if (!is.null(color_by) && color_by %in% names(tip_df)) {

    p <- p + geom_tippoint(aes(color = !!rlang::sym(color_by)), size = point_size)

    # ---- scale logic ----

    # 1) numeric : gradient
    if (is.numeric(tip_df[[color_by]])) {

      p <- p + scale_color_gradient(
        low = "white",
        high = "#08306B",
        name = paste0(color_by, " (%)")
      )

    # 2) categorical + palette provided
    } else if (!is.null(palette)) {

      p <- p + scale_color_manual(
        values = palette,
        name   = color_by,
        drop   = FALSE   # keep all levels in legend
      )

    # 3) default custom palettes from environment
    } else if (identical(color_by, "source") && exists("source_colors")) {

      p <- p + scale_color_manual(values = source_colors, name = "Source", drop = FALSE)

    } else if (identical(color_by, "host_order") && exists("host_order_colors")) {

      p <- p + scale_color_manual(values = host_order_colors, name = "Host order", drop = FALSE)

    } else if (identical(color_by, "country_normalized") && exists("country_palette")) {

      p <- p + scale_color_manual(values = country_palette, name = "Country", drop = FALSE)

    }

  } else {
    p <- p + geom_tippoint(size = point_size)
  }

  # tip labels
  if (label_tips) {
    lab_col <- if ("label_clean" %in% names(tip_df)) "label_clean" else "label"
    p <- p + geom_tiplab(size = 3, hjust = -0.1, aes(label = .data[[lab_col]]))
  }

  ani_axis(p, tree, show_threshold = show_threshold)
}
```




```{r testing phocaeicola, fig.height= 14, fig.width = 10}
# phocaeicola_results <- drep_tree_metadata(genome_metadata, ndb, drop_ids = c("GCA_040912025.1", "GCF_048453085.1", "GCA_048453085.1"))
# 
# tree <- phocaeicola_results$tree
# tip_df <- phocaeicola_results$tip_df
# ani_matrix <- phocaeicola_results$ani_matrix
# 
# 
# p_host_order <- plot_tree_basic(tree, tip_df, color_by = "host_order",
#                           label_tips = TRUE, show_threshold = 99.5)
# p_host_order

```

### PERMANOVA, PcoA

```{r functions for drep analysis}
#-------------1) Filter out unkown host type ------------------
filter_metadata <- function(tip_df) {
  tip_df %>%
    dplyr::filter(!is.na(host_type) & host_type != "")
}

#-------------2) Subset and align (same order) ------------------
subset_and_align_ani <- function(ani_matrix, meta_df) {
  keep_ids <- intersect(rownames(ani_matrix), meta_df$label)
  
  ani_mat_f <- ani_matrix[keep_ids, keep_ids, drop = FALSE]
  
  meta_f <- meta_df %>%
    dplyr::filter(label %in% keep_ids) %>%
    dplyr::arrange(match(label, rownames(ani_mat_f)))
  
  stopifnot(identical(meta_f$label, rownames(ani_mat_f)))
  
  list(ani = ani_mat_f, meta = meta_f)
}

#-------------3) Run PERMANOVA ------------------
run_permanova <- function(dist_mat, meta_f) {
  set.seed(1)
  ad <- vegan::adonis2(dist_mat ~ host_type, data = meta_f, permutations = 999)
  bd <- vegan::betadisper(dist_mat, meta_f$host_type)
  list(adonis = ad, dispersion = anova(bd))
}

#-------------4) Run PCOA ------------------
run_pcoa <- function(dist_mat, meta_f) {
  pcoa_res <- ape::pcoa(dist_mat)
  data.frame(
    PC1 = pcoa_res$vectors[, 1],
    PC2 = pcoa_res$vectors[, 2],
    host_type = meta_f$host_type,
    host_order = meta_f$host_order
  )
}

#-------------5) Plots PCOA ------------------
plot_pcoa_host_type <- function(pcoa_df, host_type_colors) {
  ggplot2::ggplot(pcoa_df, ggplot2::aes(PC1, PC2, color = host_type)) +
    ggplot2::geom_point(size = 2, alpha = 0.9) +
    ggplot2::scale_color_manual(values = host_type_colors, name = "Host type") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(title = "PCoA of ANI distances (filtered to known host_type)")
}

plot_pcoa_host_order <- function(pcoa_df, host_order_colors) {
  ggplot2::ggplot(pcoa_df, ggplot2::aes(PC1, PC2, color = host_order)) +
    ggplot2::geom_point(size = 2, alpha = 0.9) +
    ggplot2::scale_color_manual(values = host_order_colors, name = "Host order") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(title = "PCoA of ANI distances")
}

#-------------5) Pairwise ANI ------------------
pairwise_ani <- function(ani_mat_f, meta_f) {
  ani_mat_f %>%
    as.data.frame() %>%
    tibble::rownames_to_column("ref") %>%
    tidyr::pivot_longer(-ref, names_to = "qry", values_to = "ani") %>%
    dplyr::filter(ref < qry) %>%  # unique pairs
    dplyr::mutate(
      ref_host = meta_f$host_type[match(ref, meta_f$label)],
      qry_host = meta_f$host_type[match(qry, meta_f$label)],
      pair_type = dplyr::case_when(
        ref_host == "human"  & qry_host == "human"  ~ "human–human",
        ref_host == "animal" & qry_host == "animal" ~ "animal–animal",
        TRUE ~ "animal–human"
      )
    )
}


#-------------6) Plot pairwise ANI boxplot ------------------
plot_pairwise_boxplot <- function(ani_long_f) {
  ggplot2::ggplot(ani_long_f, ggplot2::aes(pair_type, ani, fill = pair_type)) +
    ggplot2::geom_boxplot(outlier_size = 0.5, width = 0.7) +
    ggplot2::scale_fill_viridis_d(guide = "none") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(x = NULL, y = "ANI", title = "Within- and between-group ANI (known host_type only)")
}


#-------------7) Wilcoxon tests ------------------
wilcox_within <- function(ani_long_f) {
  subset_within <- ani_long_f %>%
    dplyr::filter(pair_type %in% c("animal–animal", "human–human"))

  wt <- stats::wilcox.test(ani ~ pair_type, data = subset_within, exact = FALSE)

  summary_df <- subset_within %>%
    dplyr::group_by(pair_type) %>%
    dplyr::summarise(n = dplyr::n(), median_ani = stats::median(ani), .groups = "drop")

  list(test = wt, summary = summary_df)
}

wilcox_cross_vs_within <- function(ani_long_f) {
  # Compare cross-host to pooled within (animal–animal + human–human)
  df <- ani_long_f %>%
    dplyr::mutate(is_cross = ifelse(pair_type == "animal–human", "cross", "within"))

  wt <- stats::wilcox.test(ani ~ is_cross, data = df, exact = FALSE, alternative = "less")
  # 'less' tests if cross < within (lower ANI between hosts)

  summary_df <- df %>%
    dplyr::group_by(is_cross) %>%
    dplyr::summarise(n = dplyr::n(), median_ani = stats::median(ani), .groups = "drop")

  list(test = wt, summary = summary_df)
}



#----------------------------------------------------#

#-------------- DREP STATS ANALYSIS ------------------
# combines the functions above into a big one


drep_stat_analysis <- function(tip_df, ani_matrix, host_type_colors, host_order_colors) {
  tip_df_f <- tip_df %>%
    dplyr::filter(!is.na(host_type) & host_type != "")

  aligned <- subset_and_align_ani(ani_matrix, tip_df_f)
  ani_mat_f <- aligned$ani
  meta_f <- aligned$meta

  if (is.factor(meta_f$host_type)) meta_f$host_type <- droplevels(meta_f$host_type)

  dist_mat_f <- stats::as.dist(1 - ani_mat_f)

  permanova_results <- run_permanova(dist_mat_f, meta_f)

  pcoa_df <- run_pcoa(dist_mat_f, meta_f)
  pcoa_host_type_plot <- plot_pcoa_host_type(pcoa_df, host_type_colors)
  pcoa_host_order_plot <- plot_pcoa_host_order(pcoa_df, host_order_colors)

  ani_long_f <- pairwise_ani(ani_mat_f, meta_f)
  boxplot_pairs <- plot_pairwise_boxplot(ani_long_f)

  w_within <- wilcox_within(ani_long_f)
  w_cross  <- wilcox_cross_vs_within(ani_long_f)

  list(
    filtered_metadata = meta_f,
    ani = ani_mat_f,
    dist = dist_mat_f,
    permanova = permanova_results,
    pcoa = list(data = pcoa_df,
                by_host_type = pcoa_host_type_plot,
                by_host_order = pcoa_host_order_plot),
    pairwise = list(data = ani_long_f,
                    boxplot = boxplot_pairs,
                    wilcox_within = w_within,
                    wilcox_cross_vs_within = w_cross)
  )
}
```


```{r testing drep_stat_analysis function}
# res <- drep_stat_analysis(
#   tip_df = tip_df,
#   ani_matrix = ani_matrix,
#   host_type_colors = host_type_colors,
#   host_order_colors = host_order_colors
# )
# 
# # Print stats
# res$permanova$adonis
# res$permanova$dispersion
# res$pairwise$wilcox_within$test
# res$pairwise$wilcox_cross_vs_within$test
# 
# # Plot
# print(res$pcoa$by_host_type)
# print(res$pcoa$by_host_order)
# print(res$pairwise$boxplot)

# Save plots:
# ggplot2::ggsave("pcoa_host_type.png", res$pcoa$by_host_type, width = 6, height = 5, dpi = 300)


```

## FUNCTIONAL ANALYSIS: KEGG, AMR, VF, CAZY
### Create annotation matrices
```{r make_annotation_matrices}
#testing with hafnia
sp <- "citrobacter_braakii"
res <- process_one_species(sp)

hp <- res$metadata
genome_annotations <- res$annotations
genome_metadata <- hp

# Make annotation matrix - checks if annotation columns are empty and keeps running

make_annotation_matrix <- function(genome_annotations, column) {
  # Basic validation 
  if (is.null(genome_annotations) || !is.data.frame(genome_annotations) || nrow(genome_annotations) == 0) {
    message(sprintf("make_annotation_matrix(): annotations are empty or NULL; returning empty matrix for column '%s'.", column))
    return(tibble::tibble(genome = character()))
  }

  if (!column %in% names(genome_annotations)) {
    message(sprintf("Column '%s' not found in annotations; skipping and returning genome-only matrix.", column))
    # Try to extract genomes if present; else return empty
    genomes <- if ("genome" %in% names(genome_annotations)) unique(genome_annotations$genome) else character()
    return(tibble::tibble(genome = genomes))
  }

  # Filter to non-NA in the requested column
  out <- genome_annotations %>%
    dplyr::filter(!is.na(.data[[column]]))

  if (nrow(out) == 0) {
    message(sprintf("Column '%s' exists but has no non-NA values; returning genome-only matrix.", column))
    genomes <- if ("genome" %in% names(genome_annotations)) unique(genome_annotations$genome) else character()
    return(tibble::tibble(genome = genomes))
  }

  # Build the wide binary matrix (counts; you can coerce to 0/1 later if needed)
  out %>%
    dplyr::count(genome, annotation = .data[[column]]) %>%
    tidyr::pivot_wider(
      names_from  = annotation,
      values_from = n,
      values_fill = 0
    )
}


#testing
kegg_mat <- make_annotation_matrix(genome_annotations, "kegg")
amr_mat  <- make_annotation_matrix(genome_annotations, "resistance_target")
vf_mat   <- make_annotation_matrix(genome_annotations, "vf")
```

### Run Fisher tests on proportions
```{r fisher}
## general fisher function
## Runs Fisher tests comparing the proportion of MAGs containing a gene (KEGG, AMR, CAZy, VF) between two groups (ie: human vs wild or any other)

  # mat: wide matrix with columns: genome, <annotation columns...>
  # metadata: data frame with columns ID, host_type
  # annotation_name: name for the long column (e.g., "kegg", "amr")

run_fisher_tests_general <- function(
  mat,
  metadata,
  group_col = "host_type",   
  annotation_name = "feature"
) {
  # --- Safety checks ---------------------------------------------------------
  if (is.null(mat) || !is.data.frame(mat) || nrow(mat) == 0) return(tibble::tibble())
  if (!"genome" %in% names(mat)) return(tibble::tibble())
  feature_cols <- setdiff(names(mat), "genome")
  if (length(feature_cols) == 0) return(tibble::tibble())

  if (is.null(metadata) || !is.data.frame(metadata) || nrow(metadata) == 0) return(tibble::tibble())
  if (!all(c("ID_clean", group_col) %in% names(metadata))) {
    stop(paste0("metadata must contain ID_clean and ", group_col))
  }

  # Require exactly two groups (e.g., human_adj vs wild)
  host_groups <- unique(metadata[[group_col]])
  if (length(host_groups) != 2) {
    stop(paste0("'", group_col, "' must contain exactly 2 groups for Fisher testing. Found: ",
                paste(host_groups, collapse = ", ")))
  }
  g1 <- host_groups[1]
  g2 <- host_groups[2]

  # --- Presence/absence ------------------------------------------------------
  pa <- mat %>%
    dplyr::mutate(dplyr::across(-genome, ~ as.integer(!is.na(.) & . != 0)))

  pa_long <- pa %>%
    tidyr::pivot_longer(
      cols = -genome,
      names_to = annotation_name,
      values_to = "present"
    ) %>%
    dplyr::filter(present == 1L) %>%
    dplyr::distinct(genome, .data[[annotation_name]])

  if (nrow(pa_long) == 0) return(tibble::tibble())

  # Attach group column
  pa_with_group <- pa_long %>%
    dplyr::left_join(metadata %>% dplyr::select(ID_clean, all_of(group_col)),
                     by = c("genome" = "ID_clean"))

  if (all(is.na(pa_with_group[[group_col]]))) return(tibble::tibble())

  # Count presence by group
  pa_counts <- pa_with_group %>%
    dplyr::group_by(.data[[group_col]], .data[[annotation_name]]) %>%
    dplyr::summarise(n_mags = dplyr::n(), .groups = "drop")

  # Totals per group
  totals <- metadata %>%
    dplyr::group_by(.data[[group_col]]) %>%
    dplyr::summarise(total_mags = dplyr::n_distinct(ID_clean), .groups = "drop")

  # Add proportions + absences
  pa_props <- pa_counts %>%
    dplyr::left_join(totals, by = group_col) %>%
    dplyr::mutate(absent = total_mags - n_mags)

  # Spread into contingency table
  wide <- pa_props %>%
    dplyr::select(.data[[annotation_name]], all_of(group_col), n_mags, absent) %>%
    tidyr::pivot_wider(
      names_from  = all_of(group_col),
      values_from = c(n_mags, absent),
      values_fill = 0
    )

  # Expect: n_mags_<g1>, absent_<g1>, n_mags_<g2>, absent_<g2>
  expected_cols <- c(annotation_name,
                     paste0("n_mags_", g1), paste0("absent_", g1),
                     paste0("n_mags_", g2), paste0("absent_", g2))
  if (!all(expected_cols %in% names(wide))) {
    missing <- setdiff(expected_cols, names(wide))
    stop("Missing expected columns: ", paste(missing, collapse=", "))
  }

  # Fisher test
  results <- wide %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      ft = list(stats::fisher.test(
        matrix(
          c(.data[[paste0("n_mags_", g1)]],
            .data[[paste0("absent_", g1)]],
            .data[[paste0("n_mags_", g2)]],
            .data[[paste0("absent_", g2)]]),
          nrow = 2, byrow = TRUE
        )
      )),
      p_value    = ft$p.value,
      odds_ratio = unname(ft$estimate),
      conf_low   = ft$conf.int[1],
      conf_high  = ft$conf.int[2],
      log_or     = log(odds_ratio),
      log2_or    = log2(odds_ratio),
      se_log_or  = (log(conf_high) - log(conf_low)) / (2 * 1.96),
      se_log2_or = se_log_or / log(2)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-ft) %>%
    dplyr::mutate(
      p_adj       = p.adjust(p_value, method = "BH"),
      prop_g1     = .data[[paste0("n_mags_", g1)]] /
                    (.data[[paste0("n_mags_", g1)]] + .data[[paste0("absent_", g1)]]),
      prop_g2     = .data[[paste0("n_mags_", g2)]] /
                    (.data[[paste0("n_mags_", g2)]] + .data[[paste0("absent_", g2)]]),
      diff_prop   = prop_g2 - prop_g1
    )

  return(results)
}

```



### Volcano plot
```{r plotting volcano plot}
plot_volcano <- function(df, feature_col) {
  df %>%
    mutate(
      p_adj_capped = pmax(p_adj, 1e-300),
      nl10 = -log10(p_adj_capped),
      sig = p_adj < 0.05
    ) %>%
    ggplot(aes(x = log2_or, y = nl10)) +
    geom_point(aes(color = sig), size = 2) +
    scale_color_manual(values = c("grey70", "blue")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    ggrepel::geom_text_repel(
      data = . %>% filter(sig) %>% slice_max(nl10, n = 10),
      aes(label = .data[[feature_col]])
    ) +
    labs(
      title = paste("Volcano plot -", feature_col) ,
      x = "log2(Odds Ratio)",
      y = "-log10(adj. p-value)"
    ) +
    theme_minimal()

}


```

```{r pca_test}

run_pca <- function(mat) {
  m <- mat %>% column_to_rownames("genome")
  m <- m / rowSums(m)
  m <- m[, apply(m, 2, sd) > 0]
  
  prcomp(m, scale. = TRUE)
}

pca_kegg <- run_pca(kegg_mat)
pca_amr  <- run_pca(amr_mat)

```

### PERMANOVA on presence/absence matrix
To check if the groups (ie: wild animal vs human-sourced MAGs) have differences in overall presence/absence composition.

```{r permanova functional matrix}
pa_permanova <- function(mat, metadata, host_levels = NULL) {
  
  #mat - which functional matrix we want to test (ie:KEGG, VF...)
  #metadata- the genome metadata
  #host_levels- which host_type_2 we are comparing (ie: wild, human)
  
  mat <- mat %>%
  column_to_rownames("genome")

  #convert count matrx to presence/absence
  pa <- (mat > 0) * 1

  
  # remove zero-variance KOs
  pa_nz <- pa[, colSums(pa) > 0 & colSums(pa) < nrow(pa)]
  
  
  # restrict to selected hosts
  if (!is.null(host_levels)) {
      metadata <- metadata %>%
        dplyr::filter(host_type_2 %in% host_levels) %>%
        dplyr::mutate(host_type_2 = droplevels(factor(host_type_2)))
    }
  
    
  # determine the common genomes
  common_ids <- base::intersect(rownames(pa_nz), metadata$ID_clean)
  
  # Report what will be kept/dropped
  message("# common: ", length(common_ids))
  message("# in KEGG only: ", length(setdiff(rownames(pa_nz), metadata$ID_clean)))
  message("# in metadata only: ", length(setdiff(metadata$ID_clean, rownames(pa_nz))))
  
  
  
  if (length(common_ids) < 3) {
    message("Too few genomes after filtering")
    return(NULL)
  }

  
  # Subset to the intersection (and keep order identical)
  pa_nz <- pa_nz[common_ids, , drop = FALSE]
  
  meta <- metadata %>%
    dplyr::filter(ID_clean %in% common_ids) %>%
    dplyr::distinct(ID_clean, .keep_all = TRUE) %>%
    tibble::column_to_rownames("ID_clean") %>%
    .[common_ids, , drop = FALSE]
  
  stopifnot(identical(rownames(meta), rownames(pa_nz)))
  
  # Prepare variables for PERMANOVA
  required_vars <- c("genome_size", "completeness", "host_type_2")
  
  # Coerce types
  meta <- meta %>%
    dplyr::mutate(
      genome_size  = as.numeric(genome_size),
      completeness = as.numeric(completeness),
      host_type_2    = as.factor(host_type_2) 
      #country_normalized = as.factor(country_normalized)
    )
  
  # Align on complete cases (adonis2 drops NAs otherwise)
  ok <- stats::complete.cases(meta[, required_vars, drop = FALSE])
  pa_nz <- pa_nz[ok, , drop = FALSE]
  meta       <- meta[ok, , drop = FALSE]
  stopifnot(identical(rownames(meta), rownames(pa_nz)))
  
  # Remove genomes with no presence after filtering
  empty <- rowSums(pa_nz) == 0
  if (any(empty)) {
  message("Removing ", sum(empty), " empty genomes (all zeros).")
  pa_nz <- pa_nz[!empty, , drop = FALSE]
  meta  <- meta[!empty, , drop = FALSE]
}
  message("Remaining genomes: ", nrow(pa_nz))
  
  
  #####-----------------------------------------------------------
  ### Distance, dispersion, PERMANOVA
  
  ## Distance
  dist_pa <- suppressWarnings(
  suppressMessages(
    vegan::vegdist(pa_nz, method = "jaccard", binary = TRUE)
  )
)

  ## Beta-dispersion
  
  betadisp_results <- suppressWarnings(
    suppressMessages(
      vegan::betadisper(dist_pa, meta$host_type_2)
    )
  )

  ## Tidy so it prints nice :)
  tidy_betadisper <- function(bd) {
  # ANOVA table
  an <- as.data.frame(anova(bd))
  an$term <- rownames(an)
  rownames(an) <- NULL
  
  # Group mean dispersions
  means <- data.frame(
    group = names(bd$group),
    mean_distance = bd$means
  )
  
  list(
    anova_table = an,
    group_means = means
  )
}
## PERMANOVA 
###controlling for genome size and completeness, then seeing how much variance is explained by host type
 permanova_results <- vegan::adonis2(
    dist_pa ~ genome_size + completeness + host_type_2  ,
    data = meta,
    permutations = 999,
    by = "margin"
  )
 
 tidy_permanova <- function(ad) {
  df <- as.data.frame(ad)
  df$term <- rownames(df)
  rownames(df) <- NULL
  df
}
 
 
 ## return-----------------------------------
 results <- list(
  dist           = dist_pa,
  betadisper_raw = betadisp_results,
  betadisper     = tidy_betadisper(betadisp_results),
  permanova_raw  = permanova_results,
  permanova      = tidy_permanova(permanova_results),
  pa_nz          = pa_nz
)
 
  return(results)
}

results <-pa_permanova(kegg_mat, genome_metadata, host_levels=c("human", "wild"))
results$permanova

```

## GIFTs

### Plotting GIFTs
```{r gift pcoa plots}
gift_pcoa_plots <- function(genome_gifts, genome_metadata) {
  # Convert the GIFTs into a matrix of functional elements per genome (row)
  gift_pcoa <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    vegdist(method="euclidean") %>%
    pcoa() #principal component analysis
    
    #Extract eigenvalues (variance explained by first 10 axes)
    gift_pcoa_rel_eigen <- gift_pcoa$values$Relative_eig[1:10]
    
    
    # Get genome positions
    gift_pcoa_vectors <- gift_pcoa$vectors %>% #extract vectors
      as.data.frame() %>% 
     dplyr::select(Axis.1,Axis.2) # keep the first 2 axes
    
    
    gift_pcoa_eigenvalues <- gift_pcoa$values$Eigenvalues[c(1,2)]
    
    
    #For the black arrows: Functional group loadings
    # covariance between each functional trait and pcoa axis scores
    #scale with the eigenvectors
    gift_pcoa_gifts <- cov(genome_gifts, scale(gift_pcoa_vectors)) %*%
      diag((gift_pcoa_eigenvalues/(nrow(genome_gifts)-1))^(-0.5)) %>%
      as.data.frame() %>% 
      dplyr::rename(Axis.1=1,Axis.2=2) %>% 
      rownames_to_column(var="label") %>% 
      #get function summary vectors
      mutate(func=substr(label,1,3)) %>% 
      group_by(func) %>% #grouping by function
      summarise(Axis.1=mean(Axis.1),
                Axis.2=mean(Axis.2)) %>% 
      dplyr::rename(label=func) %>% 
      filter(!label %in% c("S01","S02","S03"))
      
  
    set.seed(101)
scale <- 20 # scale for vector loadings (to make arrows visible)
arrows_plot <- gift_pcoa_vectors %>% 
  rownames_to_column(var="ID") %>% 
  left_join(genome_metadata, by="ID") %>%
  ggplot() +
      #genome positions
      scale_color_manual(values=host_type_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color = host_type), 
                 alpha=0.9, shape=16) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen[1]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA1")
            ) +
     scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen[2]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA2")
            ) +
    geom_label_repel(data = gift_pcoa_gifts,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
    xlim(-0.8,1) + 
    ylim(-1,1.5) +
    theme_minimal() + 
  
  labs(
    x = paste0("PCoA1 (", round(gift_pcoa_rel_eigen[1]*100, 2), " %)"),
    y = paste0("PCoA2 (", round(gift_pcoa_rel_eigen[2]*100, 2), " %)"))
  
  centroids <- gift_pcoa_vectors %>%
  rownames_to_column(var = "ID") %>%
  left_join(genome_metadata, by = "ID") %>%
  group_by(host_type) %>%
  summarise(
    Axis.1 = mean(Axis.1, na.rm = TRUE),
    Axis.2 = mean(Axis.2, na.rm = TRUE),
    .groups = "drop"
  )

big_plot <- gift_pcoa_vectors %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(genome_metadata, by = "ID") %>%
  ggplot(aes(x = Axis.1, y = Axis.2, color = host_type)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = host_type_colors) +

  labs(
    x = paste0("PCoA1 (", round(gift_pcoa_rel_eigen[1] * 100, 2), " %)"),
    y = paste0("PCoA2 (", round(gift_pcoa_rel_eigen[2] * 100, 2), " %)")
  ) +

  coord_cartesian(xlim = c(-0.6, 1), ylim = c(-1, 1)) +

  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(size = 14, face = "bold"),
    axis.text  = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14),
    panel.grid.minor = element_blank()
  )

big_order_plot <- gift_pcoa_vectors %>% 
  rownames_to_column(var = "ID") %>% 
  left_join(genome_metadata, by = "ID") %>%
  ggplot(aes(x = Axis.1, y = Axis.2, color = host_order)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_manual(values = host_order_colors) +

  labs(
    x = paste0("PCoA1 (", round(gift_pcoa_rel_eigen[1] * 100, 2), " %)"),
    y = paste0("PCoA2 (", round(gift_pcoa_rel_eigen[2] * 100, 2), " %)")
  ) +

  coord_cartesian(xlim = c(-0.6, 1), ylim = c(-1, 1)) +

  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_text(size = 14, face = "bold"),
    axis.text  = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14),
    panel.grid.minor = element_blank()
  )

results <- list(
    arrows_plot     = arrows_plot,
    big_plot        = big_plot,
    big_order_plot  = big_order_plot,
    gift_pcoa_vectors = gift_pcoa_vectors)
 
  return(results)
}
```


```{r gift plots}
plot_gifts <- function(GIFTs_elements, metadata_with_cluster) { 
cluster <- GIFTs_elements %>%
    as.data.frame() %>%
    rownames_to_column(var="MAG")%>%
    pivot_longer(!MAG,names_to="trait",values_to="gift") %>%
    left_join(metadata_with_cluster, by = join_by(MAG == ID_clean)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=MAG,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_grid(functionid ~ cluster, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=6),
              strip.text.y = element_text(angle = 0)
              ) +
        labs(y="Traits",x="Samples",fill="GIFT")
  
  
  
  host_type <- GIFTs_elements %>%
    as.data.frame() %>%
    rownames_to_column(var="MAG")%>%
    pivot_longer(!MAG,names_to="trait",values_to="gift") %>%
    left_join(metadata_with_cluster, by = join_by(MAG == ID_clean)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=MAG,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_grid(functionid ~ host_type, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=6),
              strip.text.y = element_text(angle = 0)
              ) +
        labs(y="Traits",x="Samples",fill="GIFT")

  antibiotic <-
    GIFTs_elements %>%
  melt() %>%
  dplyr::rename(Genome = Var1, Code_element = Var2, GIFT = value) %>%
  inner_join(GIFT_db, by = "Code_element") %>%
  filter(str_starts(Code_element, "D09")) %>%
  left_join(metadata_with_cluster, by = join_by(Genome == ID)) %>%
  mutate(cluster = factor(cluster)) %>%
  droplevels() %>%
  arrange(cluster, Genome) %>%
  mutate(trait = Element) %>%
  # Ensure trait is a factor to maintain order
  mutate(trait = factor(trait, levels = rev(unique(GIFT_db$Element)))) %>%
  ggplot( aes(x = Genome, y = trait)) +
  # Heatmap
  geom_tile(aes(fill = GIFT), color = "white") +
  scale_fill_gradientn(colours = brewer.pal(7, "YlGnBu"), name = "GIFT Score") +
  
  # Start a new fill scale for the host_type bar
  new_scale_fill() +
  
  # Add the host_type bar at the very top or bottom
  geom_tile(aes(y = -0.5, fill = host_type), height = 0.5) +
  scale_fill_manual(values = host_type_colors) +
  
  facet_grid(~ cluster, scales = "free_x", space = "free_x") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  results <- list(
    host_type   = host_type,
    cluster     = cluster,
    antibiotic  = antibiotic)
 
  return(results)
  
}
```

### Differences in GIFTs
```{r checking_diff_GIFTs}
#### Differences between clusters formed in PCoA

gift_stats <- function(GIFTs_elements,
                       metadata_with_cluster,
                       GIFT_db,
                       host_type_colors,
                       host_class_colors) {

  suppressWarnings({
    
    
    # PREPARE DATA –––––––––––––––––––––––––––––––
  
    gift_cols <- setdiff(colnames(GIFTs_elements),
                         c("genome", "cluster", "host_type"))
    
    gift_df <- GIFTs_elements %>%
      as.data.frame() %>%
      tibble::rownames_to_column("ID_clean") %>%
      left_join(
        metadata_with_cluster %>% distinct(ID_clean, .keep_all = TRUE),
        by = "ID_clean"
      )
    
    
   
    # KRUSKAL-WALLIS (CLUSTERS) ––––––––––––––––––––––––––––––––
    
    kw_p <- sapply(gift_cols, function(g) {
      kruskal.test(gift_df[[g]] ~ gift_df$cluster)$p.value
    })
    
    kruskal_table <- tibble(
      GIFT = gift_cols,
      p_value = kw_p,
      p_adj = p.adjust(kw_p, "BH")
    )
    
    
   
    # PAIRWISE WILCOXON (CLUSTERS) –––––––––––––––––––––––––––––––

    pairwise_sig_table <- map_df(gift_cols, function(g) {
      pw <- pairwise.wilcox.test(gift_df[[g]],
                                 gift_df$cluster,
                                 p.adjust.method = "BH",
                                 exact = FALSE)$p.value
      
      as.data.frame(as.table(pw)) %>%
        filter(!is.na(Freq), Freq < 0.05) %>%
        rename(group1 = Var1, group2 = Var2, p_adj = Freq) %>%
        mutate(GIFT = g)
    })
    
    

    # CLUSTER DIFFERENCE PLOT ––––––––––––––––––––––––––––––––
    
    cluster_diff_plot <- NULL
    
    if (nrow(pairwise_sig_table) > 0) {
      cluster_diff_plot <- GIFTs_elements %>%
        melt() %>%
        rename(Genome = Var1, Code_element = Var2, GIFT = value) %>%
        inner_join(GIFT_db, by = "Code_element") %>%
        filter(Code_element %in% pairwise_sig_table$GIFT) %>%
        distinct(Genome, Code_element, .keep_all = TRUE) %>%
        left_join(metadata_with_cluster, by = join_by(Genome == ID_clean)) %>%
        ggplot(aes(x = Genome, y = Element, fill = GIFT)) +
        geom_tile(color = "white", linewidth = 0.2) +
        scale_fill_gradientn(limits = c(0, 1),
          colours = RColorBrewer::brewer.pal(7, "YlGnBu")
        ) +
        new_scale_fill() +
        geom_tile(aes(y = -0.5, fill = host_type), height = 0.3) +
        scale_fill_manual(values = host_type_colors) +
        facet_grid(~ cluster, scales = "free_x", space = "free_x") +
        theme(axis.text.x =
                element_text(angle = 45, hjust = 1, size = 4)) +
        labs(y = "Trait", x = "Genome", fill = "GIFT")
    }
    
    
    
    # HOST-TYPE WILCOXON  ––––––––––––––––––––––––––––––––
    
    w_p <- sapply(gift_cols, function(g) {
      wilcox.test(gift_df[[g]] ~ gift_df$host_type,
                  exact = FALSE)$p.value
    })
    
    w_adj <- p.adjust(w_p, "BH")
    
    # Effect size (median difference)
    effect_size <- sapply(gift_cols, function(g) {
      median(gift_df[[g]][gift_df$host_type == "human"]) -
        median(gift_df[[g]][gift_df$host_type == "animal"])
    })
    
    wilcox_res_host_type <- tibble(
      GIFT = gift_cols,
      p_value = w_p,
      p_adj = w_adj,
      effect = effect_size
    ) %>%
      filter(p_adj < 0.05)
    
    
    
    # HOST TYPE DIFFERENCE PLOT --------------------
    
    host_type_diff_plot <- NULL
    
    if (nrow(wilcox_res_host_type) > 0) {
      host_type_diff_plot <- GIFTs_elements %>%
        melt() %>%
        rename(Genome = Var1, Code_element = Var2, GIFT = value) %>%
        inner_join(GIFT_db, by = "Code_element") %>%
        filter(Code_element %in% wilcox_res_host_type$GIFT) %>%
        distinct(Genome, Code_element, .keep_all = TRUE) %>%
        left_join(metadata_with_cluster, by = join_by(Genome == ID_clean)) %>%
        ggplot(aes(x = Genome, y = Element, fill = GIFT)) +
        geom_tile(color = "white", linewidth = 0.2) +
        scale_fill_gradientn(limits = c(0, 1),
          colours = RColorBrewer::brewer.pal(7, "YlGnBu")
        ) +
        new_scale_fill() +
        geom_tile(aes(y = -0.5, fill = host_class), height = 0.3) +
        scale_fill_manual(values = host_class_colors) +
        facet_grid(~ host_type, scales = "free_x", space = "free_x") +
        theme(axis.text.x = element_blank()) +
        labs(y = "Trait", x = "Genome", fill = "GIFT")
    }
    
    
    #–––––––––––––––––––––––––––––––––
    # RETURN ALL RESULTS
  
    list(
      # cluster analysis
      kruskal_table = kruskal_table,
      pairwise_sig_table = pairwise_sig_table,
      cluster_diff_plot = cluster_diff_plot,
      
      # host-type analysis
      wilcox_res_host_type = wilcox_res_host_type,
      host_type_diff_plot = host_type_diff_plot
    )
  })
}

```




