# Data preparation and statistics


## Load all the species data
```{r Define species and mags to remove list_2}
# ALL SPECIES
species_list <- c(
  "lactococcus_lactis",
  "parabacteroides_goldsteinii",
  "enterococcus_faecalis",
  "bacteroides_uniformis",
  "phocaeicola_vulgatus",
  "hafnia_paralvei",
  "citrobacter_braakii",
  "enterococcus_hirae",
  "akkermansia_muciniphila",
  "bacteroides_fragilis"
)


mags_to_remove_list <- list(
  lactococcus_lactis = c(
    "GCA_018369575.1.fna","GCA_947063445.1.fna","GCA_947101685.1.fna",
    "GCA_948698275.1.fna","GCA_937910935.1.fna","GCA_947072755.1.fna",
    "GCA_948655095.1.fna","GCA_948703095.1.fna","GCA_947041925.1.fna",
    "GCA_947073355.1.fna","GCA_948675165.1.fna","GCA_948718815.1.fna"
  ),
  parabacteroides_goldsteinii = character(0), # none to remove
  enterococcus_faecalis = character(0),
  hafnia_paralvei = character(0),
  bacteroides_uniformis = character(0),
  phocaeicola_vulgatus = c("GCA_040912025.1", "GCF_048453085.1", "GCA_048453085.1"),
  citrobacter_braakii = character(0),
  enterococcus_hirae = character(0),
  akkermansia_muciniphila = character(0),
  bacteroides_fragilis = character(0)
)

species_data <- list()

for (sp in species_list) {
  species_data[[sp]] <- process_one_species(
    species        = sp,
    mags_to_remove = mags_to_remove_list[[sp]]
  )
  
  #recreate_drep_tree(species_data[[sp]]$ndb)
}

pg <- species_data[["parabacteroides_goldsteinii"]]$metadata
eh <- species_data[["enterococcus_hirae"]]$metadata
ef <- species_data[["enterococcus_faecalis"]]$metadata
pv <- species_data[["phocaeicola_vulgatus"]]$metadata
am <- species_data[["akkermansia_muciniphila"]]$metadata
bf <- species_data[["bacteroides_fragilis"]]$metadata
bu <- species_data[["bacteroides_uniformis"]]$metadata
ll <- species_data[["lactococcus_lactis"]]$metadata
hp <- species_data[["hafnia_paralvei"]]$metadata
cb <- species_data[["citrobacter_braakii"]]$metadata


named_list <- list(
  parabacteroides_goldsteinii = pg,
  enterococcus_hirae          = eh,
  enterococcus_faecalis       = ef,
  phocaeicola_vulgatus        = pv,
  akkermansia_muciniphila     = am,
  bacteroides_fragilis        = bf,
  bacteroides_uniformis       = bu,
  lactococcus_lactis          = ll,
  hafnia_paralvei             = hp,
  citrobacter_braakii         = cb
)

total_genome_metadata <- dplyr::bind_rows(named_list, .id = "bacterial_species")


#total_genome_annotations <- dplyr::bind_rows(
#   lapply(species_data, `[[`, "annotations"),
#   .id = "bacterial_species"
# )


```



## Phylogenetic tree of the 10 bacterial species

```{r load tree and metadata, eval = FALSE}
# Tree and metadata from the GTDB Release 226

tree <- read.tree("data/bac120.tree")
meta <- read_tsv("data/bac120_metadata.tsv.gz")
```

```{r gtdb tree, eval = FALSE}
# Species list
my_species <- c(
  "Akkermansia muciniphila",
  "Bacteroides fragilis",
  "Bacteroides uniformis",
  "Citrobacter braakii",
  "Enterococcus_B hirae",
  "Enterococcus faecalis",
  "Hafnia paralvei",
  "Parabacteroides goldsteinii",
  "Phocaeicola vulgatus",
  "Lactococcus lactis")


meta <- meta %>%
  mutate(
    gtdb_species = str_extract(gtdb_taxonomy, "s__[^;]+") %>% 
                   str_replace("s__", "")
  )



selected_accessions <- meta %>%
  filter(gtdb_species %in% my_species) %>%
  pull(accession)


```

```{r pruning the tree, eval = FALSE}
# prune the GTDB tree to contain only our species
pruned <- drop.tip(tree, setdiff(tree$tip.label, selected_accessions))

label_map <- meta %>%
  filter(accession %in% pruned$tip.label) %>%
  select(accession, gtdb_species) %>%
  deframe()

pruned$tip.label <- label_map[pruned$tip.label]

pruned$tip.label <- pruned$tip.label %>%
  str_to_lower() %>%
  str_replace_all(" ", "_")



ggtree(pruned) + geom_tiplab(size = 3)
```

```{r load the bacterial tree}
load("data/pruned_tree.RData")
ggtree(pruned) + geom_tiplab(size = 3)
```



## Phylogenetic tree of the host species
```{r host tree, fig.height= 8, fig.width = 10}

host_tree <- read.tree("data/host_species_tree.nwk")
plot(host_tree, cex = 0.5)
axisPhylo()

```
```{r host species tree pretty, fig.height= 8, fig.width = 8}
host_meta <- total_genome_metadata %>%
  dplyr::select(host_species, host_class) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    tip.label = gsub(" ", "_", host_species)
  ) %>%
  dplyr::select(tip.label, host_class)



p <- ggtree(host_tree) %<+% host_meta +
  geom_tiplab(
    aes(color = host_class),
    size = 2.3
  ) +
  xlim(0, max(nodeHeights(host_tree)) + 60)+
  scale_color_manual(
    values = host_class_colors,
    name = "Host class"
  ) +
  theme_tree2() +
  xlab("Million years ago (MYA)") +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  )

p

host_meta_2 <- total_genome_metadata %>%
  dplyr::select(host_species, host_type_2) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    tip.label = gsub(" ", "_", host_species)
  ) %>%
  dplyr::select(tip.label, host_type_2)


p_host_type <- ggtree(host_tree) %<+% host_meta_2 +
  geom_tiplab(
    aes(color = host_type_2),
    size = 2.3
  ) +
  xlim(0, max(nodeHeights(host_tree)) + 60)+
  scale_color_manual(
    values = host_type_2_colors,
    name = "Host Type"
  ) +
  theme_tree2() +
  xlab("Million years ago (MYA)") +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  )

p_host_type

```


## Data statistics

How many host species, order and classes in the dataset?
```{r counting species order and class_2}
total_genome_metadata %>% dplyr::count(host_species)

total_genome_metadata %>% dplyr::count(host_order)

total_genome_metadata %>% dplyr::count(host_class)
```
```{r plotting n mags per bacterial species split human and animal}
counts_mags <- total_genome_metadata %>%
  filter(!is.na(host_type_2))%>%
  group_by(bacterial_species, host_type_2)%>%
  summarize(total = n(),
            n_mags_domestic = sum(host_type_2 == "domestic"), 
            n_mags_human = sum(host_type_2 == "human"),
            n_mags_wild = sum(host_type_2 == "wild"),)

total_genome_metadata$host_type_2 <- factor(
  total_genome_metadata$host_type_2,
  levels = c("human", "domestic", "wild")
)


ggplot(counts_mags, aes(x = bacterial_species, y = total, fill = host_type_2))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = host_type_2_colors)+
  labs(
    title = "Number of MAGs per Bacterial Species (animal vs human hosts)",
    x = "Bacterial species",
    y = "N of MAGs"
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```
```{r total_counts_per_species}
counts_mags_total <- total_genome_metadata %>%
  group_by(bacterial_species)%>%
  summarize(total = n())


counts_mags_total
```



```{r plotting with tree}
counts_mags <- counts_mags %>%
  mutate(bacterial_species = bacterial_species %>%
           str_replace("^enterococcus_hirae$", "enterococcus_b_hirae"))



tree_data <- ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

counts_mags <- counts_mags %>%
  mutate(bacterial_species = factor(bacterial_species, levels = tree_order),
         host_type_2 = factor(host_type_2, levels = c("wild", "domestic", "human")))


p_tree <- ggtree(pruned) +
  theme_tree2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p_bar <- ggplot(counts_mags, aes(x = bacterial_species, y = total, fill = host_type_2))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = host_type_2_colors, name = "Host Type")+
  scale_y_continuous(breaks = seq(0, 500, by = 50))+
  labs(
    x = NULL,
    y = "Number of MAGs"
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
  coord_flip()

p_final <- (p_tree | p_bar) +
  plot_layout(widths = c(1, 2.3)) +
  plot_annotation(
    title = "Number of MAGs per Bacterial Species (colored by host type)"
  )


p_final
```


```{r plotting host order_2}
df_host_species <- total_genome_metadata %>%
  filter(!is.na(bacterial_species), !is.na(host_species)) %>%
  distinct(bacterial_species, host_species) %>%
  count(bacterial_species, name = "n_host_species") %>%
  mutate(bacterial_species = fct_reorder(bacterial_species, n_host_species))


df_host_species_wild <- total_genome_metadata %>%
  filter(!is.na(bacterial_species), !is.na(host_species), host_type_2 =="wild") %>%
  distinct(bacterial_species, host_species) %>%
  count(bacterial_species, name = "n_host_species") %>%
  mutate(bacterial_species = fct_reorder(bacterial_species, n_host_species))

df_host_species_domestic <- total_genome_metadata %>%
  filter(!is.na(bacterial_species), !is.na(host_species), host_type_2 =="domestic") %>%
  distinct(bacterial_species, host_species) %>%
  count(bacterial_species, name = "n_host_species") %>%
  mutate(bacterial_species = fct_reorder(bacterial_species, n_host_species))



ggplot(df_host_species,
       aes(x = bacterial_species, y = n_host_species)) +
  geom_col(fill = "#355C7D") +
  geom_text(aes(label = n_host_species), vjust = -0.3, size = 3) +
  labs(
    title = "Number of Host Species per Bacterial Species",
    x = "Bacterial species",
    y = "Distinct host species"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))



df_order_counts <- total_genome_metadata %>%
  filter(!is.na(host_order)) %>%
  count(host_order, name = "n_mags") %>%
  mutate(host_order = fct_reorder(host_order, n_mags))

ggplot(df_order_counts,
       aes(x = host_order, y = n_mags)) +
  geom_col() +
  scale_fill_manual(values = host_order_colors)+
  geom_text(aes(label = n_mags), vjust = -0.3, size = 3) +
  labs(
    title = "Number of MAGs per Host Order",
    x = "Host order",
    y = "MAG count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

```
```{r average n host species}
df_host_species %>% 
  summarise(
    mean_n = mean(n_host_species, na.rm = TRUE) %>% round(2),
    sd_n = sd(n_host_species, na.rm = TRUE) %>% round(2),
  ) %>%
  unite("Mean number of host species per bacteria", mean_n, sd_n, sep = " ± ") %>%
  tt()


df_host_species_wild %>% 
  summarise(
    mean_n = mean(n_host_species, na.rm = TRUE) %>% round(2),
    sd_n = sd(n_host_species, na.rm = TRUE) %>% round(2),
  ) %>%
  unite("Mean number of host species per bacteria", mean_n, sd_n, sep = " ± ") %>%
  tt()

df_host_species_domestic %>% 
  summarise(
    mean_n = mean(n_host_species, na.rm = TRUE) %>% round(2),
    sd_n = sd(n_host_species, na.rm = TRUE) %>% round(2),
  ) %>%
  unite("Mean number of host species per bacteria", mean_n, sd_n, sep = " ± ") %>%
  tt()
```


```{r how many host species total}
total_genome_metadata %>% filter(!is.na(host_species))%>%
  pull(host_species) %>% unique() 

total_genome_metadata %>% filter(!is.na(host_class))%>%
  pull(host_class) %>% unique() 
```



```{r map_2}
library(rnaturalearth)
library(rnaturalearthdata)
library(sf) 


world <- ne_countries(scale = "medium", returnclass = "sf")

mag_locations <- total_genome_metadata %>%
  filter(!is.na(country_normalized)) %>%
  left_join(
    world %>%
      dplyr::select(name, geometry) %>%
      st_centroid() %>%
      st_coordinates() %>%
      as_tibble() %>%
      bind_cols(world %>% dplyr::select(name)) ,
    by = c("country" = "name")
  )

mag_locations%>% dplyr::select(ID, host_type_2, country_normalized)

ggplot(world) +
  geom_sf(fill = "gray95", color = "gray80") +
  geom_point(
    data = mag_locations,
    aes(x = X, y = Y, color = host_type_2),
    size = 2.5,
    alpha = 0.5
  ) +
  scale_color_manual(values = host_type_2_colors) +
  theme_minimal() +
  labs(color = "Host Type")

```


### Mean completeness and contamination
```{r mean completeness and contamination final_2}
total_genome_metadata %>% 
  summarise(
    mean_c = mean(completeness, na.rm = TRUE) %>% round(2),
    sd_c = sd(completeness, na.rm = TRUE) %>% round(2),
    mean_con = mean(contamination, na.rm = TRUE) %>% round(2),
    sd_con = sd(contamination, na.rm = TRUE) %>% round(2)
  ) %>%
  unite("Completeness", mean_c, sd_c, sep = " ± ") %>%
  unite("Contamination", mean_con, sd_con, sep = " ± ") %>%
  tt()
```



```{r plot completeness}
ggplot(total_genome_metadata, aes(x= bacterial_species, y = completeness))+
  geom_violin()
  #facet_grid(~bacterial_species)

ggplot(total_genome_metadata,
       aes(x = host_type_2, y = completeness, color = host_type_2)) +

  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.3,
    width = 0.55
  ) +

  geom_jitter(
    width = 0.18,
    alpha = 0.35,
    size = 1
  ) +

  scale_color_manual(values = host_type_2_colors) +

  labs(
    title = "MAG Completeness by Host Type",
    x = "Host Type",
    y = "Completeness (%)"
  ) +

  theme_minimal(base_size = 14) +
  theme(legend.position = "none")


ggplot(total_genome_metadata,
       aes(x = host_type_2, y = contamination, color = host_type_2)) +

  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.3,
    width = 0.55
  ) +

  geom_jitter(
    width = 0.18,
    alpha = 0.35,
    size = 1
  ) +

  scale_color_manual(values = host_type_2_colors) +

  labs(
    title = "MAG Contamination by Host Type",
    x = "Host Type",
    y = "Contamination (%)"
  ) +

  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```



```{r is completeness and contamination different between bacteria}
kruskal_test(total_genome_metadata,
                   completeness ~ bacterial_species)

pairwise.wilcox.test(total_genome_metadata$completeness, total_genome_metadata$bacterial_species, p.adjust.method = "fdr")

kruskal_test(total_genome_metadata,
                   contamination ~ bacterial_species)
```
```{r general completeness contaminaton }
#Is completeness and contamination different across host types?

kruskal.test(completeness ~ host_type_2, data = total_genome_metadata)
kruskal.test(contamination ~ host_type_2, data = total_genome_metadata)

#Between which host types?


pairwise.wilcox.test(
  total_genome_metadata$completeness,
  total_genome_metadata$host_type_2,
  p.adjust.method = "BH"
)

pairwise.wilcox.test(
  total_genome_metadata$contamination,
  total_genome_metadata$host_type_2,
  p.adjust.method = "BH"
)


#How much different?


summary_tbl <- total_genome_metadata %>%
  group_by(host_type_2) %>%
  summarise(
    n = n(),
    completeness_median = median(completeness, na.rm = TRUE),
    completeness_IQR = IQR(completeness, na.rm = TRUE),
    contamination_median = median(contamination, na.rm = TRUE),
    contamination_IQR = IQR(contamination, na.rm = TRUE)
  ) %>%
  arrange(host_type_2)

summary_tbl



```

```{r general completeness contaminaton human vs wild}
total_genome_metadata_2 <- total_genome_metadata %>% filter(host_type_2 %in% c("wild", "human"))

wilcox.test(contamination ~ host_type_2, data = total_genome_metadata_2)
wilcox.test(completeness   ~ host_type_2, data = total_genome_metadata_2)


summary_tbl <- total_genome_metadata_2 %>%
  group_by(host_type_2) %>%
  summarise(
    n = n(),
    completeness_median = median(completeness, na.rm = TRUE),
    completeness_IQR    = IQR(completeness, na.rm = TRUE),
    contamination_median = median(contamination, na.rm = TRUE),
    contamination_IQR    = IQR(contamination, na.rm = TRUE)
  ) %>%
  arrange(host_type_2)

summary_tbl

```



```{r violin plots bacteria completeness}
total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = completeness))+
  geom_violin()+ 
  geom_point()+
  labs(y = "Completeness", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))


total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = contamination))+
  geom_violin()+ 
  geom_point()+
  labs(y = "Contamination", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = completeness))+
  geom_violin()+ 
  geom_point()+
  facet_grid(~host_type)+
  labs(y = "Completeness", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

Testing the differences in completeness and contamination between host types taking the different counts of bacterial species into account:

Does host type explain genome quality differences after accounting for species-level differences?

```{r statistics testing host type compelteness}
# Completeness model
m_comp <- lmer(completeness ~ host_type_2 + (1 | bacterial_species),
               data = total_genome_metadata)

summary(m_comp)
anova(m_comp)

# Contamination model
m_cont <- lmer(contamination ~ host_type_2 + (1 | bacterial_species),
               data = total_genome_metadata)

summary(m_cont)
anova(m_cont)

```
```{r means completeness and contamination}
emmeans(m_comp, pairwise ~ host_type_2)
emmeans(m_cont, pairwise ~ host_type_2)
```


Prepare the order of the host for the plots (so that orders are grouped by class):

```{r order_preparation}

  tax_map <- tibble::tibble(
  host_species = c(
    # Humans / lab / livestock / pets
    "Homo sapiens",
    "Mus musculus",
    "Rattus norvegicus",
    "Sus scrofa",
    "Bos taurus",
    "Canis lupus familiaris",
    "Equus caballus",
    "Oryctolagus cuniculus",
    "Capra hircus",

    # Birds
    "Gallus gallus",
    "Cathartes aura",
    "Columba livia",
    "Spheniscus magellanicus",

    # Insects
    "Drosophila melanogaster",
    "Bombyx mori",

    # Fish
    "Danio rerio",
    "Oncorhynchus mykiss",
    "Oreochromis niloticus",

    # Marine mammals / pinnipeds
    "Balaenoptera acutorostrata",
    "Arctocephalus australis",

    # Extinct
    "Mammuthus primigenius",
    
    #More
    "Loxodonta africana",
    "Elephas maximus",
    "Felis catus",
    "Canis lupus",
    "Phascolarctos cinereus",
    
    "Grasshopper",
    "Termite",
    "Macaca fascicularis",
    "Rhinolophus capensis",
    "Fish",
    
    #Final (from total_genome_metadata)
    "Zophobas morio",
    "Mustela putorius",
    "Ovis aries",
    "Corallus caninus",
    "Meles meles",
    "Clanga pomarina",
    "Buteo buteo",
    "Accipiter gentilis",
    "Chelonia mydas",
    "Grampus griseus",
    "Sterna trudeaui",
    "Rhodnius prolixus",
    "Turdus merula",
    "Curruca curruca",
    "Columba palumbus",
    "Delichon urbicum",
    "Chloris chloris",
    "Hirundo rustica",
    "Ciconia ciconia",
    "Anas platyrhynchos",
    "Grus grus",
    "Apis mellifera",
    "Coragyps atratus",
    "Pseudoscops clamator",
    "Pandion haliaetus",
    "Sycanus bifidus",
    "Eocanthecona furcellata",
    "Arma custos",
    "Corvus frugilegus",
    "Milvus milvus",
    "Vulpes zerda"


  ),
  tax_host_order = c(
    # Humans / lab / livestock / pets
    "Primates",
    "Rodentia",
    "Rodentia",
    "Artiodactyla",
    "Artiodactyla",
    "Carnivora",
    "Perissodactyla",
    "Lagomorpha",
    "Artiodactyla",

    # Birds
    "Galliformes",
    "Accipitriformes",
    "Columbiformes",
    "Sphenisciformes",

    # Insects
    "Diptera",
    "Lepidoptera",

    # Fish
    "Cypriniformes",
    "Salmoniformes",
    "Cichliformes",

    # Marine mammals / pinnipeds
    "Artiodactyla",      # (Cetartiodactyla)
    "Carnivora",

    # Extinct
    "Proboscidea",
    
    #More
    "Proboscidea",
    "Proboscidea",
    "Carnivora",
    "Carnivora",
    "Diprotodontia",
    
    
    "Orthoptera",     # Grasshopper
    "Blattodea",      # Termite (termites now placed within Blattodea; Isoptera as infraorder)
    "Primates",       # Macaca fascicularis
    "Chiroptera",     # Rhinolophus capensis (horseshoe bat)
    NA_character_,    # Fish (no single order fits all fish)
    
    #Final (from total_genome_metadata)
    "Coleoptera",          # Zophobas morio
    "Carnivora",           # Mustela putorius
    "Artiodactyla",        # Ovis aries
    "Squamata",            # Corallus caninus
    "Carnivora",           # Meles meles
    "Accipitriformes",     # Clanga pomarina
    "Accipitriformes",     # Buteo buteo
    "Accipitriformes",     # Accipiter gentilis
    "Testudines",          # Chelonia mydas
    "Artiodactyla",        # Grampus griseus
    "Charadriiformes",     # Sterna trudeaui
    "Hemiptera",           # Rhodnius prolixus
    "Passeriformes",       # Turdus merula
    "Passeriformes",       # Curruca curruca
    "Columbiformes",       # Columba palumbus
    "Passeriformes",       # Delichon urbicum
    "Passeriformes",       # Chloris chloris
    "Passeriformes",       # Hirundo rustica
    "Ciconiiformes",       # Ciconia ciconia
    "Anseriformes",        # Anas platyrhynchos
    "Gruiformes",          # Grus grus
    "Hymenoptera",         # Apis mellifera
    "Cathartiformes",      # Coragyps atratus
    "Strigiformes",        # Pseudoscops clamator
    "Accipitriformes",     # Pandion haliaetus
    "Hemiptera",           # Sycanus bifidus
    "Hemiptera",           # Eocanthecona furcellata
    "Hemiptera",           # Arma custos
    "Passeriformes",       # Corvus frugilegus
    "Accipitriformes",     # Milvus milvus
    "Carnivora"            # Vulpes zerda


  ),
  tax_host_class = c(
    # Humans / lab / livestock / pets
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",

    # Birds
    "Aves",
    "Aves",
    "Aves",
    "Aves",

    # Insects
    "Insecta",
    "Insecta",

    # Fish
    "Actinopterygii",
    "Actinopterygii",
    "Actinopterygii",

    # Marine mammals / pinnipeds
    "Mammalia",
    "Mammalia",

    # Extinct and more
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    
        
    "Insecta",        # Grasshopper
    "Insecta",        # Termite
    "Mammalia",       # Macaca fascicularis
    "Mammalia",       # Rhinolophus capensis
    "Actinopterygii",  # Fish (bony fishes; safe default for generic "fish")

    #Final (from total_genome_metadata)
    "Insecta",        # Zophobas morio
    "Mammalia",       # Mustela putorius
    "Mammalia",       # Ovis aries
    "Reptilia",       # Corallus caninus
    "Mammalia",       # Meles meles
    "Aves",           # Clanga pomarina
    "Aves",           # Buteo buteo
    "Aves",           # Accipiter gentilis
    "Reptilia",       # Chelonia mydas
    "Mammalia",       # Grampus griseus
    "Aves",           # Sterna trudeaui
    "Insecta",        # Rhodnius prolixus
    "Aves",           # Turdus merula
    "Aves",           # Curruca curruca
    "Aves",           # Columba palumbus
    "Aves",           # Delichon urbicum
    "Aves",           # Chloris chloris
    "Aves",           # Hirundo rustica
    "Aves",           # Ciconia ciconia
    "Aves",           # Anas platyrhynchos
    "Aves",           # Grus grus
    "Insecta",        # Apis mellifera
    "Aves",           # Coragyps atratus
    "Aves",           # Pseudoscops clamator
    "Aves",           # Pandion haliaetus
    "Insecta",        # Sycanus bifidus
    "Insecta",        # Eocanthecona furcellata
    "Insecta",        # Arma custos
    "Aves",           # Corvus frugilegus
    "Aves",           # Milvus milvus
    "Mammalia"        # Vulpes zerda

  )
)

order_to_class <- tax_map %>%
  dplyr::select(host_order = tax_host_order,
                host_class = tax_host_class) %>%
  distinct()

legend_order <- order_to_class %>%
  dplyr::mutate(
    host_class = factor(host_class,
      levels = c("Mammalia", "Aves", "Reptilia",
                 "Actinopterygii", "Insecta", "Amphibia", NA)
    )
  ) %>%
  dplyr::arrange(host_class) %>%
  dplyr::pull(host_order)

```




```{r genome_quality_plot_3, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Generate quality biplot
genome_biplot <- total_genome_metadata %>%
  dplyr::select(c(ID,completeness,contamination, host_type_2)) %>%
  ggplot(aes(x=completeness, y=contamination, color = host_type_2)) +
  scale_color_manual(values = host_type_2_colors, name = "Host Type") +
  geom_point(alpha=0.7, size = 4) +
  xlim(c(90,100)) +
  ylim(c(2.5,0)) +
  labs(y= "Contamination", x = "Completeness") +
  theme_classic() +
  theme(
    legend.position = "left",
    axis.text.y = element_text(size=12),
    axis.text.x = element_text(size=12),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold")
  )

#Generate contamination boxplot
genome_contamination <- total_genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(2.5,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_classic() +
                    theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- total_genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(90,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_classic() +
                theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
#pdf("figures/completeness_contamination.pdf",width=10, height=5)
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))
#dev.off()
```

```{r genome_quality_plot_colored_by_bact, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Generate quality biplot
genome_biplot <- total_genome_metadata %>%
  dplyr::select(c(ID,completeness,contamination, bacterial_species)) %>%
  ggplot(aes(x=completeness, y=contamination, color = bacterial_species)) +
  geom_point(alpha=0.7, size = 4) +
  xlim(c(90,100)) +
  ylim(c(2.5,0)) +
  labs(y= "Contamination", x = "Completeness") +
  theme_classic() +
  theme(
    legend.position = "left",
    axis.text.y = element_text(size=12),
    axis.text.x = element_text(size=12),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold")
  )

#Generate contamination boxplot
genome_contamination <- total_genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(2.5,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_classic() +
                    theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- total_genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(90,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_classic() +
                theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
#pdf("figures/completeness_contamination.pdf",width=10, height=5)
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))
#dev.off()
```


```{r plotting completeness means}
# Estimated marginal means 
emm_comp <- emmeans(m_comp, "host_type_2")

df_comp <- as.data.frame(emm_comp)

comp_plot <- ggplot(df_comp, aes(x = host_type_2, y = emmean, color = host_type_2)) +
  geom_point(size = 4) +
  scale_color_manual(values = host_type_2_colors, name = "Host Type") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, size = 1) +
  labs(
    title = "Species-adjusted MAG Completeness by Host Type",
    x = "Host Type",
    y = "Adjusted Completeness (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")


emm_cont <- emmeans(m_cont, "host_type_2")
df_cont <- as.data.frame(emm_cont)

cont_plot <- ggplot(df_cont, aes(x = host_type_2, y = emmean, color = host_type_2)) +
  geom_point(size = 4) +
  scale_color_manual(values = host_type_2_colors, name = "Host Type") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.15, size = 1) +
  labs(
    title = "Species-adjusted MAG Contamination by Host Type",
    x = "Host Type",
    y = "Adjusted Contamination (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")


comp_plot
cont_plot


library(patchwork)

combined_plot <- comp_plot + cont_plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Species-adjusted MAG Quality by Host Type"
  )

combined_plot

```


```{r countries}
total_genome_metadata <- total_genome_metadata %>%
  mutate(
    country_normalized = case_when(
      is.na(country_normalized) ~ "None",
      tolower(country_normalized) %in% c("none", "na", "n/a", "Not Provided", 
                                         "Not Collected", "Not Applicable") ~ "None",
      TRUE ~ country_normalized
    )
  )

country_counts <- total_genome_metadata %>%
  count(country_normalized)

small_countries <- country_counts %>%
  filter(n < 20) %>%
  pull(country_normalized)

total_genome_metadata <- total_genome_metadata %>%
  mutate(country_grouped = ifelse(country_normalized %in% small_countries,
                                  "Other",
                                  country_normalized))

country_order <- total_genome_metadata %>%
  filter(country_grouped != "Other") %>%
  distinct(country_grouped, continent) %>%
  arrange(continent, country_grouped) %>%
  pull(country_grouped)

country_order <- unique(country_order)

# Add "Other" as the last level
country_order <- c(country_order, "Other")

total_genome_metadata <- total_genome_metadata %>%
  mutate(country_grouped = factor(country_grouped, levels = country_order))

country_palette <- country_palette[c(country_order)]

ggplot(total_genome_metadata, aes(x = bacterial_species, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(total_genome_metadata, aes(x = bacterial_species, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  facet_grid(~host_type)+
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



ggplot(total_genome_metadata, aes(x = bacterial_species, fill = continent)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = continent_palette) +
  facet_grid(~host_type)+
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r countryplots}
#order
# total_genome_metadata$country_grouped <- factor(
#   total_genome_metadata$country_grouped,
#   levels = names(country_palette))

#order also host type 2
total_genome_metadata$host_type_2 <- factor(
  total_genome_metadata$host_type_2,
  levels = c("human", "domestic", "wild")
)

total_genome_metadata%>%
  filter(!is.na(host_type_2))%>%
ggplot( aes(x = host_type_2, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Host type",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


total_genome_metadata%>%
  filter(!is.na(host_type_2))%>%
ggplot( aes(x = host_type_2, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Host type",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
```{r country_counts}
ggplot(country_counts, aes(x= country_normalized, y = n))+
  geom_col()+
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```




```{r wilcoxon tests for host types}


results_tbl <- map_dfr(species_list, function(sp) {
  
  genome_metadata <- species_data[[sp]]$metadata
  
  # Variables to test
  vars_to_test <- c("contamination", "completeness", "genome_size")
  
  map_dfr(vars_to_test, function(v) {
    
    # Build formula: variable ~ host_type
    fml <- as.formula(paste0(v, " ~ host_type"))
    
    # Run Wilcoxon
    test_res <- wilcox.test(fml, data = genome_metadata)
    
    # Tidy results + add species & variable info
    tidy(test_res) %>%
      mutate(
        species = sp,
        variable = v
      )
  })
})

```
```{r stats}
results_tbl %>% dplyr::select(p.value, species, variable) %>% filter(p.value<0.05)
```


## Genomic similarity (drep)
```{r example_drep_final, fig.height = 10, fig_width = 8}
for (sp in species_list) {

  # Run drep tree construction for THIS species
  drep_results <- drep_tree_metadata(
    genome_metadata = species_data[[sp]]$metadata,
    ndb = species_data[[sp]]$ndb
  )

  p_host_order <- plot_tree_basic(
    tree       = drep_results$tree,
    tip_df     = drep_results$tip_df,
    color_by   = "host_order",
    palette = host_order_colors,
    label_tips = TRUE,
    show_threshold = 99.5
  )
  
  p_host_type_2 <- plot_tree_basic(
    tree       = drep_results$tree,
    tip_df     = drep_results$tip_df,
    color_by   = "host_type_2",
    palette = host_type_2_colors,
    label_tips = TRUE,
    show_threshold = 99.5
  )
  
#   p_country <- plot_tree_basic(
#   tree       = drep_results$tree,
#   tip_df     = drep_results$tip_df,
#   color_by   = "country_normalized",
#   palette    = country_palette
# )

  #print(p_host_order)  
  print(p_host_type_2) 
  #print(p_country) 

}
```


```{r drep stats}
# for (sp in species_list) {
#     # Stats
#   res <- drep_stat_analysis(
#   tip_df = drep_results$tip_df,
#   ani_matrix = species_data[[sp]]$ndb,
#   host_type_colors = host_type_colors,
#   host_order_colors = host_order_colors
# )
#   
#   # Print stats
#   print(res$permanova$adonis)
#   print(res$permanova$dispersion)
#   print(res$pairwise$wilcox_within$test)
#   print(res$pairwise$wilcox_cross_vs_within$test)
#   
#   # Plot
#   print(res$pcoa$by_host_type)
#   print(res$pcoa$by_host_order)
#   print(res$pairwise$boxplot)
#  
# 
# }
```


## Pangenome analysis
```{r load partitions}
partitions <- read_csv("data/pangenome.csv")
```

```{r order}
partitions$partition <- factor(
  partitions$partition,
  levels = c("Cloud", "Shell", "Persistent")
)

tree_data <- ggtree::ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

partitions <- partitions %>%
  mutate(species = species %>%
           str_replace("^enterococcus_hirae$", "enterococcus_b_hirae"))

partitions <- partitions %>%
  mutate(
    species = factor(species, levels = tree_order)
  )

```


```{r plot_pangenome}
ggplot(partitions,
       aes(x = species, y = families, fill = partition)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(
    values = c(
      Persistent = "#F7A507",
      Shell = "#01D960",
      Cloud = "#78DFFF"
    )
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Species",
    y = "Number of gene families",
    fill = "Pangenome partition"
  )

```


```{r with tree}
p_tree <- ggtree(pruned) +
  theme_tree2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p_bar_pangenome <- ggplot(partitions, aes(x = species, y = families, fill = partition))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c(
      Persistent = "#F7A507",
      Shell = "#01D960",
      Cloud = "#78DFFF"
    ))+
  scale_y_continuous(breaks = seq(0, 50000, by = 5000))+
  labs(
    x = NULL,
    y = "Number of gene families"
  ) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))+
  coord_flip()

p_final <- (p_tree | p_bar_pangenome) +
  plot_layout(widths = c(1, 2.3)) +
  plot_annotation(
    title = "Pangenome partitions per species"
  )

p_final
```

```{r proportions}
p_bar_pangenome <- ggplot(
  partitions,
  aes(x = species, y = families, fill = partition)
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c(
    Persistent = "#F7A507",
    Shell = "#01D960",
    Cloud = "#78DFFF"
  )) +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    x = NULL,
    y = "Proportion of gene families"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  coord_flip()

p_final <- (p_tree | p_bar_pangenome) +
  plot_layout(widths = c(1, 2.3)) +
  plot_annotation(
    title = "Pangenome partitions per species"
  )

p_final
```


```{r save data, eval = FALSE}
save(total_genome_metadata,
     species_list,
     species_data,
     pruned, # bacterial species tree
     host_tree,
     file = "data/intermediate_analysis/data.Rdata")
```


```{r save separately, eval = FALSE}
save(pruned, file = "data/pruned_tree.RData")
save(host_tree, file = "data/host_tree.RData")
```















