# Functional analysis

```{r load data funct}
load("data/intermediate_analysis/data.Rdata")


analysis_results <- readRDS("data/intermediate_analysis/functional_analysis_results.rds")

permanova_all      <- analysis_results$permanova
betadisper_all     <- analysis_results$betadisper
kegg_fisher_total  <- analysis_results$fisher$kegg
amr_fisher_total   <- analysis_results$fisher$amr
vf_fisher_total    <- analysis_results$fisher$vf
cazy_fisher_total  <- analysis_results$fisher$cazy



functional_results <- readRDS( "data/intermediate_analysis/fisher_results.rds")



kegg_annot_df <- functional_results$kegg
amr_df        <- functional_results$amr
vf_df_annot   <- functional_results$vf
cazy_annot    <- functional_results$cazy
```



```{r total_genome_annotations}
total_genome_annotations <- purrr::map_df(
  names(species_data),
  ~ species_data[[.x]]$annotations %>%
      dplyr::select(-start, -end) %>%   # remove problematic columns
      dplyr::mutate(bacterial_species = .x)
)
```

```{r vf_data}
vf_data <- total_genome_annotations %>% dplyr::select(vf, vf_type) %>% unique()
```




```{r testing kegg count between hosts}
#count KEGG annotations per MAG within each species
kegg_counts_meta <- total_genome_annotations %>%
  dplyr::filter(!is.na(kegg))%>%
  dplyr::left_join(
    total_genome_metadata %>%
      dplyr::select(ID, host_type),         
    join_by(genome == ID)
  ) %>%
  dplyr::group_by(bacterial_species, genome, host_type)%>%
  dplyr::summarise(n_kegg = dplyr::n(), .groups = "drop") %>%
  filter(!is.na(host_type))

# Run test per species
kegg_host_tests <- kegg_counts_meta %>%
  split(.$bacterial_species) %>%
  map_df(function(df) {
    if (length(unique(df$host_type)) < 2) {
      return(tibble(
        bacterial_species = unique(df$bacterial_species),
        p_value = NA,
        test = "Not enough groups"
      ))
    }
    wt <- wilcox.test(n_kegg ~ host_type, data = df)

    tibble(
      species = unique(df$bacterial_species),
      p_value = wt$p.value,
      median_human  = median(df$n_kegg[df$host_type == "human"]),
      median_animal = median(df$n_kegg[df$host_type == "animal"]),
      test = "Wilcoxon rank-sum"
    )
  })


kegg_host_tests

```


```{r kegg_hosts_p }
kegg_host_tests %>% filter(p_value < 0.05)
```




### PERMANOVA
```{r functional analysis permanova pcoa}
## ============================================================
## Host contrasts (pairwise PERMANOVA)
## ============================================================

host_contrasts <- list(
  human_vs_wild     = c("human", "wild"),
  wild_vs_domestic  = c("domestic", "wild"),
  human_vs_domestic = c("human", "domestic")
)

## ===========================================================
## Helpers to tidy PERMANOVA / betadisper
## ===========================================================

.extract_permanova_rows <- function(permanova_obj, species, matrix_name) {
  if (is.null(permanova_obj)) return(tibble::tibble())
  df <- as.data.frame(permanova_obj)
  df$term <- rownames(df)
  rownames(df) <- NULL
  df$species <- species
  df$matrix_name <- matrix_name
  tibble::as_tibble(df)
}

.extract_betadisper_rows <- function(betadisper_obj, species, matrix_name) {
  if (is.null(betadisper_obj)) return(tibble::tibble())

  an <- as.data.frame(anova(betadisper_obj))
  an$term <- rownames(an)
  rownames(an) <- NULL

  an$species     <- species
  an$matrix_name <- matrix_name
  an$type        <- "anova"

  tibble::as_tibble(an)
}

## ============================================================
## Accumulators
## ============================================================

permanova_rows  <- list()
betadisper_rows <- list()

## ============================================================
## MAIN LOOP
## ============================================================

for (sp in species_list) {

  message("Processing species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  ## ------------------------------------------------------------
  ## Build functional matrices
  ## ------------------------------------------------------------

  mats <- list(
    KEGG = make_annotation_matrix(genome_annotations, "kegg"),
    AMR  = make_annotation_matrix(genome_annotations, "resistance_target"),
    VF   = make_annotation_matrix(genome_annotations, "vf"),
    CAZY = make_annotation_matrix(genome_annotations, "cazy")
  )

  ## ============================================================
  ## PERMANOVA: pairwise host contrasts
  ## ============================================================

  for (contrast_name in names(host_contrasts)) {

    host_levels <- host_contrasts[[contrast_name]]

    for (nm in names(mats)) {

      obj <- tryCatch(
        pa_permanova(
          mats[[nm]],
          genome_metadata,
          host_levels = host_levels
        ),
        error = function(e) {
          message("[", sp, " | ", nm, " | ", contrast_name, "] failed: ", e$message)
          NULL
        }
      )

      if (is.null(obj)) next

      ## ---- PERMANOVA ------------------------------------------
      pr <- .extract_permanova_rows(obj$permanova, sp, nm)
      if (nrow(pr) > 0) {
        pr$contrast <- contrast_name
        permanova_rows[[length(permanova_rows) + 1]] <- pr
      }

      ## ---- Betadisper -----------------------------------------
      br <- .extract_betadisper_rows(obj$betadisper_raw, sp, nm)
      if (nrow(br) > 0) {
        br$contrast <- contrast_name
        betadisper_rows[[length(betadisper_rows) + 1]] <- br
      }
    }
  }

 
}

## ============================================================
## Bind results
## ============================================================

permanova_all  <- dplyr::bind_rows(permanova_rows)
betadisper_all <- dplyr::bind_rows(betadisper_rows)

## Quick check
print(permanova_all, n = 20)
print(betadisper_all, n = 20)


```

```{r pcoas, eval = FALSE}
 ## ========================================
  ## PCoA: One PCoA per matrix (all hosts)
  ## =======================================

  for (nm in names(mats)) {

    obj_all <- tryCatch(
      pa_permanova(
        mats[[nm]],
        genome_metadata,
        host_levels = NULL   # <- ALL hosts
      ),
      error = function(e) NULL
    )

    if (is.null(obj_all)) next

    dist <- obj_all$dist
    meta <- obj_all$metadata

    ## ---- PCoA ------------------------------------------------
    pcoa_res <- ape::pcoa(dist, correction = "cailliez")

    # Number of available axes
    n_axes <- ncol(pcoa_res$vectors)
    
    message("[", sp, " | ", nm, "] PCoA axes available: ", n_axes)
    
    if (n_axes < 2) {
      message("[", sp, " | ", nm, "] Only one PCoA axis; skipping ordination plot")
      next
    }
    
    # If we are here, PC1 and PC2 DEFINITELY exist
    var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
    
    pcoa_df <- data.frame(
      PC1 = pcoa_res$vectors[, 1],
      PC2 = pcoa_res$vectors[, 2],
      host_type_2 = meta$host_type_2,
      stringsAsFactors = FALSE
    )
    
    p <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_2)) +
      geom_point(size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(
        title = paste("PCoA of", nm, "matrix –", sp),
        x = paste0("PC1 (", var_exp[1], "%)"),
        y = paste0("PC2 (", var_exp[2], "%)")
      )
    
    print(p)
  }
```



```{r allpermanova_host_type}
permaniva_sig <- permanova_all %>%filter(term == "3") %>% dplyr::select(-Df, -SumOfSqs)

permaniva_sig
```

```{r tree_order_assigning}
tree_data <- ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

```



```{r figure_permanova, fig.height = 4, fig.width = 10}


permaniva_sig_hw <- permaniva_sig %>%
  mutate(species = species %>%
           str_replace("^enterococcus_hirae$", "enterococcus_b_hirae"))

permaniva_sig_hw <- permaniva_sig_hw %>%
  mutate(
    contrast = factor(
      contrast,
      levels = c("human_vs_wild", "wild_vs_domestic", "human_vs_domestic"),
      labels = c("H vs W", "W vs D", "H vs D")
    ),
    matrix_name = factor(
      matrix_name,
      levels = c("KEGG", "AMR", "VF", "CAZY")
    )
  )



tree_data <- ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

permaniva_sig_hw <- permaniva_sig_hw %>%
  mutate(species = factor(species, levels = tree_order))


p_tree <- ggtree(pruned) +
  theme_tree2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p_dot_hw <- permaniva_sig_hw %>%
  mutate(p_sig = ifelse(`Pr(>F)` <= 0.05, `Pr(>F)`, NA_real_)) %>%
  ggplot(aes(
    x = matrix_name,
    y = species,
    color = p_sig,
    size  = R2
  )) +

  geom_point() +

  facet_grid(
    . ~ contrast,
    scales = "free_x",
    space  = "free_x"
  ) +

  scale_color_viridis_c(
    limits = c(0, 0.05),
    option = "plasma",
    na.value = "grey70"
  ) +

  scale_size(
    limits = c(0, 0.20),
    range = c(1.5, 6)
  ) +

  theme_minimal(base_size = 13) +
  labs(
    x = NULL,
    y = NULL,
    size  = expression(R^2),
    color = "Pr(>F) (≤ 0.05 shown)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_blank(),
    panel.spacing.x = unit(1, "lines"),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 11, face = "bold"),
    plot.margin = margin(l = 0)
  )


p_final_hw <- (p_tree | p_dot_hw) +
  plot_annotation(
    title = "PERMANOVA Host type effects"
  )


p_final_hw
```





```{r kegg count analysis}
kegg_per_genome <- total_genome_annotations %>%
  filter(!is.na(kegg)) %>%
  group_by(genome) %>%
  summarise(kegg_count = n(), .groups = "drop")

kegg_summary <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_kegg = mean(kegg_count, na.rm = TRUE),
    median_kegg = median(kegg_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

kegg_test <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

ggplot(kegg_test, aes(x = host_type_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

ggplot(kegg_test, aes(x = host_type_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = bacterial_species), width = 0.2, alpha = 0.6) 

kruskal.test(kegg_count ~ host_type_2, data = kegg_test)

pairwise.wilcox.test(
  kegg_test$kegg_count,
  kegg_test$host_type_2,
  p.adjust.method = "BH"
)

```



```{r amr count analysis}
amr_per_genome <- total_genome_annotations %>%
  filter(resistance_type == "AMR",
    !is.na(resistance_target)) %>%
  group_by(genome) %>%
  summarise(amr_count = n(), .groups = "drop")

amr_summary <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_amr = mean(amr_count, na.rm = TRUE),
    median_amr = median(amr_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

amr_test <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(amr_test, aes(x = host_type_2, y = amr_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

kruskal.test(amr_count ~ host_type_2, data = amr_test)

pairwise.wilcox.test(
  amr_test$amr_count,
  amr_test$host_type_2,
  p.adjust.method = "BH"
)

```



```{r vf_count_analysis}

vf_per_genome <- total_genome_annotations %>%
  filter(!is.na(vf)) %>%
  group_by(genome) %>%
  summarise(vf_count = n(), .groups = "drop")

vf_summary <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_kegg = mean(vf_count, na.rm = TRUE),
    median_kegg = median(vf_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

vf_test <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(vf_test, aes(x = host_type_2, y = vf_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()


kruskal.test(vf_count ~ host_type_2, data = vf_test)

pairwise.wilcox.test(
  vf_test$vf_count,
  vf_test$host_type_2,
  p.adjust.method = "BH"
)
```




```{r amr different per group}
amr_presence <- total_genome_annotations %>%
  filter(resistance_type == "AMR", !is.na(resistance_target)) %>%
  distinct(genome, resistance_target) %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

amr_sets <- amr_presence %>%
  group_by(host_type_2) %>%
  summarise(amr_list = list(unique(resistance_target)))

amr_freq <- amr_presence %>%
  group_by(host_type_2, resistance_target) %>%
  summarise(n = n(), .groups = "drop")

chisq.test(table(amr_presence$resistance_target, amr_presence$host_type_2))

human_amr     <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "human")]]
domestic_amr  <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "domestic")]]
wild_amr      <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "wild")]]

Reduce(intersect, list(human_amr, domestic_amr, wild_amr))

human_only <- setdiff(human_amr, union(domestic_amr, wild_amr))
domestic_only <- setdiff(domestic_amr, union(human_amr, wild_amr))
wild_only <- setdiff(wild_amr, union(human_amr, domestic_amr))

human_only
domestic_only
wild_only

amr_binary <- amr_presence %>%
  distinct(host_type_2, resistance_target) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = host_type_2,
    values_from = present,
    values_fill = 0
  )

UpSetR::upset(
  as.data.frame(amr_binary[, -1]),  # remove resistance_target column
  sets = c("human", "domestic", "wild"),
  order.by = "freq"
)



```





### Fisher
```{r functional analysis fisher}
# Prepare empty containers
kegg_fisher_all <- list()
amr_fisher_all  <- list()
vf_fisher_all   <- list()
cazy_fisher_all <- list()

# Volcano plots (nested by contrast)
volcano_plots <- list(
  kegg = list(),
  amr  = list(),
  vf   = list(),
  cazy = list()
)

ann_list <- vector("list", length(species_list))
names(ann_list) <- species_list


for (sp in species_list) {

  message("Running Fisher tests for species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  ann_list[[sp]] <- dplyr::mutate(genome_annotations, species = sp)

  # Build matrices
  mats <- list(
    kegg = make_annotation_matrix(genome_annotations, "kegg"),
    amr  = make_annotation_matrix(genome_annotations, "resistance_target"),
    vf   = make_annotation_matrix(genome_annotations, "vf"),
    cazy = make_annotation_matrix(genome_annotations, "cazy")
  )

  for (contrast_name in names(host_contrasts)) {

    host_levels <- host_contrasts[[contrast_name]]

    # Subset metadata for the contrast
    meta_sub <- genome_metadata %>%
      dplyr::filter(host_type_2 %in% host_levels) %>%
      dplyr::mutate(host_type_2 = droplevels(factor(host_type_2)))

    # Fisher requires exactly 2 groups
    if (nlevels(meta_sub$host_type_2) != 2) {
      message("[", sp, " | ", contrast_name, "] skipped: not exactly 2 host groups")
      next
    }

    for (nm in names(mats)) {

      fisher_res <- tryCatch(
        run_fisher_tests_general(
          mat = mats[[nm]],
          metadata = meta_sub,
          group_col = "host_type_2",
          annotation_name = nm
        ),
        error = function(e) {
          message("[", sp, " | ", nm, " | ", contrast_name, "] Fisher failed: ", e$message)
          NULL
        }
      )

      if (is.null(fisher_res) || nrow(fisher_res) == 0) next

      # Add identifiers
      fisher_res$species  <- sp
      fisher_res$contrast <- contrast_name

      # Accumulate
      if (nm == "kegg") kegg_fisher_all[[length(kegg_fisher_all) + 1]] <- fisher_res
      if (nm == "amr")  amr_fisher_all[[length(amr_fisher_all) + 1]]  <- fisher_res
      if (nm == "vf")   vf_fisher_all[[length(vf_fisher_all) + 1]]   <- fisher_res
      if (nm == "cazy") cazy_fisher_all[[length(cazy_fisher_all) + 1]] <- fisher_res

      # Volcano plot (optional)
      volcano_plots[[nm]][[paste(sp, contrast_name, sep = "_")]] <-
        plot_volcano(fisher_res, nm)
    }
  }
}



```





```{r bind the rows}
kegg_fisher_total        <- dplyr::bind_rows(kegg_fisher_all)
amr_fisher_total         <- dplyr::bind_rows(amr_fisher_all)
vf_fisher_total          <- dplyr::bind_rows(vf_fisher_all)
cazy_fisher_total        <- dplyr::bind_rows(cazy_fisher_all)

```



```{r save functional_results}
# analysis_results <- list(
#   permanova      = permanova_all,
#   betadisper     = betadisper_all,
#   fisher = list(
#     kegg = kegg_fisher_total,
#     amr  = amr_fisher_total,
#     vf   = vf_fisher_total,
#     cazy = cazy_fisher_total
#   ),
#   host_contrasts = host_contrasts
# )
# 
# saveRDS(
#   analysis_results,
#   file = "data/intermediate_analysis/functional_analysis_results.rds"
# )
# 



# analysis_results <- readRDS("data/intermediate_analysis/functional_analysis_results.rds")
# 
# permanova_all      <- analysis_results$permanova
# betadisper_all     <- analysis_results$betadisper
# kegg_fisher_total  <- analysis_results$fisher$kegg
# amr_fisher_total   <- analysis_results$fisher$amr
# vf_fisher_total    <- analysis_results$fisher$vf
# cazy_fisher_total  <- analysis_results$fisher$cazy

```




```{r testing keggs totals, fig.height = 10, fig.width = 20}
prepare_volcano_df <- function(df, tree_order) {
  df %>%
    mutate(
      ##Fix enterococcus hirae name mismatch
      species = dplyr::case_when(
        species == "enterococcus_hirae" ~ "enterococcus_b_hirae",
        TRUE ~ species
      ),

      ## VOLCANO METRICS
      neg_log10_padj = -log10(pmax(p_adj, 1e-300)),
      sig = p_adj <= 0.05,

      ## order factors
      species = factor(species, levels = tree_order),
      contrast = factor(
        contrast,
        levels = c(
          "human_vs_wild",
          "wild_vs_domestic",
          "human_vs_domestic"
        ),
        labels = c(
          "HumanvsWild",
          "WildvsDomestic",
          "HumanvsDomestic"
        )
      )
    )
}



plot_volcano_contrasts <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = sig, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = c(
        `TRUE`  = "#0072B2",  # significant
        `FALSE` = "grey75"    # not significant
      ),
      name = "Adj. p ≤ 0.05"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.4),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log[10]("adj. p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}


get_top_hits <- function(df, feature_col, n = 5) {
  df %>%
    filter(sig) %>%                       # only significant features
    group_by(species, contrast) %>%       # per species × contrast
    arrange(p_adj) %>%                   
    slice_head(n = n) %>%                
    ungroup()
}
    
    
    
kegg_df <- prepare_volcano_df(
  kegg_fisher_total,
  tree_order = tree_order
)

kegg_top <- get_top_hits(kegg_df, n = 5)

plot_volcano_contrasts(
  kegg_df,
  title_prefix = "KEGG enrichment"
) +
  ggrepel::geom_text_repel(
    data = kegg_top,
    aes(label = kegg),
    size = 3,
    max.overlaps = 8
  )
```


```{r fisher bacteroides uniformis, fig.height = 8, fig.width = 6}
species_of_interest <- "bacteroides_uniformis"

kegg_df_sp <- kegg_df %>%
  filter(species == species_of_interest)

kegg_top_sp <- get_top_hits(kegg_df_sp, "kegg")

plot_volcano_contrasts(
  kegg_df_sp,
  title_prefix = "KEGG enrichment – Bacteroides uniformis"
) +
  ggrepel::geom_text_repel(
    data = kegg_top_sp,
    aes(label = kegg),
    size = 3
  )
```


```{r amr fisher,  fig.height = 10, fig.width = 15}
amr_df <- prepare_volcano_df(
  amr_fisher_total,
  tree_order = tree_order
)

amr_top <- get_top_hits(amr_df, n = 5)

plot_volcano_contrasts(
  amr_df,
  title_prefix = "AMR enrichment"
) +
  ggrepel::geom_text_repel(
    data = amr_top,
    aes(label = amr),
    size = 3
  )
```



```{r vf fisher,  fig.height = 10, fig.width = 15}
vf_df <- prepare_volcano_df(
  vf_fisher_total,
  tree_order = tree_order
)

vf_top <- get_top_hits(vf_df, n = 5)

plot_volcano_contrasts(
  vf_df,
  title_prefix = "VF enrichment"
) +
  ggrepel::geom_text_repel(
    data = vf_top,
    aes(label = vf),
    size = 3
  )
```


```{r cazy fisher,  fig.height = 10, fig.width = 15}
cazy_df <- prepare_volcano_df(
  cazy_fisher_total,
  tree_order = tree_order
)

cazy_top <- get_top_hits(cazy_df, n = 5)

plot_volcano_contrasts(
  cazy_df,
  title_prefix = "CAZy enrichment"
) +
  ggrepel::geom_text_repel(
    data = cazy_top,
    aes(label = cazy),
    size = 3
  )
```

```{r volcano by pathways}
#keep only significant KOs 
sig_df <- kegg_df %>%
  filter(p_adj < 0.05) %>%
  mutate(ko = paste0("ko:", kegg))   # (when you download the table they are like this)

# KO -> pathway links (only for sig KOs)
ko_links <- KEGGREST::keggLink("pathway", "ko") %>%
  tibble::enframe(name = "ko", value = "pathway") %>%
  semi_join(sig_df, by = "ko") %>%
  mutate(
    pathway_map = sub("^path:(?:ko|map)(\\d{5})$", "path:map\\1", pathway)
  ) %>%
  filter(str_detect(pathway_map, "^path:map\\d{5}$")) %>%
  distinct(ko, pathway_map)

#  One pathway per KO 
ko_links_one <- ko_links %>%
  group_by(ko) %>%
  slice(1) %>%
  ungroup()

# Fetch pathway info 
paths <- unique(ko_links_one$pathway_map)

get_info <- function(pid){
  x <- try(KEGGREST::keggGet(pid)[[1]], silent = TRUE)
  if (inherits(x, "try-error")) return(tibble(pathway_map=pid, pathway_name=NA, class=NA))
  tibble(
    pathway_map = pid,
    pathway_name = x$NAME[[1]],
    class = ifelse(is.null(x$CLASS), NA, x$CLASS[[1]])
  )
}

path_info <- bind_rows(lapply(paths, get_info))

# high-level category
path_info <- path_info %>%
  mutate(category = sub(";.*$", "", class))



```

```{r pathway kegg mapping}
#join annotations to kegg fisher results
ko_links_one_clean <- ko_links_one %>%
  mutate(ko_clean = sub("^ko:", "", ko))

kegg_annot_df <- kegg_df %>%
  left_join(ko_links_one_clean, by = c("kegg" = "ko_clean")) %>%
  left_join(path_info, by = "pathway_map")

```



```{r volcano plot colorful, fig.height=6, fig.width = 15, eval = FALSE}

kegg_categories <- sort(unique(na.omit(kegg_annot_df$category)))

kegg_colors <- c(
  NS = "grey75",
  setNames(
    RColorBrewer::brewer.pal(length(kegg_categories), "Set2"),
    kegg_categories
  )
)

plot_kegg_volcano_by_pathway <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(
        color = ifelse(sig, category, "NS"),
        alpha = sig
      ),
      size = 1.6
    ) +
    scale_color_manual(
      values = kegg_colors,
      name = "KEGG class"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}

plot_kegg_volcano_by_pathway(
  kegg_annot_df,
  title_prefix = "KEGG enrichment by pathway"
)

```


```{r bacteroides colored, fig.height = 10, fig.width = 7}
kegg_annot_df %>%
  filter(species == "phocaeicola_vulgatus") %>%
  plot_kegg_volcano_by_pathway(
    title_prefix = "KEGG enrichment by pathway – Phocaeicola vulgatus"
  )
```



### Virulence factors

```{r vf_volcano, fig.height = 5, fig.width = 15}

vf_df <- prepare_volcano_df(
  vf_fisher_total,
  tree_order = tree_order
)

vf_df_annot <- vf_df %>%
  left_join(vf_data, by = "vf") %>%
  mutate(
    color_group = ifelse(sig, vf_type, "nonsig")
  )

vf_types <- sort(unique(na.omit(vf_df_annot$vf_type)))

vf_colors <- c(
  nonsig = "grey70",
  setNames(
    RColorBrewer::brewer.pal(
      min(8, length(vf_types)),
      "Set1"
    ),
    vf_types[seq_len(min(8, length(vf_types)))]
  )
)


plot_vf_volcano_by_type <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = color_group, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = vf_colors,
      name = "VF type"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}


plot_vf_volcano_by_type(
  vf_df_annot,
  title_prefix = "VF enrichment by VF type"
)

```
```{r bacteroides vf, fig.height = 9, fig.width = 6}
vf_df_annot %>%
  filter(species == "phocaeicola_vulgatus") %>%
  plot_vf_volcano_by_type(
    title_prefix = "VF enrichment by type – Phocaeicola vulgatus"
  )
```

##### CAZY
```{r cazy data}
library(readxl)
faminfo <- read_excel("data/FamInfo.txt.04232020.xls")
activities <- read_tsv("data/CAZyDB.07302020.fam-activities.txt", comment = "#",
                       col_names = c("family", "activity"))
substrates <- read_tsv("data/fam-substrate-mapping-08012023.tsv")

```

```{r cazy annot}
cazy_df <- prepare_volcano_df(
  cazy_fisher_total,
  tree_order = tree_order
)


cazy_parsed <- cazy_df %>%
  mutate(
    cazy_clean = stringr::str_trim(cazy),

    # CAZy class: GH, GT, PL, CE, AA, CBM
    cazy_class = stringr::str_extract(cazy_clean, "^[A-Za-z]+"),

    # Family number
    remainder = stringr::str_remove(cazy_clean, "^[A-Za-z]+"),
    cazy_family = stringr::str_extract(remainder, "^\\d+"),

    # Canonical family ID (e.g. GH13, PL7)
    cazy_family_id = paste0(cazy_class, cazy_family)
  )



cazy_annot <- cazy_parsed %>%
  left_join(substrates, by = c("cazy_family_id" = "Family"))
```



```{r cazy volcano coloured, fig.height = 8, fig.width = 14}


cazy_annot <- cazy_annot %>%
  mutate(
    color_group = ifelse(sig, cazy_class, "nonsig")
  )

cazy_types <- sort(unique(na.omit(cazy_annot$cazy_class)))

cazy_colors <- c(
  nonsig = "grey70",
  setNames(
    RColorBrewer::brewer.pal(
      min(8, length(cazy_types)),
      "Set2"
    ),
    cazy_types[seq_len(min(8, length(cazy_types)))]
  )
)



plot_cazy_volcano_by_class <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = color_group, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = cazy_colors,
      name = "CAZy class"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}



cazy_top <- cazy_annot %>%
  filter(sig) %>%
  group_by(species, contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

plot_cazy_volcano_by_class(
  cazy_annot,
  title_prefix = "CAZy enrichment by class"
) +
  ggrepel::geom_text_repel(
    data = cazy_top,
    aes(label = cazy),
    size = 3,
    max.overlaps = 8
  )




plot_cazy_volcano_by_class(
  cazy_annot,
  title_prefix = "CAZy enrichment by class"
)


```


```{r bacteroides_uniformis_cazy}
cazy_sp <- cazy_annot %>%
  filter(species == "bacteroides_uniformis")

plot_cazy_volcano_by_class(
  cazy_sp,
  title_prefix = "CAZy enrichment by class – Bacteroides uniformis"
)

```

```{r one species all comparisons, fig.height = 8, fig.width = 8}
species_of_interest <- "bacteroides_uniformis"

kegg_df_sp <- kegg_annot_df %>%
  filter(species == species_of_interest)

kegg_top_sp <- kegg_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_kegg <- plot_kegg_volcano_by_pathway(
  kegg_df_sp,
  title_prefix = paste("KEGG –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = kegg_top_sp,
    aes(label = kegg),
    size = 3,
    max.overlaps = 8
  )

p_kegg

amr_df_sp <- amr_df %>%
  filter(species == species_of_interest)

amr_top_sp <- amr_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_amr <- plot_volcano_contrasts(
  amr_df_sp,
  title_prefix = paste("AMR –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = amr_top_sp,
    aes(label = amr),
    size = 3
  )

p_amr


vf_df_sp <- vf_df_annot %>%
  filter(species == species_of_interest)

vf_top_sp <- vf_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_vf <- plot_vf_volcano_by_type(
  vf_df_sp,
  title_prefix = paste("VF –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = vf_top_sp,
    aes(label = vf),
    size = 3
  )

p_vf


cazy_df_sp <- cazy_annot %>%
  filter(species == species_of_interest)

cazy_top_sp <- cazy_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_cazy <- plot_cazy_volcano_by_class(
  cazy_df_sp,
  title_prefix = paste("CAZy –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = cazy_top_sp,
    aes(label = cazy),
    size = 3
  )

p_cazy



```
```{r patchworking, fig.height = 8, fig.width = 12}
small_legend <- theme(
  legend.text = element_text(size = 7),
  legend.title = element_text(size = 8),
  legend.key.size = unit(0.35, "cm"),
  legend.spacing.y = unit(0.1, "cm")
)

p_kegg <- p_kegg +
  labs(title = "KEGG") +
  small_legend

p_amr <- p_amr +
  labs(title = "AMR") +
  small_legend

p_vf <- p_vf +
  labs(title = "VF") +
  small_legend

p_cazy <- p_cazy +
  labs(title = "CAZy") +
  small_legend

(p_kegg | p_amr | p_vf | p_cazy) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.35, "cm")
  )

```
```{r all species in a loop,  fig.height = 8, fig.width = 13}

kegg_annot_df <- kegg_annot_df %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

amr_df <- amr_df %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

vf_df_annot <- vf_df_annot %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

cazy_annot <- cazy_annot %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))


make_functional_patchwork <- function(species_of_interest) {

  ## ---------- KEGG ----------
  kegg_df_sp <- kegg_annot_df %>%
    filter(species == species_of_interest)

  kegg_top_sp <- kegg_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_kegg <- plot_kegg_volcano_by_pathway(
    kegg_df_sp,
    title_prefix = paste("KEGG –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = kegg_top_sp,
      aes(label = kegg),
      size = 3,
      max.overlaps = 8
    )


  ## ---------- AMR ----------
  amr_df_sp <- amr_df %>%
    filter(species == species_of_interest)

  amr_top_sp <- amr_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_amr <- plot_volcano_contrasts(
    amr_df_sp,
    title_prefix = paste("AMR –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = amr_top_sp,
      aes(label = amr),
      size = 3
    )


  ## ---------- VF ----------
  vf_df_sp <- vf_df_annot %>%
    filter(species == species_of_interest)

  vf_top_sp <- vf_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_vf <- plot_vf_volcano_by_type(
    vf_df_sp,
    title_prefix = paste("VF –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = vf_top_sp,
      aes(label = vf),
      size = 3
    )


  ## ---------- CAZy ----------
  cazy_df_sp <- cazy_annot %>%
    filter(species == species_of_interest)

  cazy_top_sp <- cazy_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_cazy <- plot_cazy_volcano_by_class(
    cazy_df_sp,
    title_prefix = paste("CAZy –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = cazy_top_sp,
      aes(label = cazy),
      size = 3
    )


  ## ---------- Legend theme ----------
  small_legend <- theme(
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.35, "cm")
  )

    p_kegg <- p_kegg +
      labs(title = "KEGG") +
      small_legend
    
    p_amr <- p_amr +
      labs(title = "AMR") +
      small_legend
    
    p_vf <- p_vf +
      labs(title = "VF") +
      small_legend
    
    p_cazy <- p_cazy +
      labs(title = "CAZy") +
      small_legend


  ## ---------- Final patchwork ----------
  final_plot <-
    (p_kegg | p_amr | p_vf | p_cazy) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "right",
      plot.title = element_text(size = 14, face = "bold")
    )

  final_plot
}


# patchwork_plots <- purrr::set_names(
#   lapply(species_list, make_functional_patchwork),
#   species_list
# )
# 
# purrr::walk(patchwork_plots, print)
# 


for (sp in species_list) {

  p <- make_functional_patchwork(sp)

  print(p)

  ggsave(
    paste0("plots/functional_enrichment_", sp, ".pdf"),
    plot = p,
    width = 13,
    height = 8
  )
}


```



```{r save functional_results_2}
# fisher_results <- list(
#   kegg = kegg_annot_df,
#   amr  = amr_df,
#   vf   = vf_df_annot,
#   cazy = cazy_annot,
#   species = species_list,
#   info = list(
#     created = Sys.time(),
#     description = "Functional enrichment input tables for volcano plots"
#   )
# )
# 
# saveRDS(
#   fisher_results,
#   file = "data/intermediate_analysis/fisher_results.rds"
# )





# functional_results <- readRDS( "data/intermediate_analysis/fisher_results.rds")
# 
# 
# 
# kegg_annot_df <- functional_results$kegg
# amr_df        <- functional_results$amr
# vf_df_annot   <- functional_results$vf
# cazy_annot    <- functional_results$cazy

```



### ZOOM-IN - quorum sensing
```{r quorum sensing}
qs_df <- kegg_annot_df %>%
  filter(
    pathway_map == "path:map02024"
  ) %>%
  mutate(
    sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
  )


ggplot(qs_df, aes(
  x = log2_or,
  y = ko,
  color = species
)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(alpha = sig), size = 2.5) +
  facet_wrap(~ contrast) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.05)) +
  labs(
    title = "Quorum sensing genes show host-associated enrichment",
    x = "log2(odds ratio)",
    y = "Quorum sensing KOs",
    alpha = "Significant"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 7)
  )

```
```{r top pathways for zoomin, fig.width = 10, fig.height = 7}
top_pathways <- kegg_annot_df %>%
  filter(
    sig,
    !is.na(pathway_map)
  ) %>%
  group_by(pathway_map, pathway_name) %>%
  summarise(
    n_sig = n(),
    mean_abs_log2_or = mean(abs(log2_or), na.rm = TRUE),
    score = n_sig * mean_abs_log2_or,
    .groups = "drop"
  ) %>%
  arrange(desc(score)) %>%
  slice_head(n = 20)

top20_df <- kegg_annot_df %>%
  semi_join(top_pathways, by = "pathway_map") %>%
  mutate(
    sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
  )

top20_summary <- top20_df %>%
  group_by(pathway_name, species, contrast) %>%
  summarise(
    mean_log2_or = mean(log2_or, na.rm = TRUE),
    n_sig = sum(sig),
    .groups = "drop"
  )


ggplot(top20_summary, aes(
  x = mean_log2_or,
  y = pathway_name,
  color = species,
  size = n_sig
)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.8) +
  facet_wrap(~ contrast) +
  labs(
    title = "Top 20 KEGG pathways enriched across host contrasts",
    x = "Mean log2(odds ratio)",
    y = NULL,
    size = "Significant KOs"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 7)
  )
```
```{r for_20_pathways}
plot_kegg_pathway_breakdown <- function(pathway_id, pathway_label) {

  df <- kegg_annot_df %>%
    filter(pathway_map == pathway_id) %>%
    mutate(
      sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
    )

  # Guard against empty or tiny pathways
  if (nrow(df) == 0) {
    return(NULL)
  }

  ggplot(df, aes(
    x = log2_or,
    y = ko,
    color = species
  )) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(aes(alpha = sig), size = 2.5) +
    facet_wrap(~ contrast) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.05)) +
    labs(
      title = pathway_label,
      subtitle = pathway_id,
      x = "log2(odds ratio)",
      y = "KEGG ortholog",
      alpha = "Significant"
    ) +
    theme_bw() +
    theme(
      legend.position = "right",
      axis.text.y = element_text(size = 7)
    )
}

plots_by_pathway <- purrr::pmap(
  list(
    top_pathways$pathway_map,
    top_pathways$pathway_name
  ),
  plot_kegg_pathway_breakdown
)

# Drop NULLs just in case
plots_by_pathway <- purrr::compact(plots_by_pathway)

purrr::walk(plots_by_pathway, print)

```



## META-ANALYSIS

### Human vs wild
```{r metaanalysis for all keggs_hw, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 8
min_total_events <- 10

kegg_keep_hw <- df_hw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_hw_screened <- df_hw %>%
  inner_join(kegg_keep_hw, by = "kegg")

# Split by KEGG
kegg_split_hw <- split(df_hw_screened, df_hw_screened$kegg)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0",
    incr =0.5
  )
}

mh_hw <- lapply(kegg_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



# select keggs where confidence interval does not include 1 
#these are consistent in the direction of prevalence
mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```
As most of the identified KOs are ribosomal subunits, we try again, filtering these before the meta-analysis.

```{r metaanalysis for all keggs_remove_ribosomal, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs

#  Low-bias screen 
min_informative_species <- 8
min_total_events <- 10

kegg_keep_hw <- df_hw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_hw_screened <- df_hw %>%
  inner_join(kegg_keep_hw, by = "kegg")

# Split by KEGG
kegg_split_hw <- split(df_hw_screened, df_hw_screened$kegg)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0",
    incr =0.5
  )
}

mh_hw <- lapply(kegg_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



# select keggs where confidence interval does not include 1 
#these are consistent in the direction of prevalence
mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```

Plotting the individual KOs to check if all species do follow same pattern

```{r forest plot individual keggs_human_wild_after_ribosomal, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_kegg <- function(kegg_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[kegg_id]]
  if (is.null(m)) stop("KEGG not found in mh_hw: ", kegg_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("KEGG: ", kegg_id)
  )
}

#individual analyses of each of these keggs:
keggs <- mh_re_table_ci %>% dplyr::pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```

Manual search table of the KOs and information for plotting.

```{r manual lookup of keggs}
kegg_search_info <- tibble::tibble(
  kegg = c(
    
    "K18799","K03436","K01356","K07461","K13573","K03580","K09144",
    "K07497","K03798","K07317","K03427","K06871","K02935","K02864",
    "K07485","K07493","K01183","K22886","K06113","K02965","K02890",
    "K01580","K02886","K02994","K02933","K02982","K02946","K02892",
    "K02926","K02907","K01154","K02954","K03279","K03046","K00571",

    
    "K07459","K03741","K07448","K08884","K08217","K07482","K07455",
    "K18220","K20922","K07339",
    "K21029","K18909","K16566","K00980","K16702","K12453","K06919",
    
    "K18234", "K03076", "K15773"
  ),

  description = c(
    
    "O-antigen flippase wzxB",
    "DeoR family transcriptional regulator",
    "Repressor LexA",
    "Putative endonuclease",
    "Proteasome accessory factor C",
    "ATP-dependent helicase HepA",
    "Nucleotidyl transferase AbiEii/AbiGii toxin-family protein",
    "Putative transposase",
    "Cell division protease FtsH",
    "Adenine-specific DNA methyltransferase",
    "Type I restriction enzyme M protein",
    "Uncharacterized protein",
    "Ribosomal protein L7/L12",
    "Ribosomal protein L10",
    "Transposase",
    "Putative transposase",
    "Chitinase",
    "Protein lysine acetyltransferase",
    "Arabinan endo-1,5-alpha-L-arabinosidase",
    "Small subunit ribosomal protein S19",
    "Large subunit ribosomal protein L22",
    "Glutamate decarboxylase",
    "Large subunit ribosomal protein L2",
    "Small subunit ribosomal protein S8",
    "Large subunit ribosomal protein L6",
    "Small subunit ribosomal protein S3",
    "Small subunit ribosomal protein S10",
    "Large subunit ribosomal protein L23",
    "Large subunit ribosomal protein L4",
    "Large subunit ribosomal protein L30",
    "Type I restriction enzyme, S subunit",
    "Small subunit ribosomal protein S14",
    "UDP-glucose:(galactosyl)LPS alpha-1,2-glucosyltransferase",
    "DNA-directed RNA polymerase subunit beta'",
    "Site-specific DNA-methyltransferase",
    "ATP-dependent endonuclease OLD family protein",
    "Arsenate reductase",
    "Type IV restriction system protein Mrr",
    "Serine/threonine protein kinase",
    "Macrolide efflux protein",
    "IS30 family transposase",
    "Recombination protein RecT",
    "Ribosomal protection tetracycline resistance protein",
    "Polysaccharide biosynthesis protein",
    "mRNA interferase toxin HicA",
    "Uncharacterized protein",
    "MarR family transcriptional regulator",
    "Exopolysaccharide biosynthesis protein ExoY",
    "Glycerol-3-phosphate cytidylyltransferase",
    "Exopolysaccharide biosynthesis glycosyltransferase",
    "CDP-paratose synthetase",
    "Putative DNA replication protein",
    "Virginiamycin A acetyltransferase",
    "Preprotein translocase subunit SecY",
    "HTH-type transcriptional regulator / antitoxin HipB"
  ),

  extra_info = c(
    "",
    "fructose operon transcriptional repressor",
    "serine endopeptidase, SOS response regulator",
    "",
    "pafC",
    "",
    "Type IV TA system toxin",
    "",
    "metallo-endopeptidase",
    "Type II R-M system",
    "adenine-specific DNA methyltransferase",
    "function unknown",
    "large subunit",
    "large subunit",
    "",
    "",
    "",
    "",
    "abnA",
    "RP-S19, RSM19, rpsS",
    "RP-L22, MRPL22, rplV",
    "Early infantile epileptic encephalopathy",
    "ribosome",
    "RP-S8, MRPS8, rpsH",
    "RP-L6, MRPL6, rplF",
    "RP-S3, rpsC",
    "RP-S10, rpsJ",
    "RP-L23, MRPL23, rplW",
    "RP-L4, MRPL4, rplD",
    "RP-L30, MRPL30, rpmD",
    "Type I R-M system specificity subunit",
    "RP-S14, MRPS14, rpsN",
    "waaJ, rfaJ",
    "rpoC",
    "adenine-specific",
    "DNA damage response",
    "arsC, thioredoxin-dependent",
    "restriction-modification system",
    "protein phosphorylation",
    "MFS transporter",
    "mobile genetic element",
    "homologous recombination",
    "tetM / tetO",
    "biofilm-associated",
    "Type II toxin–antitoxin system",
    "function unknown",
    "mepR, antibiotic resistance regulation",
    "exopolysaccharide synthesis",
    "tagD",
    "amsE",
    "rfbS",
    "replication / repair",
    "vat",
    "",
    ""
    
  )
)


kegg_search_info
```
```{r wild human with kegginfo after ribosomal, fig.height = 6, fig.width = 10}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```








```{r forest plot individual keggs_again, fig.height = 6, fid.width = 10}
#individual analyses of each of these keggs:
keggs <- mh_re_table_ci %>% dplyr::pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```




Testing risk difference
### Risk difference meta-analysis
```{r risk difference}
run_random_hw_rd <- function(dat) {
  meta::metabin(
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,

    sm      = "RD",        #  RISK DIFFERENCE
    method  = "MH",

    common  = FALSE,
    random  = TRUE,

    method.tau       = "REML",
    method.random.ci = "HK",

    prediction = TRUE
  )
}

mh_hw_rd <- lapply(kegg_split_hw, run_random_hw_rd)

mh_hw_rd_table <- bind_rows(lapply(names(mh_hw_rd), function(k) {
  m <- mh_hw_rd[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    RD_random      = m$TE.random,
    se_random      = m$seTE.random,
    ci_low_random  = m$lower.random,
    ci_high_random = m$upper.random,
    p_random       = m$pval.random,
    I2_percent     = m$I2,
    tau2_REML      = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))


mh_hw_rd_ci <- mh_hw_rd_table %>%
  filter(ci_low_random > 0 | ci_high_random < 0)



mh_hw_rd_ci$kegg_I2 <- paste0(
  mh_hw_rd_ci$kegg,
  "  (I²=",
  round(100* mh_hw_rd_ci$I2_percent, 2),
  "%)"
)

mgen_rd <- meta::metagen(
  TE   = mh_hw_rd_ci$RD_random,
  seTE = mh_hw_rd_ci$se_random,
  studlab = mh_hw_rd_ci$kegg_I2,
  sm = "RD",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_rd,
  backtransf = FALSE,
  xlab = "Prevalence difference (Human − Wild)"
)

# meta::forest(
#   mgen_rd,
#   backtransf = FALSE,
#   xlab = "Prevalence difference (Human − Wild)",
#   leftcols = c("studlab", "I2"),
#   leftlabs = c("KEGG", expression(I^2~"%")),
#   col.square = "black",
#   col.diamond = "black",
#   col.diamond.lines = "black"
# )

```





### Domestic vs wild

```{r metaanalysis for all keggs_w_d_removing_ribosomal}
# Prep totals and basic filters 
df_dw <- kegg_fisher_total %>%
  filter(contrast == "wild_vs_domestic") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_domestic = n_mags_domestic + absent_domestic
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs


#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

kegg_keep_dw <- df_dw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_domestic),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_dw_screened <- df_dw %>%
  inner_join(kegg_keep_dw, by = "kegg")

# Split by KEGG
kegg_split_dw <- split(df_dw_screened, df_dw_screened$kegg)

# Effect size pooling
# MH with random effects; domestic vs wild 
run_mh_random_dw <- function(dat) {
  meta::metabin(
    # domestic as experimental (so OR>1 => more prevalent in domestic)
    event.e = dat$n_mags_domestic,
    n.e     = dat$tot_domestic,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_dw <- lapply(kegg_split_dw, run_mh_random_dw)

# Collect pooled results + FDR 
mh_dw_table <- bind_rows(lapply(names(mh_dw), function(k) {
  m <- mh_dw[[k]]
  data.frame(
    kegg = k,
    contrast = "domestic_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))

```



```{r keggs where ci does not include_1_h_d_r_r, fig.height = 5, fig.width = 7}
mh_re_table_ci <- mh_dw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (domestic vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)


```

```{r wild domestic with kegginfo, fig.height = 6, fig.width = 10}
mh_dw_table_annot <- mh_dw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_dw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Domestic vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```


### Human vs domestic

```{r metaanalysis for all keggs_hd_rr}
# Prep totals and basic filters 
df_dw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_domestic") %>%
  mutate(
    tot_domestic  = n_mags_domestic  + absent_domestic,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs

#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

kegg_keep_dw <- df_dw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_domestic + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_dw_screened <- df_dw %>%
  inner_join(kegg_keep_dw, by = "kegg")

# Split by KEGG
kegg_split_dw <- split(df_dw_screened, df_dw_screened$kegg)

# Effect size pooling
# MH with random effects; human vs domestic 
run_mh_random_dw <- function(dat) {
  meta::metabin(
    # human as experimental (so OR>1 => more prevalent in human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_domestic,
    n.c     = dat$tot_domestic,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_dw <- lapply(kegg_split_dw, run_mh_random_dw)

# Collect pooled results + FDR 
mh_dw_table <- bind_rows(lapply(names(mh_dw), function(k) {
  m <- mh_dw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_domestic",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))


mh_re_table_ci <- mh_dw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (human vs domestic)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```




```{r human domestic with kegginfo_3, fig.height = 6, fig.width = 10}
mh_dw_table_annot <- mh_dw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_dw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Domestic)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```
### Meta-analysis with VF
#### Human vs wild
```{r metaanalysis for all vfs, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  )%>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

   
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$vf,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("VF"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```

```{r check vf name}
vf_data <- read_excel("data/VFs.xls")
```


```{r human wild with vfinfo, fig.height = 6, fig.width = 10}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    join_by ("vf" == "VFID")
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(VF_Name),
      vf,
      paste0(vf, ": ", VF_Name)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("Virulence Factor"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```

#### Human vs domestic
```{r metaanalysis for all vfs_hd}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "human_vs_domestic") %>%
  mutate(
    tot_domestic  = n_mags_domestic  + absent_domestic,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_domestic + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; Human vs domestic 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_domestic,
    n.c     = dat$tot_domestic,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "human_vs_domestic",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    join_by ("vf" == "VFID")
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(VF_Name),
      vf,
      paste0(vf, ": ", VF_Name)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Domestic)",
  leftlabs = c("Virulence Factor"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```



```{r forest plot individual vfs_hd, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_vf <- function(vf_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[vf_id]]
  if (is.null(m)) stop("vf not found in mh_hw: ", vf_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("VF: ", vf_id)
  )
}

#individual analyses of each of these vf:
vfs <- mh_re_table_ci %>% dplyr::pull(vf)

for (k in vfs) {
  forest_one_vf(
    vf_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```




#### Domestic vs wild
```{r metaanalysis for all vfs_dw_2}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "wild_vs_domestic") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_domestic = n_mags_domestic + absent_domestic
  )  %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)

#  Low-bias screen 
min_informative_species <- 3
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_domestic),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; domestic vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # domestic as experimental (so OR>1 => more prevalent in domestic)
    event.e = dat$n_mags_domestic,
    n.e     = dat$tot_domestic,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "domestic_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))

```



```{r vfs where ci does not include_1_vf_dom_wild, fig.height = 10, fig.width = 13}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    join_by ("vf" == "VFID")
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(VF_Name),
      vf,
      paste0(vf, ": ", VF_Name)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$studlab_desc,
  sm = "OR",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (domestic vs Wild)",
  leftlabs = c("Virulence Factor"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```
```{r forest plot individual vfs_dom_wild, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_vf <- function(vf_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[vf_id]]
  if (is.null(m)) stop("vf not found in mh_hw: ", vf_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("VF: ", vf_id)
  )
}

#individual analyses of each of these vf:
vfs <- mh_re_table_ci %>% dplyr::pull(vf)

for (k in vfs) {
  forest_one_vf(
    vf_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```

## Hierarchical GLMM

```{r make the matrices}
make_glmm_long <- function(mat, metadata, species_name, annotation_type) {

  # mat: wide matrix from make_annotation_matrix()
  # metadata: genome metadata table
  # annotation_col: column name for annotation ("kegg", "amr", etc.)
  # annotation_type: label ("KEGG", "AMR", "VF", "CAZy")

  if (ncol(mat) <= 1) return(tibble::tibble())

  # 1. Convert counts to presence/absence
  pa <- mat %>%
    dplyr::mutate(
      dplyr::across(-genome, ~ as.integer(. != 0))
    )

  # 2. Pivot to long format 
  pa_long <- pa %>%
    tidyr::pivot_longer(
      cols = -genome,
      names_to  = "annotation",
      values_to = "present"
    )

  # 3. Join metadata

   out <- pa_long %>%
      dplyr::left_join(
        metadata %>%
          dplyr::select(
            genome        = ID_clean,
            host_type_2,
            completeness
          ),
        by = "genome"
      ) %>%
      dplyr::mutate(
        species = species_name,
        annotation_type = annotation_type
      )


  # 4. Drop rows with missing critical info
  out %>%
    dplyr::filter(
      !is.na(host_type_2),
      !is.na(completeness)
    )
}


glmm_tables <- list()

for (sp in species_list) {

  message("Processing species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  mats <- list(
    KEGG = make_annotation_matrix(genome_annotations, "kegg"),
    AMR  = make_annotation_matrix(genome_annotations, "resistance_target"),
    VF   = make_annotation_matrix(genome_annotations, "vf"),
    CAZy = make_annotation_matrix(genome_annotations, "cazy")
  )

  glmm_tables[[sp]] <- purrr::imap(
    mats,
    ~ make_glmm_long(
        mat            = .x,
        metadata       = genome_metadata,
        species_name   = sp,
        annotation_type = .y
      )
  )
}
```
```{r combine}
kegg_glmm <- purrr::map_dfr(glmm_tables, "KEGG")
amr_glmm  <- purrr::map_dfr(glmm_tables, "AMR")
vf_glmm   <- purrr::map_dfr(glmm_tables, "VF")
cazy_glmm <- purrr::map_dfr(glmm_tables, "CAZy")
```



```{r check distribution}

ggplot(kegg_glmm, aes(x = factor(present))) +
  geom_bar() +
  labs(
    x = "Presence / Absence",
    y = "Number of observations",
    title = "Distribution of KEGG presence (row-wise)"
  )+theme_bw()

ggplot(vf_glmm, aes(x = factor(present))) +
  geom_bar() +
  labs(
    x = "Presence / Absence",
    y = "Number of observations",
    title = "Distribution of VF presence (row-wise)"
  )+theme_bw()


```
```{r kegg checks}
kegg_by_genome <- kegg_glmm %>%
  dplyr::group_by(genome, host_type_2, species) %>%
  dplyr::summarise(
    n_kegg = sum(present),
    .groups = "drop"
  )

ggplot(kegg_by_genome, aes(x = n_kegg)) +
  geom_histogram(bins = 50) +
  labs(
    x = "Number of KEGG annotations per genome",
    y = "Number of genomes",
    title = "KEGG richness per genome"
  )+ theme_bw()



vf_by_genome <- vf_glmm %>%
  dplyr::group_by(genome, host_type_2, species) %>%
  dplyr::summarise(
    n_vf = sum(present),
    .groups = "drop"
  )

ggplot(vf_by_genome, aes(x = n_vf)) +
  geom_histogram(bins = 50) +
  labs(
    x = "Number of VF annotations per genome",
    y = "Number of genomes",
    title = "VF richness per genome"
  )+ theme_bw()

```


For each KO, does the probability of being present in a MAG differ by host type (wild / domestic / human), consistently across bacterial species, after correcting for MAG completeness?

-Done in cluster

```{r load_vf_glmm_resutls}
vf_fixed  <- read_csv("data/glmm_results/vf_glmm_fixed_effects.csv")
vf_contr  <- read_csv("data/glmm_results/vf_glmm_host_contrasts.csv")
vf_effect <- read_csv("data/glmm_results/vf_glmm_annotation_effects.csv")
m_vf_2 <- readRDS("data/glmm_results/vf_glmm_model.rds")
```
```{r vf_fixed}
vf_fixed
```
```{r vf_host_contrasts}
vf_contr
```

```{r vf_annot}
vf_effect %>% head(10)
```


```{r selecting top vfs}
vf_top_human <- vf_effect %>%
  arrange(host_type_2wild) %>%   # more negative = more human-enriched
  slice_head(n = 15)

vf_top_human

vf_top_wild <- vf_effect %>%
  arrange(desc(host_type_2wild)) %>%
  slice_head(n = 15)

vf_top_wild

```


```{r load random effects model annotations}
re_vf <- ranef(m_vf_2, condVar = TRUE)$annotation
```


```{r vf_effects_with_uncertainty}

vf_df <- as.data.frame(re_vf) %>%
  tibble::rownames_to_column("vf")

postvar <- attr(re_vf, "postVar")

pv_slopes <- postvar[['c("host_type_2human", "host_type_2domestic", "host_type_2wild")']]


vf_df <- vf_df %>%
  mutate(
    se_human    = sqrt(pv_slopes[1, 1, ]),
    se_domestic = sqrt(pv_slopes[2, 2, ]),
    se_wild     = sqrt(pv_slopes[3, 3, ])
  )


vf_df <- vf_df %>%
  mutate(
    lo_human = host_type_2human - 1.96 * se_human,  #95% uncertainty interval
    hi_human = host_type_2human + 1.96 * se_human,

    lo_dom   = host_type_2domestic - 1.96 * se_domestic,
    hi_dom   = host_type_2domestic + 1.96 * se_domestic,

    lo_wild  = host_type_2wild - 1.96 * se_wild,
    hi_wild  = host_type_2wild + 1.96 * se_wild
  )

vf_df2 <- vf_df %>%
  left_join(vf_data, by = c("vf" = "vf"))


vf_human <- vf_df2 %>%
  filter(hi_wild < 0) %>%
  arrange(host_type_2wild)


vf_wild <- vf_df2 %>%
  filter(lo_wild > 0) %>%
  arrange(desc(host_type_2wild))

vf_domestic <- vf_df2 %>%
  filter(
    lo_dom > 0,
    hi_wild > 0,     # optional: distinct from wild
    hi_human > 0    # optional: distinct from human
  ) %>%
  arrange(desc(host_type_2domestic))


vf_gradient <- vf_df2  %>%
  filter(
    lo_wild > 0,
    lo_dom  > 0,
    host_type_2wild > host_type_2domestic,
    host_type_2domestic > host_type_2human
  )  %>%
  arrange(desc(host_type_2wild))



top_wild <- vf_wild  %>% slice_head(n = 12)

ggplot(top_wild,
       aes(y = reorder(vf, host_type_2wild),
           x = exp(host_type_2wild))) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(
      xmin = exp(lo_wild),
      xmax = exp(hi_wild)
    ),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Wild vs Human)",
    y = "Virulence factor"
  ) +
  theme_bw()


ggplot(top_wild,
       aes(y = reorder(vf, host_type_2wild),
           x = exp(host_type_2wild),
           color = vf_type)) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = exp(lo_wild), xmax = exp(hi_wild)),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Wild vs Human)",
    y = "Virulence factor",
    color = "VF functional class"
  ) +
  theme_bw()


vf_domestic <- vf_df2 %>%
  filter(lo_dom > 0) %>%
  arrange(desc(host_type_2domestic))

top_domestic <- vf_domestic %>% slice_head(n = 12)

ggplot(top_domestic,
       aes(y = reorder(vf, host_type_2domestic),
           x = exp(host_type_2domestic),
           color = vf_type)) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = exp(lo_dom), xmax = exp(hi_dom)),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Domestic vs Human)",
    y = "Virulence factor",
    color = "VF functional class"
  ) +
  theme_bw()

vf_human <- vf_df2 %>%
  filter(hi_wild < 0) %>%
  arrange(host_type_2wild)

top_human <- vf_human %>% slice_head(n = 12)


ggplot(top_human,
       aes(y = reorder(vf, -host_type_2wild),
           x = exp(-host_type_2wild),
           color = vf_type)) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = exp(-hi_wild), xmax = exp(-lo_wild)),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Human vs Wild)",
    y = "Virulence factor",
    color = "VF functional class"
  ) +
  theme_bw()



vf_df2 <- vf_df2 %>%
  mutate(
    wild_vs_domestic = host_type_2wild - host_type_2domestic
  )

```
```{r human and wild together}
top_wild <- vf_df2 %>%
  arrange(desc(host_type_2wild)) %>%
  slice_head(n = 10)

top_human <- vf_df2 %>%
  arrange(host_type_2wild) %>%
  slice_head(n = 10)

top_both <- bind_rows(top_wild, top_human)

ggplot(top_both,
       aes(
         y = reorder(vf, host_type_2wild),
         x = exp(host_type_2wild),
         color = vf_type
       )) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(
      xmin = exp(lo_wild),
      xmax = exp(hi_wild)
    ),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Wild vs Human, relative to average VF)",
    y = "Virulence factor",
    color = "VF functional class"
  ) +
  theme_bw()


top_domestic <- vf_df2 %>%
  arrange(desc(host_type_2domestic)) %>%
  slice_head(n = 10)

top_human_dom <- vf_df2 %>%
  arrange(host_type_2domestic) %>%
  slice_head(n = 10)

top_both_domestic <- bind_rows(top_domestic, top_human_dom)

ggplot(top_both_domestic,
       aes(
         y = reorder(vf, host_type_2domestic),
         x = exp(host_type_2domestic),
         color = vf_type
       )) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(
      xmin = exp(lo_dom),
      xmax = exp(hi_dom)
    ),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Odds ratio (Domestic vs Human, relative to average VF)",
    y = "Virulence factor",
    color = "VF functional class"
  ) +
  theme_bw()
```

```{r plotting histograms}
ggplot(vf_df, aes(x = host_type_2wild)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "VF deviation (wild vs human)",
    y = "Number of virulence factors"
  ) +
  theme_bw()


ggplot(vf_df2,
       aes(x = wild_vs_domestic)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "VF deviation (Wild vs Domestic)",
    y = "Number of virulence factors"
  ) +
  theme_bw()


```

```{r scattreplot}
ggplot(vf_df,
       aes(x = host_type_2wild, y = host_type_2domestic)) +
  geom_point(alpha = 0.3) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  labs(
    x = "VF deviation: Wild vs Human",
    y = "VF deviation: Domestic vs Human"
  ) +
  theme_bw()


ggplot(vf_df2,
       aes(x = host_type_2wild,
           y = host_type_2domestic,
           color = vf_type)) +
  geom_point(alpha = 0.6) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  labs(
    x = "VF deviation: Wild vs Human",
    y = "VF deviation: Domestic vs Human",
    color = "VF functional class"
  ) +
  theme_bw()
```





## GLMM with glmmTMB per species
```{r glmmTMB}
vf_models <- readRDS("data/glmm_results/vf_glmmTMB_models_by_species.rds")
vf_effects <- read_csv("data/glmm_results/vf_host_effects_by_species.csv")
vf_annotation_slopes <- read_csv("data/glmm_results/vf_annotation_random_slopes.csv")
```

```{r vf_effects}
vf_effects
```

```{r host_effects}
host_effects <- vf_effects %>%
  filter(term %in% c("host_type_2domestic", "host_type_2wild")) %>%
  mutate(
    odds_ratio = exp(estimate),
    ci_low = exp(estimate - 1.96 * std.error),
    ci_high = exp(estimate + 1.96 * std.error)
  )

```

```{r plot_host_efefcts}
ggplot(host_effects,
       aes(y = species, x = odds_ratio)) +
  geom_point() +
  geom_errorbarh(
    aes(xmin = ci_low, xmax = ci_high),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  facet_wrap(~ term) +
  scale_x_log10() +
  labs(
    x = "Odds ratio (log scale)",
    y = "Species",
    title = "Host-associated VF presence by species"
  ) +
  theme_bw()

```

```{r random_slopes}
vf_annotation_slopes
```




For the VF per-spceis model - not enough information to estimate the random slopes.

```{r vf per species}

vf_presence <- vf_fisher_total %>%
  mutate(present = n_mags_human > 0 | n_mags_wild > 0 | n_mags_domestic > 0)

# 2. Count number of species per VF
vf_counts <- vf_presence %>%
  filter(present) %>%
  dplyr::select(vf, species) %>%
  distinct() %>%
  group_by(vf) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

# 3. Count how many VFs fall in each category
vf_distribution <- vf_counts %>%
  count(n_species)

# 4. Plot
ggplot(vf_distribution, aes(x = n_species, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Number of species where VF is present",
    y = "Number of VFs",
    title = "Distribution of VF presence across species"
  ) +
  theme_minimal()




# Count number of species per VF per contrast
vf_counts <- vf_fisher_total %>%
  dplyr::select(vf, species, contrast) %>%
  distinct() %>%
  group_by(contrast, vf) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

# Distribution: how many VFs per species-count, per contrast
vf_distribution <- vf_counts %>%
  count(contrast, n_species) %>%
  complete(contrast, n_species = 1:10, fill = list(n = 0))

# Plot (faceted by contrast)
ggplot(vf_distribution, aes(x = n_species, y = n)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ contrast) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Number of species where VF is present",
    y = "Number of VFs",
    title = "VF distribution across species (split by contrast)"
  ) +
  theme_minimal()

```

## VF glmm TMB model
```{r vf model tmb}
vf_model <- readRDS("data/glmm_results/vf_2/vf_glmmTMB_global_model.rds")
vf_fixed_effects <- read_csv("data/glmm_results/vf_2/vf_glmm_fixed_effects_2.csv")
vf_annotation_host <- read_csv("data/glmm_results/vf_2/vf_annotation_host_OR_2.csv")
```

```{r fixed_effects_vf}
vf_fixed_effects
```

```{r vf_model_again}
library(glmmTMB)

vf_model <- up2date(vf_model)

summary(vf_model)

ranef_out <- ranef(vf_model)

vf_re <- ranef_out$cond$annotation %>%
  as.data.frame() %>%
  tibble::rownames_to_column("annotation")

head(vf_re)
```



```{r fixed_effects_checking_host}
fixef_host <- fixef(vf_model)$cond

vf_plot_df <- vf_re %>%
  dplyr::mutate(
    beta_domestic = fixef_host["host_type_2domestic"] + host_type_2domestic,
    beta_wild     = fixef_host["host_type_2wild"]     + host_type_2wild,
    OR_domestic = exp(beta_domestic),
    OR_wild     = exp(beta_wild)
  )

```


```{r selecting top20}
top_domestic <- vf_plot_df %>%
  dplyr::arrange(desc(OR_domestic)) %>%
  dplyr::slice_head(n = 20)


top_wild <- vf_plot_df %>%
  dplyr::arrange(desc(OR_wild)) %>%
  dplyr::slice_head(n = 20)
```





```{r plottingtop20}

ggplot(top_domestic,
       aes(x = reorder(annotation, OR_domestic),
           y = OR_domestic)) +
  geom_point(size = 2) +
  coord_flip() +
  geom_hline(yintercept = 1,
             linetype = "dashed",
             color = "red") +
  labs(
    title = "Top enriched VFs in domestic hosts",
    x = "Virulence factor (annotation)",
    y = "Odds ratio (vs human)"
  ) +
  theme_minimal()

ggplot(top_wild,
       aes(x = reorder(annotation, OR_wild),
           y = OR_wild)) +
  geom_point(size = 2) +
  coord_flip() +
  geom_hline(yintercept = 1,
             linetype = "dashed",
             color = "red") +
  labs(
    title = "Top enriched VFs in wild hosts",
    x = "Virulence factor (annotation)",
    y = "Odds ratio (vs human)"
  ) +
  theme_minimal()

```
```{r vflong}

vf_long <- vf_plot_df %>%
  dplyr::select(annotation, OR_domestic, OR_wild) %>%
  pivot_longer(
    cols = c(OR_domestic, OR_wild),
    names_to = "host",
    values_to = "OR"
  ) %>%
  dplyr::mutate(
    host = dplyr::recode(host,
                         OR_domestic = "domestic",
                         OR_wild = "wild")
  )

top_combined <- vf_long %>%
  dplyr::group_by(host) %>%
  dplyr::slice_max(OR, n = 15) %>%
  dplyr::ungroup()

ggplot(top_combined,
       aes(x = reorder(annotation, OR),
           y = OR,
           color = host)) +
  geom_point(size = 2,
             position = position_dodge(width = 0.6)) +
  coord_flip() +
  geom_hline(yintercept = 1,
             linetype = "dashed") +
  labs(
    title = "Top enriched virulence factors by host type",
    x = "Annotation",
    y = "Odds ratio (vs human)"
  ) +
  theme_minimal()
```

```{r randomeffects}
ranef_out <- ranef(vf_model, condVar = TRUE)
vf_re <- ranef_out$cond$annotation

post_var <- attr(vf_re, "condVar")

vf_re_df <- as.data.frame(vf_re) %>%
  tibble::rownames_to_column("annotation")

term_names <- colnames(vf_re)

idx_dom <- which(term_names == "host_type_2domestic")
idx_wild <- which(term_names == "host_type_2wild")

get_se <- function(i, term) {
  sqrt(post_var[term, term, i])
}

vf_re_df <- vf_re_df %>%
  dplyr::mutate(
    se_domestic = sapply(seq_len(nrow(vf_re_df)),
                         function(i) sqrt(post_var[idx_dom, idx_dom, i])),

    se_wild = sapply(seq_len(nrow(vf_re_df)),
                     function(i) sqrt(post_var[idx_wild, idx_wild, i]))
  )

fixef_host <- fixef(vf_model)$cond

vf_plot_df <- vf_re_df %>%
  dplyr::mutate(
    beta_domestic = fixef_host["host_type_2domestic"] + host_type_2domestic,
    beta_wild     = fixef_host["host_type_2wild"]     + host_type_2wild,

    # ORs
    OR_domestic = exp(beta_domestic),
    OR_wild     = exp(beta_wild),

    # CIs
    lower_domestic = exp(beta_domestic - 1.96 * se_domestic),
    upper_domestic = exp(beta_domestic + 1.96 * se_domestic),

    lower_wild = exp(beta_wild - 1.96 * se_wild),
    upper_wild = exp(beta_wild + 1.96 * se_wild)
  )


#deriving domestic vs wild

vf_plot_df <- vf_plot_df %>%
  dplyr::mutate(
    beta_dw = beta_domestic - beta_wild,
    OR_dw = exp(beta_dw),

    # Approximate SE (independent assumption)
    se_dw = sqrt(se_domestic^2 + se_wild^2),

    lower_dw = exp(beta_dw - 1.96 * se_dw),
    upper_dw = exp(beta_dw + 1.96 * se_dw)
  )

top_domestic <- vf_plot_df %>%
  dplyr::arrange(desc(OR_domestic)) %>%
  dplyr::slice_head(n = 20)

ggplot(top_domestic,
       aes(x = reorder(annotation, OR_domestic),
           y = OR_domestic)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_domestic, ymax = upper_domestic),
                width = 0.2) +
  coord_flip() +
  scale_y_log10() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Domestic vs Human",
       y = "Odds ratio",
       x = "Annotation") +
  theme_minimal()
```
```{r onlysig}
vf_sig <- vf_plot_df %>%
  dplyr::mutate(
    sig_domestic = (lower_domestic > 1 | upper_domestic < 1),
    sig_wild     = (lower_wild > 1 | upper_wild < 1),
    sig_dw       = (lower_dw > 1 | upper_dw < 1)
  )

vf_sig %>%
  summarise(
    domestic_sig = sum(sig_domestic, na.rm = TRUE),
    wild_sig = sum(sig_wild, na.rm = TRUE),
    domestic_vs_wild_sig = sum(sig_dw, na.rm = TRUE)
  )


vf_sig %>%
  summarise(
    domestic_enriched = sum(lower_domestic > 1),
    domestic_depleted = sum(upper_domestic < 1),

    wild_enriched = sum(lower_wild > 1),
    wild_depleted = sum(upper_wild < 1)
  )
```

```{r domesticwild}
vf_domestic_sig <- vf_plot_df %>%
  dplyr::filter(lower_domestic > 1)

vf_wild_sig <- vf_plot_df %>%
  dplyr::filter(lower_wild > 1)

top_domestic_sig <- vf_domestic_sig %>%
  dplyr::arrange(desc(OR_domestic)) %>%
  dplyr::slice_head(n = 20)

top_wild_sig <- vf_wild_sig %>%
  dplyr::arrange(desc(OR_wild)) %>%
  dplyr::slice_head(n = 20)

ggplot(top_domestic_sig,
       aes(x = reorder(annotation, OR_domestic),
           y = OR_domestic)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower_domestic, ymax = upper_domestic),
                width = 0.2) +
  coord_flip() +
  scale_y_log10() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Significantly enriched VFs (Domestic vs Human)",
       y = "Odds ratio",
       x = "Annotation") +
  theme_minimal()


ggplot(top_wild_sig,
       aes(x = reorder(annotation, OR_wild),
           y = OR_wild)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower_wild, ymax = upper_wild),
                width = 0.2) +
  coord_flip() +
  scale_y_log10() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(title = "Significantly enriched VFs (Wild vs Human)",
       y = "Odds ratio",
       x = "Annotation") +
  theme_minimal()
```
```{r vfs enriched in both}
#vfs enriched in domestic and wild vs human

vf_both <- vf_plot_df %>%
  dplyr::filter(
    lower_domestic > 1,
    lower_wild > 1
  )
nrow(vf_both)

top_both <- vf_both %>%
  dplyr::arrange(desc(OR_domestic + OR_wild)) %>%
  dplyr::slice_head(n = 20)



vf_both_long <- top_both %>%
  dplyr::select(annotation,
                OR_domestic, lower_domestic, upper_domestic,
                OR_wild, lower_wild, upper_wild) %>%
  pivot_longer(
    cols = -annotation,
    names_to = c(".value", "host"),
    names_pattern = "(OR|lower|upper)_(.*)"
  )

ggplot(vf_both_long,
       aes(x = reorder(annotation, OR),
           y = OR,
           color = host)) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.6),
                width = 0.2) +
  coord_flip() +
  scale_y_log10() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(
    title = "VFs enriched in BOTH domestic and wild (vs human)",
    y = "Odds ratio",
    x = "Annotation"
  ) +
  theme_minimal()

```
```{r plotting vfs again}

vf_plot_df <- vf_plot_df %>%
  dplyr::mutate(
    z_domestic = beta_domestic / se_domestic,
    p_domestic = 2 * pnorm(-abs(z_domestic)),

    z_wild = beta_wild / se_wild,
    p_wild = 2 * pnorm(-abs(z_wild)),

    z_dw = beta_dw / se_dw,
    p_dw = 2 * pnorm(-abs(z_dw))
  )

vf_plot_df <- vf_plot_df %>%
  dplyr::mutate(
    padj_domestic = p.adjust(p_domestic, method = "BH"),
    padj_wild     = p.adjust(p_wild, method = "BH"),
    padj_dw       = p.adjust(p_dw, method = "BH")
  )

vf_plot_df <- vf_plot_df %>%
  dplyr::mutate(
    sig_domestic = padj_domestic < 0.05,
    sig_wild     = padj_wild < 0.05,
    sig_dw       = padj_dw < 0.05
  )

```

```{r plottingtogether}
plot_domestic <- vf_plot_df %>%
  dplyr::filter(sig_domestic) %>%
  dplyr::mutate(
    logOR = log(OR_domestic),
    lower_log = log(lower_domestic),
    upper_log = log(upper_domestic),
    comparison = "Domestic vs Human"
  )

plot_wild <- vf_plot_df %>%
  dplyr::filter(sig_wild) %>%
  dplyr::mutate(
    logOR = log(OR_wild),
    lower_log = log(lower_wild),
    upper_log = log(upper_wild),
    comparison = "Wild vs Human"
  )

plot_all <- dplyr::bind_rows(plot_domestic, plot_wild)

ggplot(plot_all,
       aes(x = logOR,
           y = reorder(annotation, logOR),
           color = comparison)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lower_log, xmax = upper_log),
                 height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Host-associated virulence factors",
    x = "Log odds ratio (0 = no difference)",
    y = "Annotation"
  ) +
  theme_minimal()

```
```{r more plottn}
ggplot(plot_all,
       aes(x = logOR,
           y = reorder(annotation, logOR))) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = lower_log, xmax = upper_log),
                 height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~comparison, scales = "free_y") +
  labs(
    title = "Host-specific enrichment (significant VFs only)",
    x = "Log odds ratio",
    y = "Annotation"
  ) +
  theme_minimal()
```

## kegg-per-spceis
```{r kegg per species}
kegg_models <- readRDS("data/glmm_results/kegg/kegg_glmmTMB_models_by_species.rds")
kegg_host <- read_csv("data/glmm_results/kegg/kegg_host_effects_by_species.csv")
kegg_annotation_random <- read_csv("data/glmm_results/kegg/kegg_annotation_random_slopes.csv")
```



```{r load kegg host}
kegg_host
```

```{r checkrandomkegg}
head(kegg_annotation_random )


ggplot(kegg_annotation_random, aes(x = host_type_2domestic)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ species, scales = "free") +
  labs(
    title = "Distribution of annotation-level random slopes",
    x = "Random slope (host_type effect deviation)",
    y = "Count"
  )
ggplot(kegg_annotation_random, aes(x = host_type_2wild)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ species, scales = "free") +
  labs(
    title = "Distribution of annotation-level random slopes",
    x = "Random slope (host_type effect deviation)",
    y = "Count"
  )
```


```{r forests, fig.height = 10, fig.width =15 }
kegg_fixed_wild <- kegg_host %>%
  dplyr::filter(term == "host_type_2wild") %>%
  dplyr::select(species, fixed_wild = estimate)



kegg_or <- kegg_annotation_random %>%
  dplyr::left_join(kegg_fixed_wild, by = "species") %>%
  dplyr::mutate(
    log_odds_wild = fixed_wild + host_type_2wild,
    OR_wild_vs_human = exp(log_odds_wild)
  )

kegg_top <- kegg_or %>%
  dplyr::mutate(diff_strength = abs(log(OR_wild_vs_human))) %>%
  dplyr::group_by(species) %>%
  dplyr::slice_max(diff_strength, n = 15) %>%
  dplyr::ungroup()


ggplot(
  kegg_top,
  aes(
    x = OR_wild_vs_human,
    y = reorder(annotation, OR_wild_vs_human)
  )
) +
  geom_point() +
  facet_wrap(~ species, scales = "free_y") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    title = "Top KEGG annotations differing between wild and human",
    x = "Odds ratio (wild vs human, log scale)",
    y = "KEGG annotation"
  )

```


```{r kegg wild effects}
kegg_wild_effects <- purrr::imap_dfr(
  kegg_models,
  ~ {
    re <- ranef(.x)$cond$annotation %>% as.data.frame()
    
    fixed <- kegg_host %>%
      filter(term == "host_type_2wild", species == .y) %>%
      pull(estimate)
    
    tibble(
      annotation = rownames(re),
      log_odds = fixed + re$host_type_2wild,
      species = .y
    )
  }
)

kegg_meta <- kegg_wild_effects %>%
  group_by(annotation) %>%
  summarise(
    mean_log_odds = mean(log_odds, na.rm = TRUE),
    sd_log_odds = sd(log_odds, na.rm = TRUE),
    n_species = n(),
    consistency = mean(log_odds > 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    OR = exp(mean_log_odds)
  )

kegg_top <- kegg_meta %>%
  filter(n_species >= 3) %>%   # keep reliable ones
  mutate(strength = abs(mean_log_odds)) %>%
  arrange(desc(strength)) %>%
  slice_head(n = 30)



ggplot(
  kegg_top,
  aes(
    x = OR,
    y = reorder(annotation, OR),
    color = consistency
  )
) +
  geom_point(size = 2) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0.5
  ) +
  labs(
    title = "Consistent KEGG enrichment across species (wild vs human)",
    x = "Odds ratio (log scale)",
    y = "KEGG annotation",
    color = "Consistency\n(>0 = wild enriched)"
  )


kegg_top_2 <- kegg_meta %>%
  filter(n_species >= 3) %>%   # keep reliable ones
  mutate(strength = abs(mean_log_odds)) %>%
  arrange(desc(strength)) %>%
  slice_head(n = 50)


ggplot(
  kegg_top_2,
  aes(
    x = OR,
    y = reorder(annotation, OR),
    color = consistency
  )
) +
  geom_point(size = 2) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0.5
  ) +
  labs(
    title = "Consistent KEGG enrichment across species (wild vs human)",
    x = "Odds ratio (log scale)",
    y = "KEGG annotation",
    color = "Consistency\n(>0 = wild enriched)"
  )

```

