# Functional analysis

```{r load data funct}
load("data/intermediate_analysis/data.Rdata")


analysis_results <- readRDS("data/intermediate_analysis/functional_analysis_results.rds")

permanova_all      <- analysis_results$permanova
betadisper_all     <- analysis_results$betadisper
kegg_fisher_total  <- analysis_results$fisher$kegg
amr_fisher_total   <- analysis_results$fisher$amr
vf_fisher_total    <- analysis_results$fisher$vf
cazy_fisher_total  <- analysis_results$fisher$cazy



functional_results <- readRDS( "data/intermediate_analysis/fisher_results.rds")



kegg_annot_df <- functional_results$kegg
amr_df        <- functional_results$amr
vf_df_annot   <- functional_results$vf
cazy_annot    <- functional_results$cazy
```



```{r total_genome_annotations}
total_genome_annotations <- purrr::map_df(
  names(species_data),
  ~ species_data[[.x]]$annotations %>%
      dplyr::select(-start, -end) %>%   # remove problematic columns
      dplyr::mutate(bacterial_species = .x)
)
```

```{r vf_data}
vf_data <- total_genome_annotations %>% dplyr::select(vf, vf_type) %>% unique()
```




```{r testing kegg count between hosts}
#count KEGG annotations per MAG within each species
kegg_counts_meta <- total_genome_annotations %>%
  dplyr::filter(!is.na(kegg))%>%
  dplyr::left_join(
    total_genome_metadata %>%
      dplyr::select(ID, host_type),         
    join_by(genome == ID)
  ) %>%
  dplyr::group_by(bacterial_species, genome, host_type)%>%
  dplyr::summarise(n_kegg = dplyr::n(), .groups = "drop") %>%
  filter(!is.na(host_type))

# Run test per species
kegg_host_tests <- kegg_counts_meta %>%
  split(.$bacterial_species) %>%
  map_df(function(df) {
    if (length(unique(df$host_type)) < 2) {
      return(tibble(
        bacterial_species = unique(df$bacterial_species),
        p_value = NA,
        test = "Not enough groups"
      ))
    }
    wt <- wilcox.test(n_kegg ~ host_type, data = df)

    tibble(
      species = unique(df$bacterial_species),
      p_value = wt$p.value,
      median_human  = median(df$n_kegg[df$host_type == "human"]),
      median_animal = median(df$n_kegg[df$host_type == "animal"]),
      test = "Wilcoxon rank-sum"
    )
  })


kegg_host_tests

```


```{r kegg_hosts_p }
kegg_host_tests %>% filter(p_value < 0.05)
```




### PERMANOVA
```{r functional analysis permanova pcoa}
## ============================================================
## Host contrasts (pairwise PERMANOVA)
## ============================================================

host_contrasts <- list(
  human_vs_wild     = c("human", "wild"),
  wild_vs_domestic  = c("domestic", "wild"),
  human_vs_domestic = c("human", "domestic")
)

## ===========================================================
## Helpers to tidy PERMANOVA / betadisper
## ===========================================================

.extract_permanova_rows <- function(permanova_obj, species, matrix_name) {
  if (is.null(permanova_obj)) return(tibble::tibble())
  df <- as.data.frame(permanova_obj)
  df$term <- rownames(df)
  rownames(df) <- NULL
  df$species <- species
  df$matrix_name <- matrix_name
  tibble::as_tibble(df)
}

.extract_betadisper_rows <- function(betadisper_obj, species, matrix_name) {
  if (is.null(betadisper_obj)) return(tibble::tibble())

  an <- as.data.frame(anova(betadisper_obj))
  an$term <- rownames(an)
  rownames(an) <- NULL

  an$species     <- species
  an$matrix_name <- matrix_name
  an$type        <- "anova"

  tibble::as_tibble(an)
}

## ============================================================
## Accumulators
## ============================================================

permanova_rows  <- list()
betadisper_rows <- list()

## ============================================================
## MAIN LOOP
## ============================================================

for (sp in species_list) {

  message("Processing species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  ## ------------------------------------------------------------
  ## Build functional matrices
  ## ------------------------------------------------------------

  mats <- list(
    KEGG = make_annotation_matrix(genome_annotations, "kegg"),
    AMR  = make_annotation_matrix(genome_annotations, "resistance_target"),
    VF   = make_annotation_matrix(genome_annotations, "vf"),
    CAZY = make_annotation_matrix(genome_annotations, "cazy")
  )

  ## ============================================================
  ## PERMANOVA: pairwise host contrasts
  ## ============================================================

  for (contrast_name in names(host_contrasts)) {

    host_levels <- host_contrasts[[contrast_name]]

    for (nm in names(mats)) {

      obj <- tryCatch(
        pa_permanova(
          mats[[nm]],
          genome_metadata,
          host_levels = host_levels
        ),
        error = function(e) {
          message("[", sp, " | ", nm, " | ", contrast_name, "] failed: ", e$message)
          NULL
        }
      )

      if (is.null(obj)) next

      ## ---- PERMANOVA ------------------------------------------
      pr <- .extract_permanova_rows(obj$permanova, sp, nm)
      if (nrow(pr) > 0) {
        pr$contrast <- contrast_name
        permanova_rows[[length(permanova_rows) + 1]] <- pr
      }

      ## ---- Betadisper -----------------------------------------
      br <- .extract_betadisper_rows(obj$betadisper_raw, sp, nm)
      if (nrow(br) > 0) {
        br$contrast <- contrast_name
        betadisper_rows[[length(betadisper_rows) + 1]] <- br
      }
    }
  }

 
}

## ============================================================
## Bind results
## ============================================================

permanova_all  <- dplyr::bind_rows(permanova_rows)
betadisper_all <- dplyr::bind_rows(betadisper_rows)

## Quick check
print(permanova_all, n = 20)
print(betadisper_all, n = 20)


```

```{r pcoas, eval = FALSE}
 ## ========================================
  ## PCoA: One PCoA per matrix (all hosts)
  ## =======================================

  for (nm in names(mats)) {

    obj_all <- tryCatch(
      pa_permanova(
        mats[[nm]],
        genome_metadata,
        host_levels = NULL   # <- ALL hosts
      ),
      error = function(e) NULL
    )

    if (is.null(obj_all)) next

    dist <- obj_all$dist
    meta <- obj_all$metadata

    ## ---- PCoA ------------------------------------------------
    pcoa_res <- ape::pcoa(dist, correction = "cailliez")

    # Number of available axes
    n_axes <- ncol(pcoa_res$vectors)
    
    message("[", sp, " | ", nm, "] PCoA axes available: ", n_axes)
    
    if (n_axes < 2) {
      message("[", sp, " | ", nm, "] Only one PCoA axis; skipping ordination plot")
      next
    }
    
    # If we are here, PC1 and PC2 DEFINITELY exist
    var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
    
    pcoa_df <- data.frame(
      PC1 = pcoa_res$vectors[, 1],
      PC2 = pcoa_res$vectors[, 2],
      host_type_2 = meta$host_type_2,
      stringsAsFactors = FALSE
    )
    
    p <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_2)) +
      geom_point(size = 2, alpha = 0.8) +
      theme_minimal() +
      labs(
        title = paste("PCoA of", nm, "matrix –", sp),
        x = paste0("PC1 (", var_exp[1], "%)"),
        y = paste0("PC2 (", var_exp[2], "%)")
      )
    
    print(p)
  }
```



```{r allpermanova_host_type}
permaniva_sig <- permanova_all %>%filter(term == "3") %>% dplyr::select(-Df, -SumOfSqs)

permaniva_sig
```

```{r tree_order_assigning}
tree_data <- ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

```



```{r figure_permanova, fig.height = 4, fig.width = 10}


permaniva_sig_hw <- permaniva_sig %>%
  mutate(species = species %>%
           str_replace("^enterococcus_hirae$", "enterococcus_b_hirae"))

permaniva_sig_hw <- permaniva_sig_hw %>%
  mutate(
    contrast = factor(
      contrast,
      levels = c("human_vs_wild", "wild_vs_domestic", "human_vs_domestic"),
      labels = c("H vs W", "W vs D", "H vs D")
    ),
    matrix_name = factor(
      matrix_name,
      levels = c("KEGG", "AMR", "VF", "CAZY")
    )
  )



tree_data <- ggtree(pruned)$data
tree_order <- tree_data %>%
  filter(isTip) %>%
  arrange(y) %>%
  pull(label)

permaniva_sig_hw <- permaniva_sig_hw %>%
  mutate(species = factor(species, levels = tree_order))


p_tree <- ggtree(pruned) +
  theme_tree2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p_dot_hw <- permaniva_sig_hw %>%
  mutate(p_sig = ifelse(`Pr(>F)` <= 0.05, `Pr(>F)`, NA_real_)) %>%
  ggplot(aes(
    x = matrix_name,
    y = species,
    color = p_sig,
    size  = R2
  )) +

  geom_point() +

  facet_grid(
    . ~ contrast,
    scales = "free_x",
    space  = "free_x"
  ) +

  scale_color_viridis_c(
    limits = c(0, 0.05),
    option = "plasma",
    na.value = "grey70"
  ) +

  scale_size(
    limits = c(0, 0.20),
    range = c(1.5, 6)
  ) +

  theme_minimal(base_size = 13) +
  labs(
    x = NULL,
    y = NULL,
    size  = expression(R^2),
    color = "Pr(>F) (≤ 0.05 shown)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_blank(),
    panel.spacing.x = unit(1, "lines"),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 11, face = "bold"),
    plot.margin = margin(l = 0)
  )


p_final_hw <- (p_tree | p_dot_hw) +
  plot_annotation(
    title = "PERMANOVA Host type effects"
  )


p_final_hw
```





```{r kegg count analysis}
kegg_per_genome <- total_genome_annotations %>%
  filter(!is.na(kegg)) %>%
  group_by(genome) %>%
  summarise(kegg_count = n(), .groups = "drop")

kegg_summary <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_kegg = mean(kegg_count, na.rm = TRUE),
    median_kegg = median(kegg_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

kegg_test <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

ggplot(kegg_test, aes(x = host_type_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

ggplot(kegg_test, aes(x = host_type_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = bacterial_species), width = 0.2, alpha = 0.6) 

kruskal.test(kegg_count ~ host_type_2, data = kegg_test)

pairwise.wilcox.test(
  kegg_test$kegg_count,
  kegg_test$host_type_2,
  p.adjust.method = "BH"
)

```



```{r amr count analysis}
amr_per_genome <- total_genome_annotations %>%
  filter(resistance_type == "AMR",
    !is.na(resistance_target)) %>%
  group_by(genome) %>%
  summarise(amr_count = n(), .groups = "drop")

amr_summary <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_amr = mean(amr_count, na.rm = TRUE),
    median_amr = median(amr_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

amr_test <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(amr_test, aes(x = host_type_2, y = amr_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

kruskal.test(amr_count ~ host_type_2, data = amr_test)

pairwise.wilcox.test(
  amr_test$amr_count,
  amr_test$host_type_2,
  p.adjust.method = "BH"
)

```



```{r vf_count_analysis}

vf_per_genome <- total_genome_annotations %>%
  filter(!is.na(vf)) %>%
  group_by(genome) %>%
  summarise(vf_count = n(), .groups = "drop")

vf_summary <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_2) %>%
  summarise(
    mean_kegg = mean(vf_count, na.rm = TRUE),
    median_kegg = median(vf_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

vf_test <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(vf_test, aes(x = host_type_2, y = vf_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()


kruskal.test(vf_count ~ host_type_2, data = vf_test)

pairwise.wilcox.test(
  vf_test$vf_count,
  vf_test$host_type_2,
  p.adjust.method = "BH"
)
```




```{r amr different per group}
amr_presence <- total_genome_annotations %>%
  filter(resistance_type == "AMR", !is.na(resistance_target)) %>%
  distinct(genome, resistance_target) %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

amr_sets <- amr_presence %>%
  group_by(host_type_2) %>%
  summarise(amr_list = list(unique(resistance_target)))

amr_freq <- amr_presence %>%
  group_by(host_type_2, resistance_target) %>%
  summarise(n = n(), .groups = "drop")

chisq.test(table(amr_presence$resistance_target, amr_presence$host_type_2))

human_amr     <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "human")]]
domestic_amr  <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "domestic")]]
wild_amr      <- amr_sets$amr_list[[which(amr_sets$host_type_2 == "wild")]]

Reduce(intersect, list(human_amr, domestic_amr, wild_amr))

human_only <- setdiff(human_amr, union(domestic_amr, wild_amr))
domestic_only <- setdiff(domestic_amr, union(human_amr, wild_amr))
wild_only <- setdiff(wild_amr, union(human_amr, domestic_amr))

human_only
domestic_only
wild_only

amr_binary <- amr_presence %>%
  distinct(host_type_2, resistance_target) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = host_type_2,
    values_from = present,
    values_fill = 0
  )

UpSetR::upset(
  as.data.frame(amr_binary[, -1]),  # remove resistance_target column
  sets = c("human", "domestic", "wild"),
  order.by = "freq"
)



```





### Fisher
```{r functional analysis fisher}
# Prepare empty containers
kegg_fisher_all <- list()
amr_fisher_all  <- list()
vf_fisher_all   <- list()
cazy_fisher_all <- list()

# Volcano plots (nested by contrast)
volcano_plots <- list(
  kegg = list(),
  amr  = list(),
  vf   = list(),
  cazy = list()
)

ann_list <- vector("list", length(species_list))
names(ann_list) <- species_list


for (sp in species_list) {

  message("Running Fisher tests for species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  ann_list[[sp]] <- dplyr::mutate(genome_annotations, species = sp)

  # Build matrices
  mats <- list(
    kegg = make_annotation_matrix(genome_annotations, "kegg"),
    amr  = make_annotation_matrix(genome_annotations, "resistance_target"),
    vf   = make_annotation_matrix(genome_annotations, "vf"),
    cazy = make_annotation_matrix(genome_annotations, "cazy")
  )

  for (contrast_name in names(host_contrasts)) {

    host_levels <- host_contrasts[[contrast_name]]

    # Subset metadata for the contrast
    meta_sub <- genome_metadata %>%
      dplyr::filter(host_type_2 %in% host_levels) %>%
      dplyr::mutate(host_type_2 = droplevels(factor(host_type_2)))

    # Fisher requires exactly 2 groups
    if (nlevels(meta_sub$host_type_2) != 2) {
      message("[", sp, " | ", contrast_name, "] skipped: not exactly 2 host groups")
      next
    }

    for (nm in names(mats)) {

      fisher_res <- tryCatch(
        run_fisher_tests_general(
          mat = mats[[nm]],
          metadata = meta_sub,
          group_col = "host_type_2",
          annotation_name = nm
        ),
        error = function(e) {
          message("[", sp, " | ", nm, " | ", contrast_name, "] Fisher failed: ", e$message)
          NULL
        }
      )

      if (is.null(fisher_res) || nrow(fisher_res) == 0) next

      # Add identifiers
      fisher_res$species  <- sp
      fisher_res$contrast <- contrast_name

      # Accumulate
      if (nm == "kegg") kegg_fisher_all[[length(kegg_fisher_all) + 1]] <- fisher_res
      if (nm == "amr")  amr_fisher_all[[length(amr_fisher_all) + 1]]  <- fisher_res
      if (nm == "vf")   vf_fisher_all[[length(vf_fisher_all) + 1]]   <- fisher_res
      if (nm == "cazy") cazy_fisher_all[[length(cazy_fisher_all) + 1]] <- fisher_res

      # Volcano plot (optional)
      volcano_plots[[nm]][[paste(sp, contrast_name, sep = "_")]] <-
        plot_volcano(fisher_res, nm)
    }
  }
}



```





```{r bind the rows}
kegg_fisher_total        <- dplyr::bind_rows(kegg_fisher_all)
amr_fisher_total         <- dplyr::bind_rows(amr_fisher_all)
vf_fisher_total          <- dplyr::bind_rows(vf_fisher_all)
cazy_fisher_total        <- dplyr::bind_rows(cazy_fisher_all)

```



```{r save functional_results}
# analysis_results <- list(
#   permanova      = permanova_all,
#   betadisper     = betadisper_all,
#   fisher = list(
#     kegg = kegg_fisher_total,
#     amr  = amr_fisher_total,
#     vf   = vf_fisher_total,
#     cazy = cazy_fisher_total
#   ),
#   host_contrasts = host_contrasts
# )
# 
# saveRDS(
#   analysis_results,
#   file = "data/intermediate_analysis/functional_analysis_results.rds"
# )
# 



# analysis_results <- readRDS("data/intermediate_analysis/functional_analysis_results.rds")
# 
# permanova_all      <- analysis_results$permanova
# betadisper_all     <- analysis_results$betadisper
# kegg_fisher_total  <- analysis_results$fisher$kegg
# amr_fisher_total   <- analysis_results$fisher$amr
# vf_fisher_total    <- analysis_results$fisher$vf
# cazy_fisher_total  <- analysis_results$fisher$cazy

```




```{r testing keggs totals, fig.height = 10, fig.width = 20}
prepare_volcano_df <- function(df, tree_order) {
  df %>%
    mutate(
      ##Fix enterococcus hirae name mismatch
      species = dplyr::case_when(
        species == "enterococcus_hirae" ~ "enterococcus_b_hirae",
        TRUE ~ species
      ),

      ## VOLCANO METRICS
      neg_log10_padj = -log10(pmax(p_adj, 1e-300)),
      sig = p_adj <= 0.05,

      ## order factors
      species = factor(species, levels = tree_order),
      contrast = factor(
        contrast,
        levels = c(
          "human_vs_wild",
          "wild_vs_domestic",
          "human_vs_domestic"
        ),
        labels = c(
          "HumanvsWild",
          "WildvsDomestic",
          "HumanvsDomestic"
        )
      )
    )
}



plot_volcano_contrasts <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = sig, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = c(
        `TRUE`  = "#0072B2",  # significant
        `FALSE` = "grey75"    # not significant
      ),
      name = "Adj. p ≤ 0.05"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.4),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log[10]("adj. p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}


get_top_hits <- function(df, feature_col, n = 5) {
  df %>%
    filter(sig) %>%                       # only significant features
    group_by(species, contrast) %>%       # per species × contrast
    arrange(p_adj) %>%                   
    slice_head(n = n) %>%                
    ungroup()
}
    
    
    
kegg_df <- prepare_volcano_df(
  kegg_fisher_total,
  tree_order = tree_order
)

kegg_top <- get_top_hits(kegg_df, n = 5)

plot_volcano_contrasts(
  kegg_df,
  title_prefix = "KEGG enrichment"
) +
  ggrepel::geom_text_repel(
    data = kegg_top,
    aes(label = kegg),
    size = 3,
    max.overlaps = 8
  )
```


```{r fisher bacteroides uniformis, fig.height = 8, fig.width = 6}
species_of_interest <- "bacteroides_uniformis"

kegg_df_sp <- kegg_df %>%
  filter(species == species_of_interest)

kegg_top_sp <- get_top_hits(kegg_df_sp, "kegg")

plot_volcano_contrasts(
  kegg_df_sp,
  title_prefix = "KEGG enrichment – Bacteroides uniformis"
) +
  ggrepel::geom_text_repel(
    data = kegg_top_sp,
    aes(label = kegg),
    size = 3
  )
```


```{r amr fisher,  fig.height = 10, fig.width = 15}
amr_df <- prepare_volcano_df(
  amr_fisher_total,
  tree_order = tree_order
)

amr_top <- get_top_hits(amr_df, n = 5)

plot_volcano_contrasts(
  amr_df,
  title_prefix = "AMR enrichment"
) +
  ggrepel::geom_text_repel(
    data = amr_top,
    aes(label = amr),
    size = 3
  )
```



```{r vf fisher,  fig.height = 10, fig.width = 15}
vf_df <- prepare_volcano_df(
  vf_fisher_total,
  tree_order = tree_order
)

vf_top <- get_top_hits(vf_df, n = 5)

plot_volcano_contrasts(
  vf_df,
  title_prefix = "VF enrichment"
) +
  ggrepel::geom_text_repel(
    data = vf_top,
    aes(label = vf),
    size = 3
  )
```


```{r cazy fisher,  fig.height = 10, fig.width = 15}
cazy_df <- prepare_volcano_df(
  cazy_fisher_total,
  tree_order = tree_order
)

cazy_top <- get_top_hits(cazy_df, n = 5)

plot_volcano_contrasts(
  cazy_df,
  title_prefix = "CAZy enrichment"
) +
  ggrepel::geom_text_repel(
    data = cazy_top,
    aes(label = cazy),
    size = 3
  )
```

```{r volcano by pathways}
#keep only significant KOs 
sig_df <- kegg_df %>%
  filter(p_adj < 0.05) %>%
  mutate(ko = paste0("ko:", kegg))   # (when you download the table they are like this)

# KO -> pathway links (only for sig KOs)
ko_links <- KEGGREST::keggLink("pathway", "ko") %>%
  tibble::enframe(name = "ko", value = "pathway") %>%
  semi_join(sig_df, by = "ko") %>%
  mutate(
    pathway_map = sub("^path:(?:ko|map)(\\d{5})$", "path:map\\1", pathway)
  ) %>%
  filter(str_detect(pathway_map, "^path:map\\d{5}$")) %>%
  distinct(ko, pathway_map)

#  One pathway per KO 
ko_links_one <- ko_links %>%
  group_by(ko) %>%
  slice(1) %>%
  ungroup()

# Fetch pathway info 
paths <- unique(ko_links_one$pathway_map)

get_info <- function(pid){
  x <- try(KEGGREST::keggGet(pid)[[1]], silent = TRUE)
  if (inherits(x, "try-error")) return(tibble(pathway_map=pid, pathway_name=NA, class=NA))
  tibble(
    pathway_map = pid,
    pathway_name = x$NAME[[1]],
    class = ifelse(is.null(x$CLASS), NA, x$CLASS[[1]])
  )
}

path_info <- bind_rows(lapply(paths, get_info))

# high-level category
path_info <- path_info %>%
  mutate(category = sub(";.*$", "", class))



```

```{r pathway kegg mapping}
#join annotations to kegg fisher results
ko_links_one_clean <- ko_links_one %>%
  mutate(ko_clean = sub("^ko:", "", ko))

kegg_annot_df <- kegg_df %>%
  left_join(ko_links_one_clean, by = c("kegg" = "ko_clean")) %>%
  left_join(path_info, by = "pathway_map")

```



```{r volcano plot colorful, fig.height=6, fig.width = 15,}

kegg_categories <- sort(unique(na.omit(kegg_annot_df$category)))

kegg_colors <- c(
  NS = "grey75",
  setNames(
    RColorBrewer::brewer.pal(length(kegg_categories), "Set2"),
    kegg_categories
  )
)

plot_kegg_volcano_by_pathway <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(
        color = ifelse(sig, category, "NS"),
        alpha = sig
      ),
      size = 1.6
    ) +
    scale_color_manual(
      values = kegg_colors,
      name = "KEGG class"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}

plot_kegg_volcano_by_pathway(
  kegg_annot_df,
  title_prefix = "KEGG enrichment by pathway"
)

```


```{r bacteroides colored, fig.height = 10, fig.width = 7}
kegg_annot_df %>%
  filter(species == "phocaeicola_vulgatus") %>%
  plot_kegg_volcano_by_pathway(
    title_prefix = "KEGG enrichment by pathway – Phocaeicola vulgatus"
  )
```



### Virulence factors

```{r vf_volcano, fig.height = 5, fig.width = 15}

vf_df <- prepare_volcano_df(
  vf_fisher_total,
  tree_order = tree_order
)

vf_df_annot <- vf_df %>%
  left_join(vf_data, by = "vf") %>%
  mutate(
    color_group = ifelse(sig, vf_type, "nonsig")
  )

vf_types <- sort(unique(na.omit(vf_df_annot$vf_type)))

vf_colors <- c(
  nonsig = "grey70",
  setNames(
    RColorBrewer::brewer.pal(
      min(8, length(vf_types)),
      "Set1"
    ),
    vf_types[seq_len(min(8, length(vf_types)))]
  )
)


plot_vf_volcano_by_type <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = color_group, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = vf_colors,
      name = "VF type"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}


plot_vf_volcano_by_type(
  vf_df_annot,
  title_prefix = "VF enrichment by VF type"
)

```
```{r bacteroides vf, fig.height = 9, fig.width = 6}
vf_df_annot %>%
  filter(species == "phocaeicola_vulgatus") %>%
  plot_vf_volcano_by_type(
    title_prefix = "VF enrichment by type – Phocaeicola vulgatus"
  )
```

##### CAZY
```{r cazy data}
faminfo <- read_excel("data/FamInfo.txt.04232020.xls")
activities <- read_tsv("data/CAZyDB.07302020.fam-activities.txt", comment = "#",
                       col_names = c("family", "activity"))
substrates <- read_tsv("data/fam-substrate-mapping-08012023.tsv")

```

```{r cazy annot}
cazy_df <- prepare_volcano_df(
  cazy_fisher_total,
  tree_order = tree_order
)


cazy_parsed <- cazy_df %>%
  mutate(
    cazy_clean = stringr::str_trim(cazy),

    # CAZy class: GH, GT, PL, CE, AA, CBM
    cazy_class = stringr::str_extract(cazy_clean, "^[A-Za-z]+"),

    # Family number
    remainder = stringr::str_remove(cazy_clean, "^[A-Za-z]+"),
    cazy_family = stringr::str_extract(remainder, "^\\d+"),

    # Canonical family ID (e.g. GH13, PL7)
    cazy_family_id = paste0(cazy_class, cazy_family)
  )



cazy_annot <- cazy_parsed %>%
  left_join(substrates, by = c("cazy_family_id" = "Family"))
```



```{r cazy volcano coloured, fig.height = 8, fig.width = 14}


cazy_annot <- cazy_annot %>%
  mutate(
    color_group = ifelse(sig, cazy_class, "nonsig")
  )

cazy_types <- sort(unique(na.omit(cazy_annot$cazy_class)))

cazy_colors <- c(
  nonsig = "grey70",
  setNames(
    RColorBrewer::brewer.pal(
      min(8, length(cazy_types)),
      "Set2"
    ),
    cazy_types[seq_len(min(8, length(cazy_types)))]
  )
)



plot_cazy_volcano_by_class <- function(df, title_prefix) {

  ggplot(df, aes(x = log2_or, y = neg_log10_padj)) +
    geom_hline(
      yintercept = -log10(0.05),
      linetype = "dashed",
      color = "grey50"
    ) +
    geom_vline(
      xintercept = 0,
      linetype = "dotted",
      color = "grey60"
    ) +
    geom_point(
      aes(color = color_group, alpha = sig),
      size = 1.6
    ) +
    scale_color_manual(
      values = cazy_colors,
      name = "CAZy class"
    ) +
    scale_alpha_manual(
      values = c(`TRUE` = 1, `FALSE` = 0.35),
      guide = "none"
    ) +
    labs(
      x = "log2(odds ratio)",
      y = expression(-log("adj p-value")),
      title = title_prefix
    ) +
    theme_minimal(base_size = 12) +
    facet_grid(
      contrast ~ species
    )
}



cazy_top <- cazy_annot %>%
  filter(sig) %>%
  group_by(species, contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

plot_cazy_volcano_by_class(
  cazy_annot,
  title_prefix = "CAZy enrichment by class"
) +
  ggrepel::geom_text_repel(
    data = cazy_top,
    aes(label = cazy),
    size = 3,
    max.overlaps = 8
  )




plot_cazy_volcano_by_class(
  cazy_annot,
  title_prefix = "CAZy enrichment by class"
)


```


```{r bacteroides_uniformis_cazy}
cazy_sp <- cazy_annot %>%
  filter(species == "bacteroides_uniformis")

plot_cazy_volcano_by_class(
  cazy_sp,
  title_prefix = "CAZy enrichment by class – Bacteroides uniformis"
)

```

```{r one species all comparisons, fig.height = 8, fig.width = 8}
species_of_interest <- "bacteroides_uniformis"

kegg_df_sp <- kegg_annot_df %>%
  filter(species == species_of_interest)

kegg_top_sp <- kegg_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_kegg <- plot_kegg_volcano_by_pathway(
  kegg_df_sp,
  title_prefix = paste("KEGG –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = kegg_top_sp,
    aes(label = kegg),
    size = 3,
    max.overlaps = 8
  )

p_kegg

amr_df_sp <- amr_df %>%
  filter(species == species_of_interest)

amr_top_sp <- amr_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_amr <- plot_volcano_contrasts(
  amr_df_sp,
  title_prefix = paste("AMR –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = amr_top_sp,
    aes(label = amr),
    size = 3
  )

p_amr


vf_df_sp <- vf_df_annot %>%
  filter(species == species_of_interest)

vf_top_sp <- vf_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_vf <- plot_vf_volcano_by_type(
  vf_df_sp,
  title_prefix = paste("VF –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = vf_top_sp,
    aes(label = vf),
    size = 3
  )

p_vf


cazy_df_sp <- cazy_annot %>%
  filter(species == species_of_interest)

cazy_top_sp <- cazy_df_sp %>%
  filter(sig) %>%
  group_by(contrast) %>%
  arrange(p_adj) %>%
  slice_head(n = 5) %>%
  ungroup()

p_cazy <- plot_cazy_volcano_by_class(
  cazy_df_sp,
  title_prefix = paste("CAZy –", species_of_interest)
) +
  ggrepel::geom_text_repel(
    data = cazy_top_sp,
    aes(label = cazy),
    size = 3
  )

p_cazy



```
```{r patchworking, fig.height = 8, fig.width = 12}
small_legend <- theme(
  legend.text = element_text(size = 7),
  legend.title = element_text(size = 8),
  legend.key.size = unit(0.35, "cm"),
  legend.spacing.y = unit(0.1, "cm")
)

p_kegg <- p_kegg +
  labs(title = "KEGG") +
  small_legend

p_amr <- p_amr +
  labs(title = "AMR") +
  small_legend

p_vf <- p_vf +
  labs(title = "VF") +
  small_legend

p_cazy <- p_cazy +
  labs(title = "CAZy") +
  small_legend

(p_kegg | p_amr | p_vf | p_cazy) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.35, "cm")
  )

```
```{r all species in a loop,  fig.height = 8, fig.width = 13}

kegg_annot_df <- kegg_annot_df %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

amr_df <- amr_df %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

vf_df_annot <- vf_df_annot %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))

cazy_annot <- cazy_annot %>%
  mutate(species = recode(
    species,
    "enterococcus_b_hirae" = "enterococcus_hirae"
  ))


make_functional_patchwork <- function(species_of_interest) {

  ## ---------- KEGG ----------
  kegg_df_sp <- kegg_annot_df %>%
    filter(species == species_of_interest)

  kegg_top_sp <- kegg_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_kegg <- plot_kegg_volcano_by_pathway(
    kegg_df_sp,
    title_prefix = paste("KEGG –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = kegg_top_sp,
      aes(label = kegg),
      size = 3,
      max.overlaps = 8
    )


  ## ---------- AMR ----------
  amr_df_sp <- amr_df %>%
    filter(species == species_of_interest)

  amr_top_sp <- amr_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_amr <- plot_volcano_contrasts(
    amr_df_sp,
    title_prefix = paste("AMR –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = amr_top_sp,
      aes(label = amr),
      size = 3
    )


  ## ---------- VF ----------
  vf_df_sp <- vf_df_annot %>%
    filter(species == species_of_interest)

  vf_top_sp <- vf_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_vf <- plot_vf_volcano_by_type(
    vf_df_sp,
    title_prefix = paste("VF –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = vf_top_sp,
      aes(label = vf),
      size = 3
    )


  ## ---------- CAZy ----------
  cazy_df_sp <- cazy_annot %>%
    filter(species == species_of_interest)

  cazy_top_sp <- cazy_df_sp %>%
    filter(sig) %>%
    group_by(contrast) %>%
    arrange(p_adj) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_cazy <- plot_cazy_volcano_by_class(
    cazy_df_sp,
    title_prefix = paste("CAZy –", species_of_interest)
  ) +
    ggrepel::geom_text_repel(
      data = cazy_top_sp,
      aes(label = cazy),
      size = 3
    )


  ## ---------- Legend theme ----------
  small_legend <- theme(
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.35, "cm")
  )

    p_kegg <- p_kegg +
      labs(title = "KEGG") +
      small_legend
    
    p_amr <- p_amr +
      labs(title = "AMR") +
      small_legend
    
    p_vf <- p_vf +
      labs(title = "VF") +
      small_legend
    
    p_cazy <- p_cazy +
      labs(title = "CAZy") +
      small_legend


  ## ---------- Final patchwork ----------
  final_plot <-
    (p_kegg | p_amr | p_vf | p_cazy) +
    plot_layout(guides = "collect") &
    theme(
      legend.position = "right",
      plot.title = element_text(size = 14, face = "bold")
    )

  final_plot
}


# patchwork_plots <- purrr::set_names(
#   lapply(species_list, make_functional_patchwork),
#   species_list
# )
# 
# purrr::walk(patchwork_plots, print)
# 


for (sp in species_list) {

  p <- make_functional_patchwork(sp)

  print(p)

  # ggsave(
  #   paste0("plots/functional_enrichment_", sp, ".pdf"),
  #   plot = p,
  #   width = 13,
  #   height = 8
  )
}


```



```{r save functional_results_2}
# fisher_results <- list(
#   kegg = kegg_annot_df,
#   amr  = amr_df,
#   vf   = vf_df_annot,
#   cazy = cazy_annot,
#   species = species_list,
#   info = list(
#     created = Sys.time(),
#     description = "Functional enrichment input tables for volcano plots"
#   )
# )
# 
# saveRDS(
#   fisher_results,
#   file = "data/intermediate_analysis/fisher_results.rds"
# )





# functional_results <- readRDS( "data/intermediate_analysis/fisher_results.rds")
# 
# 
# 
# kegg_annot_df <- functional_results$kegg
# amr_df        <- functional_results$amr
# vf_df_annot   <- functional_results$vf
# cazy_annot    <- functional_results$cazy

```



### ZOOM-IN - quorum sensing
```{r quorum sensing}
qs_df <- kegg_annot_df %>%
  filter(
    pathway_map == "path:map02024"
  ) %>%
  mutate(
    sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
  )


ggplot(qs_df, aes(
  x = log2_or,
  y = ko,
  color = species
)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(alpha = sig), size = 2.5) +
  facet_wrap(~ contrast) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.05)) +
  labs(
    title = "Quorum sensing genes show host-associated enrichment",
    x = "log2(odds ratio)",
    y = "Quorum sensing KOs",
    alpha = "Significant"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 7)
  )

```
```{r top pathways for zoomin, fig.width = 10, fig.height = 7}
top_pathways <- kegg_annot_df %>%
  filter(
    sig,
    !is.na(pathway_map)
  ) %>%
  group_by(pathway_map, pathway_name) %>%
  summarise(
    n_sig = n(),
    mean_abs_log2_or = mean(abs(log2_or), na.rm = TRUE),
    score = n_sig * mean_abs_log2_or,
    .groups = "drop"
  ) %>%
  arrange(desc(score)) %>%
  slice_head(n = 20)

top20_df <- kegg_annot_df %>%
  semi_join(top_pathways, by = "pathway_map") %>%
  mutate(
    sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
  )

top20_summary <- top20_df %>%
  group_by(pathway_name, species, contrast) %>%
  summarise(
    mean_log2_or = mean(log2_or, na.rm = TRUE),
    n_sig = sum(sig),
    .groups = "drop"
  )


ggplot(top20_summary, aes(
  x = mean_log2_or,
  y = pathway_name,
  color = species,
  size = n_sig
)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.8) +
  facet_wrap(~ contrast) +
  labs(
    title = "Top 20 KEGG pathways enriched across host contrasts",
    x = "Mean log2(odds ratio)",
    y = NULL,
    size = "Significant KOs"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 7)
  )
```
```{r for_20_pathways}
plot_kegg_pathway_breakdown <- function(pathway_id, pathway_label) {

  df <- kegg_annot_df %>%
    filter(pathway_map == pathway_id) %>%
    mutate(
      sig_lab = ifelse(sig, "Adj p ≤ 0.05", "n.s.")
    )

  # Guard against empty or tiny pathways
  if (nrow(df) == 0) {
    return(NULL)
  }

  ggplot(df, aes(
    x = log2_or,
    y = ko,
    color = species
  )) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(aes(alpha = sig), size = 2.5) +
    facet_wrap(~ contrast) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.05)) +
    labs(
      title = pathway_label,
      subtitle = pathway_id,
      x = "log2(odds ratio)",
      y = "KEGG ortholog",
      alpha = "Significant"
    ) +
    theme_bw() +
    theme(
      legend.position = "right",
      axis.text.y = element_text(size = 7)
    )
}

plots_by_pathway <- purrr::pmap(
  list(
    top_pathways$pathway_map,
    top_pathways$pathway_name
  ),
  plot_kegg_pathway_breakdown
)

# Drop NULLs just in case
plots_by_pathway <- purrr::compact(plots_by_pathway)

purrr::walk(plots_by_pathway, print)

```



## META-ANALYSIS

### Human vs wild
```{r metaanalysis for all keggs_hw, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 8
min_total_events <- 10

kegg_keep_hw <- df_hw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_hw_screened <- df_hw %>%
  inner_join(kegg_keep_hw, by = "kegg")

# Split by KEGG
kegg_split_hw <- split(df_hw_screened, df_hw_screened$kegg)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0",
    incr =0.5
  )
}

mh_hw <- lapply(kegg_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



# select keggs where confidence interval does not include 1 
#these are consistent in the direction of prevalence
mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```
As most of the identified KOs are ribosomal subunits, we try again, filtering these before the meta-analysis.

```{r metaanalysis for all keggs_remove_ribosomal, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs

#  Low-bias screen 
min_informative_species <- 8
min_total_events <- 10

kegg_keep_hw <- df_hw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_hw_screened <- df_hw %>%
  inner_join(kegg_keep_hw, by = "kegg")

# Split by KEGG
kegg_split_hw <- split(df_hw_screened, df_hw_screened$kegg)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0",
    incr =0.5
  )
}

mh_hw <- lapply(kegg_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



# select keggs where confidence interval does not include 1 
#these are consistent in the direction of prevalence
mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 



ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$kegg[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)


```




Plotting the individual KOs to check if all species do follow same pattern

```{r forest plot individual keggs_human_wild_after_ribosomal, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_kegg <- function(kegg_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[kegg_id]]
  if (is.null(m)) stop("KEGG not found in mh_hw: ", kegg_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("KEGG: ", kegg_id)
  )
}

#individual analyses of each of these keggs:
keggs <- mh_re_table_ci %>% dplyr::pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```

Manual search table of the KOs and information for plotting.

```{r manual lookup of keggs}
kegg_search_info <- tibble::tibble(
  kegg = c(
    
    "K18799","K03436","K01356","K07461","K13573","K03580","K09144",
    "K07497","K03798","K07317","K03427","K06871","K02935","K02864",
    "K07485","K07493","K01183","K22886","K06113","K02965","K02890",
    "K01580","K02886","K02994","K02933","K02982","K02946","K02892",
    "K02926","K02907","K01154","K02954","K03279","K03046","K00571",

    
    "K07459","K03741","K07448","K08884","K08217","K07482","K07455",
    "K18220","K20922","K07339",
    "K21029","K18909","K16566","K00980","K16702","K12453","K06919",
    
    "K18234", "K03076", "K15773", "K00793","K00826","K07040","K03366","K03733","K07487", "K07126",
    "K03783", "K01160", "K02518"
  ),

  description = c(
    
    "O-antigen flippase wzxB",
    "DeoR family transcriptional regulator",
    "Repressor LexA",
    "Putative endonuclease",
    "Proteasome accessory factor C",
    "ATP-dependent helicase HepA",
    "Nucleotidyl transferase AbiEii/AbiGii toxin-family protein",
    "Putative transposase",
    "Cell division protease FtsH",
    "Adenine-specific DNA methyltransferase",
    "Type I restriction enzyme M protein",
    "Uncharacterized protein",
    "Ribosomal protein L7/L12",
    "Ribosomal protein L10",
    "Transposase",
    "Putative transposase",
    "Chitinase",
    "Protein lysine acetyltransferase",
    "Arabinan endo-1,5-alpha-L-arabinosidase",
    "Small subunit ribosomal protein S19",
    "Large subunit ribosomal protein L22",
    "Glutamate decarboxylase",
    "Large subunit ribosomal protein L2",
    "Small subunit ribosomal protein S8",
    "Large subunit ribosomal protein L6",
    "Small subunit ribosomal protein S3",
    "Small subunit ribosomal protein S10",
    "Large subunit ribosomal protein L23",
    "Large subunit ribosomal protein L4",
    "Large subunit ribosomal protein L30",
    "Type I restriction enzyme, S subunit",
    "Small subunit ribosomal protein S14",
    "UDP-glucose:(galactosyl)LPS alpha-1,2-glucosyltransferase",
    "DNA-directed RNA polymerase subunit beta'",
    "Site-specific DNA-methyltransferase",
    "ATP-dependent endonuclease OLD family protein",
    "Arsenate reductase",
    "Type IV restriction system protein Mrr",
    "Serine/threonine protein kinase",
    "Macrolide efflux protein",
    "IS30 family transposase",
    "Recombination protein RecT",
    "Ribosomal protection tetracycline resistance protein",
    "Polysaccharide biosynthesis protein",
    "mRNA interferase toxin HicA",
    "Uncharacterized protein",
    "MarR family transcriptional regulator",
    "Exopolysaccharide biosynthesis protein ExoY",
    "Glycerol-3-phosphate cytidylyltransferase",
    "Exopolysaccharide biosynthesis glycosyltransferase",
    "CDP-paratose synthetase",
    "Putative DNA replication protein",
    "Virginiamycin A acetyltransferase",
    "Preprotein translocase subunit SecY",
    "HTH-type transcriptional regulator / antitoxin HipB",
"Riboflavin synthase","Branched-chain amino acid aminotransferase",
    "DUF177 domain-containing protein (ribosome biogenesis)",
    "Meso-butanediol dehydrogenase",
    "Tyrosine-type recombinase / integrase (XerC)",
    "Transposase",
"uncharacterized protein",
"Purine-nucleoside phosphorylase",
"Crossover junction endodeoxyribonuclease RusA",
"Translation initiation factor IF-1"

  ),

  extra_info = c(
    "",
    "fructose operon transcriptional repressor",
    "serine endopeptidase, SOS response regulator",
    "",
    "pafC",
    "",
    "Type IV TA system toxin",
    "",
    "metallo-endopeptidase",
    "Type II R-M system",
    "adenine-specific DNA methyltransferase",
    "function unknown",
    "large subunit",
    "large subunit",
    "",
    "",
    "",
    "",
    "abnA",
    "RP-S19, RSM19, rpsS",
    "RP-L22, MRPL22, rplV",
    "Early infantile epileptic encephalopathy",
    "ribosome",
    "RP-S8, MRPS8, rpsH",
    "RP-L6, MRPL6, rplF",
    "RP-S3, rpsC",
    "RP-S10, rpsJ",
    "RP-L23, MRPL23, rplW",
    "RP-L4, MRPL4, rplD",
    "RP-L30, MRPL30, rpmD",
    "Type I R-M system specificity subunit",
    "RP-S14, MRPS14, rpsN",
    "waaJ, rfaJ",
    "rpoC",
    "adenine-specific",
    "DNA damage response",
    "arsC, thioredoxin-dependent",
    "restriction-modification system",
    "protein phosphorylation",
    "MFS transporter",
    "mobile genetic element",
    "homologous recombination",
    "tetM / tetO",
    "biofilm-associated",
    "Type II toxin–antitoxin system",
    "function unknown",
    "mepR, antibiotic resistance regulation",
    "exopolysaccharide synthesis",
    "tagD",
    "amsE",
    "rfbS",
    "replication / repair",
    "vat",
    "",
    "",
"[EC:2.5.1.9]",
    "[EC:2.6.1.42]",
    "",
    "[EC:1.1.1.76 / 1.1.1.304]",
    "",
    "","", "", "", ""

    
  )
)


kegg_search_info
```
```{r wild human with kegginfo after ribosomal, fig.height = 6, fig.width = 10}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```








```{r forest plot individual keggs_again, fig.height = 6, fid.width = 10}
#individual analyses of each of these keggs:
keggs <- mh_re_table_ci %>% dplyr::pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```




Testing risk difference
### Risk difference meta-analysis
```{r risk difference}
run_random_hw_rd <- function(dat) {
  meta::metabin(
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,

    sm      = "RD",        #  RISK DIFFERENCE
    method  = "MH",

    common  = FALSE,
    random  = TRUE,

    method.tau       = "REML",
    method.random.ci = "HK",

    prediction = TRUE
  )
}

mh_hw_rd <- lapply(kegg_split_hw, run_random_hw_rd)

mh_hw_rd_table <- bind_rows(lapply(names(mh_hw_rd), function(k) {
  m <- mh_hw_rd[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    RD_random      = m$TE.random,
    se_random      = m$seTE.random,
    ci_low_random  = m$lower.random,
    ci_high_random = m$upper.random,
    p_random       = m$pval.random,
    I2_percent     = m$I2,
    tau2_REML      = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))


mh_hw_rd_ci <- mh_hw_rd_table %>%
  filter(ci_low_random > 0 | ci_high_random < 0)



mh_hw_rd_ci$kegg_I2 <- paste0(
  mh_hw_rd_ci$kegg,
  "  (I²=",
  round(100* mh_hw_rd_ci$I2_percent, 2),
  "%)"
)

mgen_rd <- meta::metagen(
  TE   = mh_hw_rd_ci$RD_random,
  seTE = mh_hw_rd_ci$se_random,
  studlab = mh_hw_rd_ci$kegg_I2,
  sm = "RD",
  common = FALSE,
  random = FALSE
)

meta::forest(
  mgen_rd,
  backtransf = FALSE,
  xlab = "Prevalence difference (Human − Wild)"
)

# meta::forest(
#   mgen_rd,
#   backtransf = FALSE,
#   xlab = "Prevalence difference (Human − Wild)",
#   leftcols = c("studlab", "I2"),
#   leftlabs = c("KEGG", expression(I^2~"%")),
#   col.square = "black",
#   col.diamond = "black",
#   col.diamond.lines = "black"
# )

```





### Domestic vs wild

```{r metaanalysis for all keggs_w_d_removing_ribosomal}
# Prep totals and basic filters 
df_dw <- kegg_fisher_total %>%
  filter(contrast == "wild_vs_domestic") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_domestic = n_mags_domestic + absent_domestic
  ) %>%
  filter(n_mags_domestic > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_domestic > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_domestic< tot_domestic) | (n_mags_wild < tot_wild)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs


#  Low-bias screen 
min_informative_species <- 6
min_total_events <- 10

kegg_keep_dw <- df_dw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_domestic),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_dw_screened <- df_dw %>%
  inner_join(kegg_keep_dw, by = "kegg")

# Split by KEGG
kegg_split_dw <- split(df_dw_screened, df_dw_screened$kegg)

# Effect size pooling
# MH with random effects; domestic vs wild 
run_mh_random_dw <- function(dat) {
  meta::metabin(
    # domestic as experimental (so OR>1 => more prevalent in domestic)
    event.e = dat$n_mags_domestic,
    n.e     = dat$tot_domestic,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_dw <- lapply(kegg_split_dw, run_mh_random_dw)

# Collect pooled results + FDR 
mh_dw_table <- bind_rows(lapply(names(mh_dw), function(k) {
  m <- mh_dw[[k]]
  data.frame(
    kegg = k,
    contrast = "domestic_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))




mh_re_table_ci <- mh_dw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 




ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Domestic vs Wild)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```




```{r wild domestic with kegginfo, fig.height = 6, fig.width = 10}
mh_dw_table_annot <- mh_dw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_dw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)



ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Domestic vs Wild)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```


### Human vs domestic

```{r metaanalysis for all keggs_hd_rr}
# Prep totals and basic filters 
df_dw <- kegg_fisher_total %>%
  filter(contrast == "human_vs_domestic") %>%
  mutate(
    tot_domestic  = n_mags_domestic  + absent_domestic,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_domestic > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_domestic > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_domestic < tot_domestic)) %>% #remove rows where kegg is present in all mags of both groups (at least one group shows variation)
filter(!grepl("^K02[89]", kegg))   # remove ribosomal protein KOs

#  Low-bias screen 
min_informative_species <-6
min_total_events <- 10

kegg_keep_dw <- df_dw %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_domestic + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(kegg)

df_dw_screened <- df_dw %>%
  inner_join(kegg_keep_dw, by = "kegg")

# Split by KEGG
kegg_split_dw <- split(df_dw_screened, df_dw_screened$kegg)

# Effect size pooling
# MH with random effects; human vs domestic 
run_mh_random_dw <- function(dat) {
  meta::metabin(
    # human as experimental (so OR>1 => more prevalent in human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_domestic,
    n.c     = dat$tot_domestic,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_dw <- lapply(kegg_split_dw, run_mh_random_dw)

# Collect pooled results + FDR 
mh_dw_table <- bind_rows(lapply(names(mh_dw), function(k) {
  m <- mh_dw[[k]]
  data.frame(
    kegg = k,
    contrast = "human_vs_domestic",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))


mh_re_table_ci <- mh_dw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Domestic)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```




```{r human domestic with kegginfo_3, fig.height = 6, fig.width = 12}
mh_dw_table_annot <- mh_dw_table %>%
  dplyr::left_join(
    kegg_search_info,
    by = "kegg"
  ) %>%
  dplyr::mutate(
    studlab_desc = ifelse(
      is.na(description),
      kegg,
      paste0(kegg, ": ", description)
    )
  )

mh_re_table_ci <- mh_dw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)
ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_KEGG <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Domestic)",

  leftcols = c("studlab", "I2_KEGG", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```
### Meta-analysis with VF
#### Human vs wild
```{r metaanalysis for all vfs, fig.height = 10, fig.width = 13}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "human_vs_wild") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_human = n_mags_human + absent_human
  )%>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_wild > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_wild < tot_wild)) #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; Human vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

   
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "human_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



mh_re_table_ci <- mh_hw_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$vf,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",
  leftlabs = c("VF"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```

```{r rename vf data_from_befoer}
vf_type_data <- vf_data
```


```{r check vf name}
vf_data <- read_excel("data/VFs.xls")
```


```{r human wild with vfinfo, fig.height = 6, fig.width = 10}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    dplyr::join_by(vf == VFID)
  ) %>%
  dplyr::left_join(
    vf_type_data,
    dplyr::join_by(vf)
  ) %>%
  dplyr::mutate(
    label_value = dplyr::coalesce(VF_Name, vf_type),
    studlab_desc = ifelse(
      is.na(label_value),
      vf,
      paste0(vf, ": ", label_value)
    )
  )


mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_VF <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Wild)",

  leftcols = c("studlab", "I2_VF", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```

#### Human vs domestic
```{r metaanalysis for all vfs_hd, fig.height = 6, fig.width = 12}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "human_vs_domestic") %>%
  mutate(
    tot_domestic  = n_mags_domestic  + absent_domestic,
    tot_human = n_mags_human + absent_human
  ) %>%
  filter(n_mags_human > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_domestic > 0, tot_human > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_human < tot_human) | (n_mags_domestic < tot_domestic)) #remove rows where kegg is present in all mags of both groups (at least one group shows variation)


#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_domestic + n_mags_human),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; Human vs domestic 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # Human as experimental (so OR>1 => more prevalent in Human)
    event.e = dat$n_mags_human,
    n.e     = dat$tot_human,
    event.c = dat$n_mags_domestic,
    n.c     = dat$tot_domestic,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "human_vs_domestic",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))


mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    dplyr::join_by(vf == VFID)
  ) %>%
  dplyr::left_join(
    vf_type_data,
    dplyr::join_by(vf)
  ) %>%
  dplyr::mutate(
    label_value = dplyr::coalesce(VF_Name, vf_type),
    studlab_desc = ifelse(
      is.na(label_value),
      vf,
      paste0(vf, ": ", label_value)
    )
  )



mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_VF <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Human vs Domestic)",

  leftcols = c("studlab", "I2_VF", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```



```{r forest plot individual vfs_hd, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_vf <- function(vf_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[vf_id]]
  if (is.null(m)) stop("vf not found in mh_hw: ", vf_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("VF: ", vf_id)
  )
}

#individual analyses of each of these vf:
vfs <- mh_re_table_ci %>% dplyr::pull(vf)

for (k in vfs) {
  forest_one_vf(
    vf_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```




#### Domestic vs wild
```{r metaanalysis for all vfs_dw_2}
# Prep totals and basic filters 
df_hw <- vf_fisher_total %>%
  filter(contrast == "wild_vs_domestic") %>%
  mutate(
    tot_wild  = n_mags_wild  + absent_wild,
    tot_domestic = n_mags_domestic + absent_domestic
  )  %>%
  filter(n_mags_domestic > 0 | n_mags_wild > 0) %>% # remove rows where kegg is not present in both groups
  filter(tot_domestic > 0, tot_domestic > 0) %>%  #remove rows that dont have any data for kegg
  filter((n_mags_domestic < tot_domestic) | (n_mags_wild < tot_wild)) #remove rows where kegg is present in all mags of both groups (at least one group shows variation)

#  Low-bias screen 
min_informative_species <- 4
min_total_events <- 10

vf_keep_hw <- df_hw %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = n(),
    total_events = sum(n_mags_wild + n_mags_domestic),
    .groups = "drop"
  ) %>%
  filter(
    n_species_inform >= min_informative_species,
    total_events >= min_total_events
  ) %>%
  dplyr::select(vf)

df_hw_screened <- df_hw %>%
  inner_join(vf_keep_hw, by = "vf")

# Split by vf
vf_split_hw <- split(df_hw_screened, df_hw_screened$vf)

# Effect size pooling
# MH with random effects; domestic vs wild 
run_mh_random_hw <- function(dat) {
  meta::metabin(
    # domestic as experimental (so OR>1 => more prevalent in domestic)
    event.e = dat$n_mags_domestic,
    n.e     = dat$tot_domestic,
    event.c = dat$n_mags_wild,
    n.c     = dat$tot_wild,

    studlab = dat$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # remoev continuity correction
    method.incr = "only0"
  )
}

mh_hw <- lapply(vf_split_hw, run_mh_random_hw)

# Collect pooled results + FDR 
mh_hw_table <- bind_rows(lapply(names(mh_hw), function(k) {
  m <- mh_hw[[k]]
  data.frame(
    vf = k,
    contrast = "domestic_vs_wild",
    n_species = m$k,
    logOR_random = m$TE.random,
    se_random = m$seTE.random,
    OR_random = exp(m$TE.random),
    ci_low_random = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random = m$pval.random,
    I2_percent = m$I2,
    tau2_REML = m$tau^2,
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))

```



```{r vfs where ci does not include_1_vf_dom_wild, fig.height = 10, fig.width = 13}
mh_hw_table_annot <- mh_hw_table %>%
  dplyr::left_join(
    vf_data,
    dplyr::join_by(vf == VFID)
  ) %>%
  dplyr::left_join(
    vf_type_data,
    dplyr::join_by(vf)
  ) %>%
  dplyr::mutate(
    label_value = dplyr::coalesce(VF_Name, vf_type),
    studlab_desc = ifelse(
      is.na(label_value),
      vf,
      paste0(vf, ": ", label_value)
    )
  )

mh_re_table_ci <- mh_hw_table_annot %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random < 1) %>%
  dplyr::arrange(p_adj_fdr, p_random)

ord <- order(mh_re_table_ci$p_adj_fdr)

mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random[ord],
  seTE = mh_re_table_ci$se_random[ord],
  studlab = mh_re_table_ci$studlab_desc[ord],
  sm = "OR",
  common = FALSE,
  random = FALSE
)

mgen_ci$I2_VF <- round(mh_re_table_ci$I2_percent[ord], 1)
mgen_ci$FDR     <- signif(mh_re_table_ci$p_adj_fdr[ord], 2)

meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (Domestic vs Wild)",

  leftcols = c("studlab", "I2_VF", "FDR"),
  leftlabs = c("KEGG", "I²", "FDR"),

  I2 = round(mh_re_table_ci$I2_percent, 1),
  p_fdr = signif(mh_re_table_ci$p_adj_fdr, 2),

  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```
```{r forest plot individual vfs_dom_wild, fig.height = 6, fid.width = 10}
# Forest plot (Human on right, Wild on left)
forest_one_vf <- function(vf_id, xlim = NULL, at = NULL) {

  m <- mh_hw[[vf_id]]
  if (is.null(m)) stop("vf not found in mh_hw: ", vf_id)

  xlab_txt <- "Odds ratio (Human vs Wild)"

  # Columns potentially available in metabin object
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events W", "N W")

  available <- intersect(wanted_leftcols, names(m$data))

  # Always include studlab
  leftcols_use <- c("studlab", setdiff(available, "studlab"))
  leftlabs_use <- wanted_leftlabs[
    match(leftcols_use, wanted_leftcols)
  ]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,
    at          = at,
    xlog        = TRUE,   # IMPORTANT for ORs
    digits      = 2,
    main        = paste0("VF: ", vf_id)
  )
}

#individual analyses of each of these vf:
vfs <- mh_re_table_ci %>% dplyr::pull(vf)

for (k in vfs) {
  forest_one_vf(
    vf_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}
```

## Hierarchical GLMM

```{r make the matrices for the glmms}
make_glmm_long <- function(mat, metadata, species_name, annotation_type) {

  # mat: wide matrix from make_annotation_matrix()
  # metadata: genome metadata table
  # annotation_col: column name for annotation ("kegg", "amr", etc.)
  # annotation_type: label ("KEGG", "AMR", "VF", "CAZy")

  if (ncol(mat) <= 1) return(tibble::tibble())

  # 1. Convert counts to presence/absence
  pa <- mat %>%
    dplyr::mutate(
      dplyr::across(-genome, ~ as.integer(. != 0))
    )

  # 2. Pivot to long format 
  pa_long <- pa %>%
    tidyr::pivot_longer(
      cols = -genome,
      names_to  = "annotation",
      values_to = "present"
    )

  # 3. Join metadata

   out <- pa_long %>%
      dplyr::left_join(
        metadata %>%
          dplyr::select(
            genome        = ID_clean,
            host_type_2,
            completeness
          ),
        by = "genome"
      ) %>%
      dplyr::mutate(
        species = species_name,
        annotation_type = annotation_type
      )


  # 4. Drop rows with missing critical info
  out %>%
    dplyr::filter(
      !is.na(host_type_2),
      !is.na(completeness)
    )
}


glmm_tables <- list()

for (sp in species_list) {

  message("Processing species: ", sp)

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  mats <- list(
    KEGG = make_annotation_matrix(genome_annotations, "kegg"),
    AMR  = make_annotation_matrix(genome_annotations, "resistance_target"),
    VF   = make_annotation_matrix(genome_annotations, "vf"),
    CAZy = make_annotation_matrix(genome_annotations, "cazy")
  )

  glmm_tables[[sp]] <- purrr::imap(
    mats,
    ~ make_glmm_long(
        mat            = .x,
        metadata       = genome_metadata,
        species_name   = sp,
        annotation_type = .y
      )
  )
}
```
```{r combine}
kegg_glmm <- purrr::map_dfr(glmm_tables, "KEGG")
amr_glmm  <- purrr::map_dfr(glmm_tables, "AMR")
vf_glmm   <- purrr::map_dfr(glmm_tables, "VF")
cazy_glmm <- purrr::map_dfr(glmm_tables, "CAZy")
```


```{r filtering_vf_glmm_used_in_TMB, eval = FALSE}
## This is the filtering I was using before - models 1 and 2 

vf_glmm <- vf_glmm %>%
  dplyr::mutate(
    host_type_2 = factor(host_type_2),
    host_type_2 = relevel(host_type_2, ref = "human"),
    annotation  = factor(annotation),
    species     = factor(species), 
    completeness = as.numeric(completeness)
  )


# Filter rare VFs
vf_glmm <- vf_glmm %>%
  dplyr::group_by(annotation) %>%
  dplyr::filter(sum(present) >= 10) %>%
  dplyr::ungroup()
```


```{r better_filtering_based_on_metaanalysis_filtering}
## Current filtering for models 4

vf_glmm_filtered <- vf_glmm %>%
  group_by(species, annotation, host_type_2) %>%
  summarise(
    n_present = sum(present),
    n_total = n(),
    .groups = "drop"
  ) 

vf_informative <- vf_glmm_filtered %>%
  group_by(species, annotation) %>%
  summarise(
    has_variation = n_distinct(n_present) > 1, #exclude all 0 and all 1 , model 3 was using: has_variation = any(n_present > 0) & any(n_present < n_total)
    total_events = sum(n_present),  #total events across host groups
    .groups = "drop"
  ) %>%
  filter(
    has_variation,
    total_events >= 5  #removing cases where only <5 mags across all host groups (very rare VFs)
  )


vf_keep <- vf_informative %>%
  group_by(annotation) %>%
  summarise(n_species = n(), .groups = "drop") %>%
  filter(n_species >= 3) %>%
  pull(annotation)

vf_glmm <- vf_glmm %>%
  filter(annotation %in% vf_keep)
```



```{r check distribution}

ggplot(kegg_glmm, aes(x = factor(present))) +
  geom_bar() +
  labs(
    x = "Presence / Absence",
    y = "Number of observations",
    title = "Distribution of KEGG presence (row-wise)"
  )+theme_bw()

ggplot(vf_glmm, aes(x = factor(present))) +
  geom_bar() +
  labs(
    x = "Presence / Absence",
    y = "Number of observations",
    title = "Distribution of VF presence (row-wise)"
  )+theme_bw()


```
```{r kegg checks}
kegg_by_genome <- kegg_glmm %>%
  dplyr::group_by(genome, host_type_2, species) %>%
  dplyr::summarise(
    n_kegg = sum(present),
    .groups = "drop"
  )

ggplot(kegg_by_genome, aes(x = n_kegg)) +
  geom_histogram(bins = 50) +
  labs(
    x = "Number of KEGG annotations per genome",
    y = "Number of genomes",
    title = "KEGG richness per genome"
  )+ theme_bw()



vf_by_genome <- vf_glmm %>%
  dplyr::group_by(genome, host_type_2, species) %>%
  dplyr::summarise(
    n_vf = sum(present),
    .groups = "drop"
  )

ggplot(vf_by_genome, aes(x = n_vf)) +
  geom_histogram(bins = 50) +
  labs(
    x = "Number of VF annotations per genome",
    y = "Number of genomes",
    title = "VF richness per genome"
  )+ theme_bw()

```


The following generalized linear mixed models are done in Mjolnir HPC, then the results are imported into this R Markdown.


### 1) VF glmm with lme4

This is the first model I tested, with the outdated filtering.

```{r model 1}
vf_glmm <- vf_glmm %>%
  dplyr::mutate(
    host_type_2 = factor(host_type_2),
    host_type_2 = relevel(host_type_2, ref = "human"),  #setting human as reference factor
    annotation  = factor(annotation),
    species     = factor(species)
  )

# Filter rare VFs (important for speed)
vf_glmm <- vf_glmm %>%
  dplyr::group_by(annotation) %>%
  dplyr::filter(sum(present) >= 15) %>%
  dplyr::ungroup()


# vf_glmm_1 <- glmer(
#   present ~ host_type_2 + completeness +
#     (host_type_2 || annotation) +   
#     (1 | species),
#   data = vf_glmm,
#   family = binomial(),
#   control = glmerControl(
#     optimizer = "bobyqa",
#     optCtrl = list(maxfun = 2e5)
#   )
# )
```

MODEL:

$$
\text{logit}(P(present=1)) = \beta_0 + \beta_1(host) + \beta_2(completeness) + b_{annotation} + b_{annotation, host} + b_{species}
$$

m_vf_2 <- glmer(
  present ~ host_type_2 + completeness +
    (host_type_2 || annotation) +   # || = faster, safer
    (1 | species),
  data = vf_glmm,
  family = binomial(),
  control = glmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 2e5)
  )
)



For each VF, does the probability of being present in a MAG differ by host type (wild / domestic / human), consistently across bacterial species, after correcting for MAG completeness?

Is the presence of a virulence factor (VF) associated with host type, while accounting for genome completeness and variation across annotations and species?

---

#### Fixed effects: 
1) host_type_2 - How the probability of a VF being present differs in other hosts compared to human
2) completeness - Control for differences in MAG quality

#### Random effects:
Different virulence factors behave differently across host types.


(host_type_2 || annotation) - each VF gets its own intercept (baseline) and slope (host type effect) 
The || means random intercepts and slopes are independent.

--- 

Some species have more VFs overall than others (biological variation and genome size).
(1 | species)

```{r load_vf_glmm_model}
vf_glmm_1 <- readRDS("data/glmm_results/vf_glmm_model.rds")
```

```{r summary of model}
summary(vf_glmm_1)
```

Random effects: 
Intercept variance = 2.92  (large)
Host effects variance:
  human    = 11.39
  domestic = 13.92
  wild     = 6.06
  
Very large -- Different VFs behave VERY differently per host type

Correlations: 0.90–0.94 between host effects
-VFs that respond strongly to one host type tend to respond similarly to others


Species effect: Variance = 0.57 (moderate)
- Species differ in overall VF load, but much less than the variation from VF to VF (VF-specific effects)


Residuals: Min: -42.7   Max: 14.1
-very extreme -some predictions are very wrong, heterogeneous VFs


QUESTIONS:
For each VF, does probability differ by host type? - YES: large random-effect variance for host slopes
Is VF presence associated with host type overall?  - wild: strong negative effect compared to human, domestic not so much
This is somewhat conserved across species: species variance is relatively small.


```{r vf_fixed_1}
#extract fixed effects
vf_fixed <-  broom.mixed::tidy(vf_glmm_1, effects = "fixed")
vf_fixed
```
Fixed effects:

Intercept = 0.192 (p = 0.37)  - baseline human host
Domestic = -0.129 (p = 0.072)  - VF presence in domestic animals is slightly lower than human (but not p<0.05)
Wild = -0.459 (p = 4.32e-13) - VFs are significantly less likely to be present in wild-associated genomes compared to human
Completeness = 0.0175 (p = 1.53e-15) - More complete genomes are more likely to have VFs detected 

```{r vf_emmeans_host_contrasts_1}
# What is the average probability of VF presence for each host type, and how do they compare??
# Use emmeans to calculate av prob of vf presence

emm <- emmeans(vf_glmm_1, pairwise ~ host_type_2, type = "response")


emm$contrasts

```
Human vs domestic : not significant (no clear difference between human and domestic)
Human vs wild: Human genomes have 58% higher odds of VF presence than wild (p < 0.0001)
Domestic vs Wild: Domestic genomes have 39% higher odds than wild (p < 0.0001)

Prevalence of VFs : human > domestic > wild (average, not true for every VF)


```{r vf_random_effects_per_vf_1}
#get random deviation effects of each individual VF
# values here are log-odds of deviation from the average (fixed effect)


vf_effects <- ranef(vf_glmm_1)$annotation 
head(vf_effects)
```
Intercept: baseline prevalence of VF, + : generally common
Example - VF0003
host_type_2human: +5.42 : effect of human is much stronger than average VF
host_type2domestic : same
host_type2wild: +2.89 -> VF0003 is much more strongly associated with human (and domestic) than wild




Fixed effects → the average trend across all VFs
Random effects → how each VF deviates from that average


If we want the real effect: fixed + random
```{r real effect per vf}
# real effect per VF: add fixed + random effects

vf_effects_real <- vf_effects %>%
  tibble::rownames_to_column("VF") %>%
  mutate(
    wild_real = fixef(vf_glmm_1)["host_type_2wild"] + host_type_2wild,
    domestic_real = fixef(vf_glmm_1)["host_type_2domestic"] + host_type_2domestic,
    human_real = 0 + host_type_2human  # human is reference → fixed = 0
  )

```

```{r adding comparison human_vs_wild}
vf_effects_real <- vf_effects_real %>%
  mutate(wild_vs_human = wild_real - human_real) %>%
  arrange(desc(wild_vs_human)) %>%
  dplyr::select(VF, wild_real, human_real, domestic_real, wild_vs_human) 

vf_effects_real%>%
  head(10)
```
Example: wild_real : -0.46  vs human_real: -3.22

These VFs are more present in wild than human- they are rare overall (negative values-low probability) but they are less rare in wild compared to human.



```{r heatmap of top vfs}
top_vfs <- vf_effects_real %>%
  mutate(
    wild_vs_human = wild_real - human_real
  ) %>%
  {
    bind_rows(
      arrange(., desc(wild_real)) %>% slice_head(n = 10),
      arrange(., desc(human_real)) %>% slice_head(n = 10),
      arrange(., desc(wild_vs_human)) %>% slice_head(n = 10)
    )
  } %>%
  distinct(VF, .keep_all = TRUE)


heatmap_df <- top_vfs %>%
  dplyr::select(VF, human_real, domestic_real, wild_real) %>%
  pivot_longer(
    cols = -VF,
    names_to = "host",
    values_to = "effect"
  )

ggplot(heatmap_df, aes(x = host, y = VF, fill = effect)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "VF host associations (log-odds)",
    x = "Host type",
    y = "Virulence Factor",
    fill = "Effect"
  )


```

-VFs that are generally common (more yellow) are prevalent in human/domestic and less so in wild (orange)
-VFs that are rare everywhere (purple) are less rare in wild (lighter purple) compared to human/domestic.


```{r forest_plit_glmm1, fig.height = 6, fig.width =10}
# we CANNOT reliably compute per‑VF confidence intervals from this model, we would need to fit individual models per VF
#plot direction only


vf_hw <- vf_effects_real %>%
  mutate(
    logOR = human_real - wild_real, ## logOR > 0 : human > wild
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)  ##  strongest differences, regardless of direction:

p_hw <- ggplot(vf_hw, aes(x = OR, y = reorder(VF, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal()+
labs(
    title = "Human vs Wild",
    x = "Odds ratio (human / wild, log scale)",
    y = "Virulence factor"
  )


#domestic vs human
vf_dh <- vf_effects_real %>%
  mutate(
    logOR = human_real - domestic_real,
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)

p_hd <- ggplot(vf_dh, aes(x = OR, y = reorder(VF, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal()+
labs(
    title = "Human vs Domestic",
    x = "Odds ratio (human / domestic, log scale)",
    y = "Virulence factor"
  )


#domestic vs wild
 vf_dw <- vf_effects_real %>%
  mutate(
    logOR = domestic_real - wild_real,
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)

p_dw <- ggplot(vf_dw, aes(x = OR, y = reorder(VF, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal()+
labs(
    title = "Domestic vs Wild",
    x = "Odds ratio (domestic / wild, log scale)",
    y = "Virulence factor"
  )



(p_hw +p_dw+ p_hd) +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Virulence factor enrichment across host types"
  )
```

Checking the top terms to see what the model is really doing.
1) VF0700

```{r chcecks vf0700}
## vfs
#cleaning so we can check if vf that are given by model are actually enriched or not using the fisher tables
vf_clean <- vf_fisher_total %>%
  mutate(
    total_human = n_mags_human + absent_human,
    total_wild  = n_mags_wild  + absent_wild,
    total_domestic = n_mags_domestic + absent_domestic,

    prop_human = n_mags_human / total_human,
    prop_wild  = n_mags_wild  / total_wild,
    prop_domestic = n_mags_domestic / total_domestic,

    label_human = paste0(n_mags_human, "/", total_human),
    label_wild  = paste0(n_mags_wild, "/", total_wild),
    label_domestic  = paste0(n_mags_domestic, "/", total_domestic)
  )


vf_summary_hw <- vf_clean %>%
  dplyr::select(
    vf,label_human, label_wild,prop_human, prop_wild,
    diff_prop,odds_ratio,p_value,p_adj,species,contrast)

vf_summary_hd <- vf_clean %>%
  dplyr::select(
    vf,label_human, label_domestic,prop_human, prop_domestic,
    diff_prop,odds_ratio,p_value,p_adj,species,contrast)

vf_summary_dw <- vf_clean %>%
  dplyr::select(vf,label_wild, label_domestic,
    prop_wild, prop_domestic,diff_prop,odds_ratio,
    p_value,p_adj, species,contrast)


vf_summary_hw %>% filter(vf == "VF0700", contrast == "human_vs_wild" )


## keggs
kegg_clean <- kegg_fisher_total %>%
  mutate(
    total_human = n_mags_human + absent_human,
    total_wild  = n_mags_wild  + absent_wild,
    total_domestic = n_mags_domestic + absent_domestic,

    prop_human = n_mags_human / total_human,
    prop_wild  = n_mags_wild  / total_wild,
    prop_domestic = n_mags_domestic / total_domestic,

    label_human = paste0(n_mags_human, "/", total_human),
    label_wild  = paste0(n_mags_wild, "/", total_wild),
    label_domestic  = paste0(n_mags_domestic, "/", total_domestic)
  )


kegg_summary_hw <- kegg_clean %>%
  dplyr::select(
    kegg,label_human, label_wild,prop_human, prop_wild,
    diff_prop,odds_ratio,p_value,p_adj,species,contrast)

kegg_summary_hd <- kegg_clean %>%
  dplyr::select(
    kegg,label_human, label_domestic,prop_human, prop_domestic,
    diff_prop,odds_ratio,p_value,p_adj,species,contrast)

kegg_summary_dw <- kegg_clean %>%
  dplyr::select(kegg,label_wild, label_domestic,
    prop_wild, prop_domestic,diff_prop,odds_ratio,
    p_value,p_adj, species,contrast)

```

Here it is detecting a VF that is present across all mags in both comparisons except for hafnia paralvei, where it is present in 1 wild mag and 0 human mags, and it is assigned as "enriched in human", probably because we have more human mags overall.- This is not what we want to detect.

```{r vf0141}
#VF0141 is the top human-enriched term when comparing to domestic in the plot
vf_summary_hw %>% filter(vf == "VF0141", contrast == "human_vs_wild" )
```
This one is present overall but is more represented in humans compared to wild for parabacteroides goldsteinii only- also not what we are looking for..





```{r vf0377}
#VF0377 is supposed to be enriched in wild compared to humans
vf_summary_hw %>% filter(vf == "VF0377", contrast == "human_vs_wild" )
```
This one is absent in humans, present in few wild.


### 2) Same VF global model - but this time with TMB and proper filtering

PROPER FILTERING

```{r vf model tmb}
library(glmmTMB)
vf_glmm_4_TMB <- readRDS("data/glmm_results/vf_glmmTMB_filtered_model_4.rds")

up2date(vf_glmm_4_TMB)

summary(vf_glmm_4_TMB)
```

Interpretation

Fixed effects: 
-Intercept: 0.67 (p = 0.09) -> VF presence is generally high 
-Domestic vs human : 0.107 (p=0.24) -> VF presence is NOT different between human and domestic animals
-Wild vs human: -0.287 (p=0.000269) -> Wild hosts have about 25% lower odds of VF presence than humans 
-Completeness: 0.021 (p = 7.77e-13) -> higher completeness= higher prob of VF presence


Random effects:
-Intercept: 3.55
-Domestic slope: 1.58
-Wild slope:1.38
-Species: 0.64 -> some variation between species but smaller than VF to VF variation

Huge variability between VFs, different VFs respond differently to host type.

```{r vf_emmeans_host_contrasts_3}
# What is the average probability of VF presence for each host type, and how do they compare??
# Use emmeans to calculate av prob of vf presence


emm_2 <- emmeans(
  vf_glmm_4_TMB,
  pairwise ~ host_type_2,
  type = "response",
  data = vf_glmm
)



emm_2$contrasts

```
Domestic and human hosts have significantly higher VF presence than wild hosts.


```{r randomeffects_3}
#extract random effects
re <- ranef(vf_glmm_4_TMB)$cond$annotation
head(re)

#extract fixed effects
fix <- fixef(vf_glmm_4_TMB)$cond
fix
```
```{r combine fixed and random effects to get real effect}

vf_effects <- re %>%
  tibble::rownames_to_column("annotation") %>%
  mutate(
    intercept = fix["(Intercept)"] + `(Intercept)`,
    domestic  = fix["host_type_2domestic"] + host_type_2domestic,
    wild      = fix["host_type_2wild"] + host_type_2wild
  )%>% 
  mutate(
    OR_domestic = exp(domestic),#convert to odds ratio
    OR_wild     = exp(wild)
  )
```


```{r forest_plot_glmm_filtered_4, fig.height = 6, fig.width =10}
vf_effects_plot <- vf_effects %>%
  mutate(
    human     = intercept,
    domestic  = intercept + domestic,
    wild      = intercept + wild
  )

# Human vs Wild
vf_hw <- vf_effects_plot %>%
  mutate(
    logOR = human - wild,
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)

p_hw <- ggplot(vf_hw, aes(x = OR, y = reorder(annotation, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Human vs Wild",
    x = "Odds ratio (human / wild, log scale)",
    y = "Virulence factor"
  )

# Human vs Domestic
vf_hd <- vf_effects_plot %>%
  mutate(
    logOR = human - domestic,
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)

p_hd <- ggplot(vf_hd, aes(x = OR, y = reorder(annotation, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Human vs Domestic",
    x = "Odds ratio (human / domestic, log scale)",
    y = "Virulence factor"
  )


# Domestic vs Wild
vf_dw <- vf_effects_plot %>%
  mutate(
    logOR = domestic - wild,
    OR = exp(logOR)
  ) %>%
  slice_max(abs(logOR), n = 50)

p_dw <- ggplot(vf_dw, aes(x = OR, y = reorder(annotation, OR))) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point() +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Domestic vs Wild",
    x = "Odds ratio (domestic / wild, log scale)",
    y = "Virulence factor"
  )


# Combine plots
(p_hw + p_dw + p_hd) +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Virulence factor enrichment across host types"
  )
```

```{r vf0700_again}
#VF0700 is supposed to be top term enriched in human vs wild
vf_summary_hw %>% filter(vf == "VF0700", contrast == "human_vs_wild" )
```
VF0700 -supposed to be enriched in human -> is not really...


```{r vf0594}
#VF0594 is supposed to be enriched in wild animal compared to human
vf_summary_hw %>% filter(vf == "VF0594", contrast == "human_vs_wild" )
```
VF0594->supposed to be enriched in wild -> slightly enriched.

```{r vf0033}
#VF0033 is supposed to be enriched in wild animal compared to human
vf_summary_hw %>% filter(vf == "VF0033", contrast == "human_vs_wild" )
```
Is slightly enriched in wild animals in phocaeicola vulgatus.

### 3) VF GLMM with glmmTMB per species

```{r modelling_vfs_per_species, eval = FALSE}
#### done in cluster!!

#Separate per species
vf_by_species <- split(vf_glmm, vf_glmm$species)


# -------------------------
# Fit GLMM (one per species)
# -------------------------
fit_vf_model_one_species <- function(df, min_prevalence = 10) {

  # Drop species factor (constant within subset)
  df <- df %>% dplyr::select(-species)

  # Filter rare VFs
  df <- df %>%
    dplyr::group_by(annotation) %>%
    dplyr::filter(sum(present) >= min_prevalence) %>% #filter rare vfs 
    dplyr::ungroup()

  if (nrow(df) == 0 || n_distinct(df$annotation) < 2) {
    return(NULL)
  }


  tryCatch(
  glmmTMB(
  present ~ host_type_2 + completeness +
    (host_type_2 || annotation) ,
  data = df,
  family = binomial(),
  control = glmmTMBControl(
    optimizer = optim,
    optArgs = list(method = "BFGS")
  )
), error = function(e) {
  message("Model failed  :(  : ", conditionMessage(e))
  NULL
}
  )

  
}

vf_models <- purrr::imap(
  vf_by_species,
  ~ {
    message("Fitting VF GLMM for species: ", .y)
    fit_vf_model_one_species(.x)
  }
)

# Drop failed models
vf_models <- vf_models[!purrr::map_lgl(vf_models, is.null)]

```

For a each species, what determines whether a virulence factor is present — and how does that relationship vary across different virulence factors?

(same as before but here we have 1 model per bacterial species and we are using glmmTMB-faster)


```{r glmmTMB_loading_vf_models_per_species}
library(glmmTMB)
vf_models <- readRDS("data/glmm_results/vf_glmmTMB_models_by_species.rds")

#checking one model
up2date(vf_models[["akkermansia_muciniphila"]])
summary(vf_models[["akkermansia_muciniphila"]])
```

For Akkermansia muciniphila:
Random effects :

annotation (Intercept)         SD = 7.166  - huge variation: different VFs have wildly different baseline prevalence (some all absent or all present)
annotation host_type_2domestic SD = 2.594  - large random slopes - effect of host type varies a lot across VFs
annotation host_type_2wild     SD = 2.589



```{r vf_effects_per_species_1}
vf_models <- purrr::map(vf_models, glmmTMB::up2date)


vf_effects <- purrr::imap_dfr(
  vf_models,
  ~ broom.mixed::tidy(.x, effects = "fixed") %>%
      dplyr::mutate(species = .y)
)

vf_effects
```
For Akkermansia muciniphila:
Intercept : -1.47 : baseline log-odds of VF presence in human 
host_type2domestic: -0.27  (p = 0.49) - not significant
host_type2wild: -0.12  (p = 0.77) - not significant
completeness: +0.119 : more completeness = higher probability of VF presence, for each unit increase in completeness, the odds of detecting a VF increase by: exp⁡(0.1189)≈1.13-> 13 (p=5.9e-05 )


Fixed effects dont't detect differences across host types because it is the average effect (individual VFs some very high, some very low, cancel each other out).

But this is different beteween species

```{r vf_host_effects_fixed, fig.width= 10, fig.height = 5}
vf_host_effects <- vf_effects %>%
  dplyr::filter(term %in% c("host_type_2domestic", "host_type_2wild")) %>%
  dplyr::mutate(
    # Convert to odds ratios
    odds_ratio = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    
    # Significance flag
    significant = p.value < 0.05,
    
    # Clean labels
    term = dplyr::recode(
      term,
      host_type_2domestic = "Domestic vs human",
      host_type_2wild = "Wild vs human"
    )
  )

ggplot(vf_host_effects,
       aes(x = odds_ratio, y = species,
           color = significant)) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(
    values = c(`TRUE` = "#2C7BB6",   
               `FALSE` = "grey80"),
    labels = c("Not significant", "Significant")
  ) +
  
  facet_wrap(~ term) +
  
  labs(
    x = "Odds ratio (log scale)",
    y = "Species",
    color = "Effect significance",
    title = "Host-type effects on virulence factor presence"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "top"
  )
```
Host type effects are inconsistent and species-dependent.

Human vs wild: for most species, VFs are less likely to be present in wild compared to human.

Human vs domestic: mix
Bacteroides fragilis: VFs are less likely to be present in both wild and domestic compared to human.
Parabacteroides goldsteinii and Enterococcus hirae: VFs are more likely to be in domestic than human.
The rest - not significantly different.


  
```{r annotation_slopers_per_speceis}
vf_annotation_slopes <- purrr::imap_dfr(
  vf_models,
  ~ ranef(.x)$cond$annotation %>%
      as.data.frame() %>%
      tibble::rownames_to_column("annotation") %>%
      dplyr::mutate(species = .y)
)

vf_annotation_slopes
```

```{r vf_real_effects_per_species, fig.height = 7, fig.width = 12}
vf_fixed <- vf_models %>%
  purrr::imap_dfr(~ {
    coefs <- summary(.x)$coefficients$cond
    
    tibble::tibble(
      term = rownames(coefs),
      estimate = coefs[, "Estimate"],
      species = .y
    )
  }) %>%
  dplyr::filter(term %in% c("host_type_2domestic", "host_type_2wild"))

vf_combined <- vf_annotation_slopes %>%
  tidyr::pivot_longer(
    cols = c(host_type_2domestic, host_type_2wild),
    names_to = "term",
    values_to = "random_effect"
  ) %>%
  dplyr::left_join(vf_fixed, by = c("species", "term")) %>%
  dplyr::mutate(
    total_effect = estimate + random_effect,
    odds_ratio = exp(total_effect)
  )


plots <- vf_combined %>%
  dplyr::filter(odds_ratio > 0.1, odds_ratio < 10) %>%
  dplyr::group_split(species) %>%
  purrr::map(~ {
    species_name <- unique(.x$species)
    
    top <- .x %>%
      dplyr::group_by(term) %>%
      dplyr::slice_max(abs(total_effect), n = 10)
    
    ggplot(top,
           aes(x = odds_ratio,
               y = reorder(annotation, odds_ratio),
               color = term)) +
      geom_point() +
      facet_wrap(~ term, scales = "free_y") +
      scale_x_log10() +
      geom_vline(xintercept = 1, linetype = "dashed") +
      labs(title = species_name,
           y = "VFs") +
      theme_minimal()
  })

# combine
wrap_plots(plots, guides = "collect")
```

#### Trying to do meta-analysis on the VF per species model

To check if there are any VFs that across species follow the same direction of enrichment (in terms of host type).

Random effects do not give you uncertainty (SE), so we cannot do a proper meta-analysis.


Here we do we will try to check for VF-level consistency doing a pseudo-meta-analysis, by calculating the mean and standard variation across species.

Here it is unweighted, so the same weight is given to each species:

```{r pseudometaanalysis_unweighted}
vf_meta <- vf_combined %>%
  dplyr::group_by(annotation, term) %>%
  dplyr::summarise(
    mean_effect = mean(total_effect),  #average effect across species
    sd_effect = sd(total_effect),
    n_species = n(),
    prop_positive = mean(total_effect > 0),  # consistenvy of direction; ie: 1 - always enriched, 0 -always depleted
    .groups = "drop"
  )


#select consistent VFs accross at least 3 species
consistent_vfs <- vf_meta %>%
  dplyr::filter(
    n_species >= 3,
    abs(mean_effect) > 0.5,
    prop_positive > 0.8 | prop_positive < 0.2
  )


consistent_vfs <- consistent_vfs %>%
  mutate(
    odds_ratio = exp(mean_effect),
    conf.low = exp(mean_effect - sd_effect),
    conf.high = exp(mean_effect + sd_effect)
  ) %>%
  filter((conf.low > 1 & conf.high > 1) | (conf.low < 1 & conf.high < 1))

ggplot(consistent_vfs,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point() +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (across species)",
    y = "Virulence factor",
    title = "Consistently host-associated virulence factors"
  ) +
  theme_minimal()

```

Checking for what we are actually finding with this analysis:

Supposedly enriched in domestic: VF1055, VF1299
```{r vf1055}
vf_summary_hd %>% filter(vf == "VF1055", contrast == "human_vs_domestic" )
```

VF1055 is slighlty more prevalent in domestic for lactococcus_lactis and phocaeicola vulgatus, but it is a very small difference and this is not consistent for the other species...

```{r vf1299}
vf_summary_hd %>% filter(vf == "VF1299", contrast == "human_vs_domestic" )
```
VF1299 -not really "higer in domestic", jsut present across all mags in both host types, except for one where it is actually higher in human...



Checking supposedly enriched in human: VF0946
```{r vf0946}
vf_summary_hd %>% filter(vf == "VF0946", contrast == "human_vs_domestic" )
```
Higher in human across all.



```{r top wild check }
#which vf is enriched in wild according to previous psuedo-meta-analysis?
top_wild_vfs <- consistent_vfs %>%
  dplyr::filter(term == "host_type_2wild") %>%   # only wild comparison
  dplyr::filter(odds_ratio > 1) %>%              # enriched in wild 
  dplyr::arrange(desc(odds_ratio)) %>%           # strongest first
  dplyr::select(
    annotation,
    odds_ratio,
    conf.low,
    conf.high,
    mean_effect,
    n_species,
    prop_positive
  )
  
top_wild_vfs
```
```{r vf0126}
vf_summary_hw %>% filter(vf == "VF0126", contrast == "human_vs_wild" )
```



Here it is again, but this time weighing each species by number of observations (hafnia and citrobacter have much lower n_mags than phocaeicola and others).

Correct way to say it:  weighted aggregation of BLUP-based estimates.
```{r pseudometaanalysis_weighted}
# Get species-level weights (sample size)
species_weights <- vf_models %>%
  purrr::imap_dfr(~ {
    data.frame(
      species = .y,
      n_obs = nrow(.x$frame)   # approximate sample size per species
    )
  })


# Combine VF random + fixed effects = TRUE EFFECT
vf_combined <- vf_annotation_slopes %>%
  pivot_longer(
    cols = c(host_type_2domestic, host_type_2wild),
    names_to = "term",
    values_to = "random_effect"
  ) %>%
  left_join(vf_fixed, by = c("species", "term")) %>%
  mutate(
    total_effect = estimate + random_effect   # log-odds
  ) %>%
  left_join(species_weights, by = "species")


#Weighted meta-like aggregation per VF
vf_meta <- vf_combined %>%
  group_by(annotation, term) %>%
  summarise(
    n_species = n(),
    
    # weighted mean 
    weighted_mean = weighted.mean(total_effect, w = n_obs),
    
    # weighted variance (approximate)
    weighted_var = sum(n_obs * (total_effect - weighted_mean)^2) / sum(n_obs),
    
    weighted_se = sqrt(weighted_var / n_species),
    
    # consistency of direction
    prop_positive = mean(total_effect > 0),
    
    .groups = "drop"
  )


# Build confidence intervals (log-odds scale)
vf_meta <- vf_meta %>%
  mutate(
    ci_low = weighted_mean - 1.96 * weighted_se,
    ci_high = weighted_mean + 1.96 * weighted_se,
    
    # convert to odds ratios
    odds_ratio = exp(weighted_mean),
    or_low = exp(ci_low),
    or_high = exp(ci_high)
  )


# Filter for CONSISTENT direction
# (CI does NOT cross 0 (log scale)/ OR does NOT cross 1(odds ratio scale))
vf_consistent <- vf_meta %>%
  filter(
    n_species >= 3,   # require at least 3 species
    
    # CI does NOT cross 0 (log scale)
    (ci_low > 0 & ci_high > 0) | (ci_low < 0 & ci_high < 0)
  )

# keep top VFs 
vf_top <- vf_consistent %>%
  group_by(term) %>%
  slice_max(abs(weighted_mean), n = 15) %>%
  ungroup()


# Final forest-style plot - plotting only the top 15
ggplot(vf_top,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = or_low, xmax = or_high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (weighted across species)",
    y = "Virulence factor",
    title = "Virulence factors with consistent host association across species"
  ) +
  theme_minimal()


#Plotting all of the ones where the ci does not cross
ggplot(vf_consistent,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = or_low, xmax = or_high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (weighted across species)",
    y = "Virulence factor",
    title = "Virulence factors with consistent host association across species"
  ) +
  theme_minimal()
```

```{r vf0089}
#supposedly enriched in domestic
vf_summary_hd %>% filter(vf == "VF0089", contrast == "human_vs_domestic" )
```
VF0089 is higher in domestic 2 species, not for the other 2.


```{r vf0083}
#higher in humans -yes
vf_summary_hw %>% filter(vf == "VF0083", contrast == "human_vs_wild" )
```


```{r checking sharing of vf per species}

vf_presence <- vf_fisher_total %>%
  mutate(present = n_mags_human > 0 | n_mags_wild > 0 | n_mags_domestic > 0)

# 2. Count number of species per VF
vf_counts <- vf_presence %>%
  filter(present) %>%
  dplyr::select(vf, species) %>%
  distinct() %>%
  group_by(vf) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

# 3. Count how many VFs fall in each category
vf_distribution <- vf_counts %>%
  count(n_species)

# 4. Plot
ggplot(vf_distribution, aes(x = n_species, y = n)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Number of species where VF is present",
    y = "Number of VFs",
    title = "Distribution of VF presence across species"
  ) +
  theme_minimal()




# Count number of species per VF per contrast
vf_counts <- vf_fisher_total %>%
  dplyr::select(vf, species, contrast) %>%
  distinct() %>%
  group_by(contrast, vf) %>%
  summarise(n_species = n_distinct(species), .groups = "drop")

# Distribution: how many VFs per species-count, per contrast
vf_distribution <- vf_counts %>%
  count(contrast, n_species) %>%
  complete(contrast, n_species = 1:10, fill = list(n = 0))

# Plot (faceted by contrast)
ggplot(vf_distribution, aes(x = n_species, y = n)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ contrast) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Number of species where VF is present",
    y = "Number of VFs",
    title = "VF distribution across species (split by contrast)"
  ) +
  theme_minimal()

```




## kegg-per-species
```{r load kegg per species model filtered}
kegg_models <- readRDS("data/glmm_results/kegg/kegg_glmmTMB_models_by_species_FILTERED.rds")
```


```{r kegg_effects_per_species_oo}
kegg_models <- purrr::map(kegg_models, glmmTMB::up2date)


kegg_effects <- purrr::imap_dfr(
  kegg_models,
  ~ broom.mixed::tidy(.x, effects = "fixed") %>%
      dplyr::mutate(species = .y)
)

kegg_effects
```
For Akkermansia muciniphila (first 4 lines):
Intercept : -1.04 : baseline log-odds of KEGG presence in human 
host_type2domestic: -0.75  (p = 1.22e-05) - domestic have significantly less KEGGs than human
host_type2wild: -0.40  (p = 0.021) - wild has less KEGG than human.
completeness: +0.13 : (p = 6.9 e-20	)- KEGG presence increases with completeness


```{r kegg_per_species_host_effects_fixed, fig.width= 10, fig.height = 5}
kegg_host_effects <- kegg_effects %>%
  dplyr::filter(term %in% c("host_type_2domestic", "host_type_2wild")) %>%
  dplyr::mutate(
    # Convert to odds ratios
    odds_ratio = exp(estimate),
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    
    # Significance flag
    significant = p.value < 0.05,
    
    # Clean labels
    term = dplyr::recode(
      term,
      host_type_2domestic = "Domestic vs human",
      host_type_2wild = "Wild vs human"
    )
  )

ggplot(kegg_host_effects,
       aes(x = odds_ratio, y = species,
           color = significant)) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  scale_color_manual(
    values = c(`TRUE` = "#2C7BB6",   
               `FALSE` = "grey80"),
    labels = c("Not significant", "Significant")
  ) +
  
  facet_wrap(~ term) +
  
  labs(
    x = "Odds ratio (log scale)",
    y = "Species",
    color = "Effect significance",
    title = "Host-type effects on kegg ortholog presence"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "top"
  )
```
Host type effects are inconsistent and species-dependent.

Human vs wild: for all species, KOs are less likely to be present in wild compared to human.

Human vs domestic: mix
Lactococcus lactis: KOs are less likely to be present in both domestic compared to human.
Parabacteroides goldsteinii, Citrobacter braakii and Enterococcus hirae: KOs are more likely to be in domestic than human.
The rest - not significantly different.


  
```{r annotation_slopers_per_species_kegg}
kegg_annotation_slopes <- purrr::imap_dfr(
  kegg_models,
  ~ ranef(.x)$cond$annotation %>%
      as.data.frame() %>%
      tibble::rownames_to_column("annotation") %>%
      dplyr::mutate(species = .y)
)

kegg_annotation_slopes
```

```{r kegg_real_effects_per_species, fig.height = 7, fig.width = 12}
kegg_fixed <- kegg_models %>%
  purrr::imap_dfr(~ {
    coefs <- summary(.x)$coefficients$cond
    
    tibble::tibble(
      term = rownames(coefs),
      estimate = coefs[, "Estimate"],
      species = .y
    )
  }) %>%
  dplyr::filter(term %in% c("host_type_2domestic", "host_type_2wild"))

kegg_combined <- kegg_annotation_slopes %>%
  tidyr::pivot_longer(
    cols = c(host_type_2domestic, host_type_2wild),
    names_to = "term",
    values_to = "random_effect"
  ) %>%
  dplyr::left_join(kegg_fixed, by = c("species", "term")) %>%
  dplyr::mutate(
    total_effect = estimate + random_effect,
    odds_ratio = exp(total_effect)
  )


plots <- kegg_combined %>%
  dplyr::filter(odds_ratio > 0.1, odds_ratio < 10) %>%
  dplyr::group_split(species) %>%
  purrr::map(~ {
    species_name <- unique(.x$species)
    
    top <- .x %>%
      dplyr::group_by(term) %>%
      dplyr::slice_max(abs(total_effect), n = 10)
    
    ggplot(top,
           aes(x = odds_ratio,
               y = reorder(annotation, odds_ratio),
               color = term)) +
      geom_point() +
      facet_wrap(~ term, scales = "free_y") +
      scale_x_log10() +
      geom_vline(xintercept = 1, linetype = "dashed") +
      labs(title = species_name,
           y = "KOs") +
      theme_minimal()
  })

# combine
wrap_plots(plots, guides = "collect")
```

#### Trying to do meta-analysis on the KEGG per species model

To check if there are any kos that across species follow the same direction of enrichment (in terms of host type).

Random effects do not give you uncertainty (SE), so we cannot do a proper meta-analysis.


Here we do we will try to check for KO-level consistency doing a pseudo-meta-analysis, by calculating the mean and standard variation across species.

Here it is unweighted, so the same weight is given to each species:

```{r pseudometaanalysis_unweighted_kegg}
kegg_meta <- kegg_combined %>%
  dplyr::group_by(annotation, term) %>%
  dplyr::summarise(
    mean_effect = mean(total_effect),  #average effect across species
    sd_effect = sd(total_effect),
    n_species = n(),
    prop_positive = mean(total_effect > 0),  # consistenvy of direction; ie: 1 - always enriched, 0 -always depleted
    .groups = "drop"
  )


#select consistent keggs accross at least 3 species
consistent_keggs <- kegg_meta %>%
  dplyr::filter(
    n_species >= 3,
    abs(mean_effect) > 0.5,
    prop_positive > 0.8 | prop_positive < 0.2
  )


consistent_keggs <- consistent_keggs %>%
  mutate(
    odds_ratio = exp(mean_effect),
    conf.low = exp(mean_effect - sd_effect),
    conf.high = exp(mean_effect + sd_effect)
  ) %>%
  filter((conf.low > 1 & conf.high > 1) | (conf.low < 1 & conf.high < 1))

ggplot(consistent_keggs,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point() +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (across species)",
    y = "Virulence factor",
    title = "Consistently host-associated virulence factors"
  ) +
  theme_minimal()

```

Checking for what we are actually finding with this analysis:

Supposedly enriched in domestic: K12944, K18434
```{r k12944}
kegg_summary_hd %>% filter(kegg == "K12944", contrast == "human_vs_domestic" )
```



```{r K13290}
kegg_summary_hd %>% filter(kegg == "K13290", contrast == "human_vs_domestic" )
```
K13290 is supposed to be higher in domestic, is not really...




```{r top kegg wild check }
#which vf is enriched in wild according to previous psuedo-meta-analysis?
top_wild_keggs <- consistent_keggs %>%
  dplyr::filter(term == "host_type_2wild") %>%   # only wild comparison
  dplyr::filter(odds_ratio > 1) %>%              # enriched in wild 
  dplyr::arrange(desc(odds_ratio)) %>%           # strongest first
  dplyr::select(
    annotation,
    odds_ratio,
    conf.low,
    conf.high,
    mean_effect,
    n_species,
    prop_positive
  )
  
top_wild_keggs
```
```{r K13408}
kegg_summary_hw %>% filter(kegg == "K13408", contrast == "human_vs_wild" )

kegg_summary_hw %>% filter(kegg == "K27086", contrast == "human_vs_wild" )
```
K13408 and K27086- do seem to be more enriched in wild than human across a few species.

K13408: membrane fusion protein (cvaA, mchE, raxA) - type I secretion system - plant-pathogen interaction
K27086: ESX secretion system protein EccC- Type VII secretion system.


Here it is again, but this time weighing each species by number of observations (hafnia and citrobacter have much lower n_mags than phocaeicola and others).

Correct way to say it:  weighted aggregation of BLUP-based estimates.
```{r pseudometaanalysis_weighted_kegg}
# Get species-level weights (sample size)
species_weights <- kegg_models %>%
  purrr::imap_dfr(~ {
    data.frame(
      species = .y,
      n_obs = nrow(.x$frame)   # approximate sample size per species
    )
  })


# Combine kegg random + fixed effects = TRUE EFFECT
kegg_combined <- kegg_annotation_slopes %>%
  pivot_longer(
    cols = c(host_type_2domestic, host_type_2wild),
    names_to = "term",
    values_to = "random_effect"
  ) %>%
  left_join(kegg_fixed, by = c("species", "term")) %>%
  mutate(
    total_effect = estimate + random_effect   # log-odds
  ) %>%
  left_join(species_weights, by = "species")


#Weighted meta-like aggregation per kegg
kegg_meta <- kegg_combined %>%
  group_by(annotation, term) %>%
  summarise(
    n_species = n(),
    
    # weighted mean 
    weighted_mean = weighted.mean(total_effect, w = n_obs),
    
    # weighted variance (approximate)
    weighted_var = sum(n_obs * (total_effect - weighted_mean)^2) / sum(n_obs),
    
    weighted_se = sqrt(weighted_var / n_species),
    
    # consistency of direction
    prop_positive = mean(total_effect > 0),
    
    .groups = "drop"
  )


# Build confidence intervals (log-odds scale)
kegg_meta <- kegg_meta %>%
  mutate(
    ci_low = weighted_mean - 1.96 * weighted_se,
    ci_high = weighted_mean + 1.96 * weighted_se,
    
    # convert to odds ratios
    odds_ratio = exp(weighted_mean),
    or_low = exp(ci_low),
    or_high = exp(ci_high)
  )


# Filter for CONSISTENT direction
# (CI does NOT cross 0 (log scale)/ OR does NOT cross 1(odds ratio scale))
kegg_consistent <- kegg_meta %>%
  filter(
    n_species >= 3,   # require at least 3 species
    
    # CI does NOT cross 0 (log scale)
    (ci_low > 0 & ci_high > 0) | (ci_low < 0 & ci_high < 0)
  )

# keep top keggs 
kegg_top <- kegg_consistent %>%
  group_by(term) %>%
  slice_max(abs(weighted_mean), n = 15) %>%
  ungroup()


# Final forest-style plot - plotting only the top 15
ggplot(kegg_top,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = or_low, xmax = or_high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (weighted across species)",
    y = "KO",
    title = "KEGG Orthologs with consistent host association across species"
  ) +
  theme_minimal()


#Plotting all of the ones where the ci does not cross
ggplot(kegg_consistent,
       aes(x = odds_ratio,
           y = reorder(annotation, odds_ratio))) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(xmin = or_low, xmax = or_high),
    height = 0.2
  ) +
  facet_wrap(~ term, scales = "free_y") +
  scale_x_log10() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(
    x = "Odds ratio (weighted across species)",
    y = "KO",
    title = "KEGG Orthologs with consistent host association across species"
  ) +
  theme_minimal()
```

```{r K18434}
#supposedly enriched in domestic
kegg_summary_hd %>% filter(kegg == "K18434", contrast == "human_vs_domestic" )
```
K18434 is supposed to be enriched in domestic compared to human -> is higher in domestic for 3 of the species.
K18434: type IV secretion system T-DNA border endonuclease VirD2



```{r K05601}
#supposedly enriched in human compared to wild
kegg_summary_hw %>% filter(kegg == "K18434", contrast == "human_vs_wild" )
```

###Testing model with (host_type_2 | species)?

```{r vf model tmb_5}
vf_glmm_5_TMB <- readRDS("data/glmm_results/vf_glmmTMB_filtered_model_5.rds")

up2date(vf_glmm_5_TMB)

summary(vf_glmm_5_TMB)
```

Random effects- species-level:
Intecept SD = 0.62  - species differ in baseline VF prevalence
domestic SD =0.27   - domestic host effect varies moderately across species
Wild  SD = 0        - wild host effect is constant across species


```{r extracting effects}
#fixed effects
fixef_df <- broom.mixed::tidy(
  vf_glmm_5_TMB,
  effects = "fixed"
) %>%
  dplyr::select(term, estimate)

#randome effects
vf_re <- ranef(vf_glmm_5_TMB)$cond$annotation %>%
  as.data.frame() %>%
  tibble::rownames_to_column("annotation")


#species random effects
species_re <- ranef(vf_glmm_5_TMB)$cond$species %>%
  as.data.frame() %>%
  tibble::rownames_to_column("species")


#convert to long formats
# Keep only host effects
vf_long <- vf_re %>%
  pivot_longer(
    cols = starts_with("host_type_2"),
    names_to = "term",
    values_to = "vf_dev"
  )

species_long <- species_re %>%
  pivot_longer(
    cols = starts_with("host_type_2"),
    names_to = "term",
    values_to = "species_dev"
  )

fixed_long <- fixef_df %>%
  filter(str_detect(term, "host_type_2")) %>%
  rename(fixed = estimate)


##combine all to get true estimated effect:
vf_species_effects <- vf_long %>%
  left_join(fixed_long, by = "term") %>%
  left_join(species_long, by = "term") %>%
  mutate(
    total_effect = fixed + vf_dev + species_dev,
    odds_ratio = exp(total_effect)
  )
```
Model does not give true standard error for random effects, so we will approximate SE_total ≈ SE_fixed




