```{r order_preparation}

  tax_map <- tibble::tibble(
  host_species = c(
    # human_adjs / lab / livestock / pets
    "Homo sapiens",
    "Mus musculus",
    "Rattus norvegicus",
    "Sus scrofa",
    "Bos taurus",
    "Canis lupus familiaris",
    "Equus caballus",
    "Oryctolagus cuniculus",
    "Capra hircus",

    # Birds
    "Gallus gallus",
    "Cathartes aura",
    "Columba livia",
    "Spheniscus magellanicus",

    # Insects
    "Drosophila melanogaster",
    "Bombyx mori",

    # Fish
    "Danio rerio",
    "Oncorhynchus mykiss",
    "Oreochromis niloticus",

    # Marine mammals / pinnipeds
    "Balaenoptera acutorostrata",
    "Arctocephalus australis",

    # Extinct
    "Mammuthus primigenius",
    
    #More
    "Loxodonta africana",
    "Elephas maximus",
    "Felis catus",
    "Canis lupus",
    "Phascolarctos cinereus",
    
    "Grasshopper",
    "Termite",
    "Macaca fascicularis",
    "Rhinolophus capensis",
    "Fish",
    
    #Final (from total_genome_metadata)
    "Zophobas morio",
    "Mustela putorius",
    "Ovis aries",
    "Corallus caninus",
    "Meles meles",
    "Clanga pomarina",
    "Buteo buteo",
    "Accipiter gentilis",
    "Chelonia mydas",
    "Grampus griseus",
    "Sterna trudeaui",
    "Rhodnius prolixus",
    "Turdus merula",
    "Curruca curruca",
    "Columba palumbus",
    "Delichon urbicum",
    "Chloris chloris",
    "Hirundo rustica",
    "Ciconia ciconia",
    "Anas platyrhynchos",
    "Grus grus",
    "Apis mellifera",
    "Coragyps atratus",
    "Pseudoscops clamator",
    "Pandion haliaetus",
    "Sycanus bifidus",
    "Eocanthecona furcellata",
    "Arma custos",
    "Corvus frugilegus",
    "Milvus milvus",
    "Vulpes zerda"


  ),
  tax_host_order = c(
    # human_adjs / lab / livestock / pets
    "Primates",
    "Rodentia",
    "Rodentia",
    "Artiodactyla",
    "Artiodactyla",
    "Carnivora",
    "Perissodactyla",
    "Lagomorpha",
    "Artiodactyla",

    # Birds
    "Galliformes",
    "Accipitriformes",
    "Columbiformes",
    "Sphenisciformes",

    # Insects
    "Diptera",
    "Lepidoptera",

    # Fish
    "Cypriniformes",
    "Salmoniformes",
    "Cichliformes",

    # Marine mammals / pinnipeds
    "Artiodactyla",      # (Cetartiodactyla)
    "Carnivora",

    # Extinct
    "Proboscidea",
    
    #More
    "Proboscidea",
    "Proboscidea",
    "Carnivora",
    "Carnivora",
    "Diprotodontia",
    
    
    "Orthoptera",     # Grasshopper
    "Blattodea",      # Termite (termites now placed within Blattodea; Isoptera as infraorder)
    "Primates",       # Macaca fascicularis
    "Chiroptera",     # Rhinolophus capensis (horseshoe bat)
    NA_character_,    # Fish (no single order fits all fish)
    
    #Final (from total_genome_metadata)
    "Coleoptera",          # Zophobas morio
    "Carnivora",           # Mustela putorius
    "Artiodactyla",        # Ovis aries
    "Squamata",            # Corallus caninus
    "Carnivora",           # Meles meles
    "Accipitriformes",     # Clanga pomarina
    "Accipitriformes",     # Buteo buteo
    "Accipitriformes",     # Accipiter gentilis
    "Testudines",          # Chelonia mydas
    "Artiodactyla",        # Grampus griseus
    "Charadriiformes",     # Sterna trudeaui
    "Hemiptera",           # Rhodnius prolixus
    "Passeriformes",       # Turdus merula
    "Passeriformes",       # Curruca curruca
    "Columbiformes",       # Columba palumbus
    "Passeriformes",       # Delichon urbicum
    "Passeriformes",       # Chloris chloris
    "Passeriformes",       # Hirundo rustica
    "Ciconiiformes",       # Ciconia ciconia
    "Anseriformes",        # Anas platyrhynchos
    "Gruiformes",          # Grus grus
    "Hymenoptera",         # Apis mellifera
    "Cathartiformes",      # Coragyps atratus
    "Strigiformes",        # Pseudoscops clamator
    "Accipitriformes",     # Pandion haliaetus
    "Hemiptera",           # Sycanus bifidus
    "Hemiptera",           # Eocanthecona furcellata
    "Hemiptera",           # Arma custos
    "Passeriformes",       # Corvus frugilegus
    "Accipitriformes",     # Milvus milvus
    "Carnivora"            # Vulpes zerda


  ),
  tax_host_class = c(
    # human_adjs / lab / livestock / pets
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",

    # Birds
    "Aves",
    "Aves",
    "Aves",
    "Aves",

    # Insects
    "Insecta",
    "Insecta",

    # Fish
    "Actinopterygii",
    "Actinopterygii",
    "Actinopterygii",

    # Marine mammals / pinnipeds
    "Mammalia",
    "Mammalia",

    # Extinct and more
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    "Mammalia",
    
        
    "Insecta",        # Grasshopper
    "Insecta",        # Termite
    "Mammalia",       # Macaca fascicularis
    "Mammalia",       # Rhinolophus capensis
    "Actinopterygii",  # Fish (bony fishes; safe default for generic "fish")

    #Final (from total_genome_metadata)
    "Insecta",        # Zophobas morio
    "Mammalia",       # Mustela putorius
    "Mammalia",       # Ovis aries
    "Reptilia",       # Corallus caninus
    "Mammalia",       # Meles meles
    "Aves",           # Clanga pomarina
    "Aves",           # Buteo buteo
    "Aves",           # Accipiter gentilis
    "Reptilia",       # Chelonia mydas
    "Mammalia",       # Grampus griseus
    "Aves",           # Sterna trudeaui
    "Insecta",        # Rhodnius prolixus
    "Aves",           # Turdus merula
    "Aves",           # Curruca curruca
    "Aves",           # Columba palumbus
    "Aves",           # Delichon urbicum
    "Aves",           # Chloris chloris
    "Aves",           # Hirundo rustica
    "Aves",           # Ciconia ciconia
    "Aves",           # Anas platyrhynchos
    "Aves",           # Grus grus
    "Insecta",        # Apis mellifera
    "Aves",           # Coragyps atratus
    "Aves",           # Pseudoscops clamator
    "Aves",           # Pandion haliaetus
    "Insecta",        # Sycanus bifidus
    "Insecta",        # Eocanthecona furcellata
    "Insecta",        # Arma custos
    "Aves",           # Corvus frugilegus
    "Aves",           # Milvus milvus
    "Mammalia"        # Vulpes zerda

  )
)

order_to_class <- tax_map %>%
  dplyr::select(host_order = tax_host_order,
                host_class = tax_host_class) %>%
  distinct()

legend_order <- order_to_class %>%
  dplyr::mutate(
    host_class = factor(host_class,
      levels = c("Mammalia", "Aves", "Reptilia",
                 "Actinopterygii", "Insecta", "Amphibia", NA)
    )
  ) %>%
  dplyr::arrange(host_class) %>%
  dplyr::pull(host_order)

```


```{r loading_bateria_characteristics}
bact_metadata <- read_tsv("data/mags_metadata/bacteria_characteristics.tsv")
```


## Define species list
```{r Define species and mags to remove list_2}
# ALL SPECIES
species_list <- c(
  "lactococcus_lactis",
  "parabacteroides_goldsteinii",
  "enterococcus_faecalis",
  "bacteroides_uniformis",
  "phocaeicola_vulgatus",
  "hafnia_paralvei",
  "citrobacter_braakii",
  "enterococcus_hirae",
  "akkermansia_muciniphila",
  "bacteroides_fragilis"
)


mags_to_remove_list <- list(
  lactococcus_lactis = c(
    "GCA_018369575.1.fna","GCA_947063445.1.fna","GCA_947101685.1.fna",
    "GCA_948698275.1.fna","GCA_937910935.1.fna","GCA_947072755.1.fna",
    "GCA_948655095.1.fna","GCA_948703095.1.fna","GCA_947041925.1.fna",
    "GCA_947073355.1.fna","GCA_948675165.1.fna","GCA_948718815.1.fna"
  ),
  parabacteroides_goldsteinii = character(0), # none to remove
  enterococcus_faecalis = character(0),
  hafnia_paralvei = character(0),
  bacteroides_uniformis = character(0),
  phocaeicola_vulgatus = c("GCA_040912025.1", "GCF_048453085.1", "GCA_048453085.1"),
  citrobacter_braakii = character(0),
  enterococcus_hirae = character(0),
  akkermansia_muciniphila = character(0),
  bacteroides_fragilis = character(0)
)

results <- list()

for (sp in species_list) {
  results[[sp]] <- process_one_species(
    species        = sp,
    mags_to_remove = mags_to_remove_list[[sp]]
  )
}

pg <- results[["parabacteroides_goldsteinii"]]$metadata
eh <- results[["enterococcus_hirae"]]$metadata
ef <- results[["enterococcus_faecalis"]]$metadata
pv <- results[["phocaeicola_vulgatus"]]$metadata
am <- results[["akkermansia_muciniphila"]]$metadata
bf <- results[["bacteroides_fragilis"]]$metadata
bu <- results[["bacteroides_uniformis"]]$metadata
ll <- results[["lactococcus_lactis"]]$metadata
hp <- results[["hafnia_paralvei"]]$metadata
cb <- results[["citrobacter_braakii"]]$metadata


named_list <- list(
  parabacteroides_goldsteinii = pg,
  enterococcus_hirae          = eh,
  enterococcus_faecalis       = ef,
  phocaeicola_vulgatus        = pv,
  akkermansia_muciniphila     = am,
  bacteroides_fragilis        = bf,
  bacteroides_uniformis       = bu,
  lactococcus_lactis          = ll,
  hafnia_paralvei             = hp,
  citrobacter_braakii         = cb
)

total_genome_metadata <- dplyr::bind_rows(named_list, .id = "bacterial_species")


#total_genome_annotations <- dplyr::bind_rows()

```



## Data statistics
```{r counting species order and class_2}
total_genome_metadata %>% dplyr::count(host_species)

total_genome_metadata %>% dplyr::count(host_order)

total_genome_metadata %>% dplyr::count(host_class)
```
```{r plotting n mags per bacterial species split human_adj and wild}
counts_mags <- total_genome_metadata %>%
  filter(!is.na(host_type_3))%>%
  group_by(bacterial_species, host_type_3)%>%
  summarize(total = n(),
            n_mags_wild = sum(host_type_3 == "wild"), 
            n_mags_human_adj = sum(host_type_3 == "human_adj"))


ggplot(counts_mags, aes(x = bacterial_species, y = total, fill = host_type_3))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = host_type_3_colors)+
  labs(
    title = "Number of MAGs per Bacterial Species (wild vs human_adj hosts)",
    x = "Bacterial species",
    y = "N of MAGs"
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

```{r plotting host order_2}
df_host_species <- total_genome_metadata %>%
  filter(!is.na(bacterial_species), !is.na(host_species)) %>%
  distinct(bacterial_species, host_species) %>%
  count(bacterial_species, name = "n_host_species") %>%
  mutate(bacterial_species = fct_reorder(bacterial_species, n_host_species))

ggplot(df_host_species,
       aes(x = bacterial_species, y = n_host_species)) +
  geom_col(fill = "#355C7D") +
  geom_text(aes(label = n_host_species), vjust = -0.3, size = 3) +
  labs(
    title = "Number of Host Species per Bacterial Species",
    x = "Bacterial species",
    y = "Distinct host species"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))



df_order_counts <- total_genome_metadata %>%
  filter(!is.na(host_order)) %>%
  count(host_order, name = "n_mags") %>%
  mutate(host_order = fct_reorder(host_order, n_mags))

ggplot(df_order_counts,
       aes(x = host_order, y = n_mags)) +
  geom_col() +
  scale_fill_manual(values = host_order_colors)+
  geom_text(aes(label = n_mags), vjust = -0.3, size = 3) +
  labs(
    title = "Number of MAGs per Host Order",
    x = "Host order",
    y = "MAG count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

```
```{r average n host species}
df_host_species %>% 
  summarise(
    mean_n = mean(n_host_species, na.rm = TRUE) %>% round(2),
    sd_n = sd(n_host_species, na.rm = TRUE) %>% round(2),
  ) %>%
  unite("Mean number of host species per bacteria", mean_n, sd_n, sep = " ± ") %>%
  tt()
```


```{r how many host species total}
total_genome_metadata %>% filter(!is.na(host_species))%>%
  pull(host_species) %>% unique() 

total_genome_metadata %>% filter(!is.na(host_class))%>%
  pull(host_class) %>% unique() 
```



```{r map_2}
library(rnaturalearth)
library(rnaturalearthdata)
library(sf) 


world <- ne_countries(scale = "medium", returnclass = "sf")

mag_locations <- total_genome_metadata %>%
  filter(!is.na(country_normalized)) %>%
  left_join(
    world %>%
      dplyr::select(name, geometry) %>%
      st_centroid() %>%
      st_coordinates() %>%
      as_tibble() %>%
      bind_cols(world %>% dplyr::select(name)) ,
    by = c("country" = "name")
  )

mag_locations%>% dplyr::select(ID, host_type_3, country_normalized)

ggplot(world) +
  geom_sf(fill = "gray95", color = "gray80") +
  geom_point(
    data = mag_locations,
    aes(x = X, y = Y, color = host_type_3),
    size = 2.5,
    alpha = 0.5
  ) +
  scale_color_manual(values = host_type_3_colors) +
  theme_minimal() +
  labs(color = "Host Type")

```


### Mean completeness and contamination
```{r mean completeness and contamination final_2}
total_genome_metadata %>% 
  summarise(
    mean_c = mean(completeness, na.rm = TRUE) %>% round(2),
    sd_c = sd(completeness, na.rm = TRUE) %>% round(2),
    mean_con = mean(contamination, na.rm = TRUE) %>% round(2),
    sd_con = sd(contamination, na.rm = TRUE) %>% round(2)
  ) %>%
  unite("Completeness", mean_c, sd_c, sep = " ± ") %>%
  unite("Contamination", mean_con, sd_con, sep = " ± ") %>%
  tt()
```


```{r is completeness and contamination different between bacteria}
kruskal_test(total_genome_metadata,
                   completeness ~ bacterial_species)

pairwise.wilcox.test(total_genome_metadata$completeness, total_genome_metadata$bacterial_species, p.adjust.method = "fdr")

kruskal_test(total_genome_metadata,
                   contamination ~ bacterial_species)
```
```{r general completeness contaminaton }
wilcox.test(contamination ~ host_type_3, data = total_genome_metadata)
wilcox.test(completeness   ~ host_type_3, data = total_genome_metadata)


summary_tbl <- total_genome_metadata %>%
  group_by(host_type_3) %>%
  summarise(
    n = n(),
    completeness_median = median(completeness, na.rm = TRUE),
    completeness_IQR    = IQR(completeness, na.rm = TRUE),
    contamination_median = median(contamination, na.rm = TRUE),
    contamination_IQR    = IQR(contamination, na.rm = TRUE)
  ) %>%
  arrange(host_type_3)

summary_tbl

```




```{r violin plots bacteria completeness}
total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = completeness))+
  geom_violin()+ 
  geom_point()+
  labs(y = "Completeness", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))


total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = contamination))+
  geom_violin()+ 
  geom_point()+
  labs(y = "Contamination", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

total_genome_metadata %>%
ggplot(aes(x= bacterial_species, y = completeness))+
  geom_violin()+ 
  geom_point()+
  facet_grid(~host_type_3)+
  labs(y = "Completeness", x = "Bacterial Species")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```








```{r genome_quality_plot_3, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Generate quality biplot
genome_biplot <- total_genome_metadata %>%
  dplyr::select(c(ID,completeness,contamination, host_order)) %>%
  ggplot(aes(x=completeness, y=contamination, color = host_order)) +
  scale_color_manual(values = host_order_colors, name = "Host Order") +
  geom_point(alpha=0.7, size = 4) +
  xlim(c(90,100)) +
  ylim(c(2.5,0)) +
  labs(y= "Contamination", x = "Completeness") +
  theme_classic() +
  theme(
    legend.position = "left",
    axis.text.y = element_text(size=12),
    axis.text.x = element_text(size=12),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold")
  )

#Generate contamination boxplot
genome_contamination <- total_genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(2.5,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_classic() +
                    theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- total_genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(90,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_classic() +
                theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
#pdf("figures/completeness_contamination.pdf",width=10, height=5)
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))
#dev.off()
```

```{r genome_quality_plot_colored_by_bact, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
#Generate quality biplot
genome_biplot <- total_genome_metadata %>%
  dplyr::select(c(ID,completeness,contamination, bacterial_species)) %>%
  ggplot(aes(x=completeness, y=contamination, color = bacterial_species)) +
  geom_point(alpha=0.7, size = 4) +
  xlim(c(90,100)) +
  ylim(c(2.5,0)) +
  labs(y= "Contamination", x = "Completeness") +
  theme_classic() +
  theme(
    legend.position = "left",
    axis.text.y = element_text(size=12),
    axis.text.x = element_text(size=12),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold")
  )

#Generate contamination boxplot
genome_contamination <- total_genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(2.5,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_classic() +
                    theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- total_genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(90,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_classic() +
                theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
#pdf("figures/completeness_contamination.pdf",width=10, height=5)
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))
#dev.off()
```

```{r countries}
total_genome_metadata <- total_genome_metadata %>%
  mutate(
    country_normalized = case_when(
      is.na(country_normalized) ~ "None",
      tolower(country_normalized) %in% c("none", "na", "n/a") ~ "None",
      TRUE ~ country_normalized
    )
  )

country_counts <- total_genome_metadata %>%
  count(country_normalized)

small_countries <- country_counts %>%
  filter(n < 20) %>%
  pull(country_normalized)

total_genome_metadata <- total_genome_metadata %>%
  mutate(country_grouped = ifelse(country_normalized %in% small_countries,
                                  "Other",
                                  country_normalized))

country_order <- total_genome_metadata %>%
  filter(country_grouped != "Other") %>%
  distinct(country_grouped, continent) %>%
  arrange(continent, country_grouped) %>%
  pull(country_grouped)

country_order <- unique(country_order)

# Add "Other" as the last level
country_order <- c(country_order, "Other")

total_genome_metadata <- total_genome_metadata %>%
  mutate(country_grouped = factor(country_grouped, levels = country_order))

country_palette <- country_palette[c(country_order)]

ggplot(total_genome_metadata, aes(x = bacterial_species, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(total_genome_metadata, aes(x = bacterial_species, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  facet_grid(~host_type_3)+
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



ggplot(total_genome_metadata, aes(x = bacterial_species, fill = continent)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = continent_palette) +
  facet_grid(~host_type_3)+
  labs(
    x = "Bacterial species",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r countryplots}
#order
# total_genome_metadata$country_grouped <- factor(
#   total_genome_metadata$country_grouped,
#   levels = names(country_palette))

#order also host type 2
total_genome_metadata$host_type_2 <- factor(
  total_genome_metadata$host_type_2,
  levels = c("human_adj", "domestic", "wild")
)

total_genome_metadata%>%
  filter(!is.na(host_type_3))%>%
ggplot( aes(x = host_type_3, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Host type",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


total_genome_metadata%>%
  filter(!is.na(host_type_2))%>%
ggplot( aes(x = host_type_2, fill = country_grouped)) +
  geom_bar(color = "grey20", linewidth = 0.2) +
  scale_fill_manual(values = country_palette) +
  labs(
    x = "Host type",
    y = "Number of MAGs",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
```{r country_counts}
ggplot(country_counts, aes(x= country_normalized, y = n))+
  geom_col()+
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```




## Load data
```{r loop_1}
species_list <- c(
  "lactococcus_lactis",
  "parabacteroides_goldsteinii",
  "hafnia_paralvei",
  "citrobacter_braakii",
  "enterococcus_hirae",
  "akkermansia_muciniphila", 
  "enterococcus_faecalis",
  "phocaeicola_vulgatus",
  "bacteroides_fragilis",
  "bacteroides_uniformis"
)


mags_to_remove_list <- list(
  lactococcus_lactis = c(
    "GCA_018369575.1.fna","GCA_947063445.1.fna","GCA_947101685.1.fna",
    "GCA_948698275.1.fna","GCA_937910935.1.fna","GCA_947072755.1.fna",
    "GCA_948655095.1.fna","GCA_948703095.1.fna","GCA_947041925.1.fna",
    "GCA_947073355.1.fna","GCA_948675165.1.fna","GCA_948718815.1.fna"
  ),
  parabacteroides_goldsteinii = character(0), # none to remove
  hafnia_paralvei = character(0),
  citrobacter_braakii = character(0),
  enterococcus_hirae = character(0),
  akkermansia_muciniphila = character(0),
  enterococcus_faecalis = character(0),
  phocaeicola_vulgatus = c("GCA_040912025.1.fna", "GCA_048453085.1.fna", "GCF_048453085.1"),
  bacteroides_fragilis = character(0),
  bacteroides_uniformis = character(0)
)

species_data <- list()

for (sp in species_list) {
  species_data[[sp]] <- process_one_species(
    species        = sp,
    mags_to_remove = mags_to_remove_list[[sp]]
  )
  
  recreate_drep_tree(species_data[[sp]]$ndb)
  
}
```

```{r wilcoxon tests for host types}


results_tbl <- map_dfr(species_list, function(sp) {
  
  genome_metadata <- species_data[[sp]]$metadata
  
  # Variables to test
  vars_to_test <- c("contamination", "completeness", "genome_size")
  
  map_dfr(vars_to_test, function(v) {
    
    # Build formula: variable ~ host_type_3
    fml <- as.formula(paste0(v, " ~ host_type_3"))
    
    # Run Wilcoxon
    test_res <- wilcox.test(fml, data = genome_metadata)
    
    # Tidy results + add species & variable info
    tidy(test_res) %>%
      mutate(
        species = sp,
        variable = v
      )
  })
})

```
```{r stats}
results_tbl %>% dplyr::select(p.value, species, variable) %>% filter(p.value<0.05)
```


## Genomic similarity (drep)
```{r example_drep_final, fig.height = 10, fig_width = 8}
for (sp in species_list) {

  # Run drep tree construction for THIS species
  drep_results <- drep_tree_metadata(
    genome_metadata = species_data[[sp]]$metadata,
    ndb = species_data[[sp]]$ndb
  )

  p_host_order <- plot_tree_basic(
    tree       = drep_results$tree,
    tip_df     = drep_results$tip_df,
    color_by   = "host_order",
    palette = host_order_colors,
    label_tips = TRUE,
    show_threshold = 99.5
  )
  
  p_host_type_3_2 <- plot_tree_basic(
    tree       = drep_results$tree,
    tip_df     = drep_results$tip_df,
    color_by   = "host_type_3_2",
    palette = host_type_3_2_colors,
    label_tips = TRUE,
    show_threshold = 99.5
  )
  
#   p_country <- plot_tree_basic(
#   tree       = drep_results$tree,
#   tip_df     = drep_results$tip_df,
#   color_by   = "country_normalized",
#   palette    = country_palette
# )

  print(p_host_order)  
  print(p_host_type_3_2) 
  #print(p_country) 

}
```


```{r drep stats}
# for (sp in species_list) {
#     # Stats
#   res <- drep_stat_analysis(
#   tip_df = drep_results$tip_df,
#   ani_matrix = species_data[[sp]]$ndb,
#   host_type_3_colors = host_type_3_colors,
#   host_order_colors = host_order_colors
# )
#   
#   # Print stats
#   print(res$permanova$adonis)
#   print(res$permanova$dispersion)
#   print(res$pairwise$wilcox_within$test)
#   print(res$pairwise$wilcox_cross_vs_within$test)
#   
#   # Plot
#   print(res$pcoa$by_host_type_3)
#   print(res$pcoa$by_host_order)
#   print(res$pairwise$boxplot)
#  
# 
# }
```

## Functional analysis

### Fisher
```{r functional analysis fisher}
# Prepare empty containers for combined tables
kegg_fisher_all       <- list()
amr_fisher_all        <- list()
vf_fisher_all         <- list()
cazy_fisher_all       <- list()
defense_fisher_all    <- list()
antidefense_fisher_all<- list()

# store volcano plots 
volcano_plots <- list(
  kegg   = list(),
  amr    = list(),
  vf     = list(),
  cazy   = list()
)



ann_list <- vector("list", length(species_list))
names(ann_list) <- species_list

for (sp in species_list) {

  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata
  
  
  
 # ---- Remove genomes with NA host_type_3 ----
  valid_ids <- genome_metadata %>%
    dplyr::filter(!is.na(host_type_3)) %>%
    dplyr::pull(ID_clean)

  genome_metadata <- genome_metadata %>%
    dplyr::filter(ID_clean %in% valid_ids)

  genome_annotations <- genome_annotations %>%
    dplyr::filter(genome %in% valid_ids)

  
  
  #collect so that we can have total genome annotations
  ann_list[[sp]] <- dplyr::mutate(genome_annotations, species = sp)

  # Build matrices
  kegg_mat        <- make_annotation_matrix(genome_annotations, "kegg")
  amr_mat         <- make_annotation_matrix(genome_annotations, "resistance_target")
  vf_mat          <- make_annotation_matrix(genome_annotations, "vf")
  cazy_mat        <- make_annotation_matrix(genome_annotations, "cazy")
  defense_mat     <- make_annotation_matrix(genome_annotations, "defense")
  antidefense_mat <- make_annotation_matrix(genome_annotations, "antidefense")

  # Fisher
  kegg_fisher <- run_fisher_tests_general(
  mat = kegg_mat,
  metadata = genome_metadata,
  group_col = "host_type_3",
  annotation_name = "kegg"
  )

  amr_fisher <- run_fisher_tests_general(
    mat = amr_mat,
    metadata = genome_metadata,
    group_col = "host_type_3",
    annotation_name = "amr"
  )
  
  vf_fisher <- run_fisher_tests_general(
    mat = vf_mat,
    metadata = genome_metadata,
    group_col = "host_type_3",
    annotation_name = "vf"
  )
  
  cazy_fisher <- run_fisher_tests_general(
    mat = cazy_mat,
    metadata = genome_metadata,
    group_col = "host_type_3",
    annotation_name = "cazy"
  )
  defense_fisher     <- run_fisher_tests(defense_mat,     genome_metadata, "defense")
  #antidefense_fisher <- run_fisher_tests(antidefense_mat, genome_metadata, "antidefense")

  # Add species column before binding
  kegg_fisher$species        <- sp
  amr_fisher$species         <- sp
  vf_fisher$species          <- sp
  cazy_fisher$species        <- sp
  defense_fisher$species     <- sp
  #antidefense_fisher$species <- sp

  # Accumulate
  kegg_fisher_all[[sp]]        <- kegg_fisher
  amr_fisher_all[[sp]]         <- amr_fisher
  vf_fisher_all[[sp]]          <- vf_fisher
  cazy_fisher_all[[sp]]        <- cazy_fisher
  defense_fisher_all[[sp]]     <- defense_fisher
  #antidefense_fisher_all[[sp]] <- antidefense_fisher

  # Volcano
  volcano_kegg <- plot_volcano(kegg_fisher, "kegg")
  volcano_amr  <- plot_volcano(amr_fisher,  "amr")
  volcano_vf   <- plot_volcano(vf_fisher,   "vf")
  volcano_cazy <- plot_volcano(cazy_fisher, "cazy")

  
  volcano_plots$kegg[[sp]] <- volcano_kegg
  volcano_plots$amr[[sp]]  <- volcano_amr
  volcano_plots$vf[[sp]]   <- volcano_vf
  volcano_plots$cazy[[sp]] <- volcano_cazy

  # print(volcano_kegg)
  # print(volcano_amr)
  # print(volcano_vf)
  # print(volcano_cazy)
}



```

```{r total_genome_annotations}


total_genome_annotations <- purrr::map_df(
  names(species_data),
  ~ species_data[[.x]]$annotations %>%
      dplyr::select(-start, -end) %>%   # remove problematic columns
      dplyr::mutate(bacterial_species = .x)
)

```


```{r testing kegg count between hosts}
#count KEGG annotations per MAG within each species
kegg_counts_meta <- total_genome_annotations %>%
  dplyr::filter(!is.na(kegg))%>%
  dplyr::left_join(
    total_genome_metadata %>%
      dplyr::select(ID, host_type_3),         
    join_by(genome == ID)
  ) %>%
  dplyr::group_by(bacterial_species, genome, host_type_3)%>%
  dplyr::summarise(n_kegg = dplyr::n(), .groups = "drop") %>%
  filter(!is.na(host_type_3))

# Run test per species
kegg_host_tests <- kegg_counts_meta %>%
  split(.$bacterial_species) %>%
  map_df(function(df) {
    if (length(unique(df$host_type_3)) < 2) {
      return(tibble(
        bacterial_species = unique(df$bacterial_species),
        p_value = NA,
        test = "Not enough groups"
      ))
    }
    wt <- wilcox.test(n_kegg ~ host_type_3, data = df)

    tibble(
      species = unique(df$bacterial_species),
      p_value = wt$p.value,
      median_human_adj  = median(df$n_kegg[df$host_type_3 == "human_adj"]),
      median_wild = median(df$n_kegg[df$host_type_3 == "wild"]),
      test = "Wilcoxon rank-sum"
    )
  })


kegg_host_tests

```
```{r kegg_hosts_p }
kegg_host_tests %>% filter(p_value < 0.05)
```



```{r bind the rows}
kegg_fisher_total        <- dplyr::bind_rows(kegg_fisher_all)
amr_fisher_total         <- dplyr::bind_rows(amr_fisher_all)
vf_fisher_total          <- dplyr::bind_rows(vf_fisher_all)
cazy_fisher_total        <- dplyr::bind_rows(cazy_fisher_all)
defense_fisher_total     <- dplyr::bind_rows(defense_fisher_all)
antidefense_fisher_total <- dplyr::bind_rows(antidefense_fisher_all)
```




```{r testing keggs totals, fig.height = 12, fig.width = 15}
kegg_df <- kegg_fisher_total %>%
  mutate(
    neg_log10_padj = -log10(p_adj),
    sig = p_adj < 0.05,
    direction = case_when(
      !sig ~ "NS",
      sig & log2_or > 0 ~ "human_adj↑",
      sig & log2_or < 0 ~ "wild↑",
      TRUE ~ "NS"
    ),
    direction = factor(direction, levels = c("wild↑","NS","human_adj↑"))
  )


volcano <- ggplot(kegg_df, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = direction, alpha = sig), size = 1.6) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  scale_color_manual(values = c("human_adj↑" = "#CC3311", "NS" = "grey70", "wild↑" = "#3366CC")) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "Direction",
    title = "Volcano plots KEGG per species"
  ) +
  theme_minimal(base_size = 12) +
  facet_wrap(~ species, scales = "free")

top_hits <- kegg_df %>%
  group_by(species) %>%
  filter(p_adj < 0.05)%>%
  arrange(p_adj) %>%
  #dplyr::select(species, kegg, log2_or, neg_log10_padj, p_adj, direction) %>%
  slice_head(n = 5) %>%
  ungroup()%>%
  print()

volcano +
  ggrepel::geom_text_repel(
    data = top_hits,
    aes(label = kegg),
    size = 3,
    max.overlaps = 8
  )








vf_df <- vf_fisher_total %>%
  mutate(
    neg_log10_padj = -log10(p_adj),
    sig = p_adj < 0.05,
    direction = case_when(
      !sig ~ "NS",
      sig & log2_or > 0 ~ "human_adj↑",
      sig & log2_or < 0 ~ "wild↑",
      TRUE ~ "NS"
    ),
    direction = factor(direction, levels = c("wild↑","NS","human_adj↑"))
  )


volcano <- ggplot(vf_df, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = direction, alpha = sig), size = 1.6) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  scale_color_manual(values = c("human_adj↑" = "#CC3311", "NS" = "grey70", "wild↑" = "#3366CC")) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "Direction",
    title = "Volcano plots VF per species"
  ) +
  theme_minimal(base_size = 14) +
  facet_wrap(~ species, scales = "free")

top_hits <- vf_df %>%
  group_by(species) %>%
  filter(p_adj < 0.05)%>%
  arrange(p_adj) %>%
  #dplyr::select(species, vf, log2_or, neg_log10_padj, p_adj, direction) %>%
  slice_head(n = 5) %>%
  ungroup()%>%
  print()

volcano +
  ggrepel::geom_text_repel(
    data = top_hits,
    aes(label = vf),
    size = 3,
    max.overlaps = 8
  )





amr_df <- amr_fisher_total %>%
  mutate(
    neg_log10_padj = -log10(p_adj),
    sig = p_adj < 0.05,
    direction = case_when(
      !sig ~ "NS",
      sig & log2_or > 0 ~ "human_adj↑",
      sig & log2_or < 0 ~ "wild↑",
      TRUE ~ "NS"
    ),
    direction = factor(direction, levels = c("wild↑","NS","human_adj↑"))
  )


volcano <- ggplot(amr_df, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = direction, alpha = sig), size = 1.6) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  scale_color_manual(values = c("human_adj↑" = "#CC3311", "NS" = "grey70", "wild↑" = "#3366CC")) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "Direction",
    title = "Volcano plots AMR per species"
  ) +
  theme_minimal(base_size = 14) +
  facet_wrap(~ species, scales = "free")

top_hits <- amr_df %>%
  group_by(species) %>%
  filter(p_adj < 0.05)%>%
  arrange(p_adj) %>%
  #dplyr::select(species, amr, log2_or, neg_log10_padj, p_adj, direction) %>%
  slice_head(n = 5) %>%
  ungroup()%>%
  print()



volcano +
  ggrepel::geom_text_repel(
    data = top_hits,
    aes(label = amr),
    size = 3,
    max.overlaps = 8
  )







cazy_df <- cazy_fisher_total %>%
  mutate(
    neg_log10_padj = -log10(p_adj),
    sig = p_adj < 0.05,
    direction = case_when(
      !sig ~ "NS",
      sig & log2_or > 0 ~ "human_adj↑",
      sig & log2_or < 0 ~ "wild↑",
      TRUE ~ "NS"
    ),
    direction = factor(direction, levels = c("wild↑","NS","human_adj↑"))
  )


volcano <- ggplot(cazy_df, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = direction, alpha = sig), size = 1.6) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  scale_color_manual(values = c("human_adj↑" = "#CC3311", "NS" = "grey70", "wild↑" = "#3366CC")) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "Direction",
    title = "Volcano plots cazy per species"
  ) +
  theme_minimal(base_size = 14) +
  facet_wrap(~ species, scales = "free")

top_hits <- cazy_df %>%
  group_by(species) %>%
  filter(p_adj < 0.05)%>%
  arrange(p_adj) %>%
  #dplyr::select(species, cazy, log2_or, neg_log10_padj, p_adj, direction) %>%
  slice_head(n = 5) %>%
  ungroup()

print(top_hits)


volcano +
  ggrepel::geom_text_repel(
    data = top_hits,
    aes(label = cazy),
    size = 3,
    max.overlaps = 8
  )

```


```{r volcano by pathways, eval = FALSE}
#keep only significant KOs 
sig_df <- kegg_df %>%
  filter(p_adj < 0.05) %>%
  mutate(ko = paste0("ko:", kegg))   # (when you download the table they are like this)

# KO -> pathway links (only for sig KOs)
ko_links <- KEGGREST::keggLink("pathway", "ko") %>%
  tibble::enframe(name = "ko", value = "pathway") %>%
  semi_join(sig_df, by = "ko")

# keep only 'map' namespace
ko_links <- ko_links %>%
  mutate(pathway_map = sub("^path:(?:ko|map)(\\d{5})$", "path:map\\1", pathway)) %>%
  filter(str_detect(pathway_map, "^path:map\\d{5}$")) %>%
  distinct(ko, pathway_map)

#  One pathway per KO 
ko_links_one <- ko_links %>%
  group_by(ko) %>%
  slice(1) %>%
  ungroup()

# Fetch pathway info 
paths <- unique(ko_links_one$pathway_map)

get_info <- function(pid){
  x <- try(KEGGREST::keggGet(pid)[[1]], silent = TRUE)
  if (inherits(x, "try-error")) return(tibble(pathway_map=pid, pathway_name=NA, class=NA))
  tibble(
    pathway_map = pid,
    pathway_name = x$NAME[[1]],
    class = ifelse(is.null(x$CLASS), NA, x$CLASS[[1]])
  )
}

path_info <- bind_rows(lapply(paths, get_info))

# high-level category
path_info <- path_info %>%
  mutate(category = sub(";.*$", "", class))

# Join annotations back to significant df
annot_df <- sig_df %>%
  left_join(ko_links_one, by = "ko") %>%
  left_join(path_info, by = "pathway_map")

```

```{r volcano plot colorful, fig.height=6, fig.width = 9, eval = FALSE}

ggplot(annot_df, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = category, alpha = sig), size = 1.6) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = "-log(adj p-value)",
    color = "KEGG class",
    title = "Volcano plots KEGG per species"
  ) +
  theme_minimal(base_size = 12) +
  facet_wrap(~ species, scales = "free")

```

```{r testing again, eval = FALSE}
# 
# # 1) Normalize KO column
# kegg_df2 <- kegg_df %>%
#   mutate(ko = paste0("ko:", kegg))
# 
# # 2) Significant KOs only (for annotation)
# sig_kos <- kegg_df2 %>% filter(p_adj < 0.05)
# 
# # 3) KO → pathway links (only sig)
# ko_links <- KEGGREST::keggLink("pathway", "ko") %>%
#   tibble::enframe(name = "ko", value = "pathway") %>%
#   semi_join(sig_kos, by = "ko") %>%
#   mutate(pathway_map = sub("^path:(?:ko|map)(\\d{5})$", "path:map\\1", pathway)) %>%
#   filter(str_detect(pathway_map, "^path:map\\d{5}$")) %>%
#   distinct(ko, pathway_map)
# 
# # 4) One pathway per KO
# ko_links1 <- ko_links %>% group_by(ko) %>% slice(1) %>% ungroup()
# 
# # 5) Fetch pathway info
# paths <- unique(ko_links1$pathway_map)
# 
# get_info <- function(pid){
#   x <- try(KEGGREST::keggGet(pid)[[1]], silent = TRUE)
#   if (inherits(x, "try-error") || is.null(x))
#     return(tibble(pathway_map=pid, class=NA, category=NA))
#   tibble(
#     pathway_map = pid,
#     class = ifelse(is.null(x$CLASS), NA, x$CLASS[[1]]),
#     category = sub(";.*$", "", ifelse(is.null(x$CLASS), NA, x$CLASS[[1]]))
#   )
# }
# 
# path_info <- bind_rows(lapply(paths, get_info))

```


```{r tete,  fig.height = 10, fig.width = 14}
#annot_full <- readRDS("data/kegg_annotation.rds")

#annot_full <- annot_full  %>% tidyr::separate(class, into = c("class_higher", "class_lower"), sep = ";\\s*", fill = "right", remove = TRUE)



ggplot(annot_full, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  # non-significant in grey
  geom_point(data = annot_full %>% filter(!sig), color = "grey75", size = 1.5, alpha = 0.6) +
  # significant colored by lower class
  geom_point(data = annot_full %>% filter(sig),
             aes(color = class_lower),
             size = 1.8, alpha = 0.95) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "KEGG class (lower)",
    title = "Volcano plots KEGG per species"
  ) +
  theme_minimal(base_size = 12) +
  facet_wrap(~ species, scales = "free")


ggplot(annot_full, aes(log2_or, -log10(p_adj))) +
  geom_point(aes(color = category), alpha = 0.9, size = 1.5) +
  scale_color_manual(values = c("grey80", RColorBrewer::brewer.pal(8, "Dark2"))) +
  facet_wrap(~ species, scales = "free") +
  theme_minimal()

ggplot(annot_full, aes(log2_or, -log10(p_adj))) +
  geom_point(aes(color = class_lower), alpha = 0.9, size = 1.5) +
  facet_wrap(~ species, scales = "free") +
  theme_minimal()



topN <- 8
top_lower <- annot_full %>%
  filter(sig, !is.na(class_lower)) %>%
  count(class_lower, sort = TRUE) %>%
  slice_head(n = topN) %>%
  pull(class_lower)

annot_full2 <- annot_full %>%
  mutate(class_lower_top = ifelse(sig & class_lower %in% top_lower, class_lower,
                                  ifelse(sig & !is.na(class_lower), "Other", NA_character_)))

ggplot(annot_full2, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(data = annot_full2 %>% filter(!sig), color = "grey75", size = 1.5, alpha = 0.6) +
  geom_point(data = annot_full2 %>% filter(sig),
             aes(color = class_lower_top),
             size = 1.8, alpha = 0.95) +
  scale_color_manual(values = setNames(
    c(RColorBrewer::brewer.pal(max(3, min(8, length(top_lower) + 1)), "Set1")[seq_len(length(top_lower))],
      "#999999"),
    c(top_lower, "Other")
  )) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "KEGG class (lower)",
    title = "Volcano plots KEGG per species (top classes)"
  ) +
  theme_minimal(base_size = 12) +
  facet_wrap(~ species)

```


```{r testing sizes, fig.height = 8, fig.width = 12}
ggplot(annot_full2, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(data = annot_full2 %>% filter(!sig), color = "grey75", size = 1.5, alpha = 0.6) +
  geom_point(data = annot_full2 %>% filter(sig),
             aes(color = class_lower_top),
             size = 1.8, alpha = 0.95) +
  scale_color_manual(values = setNames(
    c(RColorBrewer::brewer.pal(max(3, min(8, length(top_lower) + 1)), "Set1")[seq_len(length(top_lower))],
      "#999999"),
    c(top_lower, "Other")
  )) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = expression(-log[10]("adj p-value")),
    color = "KEGG class (lower)",
    title = "Volcano plots KEGG per species (top classes)"
  ) +
  theme_minimal(base_size = 12) +
  facet_wrap(~ species)

```

```{r top 15 terms for each bacteria}
top15_sig_per_species <- annot_full2 %>%
  filter(p_adj < 0.05) %>%
  group_by(species) %>%
  arrange(p_adj, desc(abs(odds_ratio)), .by_group = TRUE) %>%
  slice_head(n = 15) %>%
  ungroup()
```




```{r save annot_full}
#saveRDS(annot_full, "data/kegg_annotation_host3.rds")
```




```{r kegg analysis 4}
per_kegg <- kegg_df %>%
  group_by(kegg) %>%
  summarise(
    n_species = dplyr::n_distinct(species),
    n_sig = sum(sig),
    all_sig = n_sig == n_species,
    # consistency of direction (all positive vs all negative among the significant ones)
    all_wild_up = all(log2_or[sig] < 0),
    all_human_adj_up  = all(log2_or[sig] > 0),
    .groups = "drop"
  )

per_kegg %>% arrange(-n_species, -n_sig) %>% head(12)

per_vf <- vf_df %>%
  group_by(vf) %>%
  summarise(
    n_species = dplyr::n_distinct(species),
    n_sig = sum(sig),
    all_sig = n_sig == n_species,
    # consistency of direction (all positive vs all negative among the significant ones)
    all_wild_up = all(log2_or[sig] < 0),
    all_human_adj_up  = all(log2_or[sig] > 0),
    .groups = "drop"
  )

per_vf %>% arrange(-n_species, -n_sig) %>% head(12)
```

##### Virulence factors
```{r vf_data}
vf_data <- total_genome_annotations %>% dplyr::select(vf, vf_type) %>% unique()
```

```{r vf_volcano, fig.height = 8, fig.width = 12}

vf_df_annot <- left_join(vf_df, vf_data, by = "vf")

# do not color the non-significant (keep them grey)
vf_df_annot <- vf_df_annot %>%
  mutate(color_sig = ifelse(p_adj > 0.05, "nonsig", vf_type))

# Create named colors: grey for nonsig + a palette for VF types
vf_types <- unique(vf_df_annot$vf_type)  


vf_colors <- c(
  nonsig = "grey70",
  setNames(colorRampPalette(RColorBrewer::brewer.pal(8, "Set1"))(length(vf_types)),
           vf_types)
)



volcano <- ggplot(vf_df_annot, aes(x = log2_or, y = neg_log10_padj)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(aes(color = color_sig, alpha = sig), size = 1.6) +
  scale_color_manual(values = vf_colors) +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.4)) +
  labs(
    x = "log2(odds ratio)  (wild vs human_adj)",
    y = "-log(adj p-value)",
    color = "VF type",
    title = "Volcano plots VF per species"
  ) +
  theme_minimal(base_size = 14) +
  facet_wrap(~ species, scales = "free")


top_hits <- vf_df %>%
  group_by(species) %>%
  filter(p_adj < 0.05)%>%
  arrange(p_adj) %>%
  dplyr::select(species, vf, log2_or, neg_log10_padj, p_adj, direction) %>%
  slice_head(n = 5) %>%
  ungroup()

volcano +
  ggrepel::geom_text_repel(
    data = top_hits,
    aes(label = vf),
    size = 3,
    max.overlaps = 8
  )
```




#### META-ANALYSIS
```{r metatest with one kegg}
df <- kegg_fisher_total %>% filter(kegg == "K03328")


m <- metabin(
  event.e = n_mags_human_adj,
  n.e = n_mags_human_adj + absent_human_adj,
  event.c = n_mags_wild,
  n.c = n_mags_wild + absent_wild,
  
  studlab = species,
  data = df,
  
  sm = "OR",
  method = "MH",           # Mantel-Haenszel
  random = TRUE,
  fixed = FALSE,
  
  incr = 0.5,              # correction
  allstudies = FALSE       # only apply if needed
)

summary(m)


forest(m)


# m_pre_calc <- metagen(
#   TE = log_or,
#   seTE = se_log_or,
#   studlab = species,
#   data = df,
#   sm = "OR",
#   random = TRUE,
#   fixed = FALSE,
#   method.tau = "REML"
# )
# 
# forest(m_pre_calc)
```

```{r ifluence analsysi}
i_a_m <- InfluenceAnalysis(m, random = TRUE)
plot(i_a_m, "baujat")
```


```{r metaanalysis for all keggs}
# Prep totals and basic filters 
df_prepped <- kegg_fisher_total %>%
  mutate(
    tot_wild = n_mags_wild + absent_wild,
    tot_human_adj  = n_mags_human_adj  + absent_human_adj
  ) %>%
  # Require nonzero denominators
  filter(tot_wild > 0, tot_human_adj > 0) %>%
  # Drop double-zero rows (no information for OR in IV/MH-RE)
  filter(!(n_mags_wild == 0 & n_mags_human_adj == 0))

#  Low-bias screen 
min_informative_species <- 8   # at least 8 informative species (non double-zero)
min_total_events        <- 10   # total presences across all species (human_adj + wild)

kegg_keep <- df_prepped %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = sum(!(n_mags_wild == 0 & n_mags_human_adj == 0)),
    total_events     = sum(n_mags_wild + n_mags_human_adj)
  ) %>%
  filter(n_species_inform >= min_informative_species,
         total_events >= min_total_events) %>%
  dplyr::select(kegg) %>%
  ungroup()

df_screened <- df_prepped %>% inner_join(kegg_keep, by = "kegg")


# Split by KEGG
kegg_split <- split(df_screened, df_screened$kegg)

# Effect size pooling
# MH with random effects; human_adj vs wild 
run_mh_random_one_kegg <- function(dat_kegg) {
  meta::metabin(
    # human_adj as experimental (so OR>1 => more prevalent in human_adj)
    event.e = dat_kegg$n_mags_human_adj,
    n.e     = dat_kegg$tot_human_adj,
    event.c = dat_kegg$n_mags_wild,
    n.c     = dat_kegg$tot_wild,

    studlab = dat_kegg$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # Continuity correction: add 0.5 only if a zero cell is present
    incr        = 0.5,
    method.incr = "if0"
  )
}

mh_re_list <- lapply(kegg_split, run_mh_random_one_kegg)

# Collect pooled results + FDR 
mh_re_table <- lapply(names(mh_re_list), function(k) {
  m <- mh_re_list[[k]]
  data.frame(
    kegg           = k,
    n_species      = m$k,
    logOR_random   = m$TE.random,
    se_random      = m$seTE.random,
    OR_random      = exp(m$TE.random),
    ci_low_random  = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random       = m$pval.random,
    tau2_REML      = m$tau^2,
    I2_percent     = m$I2,
    pred_low       = exp(m$lower.predict),
    pred_high      = exp(m$upper.predict),
    stringsAsFactors = FALSE
  )
}) %>% dplyr::bind_rows() %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))

```



```{r check metatable}
mh_re_table %>% arrange(-n_species, p_adj_fdr ) %>% head(10)
```


```{r manual lookup of keggs}
kegg_search_info <- tibble(
  kegg = c(
    "K18799",
    "K03436",
    "K01356",
    "K07461",
    "K13573",
    "K03580",
    "K09144",
    "K07497",
    "K03798",
    "K07317",
    "K03427",
    "K06871",
    "K02935",
    "K02864",
    "K07485"
  ),
  description = c(
    "O-antigen flippase",
    "DeoR family transcriptional regulator",
    "Repressor LexA",
    "Putative endonuclease",
    "Proteasome accessory factor C",
    "ATP-dependent helicase HepA",
    "Nucleotidyl transferase AbiEii/AbiGii toxin-family protein ",
    "Putative transposase",
    "Cell division protease FtsH",
    "Adenine-specific DNA methyltransferase",
    "Type I restriction enzyme M protein",
    "Uncharacterized protein",
    "Ribosomal protein L7/L12",
    "Ribosomal protein L10",
    "Transposase"
  ), 
  extra_info = c(
    " ",
    "fructose operon transcriptional repressor",
    "serine endopeptidase, SOS response regulator",
    " ",
    "pafC",
    " ",
    "Type IV TA system toxin",
    "",
    "metallo-endopeptidase",
    "Type II R-M system",
    "adenine-specific DNA methyltransferase",
    "function unknown",
    "large subunit",
    "large subunit",
    " "
    
  )
)

kegg_search_info
```


```{r kegg forest plot metameta}
top_n <- 15

best_tbl <- mh_re_table %>%
  dplyr::arrange(p_adj_fdr, p_random) %>%
  dplyr::slice_head(n = top_n) 

mgen <- meta::metagen(
  TE    = best_tbl$logOR_random,
  seTE  = best_tbl$se_random,
  studlab = best_tbl$kegg,
  sm = "OR",
  # No additional pooling across the KEGGs themselves — we just want rows
  common = FALSE,
  random = FALSE
)

# Forest plot of the pooled OR per KEGG (no overall diamond)
meta::forest(
  mgen,
  backtransf = TRUE,                    # show OR scale instead of log(OR)
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("KEGG pathway"),
  sortvar    = best_tbl$p_adj_fdr,      # sort by FDR 
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

# Save to a PDF
pdf("plots/kegg_top_forest_meta.pdf", width = 8, height = 7)
meta::forest(
  mgen,
  backtransf = TRUE,
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("KEGG pathway"),
  sortvar    = best_tbl$p_adj_fdr
)
dev.off()
```

```{r keggs where ci does not include_1}
mh_re_table_ci <- mh_re_table %>%
  dplyr::filter(ci_low_random > 1 | ci_high_random <1) %>%
  dplyr::arrange(p_adj_fdr, p_random) 


mgen_ci <- metagen(
  TE   = mh_re_table_ci$logOR_random,   
  seTE = mh_re_table_ci$se_random,
  studlab = mh_re_table_ci$kegg,
  sm = "OR",
  common = FALSE, 
  random = FALSE
)


meta::forest(
  mgen_ci,
  backtransf = TRUE,
  xlab = "Odds ratio (human_adj vs wild)",
  leftlabs = c("KEGG pathway"),
  sortvar = mh_re_table_ci$p_adj_fdr,
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)


```



```{r plot with info, fig.height = 6, fig.width = 10}
best_tbl <- best_tbl %>% left_join(kegg_search_info) 

mgen <- meta::metagen(
  TE    = best_tbl$logOR_random,
  seTE  = best_tbl$se_random,
  studlab = best_tbl$description,
  sm = "OR",
  # No additional pooling across the KEGGs themselves — we just want rows
  common = FALSE,
  random = FALSE
)

# Forest plot of the pooled OR per KEGG (no overall diamond)
meta::forest(
  mgen,
  backtransf = TRUE,                    # show OR scale instead of log(OR)
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("KEGG pathway"),
  sortvar    = best_tbl$p_adj_fdr,      # sort by FDR if you like
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```


```{r forest plot individual keggs, fig.height = 6, fid.width = 10}
# Forest plot (human_adj on right, wild on left)
forest_one_kegg <- function(kegg_id, xlim = NULL, at = NULL) {
  m <- mh_re_list[[kegg_id]]
  if (is.null(m)) stop("KEGG not found in mh_re_list: ", kegg_id)

  
  xlab_txt <- "Odds ratio (human_adj vs wild) "

  
 # Try to show events and Ns if they exist
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events A", "N A")

  available <- intersect(wanted_leftcols, names(m))
  leftcols_use <- if (length(available) > 0) available else "studlab"
  leftlabs_use <- wanted_leftlabs[match(leftcols_use, wanted_leftcols)]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,   # keep input order
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,    # e.g., c(0.1, 10)
    at          = at,      # e.g., c(0.1, 0.25, 1, 4, 10)
    digits      = 2,
     main        = paste0("KEGG: ", kegg_id)  #KEGG ID as title
  )

}

#individual analyses of each of these keggs:
keggs <-best_tbl %>% pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}

```
```{r what are these keggs}
kegg_data <- annot_full%>% dplyr::select(kegg, pathway_map, class_higher, 
                                         class_lower, category, class_plot) %>%
   dplyr::distinct(kegg, .keep_all = TRUE)

kegg_info_15 <- left_join(best_tbl, kegg_data)

kegg_info_15
```


##### Find outliers
```{r cheking metaanalysis of first keggs}
find.outliers(mh_re_list$K18799)
find.outliers(mh_re_list$K03436)
find.outliers(mh_re_list$K01356)
```

```{r influence analsysis}
plot(InfluenceAnalysis(mh_re_list$K18799, random = TRUE), "baujat")
plot(InfluenceAnalysis(mh_re_list$K03436, random = TRUE), "baujat")
plot(InfluenceAnalysis(mh_re_list$K01356, random = TRUE), "baujat")
```
```{r influence diagnostics}
plot(InfluenceAnalysis(mh_re_list$K18799, random = TRUE), "i2")
```

##### Subgroup analysis
Translating meta object to metafor
```{r metafor}
library(metafor)

m.rma <- rma(yi = mh_re_list$K18799$TE,
             sei = mh_re_list$K18799$seTE,
             method = mh_re_list$K18799$method.tau,
             test = "knha")
```
```{r gosh plot}
res.gosh <- gosh(m.rma)

plot(res.gosh, alpha = 0.5)
```

```{r gosh diagnostic plots}
res.gosh.diag <- gosh.diagnostics(
  res.gosh,
  km.params = list(centers = 2),
  db.params = list(eps = 0.2, MinPts = 10)
)

res.gosh.diag
```

```{r  res gosh}
plot(res.gosh.diag)
```
#### Meta-analysis BRITE
```{r testing metaanalysis for kegg class}
# get the keggs that have a class
kegg_to_brite <- annot_full %>%
   dplyr::select(kegg, class_lower) %>%
   filter(!is.na(class_lower)) %>% 
   distinct()


#sum the presence/absences of each kegg across classes
df_prepped_path <- df_prepped %>%
   inner_join(kegg_to_brite, by="kegg") %>%
   group_by(species, class_lower) %>%
   summarise(
       n_mags_human_adj  = sum(n_mags_human_adj),
       n_mags_wild = sum(n_mags_wild),
       absent_human_adj  = sum(absent_human_adj),
       absent_wild = sum(absent_wild),
       .groups="drop"
   ) %>%
   mutate(
       tot_human_adj  = n_mags_human_adj  + absent_human_adj,
       tot_wild = n_mags_wild + absent_wild
   )



#  Low-bias screen 
min_informative_species <- 8   # at least 8 informative species (non double-zero)
min_total_events        <- 10   # total presences across all species (human_adj + wild)

class_lower_keep <- df_prepped_path %>%
  group_by(class_lower) %>%
  summarise(
    n_species_inform = sum(!(n_mags_wild == 0 & n_mags_human_adj == 0)),
    total_events     = sum(n_mags_wild + n_mags_human_adj)
  ) %>%
  filter(n_species_inform >= min_informative_species,
         total_events >= min_total_events) %>%
  dplyr::select(class_lower) %>%
  ungroup()

df_screened <- df_prepped_path %>% inner_join(class_lower_keep, by = "class_lower")


# Split by class_lower
class_lower_split <- split(df_screened, df_screened$class_lower)

# Effect size pooling
# MH with random effects; human_adj vs wild 
run_mh_random_one_class_lower <- function(dat_class_lower) {
  meta::metabin(
    # human_adj as experimental (so OR>1 => more prevalent in human_adj)
    event.e = dat_class_lower$n_mags_human_adj,
    n.e     = dat_class_lower$tot_human_adj,
    event.c = dat_class_lower$n_mags_wild,
    n.c     = dat_class_lower$tot_wild,

    studlab = dat_class_lower$species,
    sm      = "OR",
    method  = "MH",

    common  = FALSE,          # do not compute fixed/common summary
    random  = TRUE,           # compute random-effects summary
    method.tau       = "REML",
    method.random.ci = "HK",  # Hartung–Knapp

    prediction = TRUE,

    # Continuity correction: add 0.5 only if a zero cell is present
    incr        = 0.5,
    method.incr = "if0"
  )
}

mh_re_list_2 <- lapply(class_lower_split, run_mh_random_one_class_lower)

# Collect pooled results + FDR 
mh_re_table_2 <- lapply(names(mh_re_list_2), function(k) {
  m <- mh_re_list_2[[k]]
  data.frame(
    class_lower           = k,
    n_species      = m$k,
    logOR_random   = m$TE.random,
    se_random      = m$seTE.random,
    OR_random      = exp(m$TE.random),
    ci_low_random  = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random       = m$pval.random,
    tau2_REML      = m$tau^2,
    I2_percent     = m$I2,
    pred_low       = exp(m$lower.predict),
    pred_high      = exp(m$upper.predict),
    stringsAsFactors = FALSE
  )
}) %>% dplyr::bind_rows() %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



top_n <- 15

best_tbl <- mh_re_table_2 %>%
  dplyr::arrange(p_adj_fdr, p_random) %>%
  dplyr::slice_head(n = top_n) 

mgen_2 <- meta::metagen(
  TE    = best_tbl$logOR_random,
  seTE  = best_tbl$se_random,
  studlab = best_tbl$class_lower,
  sm = "OR",
  # No additional pooling across the classes themselves — we just want rows
  common = FALSE,
  random = FALSE
)


```

```{r forest plot class lower, fig.height = 6, fig.width = 10}
# Forest plot of the pooled OR per class (no overall diamond)
meta::forest(
  mgen_2,
  backtransf = TRUE,                    # show OR scale instead of log(OR)
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("KEGG pathway"),
  sortvar    = best_tbl$p_adj_fdr,      # sort by FDR
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)

```
#### Meta-meta analysis without enterococcus

```{r metameta_ent}
# Prep totals and basic filters 
df_prepped <- kegg_fisher_total %>%
  filter(species != "enterococcus_faecalis")%>%
  mutate(
    tot_wild = n_mags_wild + absent_wild,
    tot_human_adj  = n_mags_human_adj  + absent_human_adj
  ) %>%
  # Require nonzero denominators
  filter(tot_wild > 0, tot_human_adj > 0) %>%
  # Drop double-zero rows (no information for OR in IV/MH-RE)
  filter(!(n_mags_wild == 0 & n_mags_human_adj == 0))

#  Low-bias screen 
min_informative_species <- 8   # at least 8 informative species (non double-zero)
min_total_events        <- 10   # total presences across all species (human_adj + wild)

kegg_keep <- df_prepped %>%
  group_by(kegg) %>%
  summarise(
    n_species_inform = sum(!(n_mags_wild == 0 & n_mags_human_adj == 0)),
    total_events     = sum(n_mags_wild + n_mags_human_adj)
  ) %>%
  filter(n_species_inform >= min_informative_species,
         total_events >= min_total_events) %>%
  dplyr::select(kegg) %>%
  ungroup()

df_screened <- df_prepped %>% inner_join(kegg_keep, by = "kegg")

# Split by KEGG
kegg_split <- split(df_screened, df_screened$kegg)


mh_re_list_went <- lapply(kegg_split, run_mh_random_one_kegg)

# Collect pooled results + FDR 
mh_re_table_went <- lapply(names(mh_re_list_went), function(k) {
  m <- mh_re_list_went[[k]]
  data.frame(
    kegg           = k,
    n_species      = m$k,
    logOR_random   = m$TE.random,
    se_random      = m$seTE.random,
    OR_random      = exp(m$TE.random),
    ci_low_random  = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random       = m$pval.random,
    tau2_REML      = m$tau^2,
    I2_percent     = m$I2,
    pred_low       = exp(m$lower.predict),
    pred_high      = exp(m$upper.predict),
    stringsAsFactors = FALSE
  )
}) %>% dplyr::bind_rows() %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



top_n <- 15

best_tbl_went <- mh_re_table_went %>%
  dplyr::arrange(p_adj_fdr, p_random) %>%
  dplyr::slice_head(n = top_n) 

mgen_went <- meta::metagen(
  TE      = best_tbl_went$logOR_random,
  seTE    = best_tbl_went$se_random,
  studlab = best_tbl_went$kegg,
  sm      = "OR",
  common  = FALSE,
  random  = FALSE
)


# Forest plot of the pooled OR per KEGG (no overall diamond)
meta::forest(
  mgen_went,
  backtransf = TRUE,                    # show OR scale instead of log(OR)
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("KEGG pathway"),
  sortvar    = best_tbl_went$p_adj_fdr,      # sort by FDR 
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```

```{r manual_keggs}
kegg_search_info_went <- tibble(
  kegg = c(
    "K18799",
    "K01356",
    "K03580",
    "K07497",
    "K07317",
    "K03427",
    "K02935",
    "K02864",
    "K02886",
    "K03046",
    "K02965",
    "K02892",
    "K02926",
    "K01154",
    "K02890"
  ),
  
  description = c(
    "O-antigen flippase",
    "Repressor LexA",
    "ATP-dependent helicase HepA",
    "Putative transposase",
    "Adenine-specific DNA methyltransferase",
    "Type I restriction enzyme M protein",
    "Ribosomal protein L7/L12",
    "Ribosomal protein L10",
    "Ribosomal protein L2",
    "RNA polymerase subunit beta'",
    "Ribosomal protein S19",
    "Ribosomal protein L23",
    "Ribosomal protein L4",
    "No KEGG entry found",
    "Ribosomal protein L22"
  ),
  
  extra_info = c(
    "wzxB / rfbX; involved in O-antigen membrane transport",                                       # K18799
    "SOS response regulator; serine-type endopeptidase",                                           # K01356
    "Transcription-associated helicase (RapA)",                                                    # K03580
    "Mobile element protein",                                                                       # K07497
    "Type II restriction–modification system methyltransferase",                                    # K07317
    "Adenine-specific DNA methyltransferase (Type I R-M system)",                                   # K03427
    "Large ribosomal subunit",                                                                      # K02935
    "Large ribosomal subunit",                                                                      # K02864
    "Large ribosomal subunit; peptidyl transferase center",                                         # K02886
    "Core enzyme of prokaryotic RNA polymerase",                                                    # K03046
    "Small ribosomal subunit",                                                                       # K02965
    "Large ribosomal subunit",                                                                      # K02892
    "Large ribosomal subunit",                                                                      # K02926
    "Not found in KEGG searches",                                                                    # K01154
    "Large ribosomal subunit"                                                                       # K02890
  )
)

kegg_search_info_went
```



```{r forest plot individual keggs_ent, fig.height = 6, fid.width = 10}
#individual analyses of each of these keggs:
keggs <-best_tbl_went %>% pull(kegg)

for (k in keggs) {
  forest_one_kegg(
    kegg_id = k,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}

```



#### Meta-meta analysis with VF


```{r metameta_vf}
# Prep totals and basic filters 
vf_prepped <- vf_fisher_total %>%
  mutate(
    tot_wild = n_mags_wild + absent_wild,
    tot_human_adj  = n_mags_human_adj  + absent_human_adj
  ) %>%
  # Require nonzero denominators
  filter(tot_wild > 0, tot_human_adj > 0) %>%
  # Drop double-zero rows (no information for OR in IV/MH-RE)
  filter(!(n_mags_wild == 0 & n_mags_human_adj == 0))

#  Low-bias screen 
min_informative_species <- 8   # at least 8 informative species (non double-zero)
min_total_events        <- 10   # total presences across all species (human_adj + wild)

vf_keep <- vf_prepped %>%
  group_by(vf) %>%
  summarise(
    n_species_inform = sum(!(n_mags_wild == 0 & n_mags_human_adj == 0)),
    total_events     = sum(n_mags_wild + n_mags_human_adj)
  ) %>%
  filter(n_species_inform >= min_informative_species,
         total_events >= min_total_events) %>%
  dplyr::select(vf) %>%
  ungroup()

vf_screened <- vf_prepped %>% inner_join(vf_keep, by = "vf")

# Split by vf
vf_split <- split(vf_screened, vf_screened$vf)


mh_re_list_vf <- lapply(vf_split, run_mh_random_one_kegg)

# Collect pooled results + FDR 
mh_re_table_vf <- lapply(names(mh_re_list_vf), function(k) {
  m <- mh_re_list_vf[[k]]
  data.frame(
    vf           = k,
    n_species      = m$k,
    logOR_random   = m$TE.random,
    se_random      = m$seTE.random,
    OR_random      = exp(m$TE.random),
    ci_low_random  = exp(m$lower.random),
    ci_high_random = exp(m$upper.random),
    p_random       = m$pval.random,
    tau2_REML      = m$tau^2,
    I2_percent     = m$I2,
    pred_low       = exp(m$lower.predict),
    pred_high      = exp(m$upper.predict),
    stringsAsFactors = FALSE
  )
}) %>% dplyr::bind_rows() %>%
  mutate(p_adj_fdr = p.adjust(p_random, method = "fdr"))



top_n <- 6

best_tbl_vf <- mh_re_table_vf %>%
  dplyr::arrange(p_adj_fdr, p_random) %>%
  dplyr::slice_head(n = top_n) 

mgen_vf <- meta::metagen(
  TE      = best_tbl_vf$logOR_random,
  seTE    = best_tbl_vf$se_random,
  studlab = best_tbl_vf$vf,
  sm      = "OR",
  common  = FALSE,
  random  = FALSE
)


# Forest plot of the pooled OR per KEGG (no overall diamond)
meta::forest(
  mgen_vf,
  backtransf = TRUE,                    # show OR scale instead of log(OR)
  xlab       = "Odds ratio (human_adj vs wild)",
  leftlabs   = c("VF"),
  sortvar    = best_tbl_vf$p_random,      # sort by FDR 
  col.square = "black",
  col.diamond = "black",
  col.diamond.lines = "black"
)
```

```{r vf_types}
best_tbl_vf %>% left_join(vf_data)
```



```{r forest plot individual keggs_vf, fig.height = 6, fid.width = 10}
forest_one_vf <- function(vf_id, xlim = NULL, at = NULL) {
  m <- mh_re_list_vf[[vf_id]]
  if (is.null(m)) stop("VF not found in mh_re_list_vf: ", vf_id)

  
  xlab_txt <- "Odds ratio (human_adj vs wild) "

  
 # Try to show events and Ns if they exist
  wanted_leftcols <- c("studlab", "event.e", "n.e", "event.c", "n.c")
  wanted_leftlabs <- c("Species", "Events H", "N H", "Events A", "N A")

  available <- intersect(wanted_leftcols, names(m))
  leftcols_use <- if (length(available) > 0) available else "studlab"
  leftlabs_use <- wanted_leftlabs[match(leftcols_use, wanted_leftcols)]

  meta::forest(
    m,
    overall     = TRUE,
    prediction  = TRUE,
    sortvar     = NULL,   # keep input order
    leftcols    = leftcols_use,
    leftlabs    = leftlabs_use,
    rightlabs   = c("OR", "95% CI"),
    xlab        = xlab_txt,
    xlim        = xlim,    # e.g., c(0.1, 10)
    at          = at,      # e.g., c(0.1, 0.25, 1, 4, 10)
    digits      = 2,
     main        = paste0("VF: ", vf_id)  #KEGG ID as title
  )

}

vfs <-best_tbl_vf %>% pull(vf)

for (vf in vfs) {
  forest_one_vf(
    vf_id = vf,
    xlim    = c(0.1, 10),
    at      = c(0.1, 0.25, 1, 4, 10)
  )
}

```


```{r testing with one species metaanalysis}
df <- kegg_fisher_total %>% filter(species == "bacteroides_uniformis",
                                   p_adj < 0.05,
                                   abs(log2_or) > 3)

```


```{r metaanalyis_2_per_kegg, fig.height = 7}


m <- metabin(
  event.e = n_mags_human_adj,
  n.e = n_mags_human_adj + absent_human_adj,
  event.c = n_mags_wild,
  n.c = n_mags_wild + absent_wild,
  
  studlab = kegg,
  data = df,
  
  sm = "OR",
  method = "MH",           # Mantel-Haenszel
  random = TRUE,
  fixed = FALSE,
  
  incr = 0.5,              # correction
  allstudies = FALSE       # only apply if needed
)

summary(m)


forest(m)
```



### PERMANOVA
```{r functional analysis permanova pcoa}

# Helper: extract PERMANOVA table into tidy rows
.extract_permanova_rows <- function(permanova_obj, species, matrix_name) {
  if (is.null(permanova_obj)) return(tibble::tibble())
  df <- tryCatch(
    {
      if (inherits(permanova_obj, "adonis") || inherits(permanova_obj, "adonis2")) {
        as.data.frame(permanova_obj)
      } else if (!is.null(permanova_obj$aov.tab)) {
        as.data.frame(permanova_obj$aov.tab)
      } else if (is.data.frame(permanova_obj)) {
        permanova_obj
      } else {
        as.data.frame(permanova_obj)
      }
    },
    error = function(e) NULL
  )
  if (is.null(df) || nrow(df) == 0) return(tibble::tibble())

  # Add the term from rownames if not already present
  if (is.null(df$term)) df$term <- rownames(df)

  df$species     <- species
  df$matrix_name <- matrix_name

  # Move species/matrix/term up front
  df <- df[, c("species", "matrix_name", "term", setdiff(names(df), c("species","matrix_name","term"))), drop = FALSE]
  tibble::as_tibble(df, .name_repair = "minimal")
}

# Helper: extract betadisper results (ANOVA table + optional permutest p-value)
.extract_betadisper_rows <- function(betadisper_obj, species, matrix_name) {
  if (is.null(betadisper_obj)) return(tibble::tibble())

  out <- list()

  # ANOVA table
  an_df <- tryCatch(
    {
      an <- stats::anova(betadisper_obj)
      df <- as.data.frame(an)
      df$term <- rownames(an)
      df
    },
    error = function(e) NULL
  )
  if (!is.null(an_df)) {
    an_df$type <- "anova"
    out[["anova"]] <- an_df
  }

  # permutest p-value (if vegan is available)
  pt_df <- tryCatch(
    {
      pt <- vegan::permutest(betadisper_obj)
      pval <- tryCatch(pt$tab[1, "Pr(>F)"], error = function(e) NA_real_)
      data.frame(term = "Groups", `Pr(>F)` = pval, stringsAsFactors = FALSE)
    },
    error = function(e) NULL
  )
  if (!is.null(pt_df)) {
    pt_df$type <- "permutest"
    out[["permutest"]] <- pt_df
  }

  if (length(out) == 0) return(tibble::tibble())

  res <- dplyr::bind_rows(out)
  res$species     <- species
  res$matrix_name <- matrix_name
  res <- res[, c("species", "matrix_name", "type", "term", setdiff(names(res), c("species","matrix_name","type","term"))), drop = FALSE]
  tibble::as_tibble(res, .name_repair = "minimal")
}




# Accumulators
permanova_rows  <- list()
betadisper_rows <- list()
row_idx <- 0  # simple counter

for (sp in species_list) {
  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata    <- species_data[[sp]]$metadata

  # Build matrices
  kegg_mat        <- make_annotation_matrix(genome_annotations, "kegg")
  amr_mat         <- make_annotation_matrix(genome_annotations, "resistance_target")
  vf_mat          <- make_annotation_matrix(genome_annotations, "vf")
  cazy_mat        <- make_annotation_matrix(genome_annotations, "cazy")
  defense_mat     <- make_annotation_matrix(genome_annotations, "defense")
  antidefense_mat <- make_annotation_matrix(genome_annotations, "antidefense")

  # PERMANOVAs 
  kegg  <- tryCatch(pa_permanova(kegg_mat,        genome_metadata), error = function(e) { message("[", sp, "] KEGG failed: ", e$message); NULL })
  amr   <- tryCatch(pa_permanova(amr_mat,         genome_metadata), error = function(e) { message("[", sp, "] AMR failed: ", e$message); NULL })
  vf    <- tryCatch(pa_permanova(vf_mat,          genome_metadata), error = function(e) { message("[", sp, "] VF failed: ", e$message); NULL })
  cazy  <- tryCatch(pa_permanova(cazy_mat,        genome_metadata), error = function(e) { message("[", sp, "] CAZY failed: ", e$message); NULL })
  defense <- tryCatch(pa_permanova(defense_mat,   genome_metadata), error = function(e) { message("[", sp, "] DEFENSE failed: ", e$message); NULL })
  antidefense <- tryCatch(pa_permanova(antidefense_mat, genome_metadata), error = function(e) { message("[", sp, "] ANTIDEFENSE failed: ", e$message); NULL })

  # Collect results into tables
  res_list <- list(
    KEGG        = kegg,
    AMR         = amr,
    VF          = vf,
    CAZY        = cazy
   
    # DEFENSE     = defense,
    # ANTIDEFENSE = antidefense
  )

  for (nm in names(res_list)) {
    obj <- res_list[[nm]]
    if (is.null(obj)) next

    # PERMANOVA rows
    pr <- .extract_permanova_rows(obj$permanova, species = sp, matrix_name = nm)
    if (nrow(pr) > 0) {
      row_idx <- row_idx + 1
      permanova_rows[[row_idx]] <- pr
    }

    # betadisper rows
    br <- .extract_betadisper_rows(obj$betadisper, species = sp, matrix_name = nm)
    if (nrow(br) > 0) {
      row_idx <- row_idx + 1
      betadisper_rows[[row_idx]] <- br
    }
  }

  # PCoA plotting
  
# Choose which PERMANOVA result objects to plot
mat_objs <- list(
  KEGG        = kegg,
  AMR         = amr,
  VF          = vf,
  CAZY        = cazy
  # You can include these too if you want PCoAs for them:
  # DEFENSE     = defense,
  # ANTIDEFENSE = antidefense
)

for (nm in names(mat_objs)) {
  obj <- mat_objs[[nm]]
  if (is.null(obj) || is.null(obj$dist) || is.null(obj$pa_nz)) {
    message(sprintf("[%s - %s] Skipping: missing 'dist' or 'pa_nz'.", sp, nm))
    next
  }

  # --- IDs from dist, pa_nz, and metadata ------------------------------------
  ids_dist <- if (inherits(obj$dist, "dist")) {
    attr(obj$dist, "Labels")
  } else {
    rownames(obj$dist)
  }
  ids_pa   <- rownames(obj$pa_nz)
  ids_meta <- genome_metadata$ID_clean

  common_ids <- Reduce(intersect, list(ids_dist, ids_pa, ids_meta))
  if (length(common_ids) < 3) {
    message(sprintf("[%s - %s] Too few common IDs (%d). Need >= 3 for PCoA.", sp, nm, length(common_ids)))
    next
  }

  # --- Subset dist to common_ids (preserve order = common_ids) ---------------
  dist_mat <- if (inherits(obj$dist, "dist")) as.matrix(obj$dist) else obj$dist
  # Ensure the distance matrix has rownames/colnames to subset by names:
  if (is.null(rownames(dist_mat)) || is.null(colnames(dist_mat))) {
    stop(sprintf("[%s - %s] Distance matrix lacks row/col names; cannot align.", sp, nm))
  }
  dist_sub <- dist_mat[common_ids, common_ids, drop = FALSE]
  dist_sub <- stats::as.dist(dist_sub)

  # --- Subset PA matrix (not strictly needed for plots, but keeps things aligned)
  pa_sub <- obj$pa_nz[common_ids, , drop = FALSE]

  # --- Subset & re-order metadata to match common_ids ------------------------
  meta_sub <- genome_metadata[match(common_ids, genome_metadata$ID_clean), , drop = FALSE]

  # --- PCoA -------------------------------------------------------------------
  pcoa_res <- ape::pcoa(dist_sub, correction = "cailliez")
  # Explained variance (%), using correction-adjusted components:
  var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)

  # Build PCoA data frame (rows are in the same order as dist_sub/common_ids)
  pcoa_df <- data.frame(
    ID         = common_ids,
    PC1        = pcoa_res$vectors[, 1],
    PC2        = pcoa_res$vectors[, 2],
    host_type_3  = meta_sub$host_type_3,
    host_order = meta_sub$host_order,
    stringsAsFactors = FALSE
  )

  # --- Plot 1: color by host_type_3 --------------------------------------------
  p1 <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
    geom_point(size = 2) +
    scale_color_manual(values = host_type_3_colors, name = "Host type") +
    theme_minimal() +
    labs(
      title = paste("PCoA of", nm, "presence/absence matrix across MAGs (colored by host type) -", sp),
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)")
    )
  print(p1)

  # --- Plot 2: color by host_order -------------------------------------------
  p2 <- ggplot(pcoa_df, aes(PC1, PC2, color = host_order)) +
    geom_point(size = 2) +
    scale_color_manual(values = host_order_colors, name = "Host order") +
    theme_minimal() +
    labs(
      title = paste("PCoA of", nm, "presence/absence matrix across MAGs (colored by host order) -", sp),
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)")
    )
  print(p2)
}



# Bind lists into final tables
permanova_all  <- if (length(permanova_rows))  dplyr::bind_rows(permanova_rows)  else tibble::tibble()
betadisper_all <- if (length(betadisper_rows)) dplyr::bind_rows(betadisper_rows) else tibble::tibble()

# Quick peek
print(permanova_all, n = 20)
print(betadisper_all, n = 20)

# Or a compact glimpse
# dplyr::glimpse(permanova_all)
# dplyr::glimpse(betadisper_all)


}
  


```
```{r allpermanova_courntry}
permaniva_sig_country <- permanova_all %>%filter(term == "country_normalized") %>% dplyr::select(-Df, -SumOfSqs) %>% filter(`Pr(>F)` < 0.05)

permaniva_sig_country
```
```{r plotting permanova results_country}
ggplot(permaniva_sig_country, aes(x = matrix_name,
             y = species,
             color = `Pr(>F)`,
             size = R2)) +
  geom_point() +
  scale_color_viridis_c(direction = 1, option = "plasma") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Association strength by species and matrix - Country effects",
    x = "Matrix",
    y = "Species",
    size = "R²",
    color = "Pr(>F)"
  )

```

```{r allpermanova_host_type_3}
permaniva_sig <- permanova_all %>%filter(term == "host_type_3") %>% dplyr::select(-Df, -SumOfSqs)

permaniva_sig
```



#### Phylogenetic tree of the 10 species
```{r tree_10}
tree <- read.tree("data/mags_metadata/gtotree_output.tre")
map <- read.table("data/mags_metadata/mapping.txt", header = TRUE, sep = "\t")

# tip labels
tree$tip.label <- map$name[ match(tree$tip.label, map$id) ]


tree_order <- tree$tip.label
permaniva_sig <- permaniva_sig %>%
  mutate(species = factor(species, levels = tree_order))
```


```{r plotting permanova results, fig.height = 7, fig.width = 7}

permaniva_sig %>%
  dplyr::mutate(p_sig = ifelse(`Pr(>F)` <= 0.05, `Pr(>F)`, NA_real_)) %>%
  ggplot(aes(x = matrix_name,
             y = species,
             color = p_sig,
             size  = R2)) +
  geom_point() +
  scale_color_viridis_c(
    direction = 1,
    option    = "plasma",
    na.value  = "grey70"   # non‑significant points become grey
  ) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Association strength by species and matrix — Host type effects",
    x = "Matrix",
    y = "Species",
    size  = "R²",
    color = "Pr(>F) (≤ 0.05 shown)"
  )


```

```{r figure}
p_tree <- ggtree(tree) +
  theme_tree2() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p_dot <- permaniva_sig %>%
  dplyr::mutate(p_sig = ifelse(`Pr(>F)` <= 0.05, `Pr(>F)`, NA_real_)) %>%
  ggplot(aes(
    x = matrix_name,
    y = species,
    color = p_sig,
    size  = R2
  )) +
  geom_point() +
  scale_color_viridis_c(
    direction = 1,
    option    = "plasma",
    na.value  = "grey70"
  ) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Association strength by species and matrix — Host type effects",
    x = NULL,       # ← remove X‑axis label
    y = "Species",
    size  = "R²",
    color = "Pr(>F) (≤ 0.05 shown)"
  ) +
  theme(
    axis.title.y = element_blank(),     # ensure blank title
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.margin  = margin(l = 0)        # pull plot closer to the tree
  )

p_final <- p_tree | p_dot
p_final
```


### GIFTs
GIFTs are being put on hold because they seem to be biased by pathways where one gene is present/absent (mainly antibiotic degradation).
```{r GIFTs_ginak, eval = FALSE}

for (sp in species_list) {
  
  genome_annotations <- species_data[[sp]]$annotations
  genome_metadata <- species_data[[sp]]$metadata
  
  genome_gifts <- distill(genome_annotations,GIFT_db,genomecol= "genome", 
                          annotcol=c("kegg", "ec", "pfam", "cazy"), verbosity = F)
  
  plots <- gift_pcoa_plots(genome_gifts, genome_metadata)

  print(plots$arrows_plot)
  print(plots$big_plot)
  print(plots$big_order_plot)

  
  #save the vectors to create functional clusters
  pcoa_gift_vectors <- plots$gift_pcoa_vectors
  
  coords <- pcoa_gift_vectors %>%
    rownames_to_column("MAG") %>%
    as_tibble()
  
  set.seed(123)
  km <- kmeans(coords[, c("Axis.1", "Axis.2")], centers = 2, nstart = 25, iter.max = 100)
  
  coords <- coords %>% mutate(cluster = factor(km$cluster))
  
  centroids <- as_tibble(km$centers) %>% mutate(cluster = factor(1:nrow(km$centers)))
  
  print(ggplot(coords, aes(x = Axis.1, y = Axis.2, color = cluster)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_point(data = centroids, aes(x = Axis.1, y = Axis.2, color = cluster),
               shape = 4, size = 5, stroke = 1.25) +   # X marks centroids
    theme_minimal() +
    labs(title = paste0("PCoA colored by kmeans (k=2)"),
         color = "cluster"))
  
  metadata_with_cluster <- genome_metadata %>%
    left_join(coords %>% dplyr::select(MAG, cluster), by = c("ID_clean" = "MAG"))
  
  #Is genome size different between clusters?
  print(kruskal.test(genome_size ~ cluster, data = metadata_with_cluster))
  
  
  print(pairwise.wilcox.test(
    x = metadata_with_cluster$genome_size,
    g = metadata_with_cluster$cluster,
    p.adjust.method = "BH"   # FDR correction
  ))

  
    #Aggregate bundle-level GIFTs into the compound level
    GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
    
    #Aggregate element-level GIFTs into the function level
    GIFTs_functions <- to.functions(GIFTs_elements,GIFT_db)
    
    #Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
    GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

    
    #community-level plots
    
    library(reshape2)
    gift_plots <- plot_gifts(GIFTs_elements, metadata_with_cluster)
    
    #print(gift_plots$cluster)
    #print(gift_plots$host_type_3)
    #print(gift_plots$antibiotic)
    
    
    # Save cluster plot
    ggsave(
      filename = file.path("plots",
                           paste0(sp, "_gift_plot_cluster.png")),
      plot = gift_plots$cluster,
      width = 14,
      height = 16,
      units = "in",
      dpi = 300
    )
    
    # Save host type plot
    ggsave(
      filename = file.path("plots",
                           paste0(sp, "_gift_plot_host_type_3.png")),
      plot = gift_plots$host_type_3,
      width = 14,
      height = 16,
      units = "in",
      dpi = 300
    )


    
    gift_stats_res <- gift_stats(GIFTs_elements,
                  metadata_with_cluster,
                  GIFT_db,
                  host_type_3_colors,
                  host_class_colors)


    # PLot statistical Differences in -
    #cluster
    #print(gift_stats_res$kruskal_table)
    #print(gift_stats_res$pairwise_sig_table)
    #print(gift_stats_res$cluster_diff_plot)
    
    #host type
    print(gift_stats_res$wilcox_res_host_type_3)
    print(gift_stats_res$host_type_3_diff_plot)

}


```












# Grouped analysis
```{r total_genome_annotations}

lactococcus_annotations    <- species_data$lactococcus_lactis$annotations
hafnia_annotations         <- species_data$hafnia_paralvei$annotations
enterococcus_annotations   <- species_data$enterococcus_hirae$annotations
citrobacter_annotations    <- species_data$citrobacter_braakii$annotations
parabacteroides_annotations<- species_data$parabacteroides_goldsteinii$annotations


ann_list <- list(
  lactococcus     = lactococcus_annotations,
  hafnia          = hafnia_annotations,
  enterococcus    = enterococcus_annotations,
  citrobacter     = citrobacter_annotations,
  parabacteroides = parabacteroides_annotations
)

# Coerce start/end to numeric wherever they exist and are character
ann_list <- purrr::map(ann_list, ~ .x %>%
  mutate(
    across(any_of(c("start", "end")),
           ~ if (is.character(.x)) readr::parse_double(.x, na = c("", "NA")) else .x)
  )
)

total_genome_annotations <- bind_rows(ann_list, .id = "species")

```

```{r kegg count analysis}
kegg_per_genome <- total_genome_annotations %>%
  filter(!is.na(kegg)) %>%
  group_by(genome) %>%
  summarise(kegg_count = n(), .groups = "drop")

kegg_summary <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_3_2) %>%
  summarise(
    mean_kegg = mean(kegg_count, na.rm = TRUE),
    median_kegg = median(kegg_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

kegg_test <- kegg_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

ggplot(kegg_test, aes(x = host_type_3_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

ggplot(kegg_test, aes(x = host_type_3_2, y = kegg_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = bacterial_species), width = 0.2, alpha = 0.6) 

kruskal.test(kegg_count ~ host_type_3_2, data = kegg_test)

pairwise.wilcox.test(
  kegg_test$kegg_count,
  kegg_test$host_type_3_2,
  p.adjust.method = "BH"
)

```
```{r amr count analysis}
amr_per_genome <- total_genome_annotations %>%
  filter(resistance_type == "AMR",
    !is.na(resistance_target)) %>%
  group_by(genome) %>%
  summarise(amr_count = n(), .groups = "drop")

amr_summary <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_3_2) %>%
  summarise(
    mean_amr = mean(amr_count, na.rm = TRUE),
    median_amr = median(amr_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

amr_test <- amr_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(amr_test, aes(x = host_type_3_2, y = amr_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()

kruskal.test(amr_count ~ host_type_3_2, data = amr_test)

pairwise.wilcox.test(
  amr_test$amr_count,
  amr_test$host_type_3_2,
  p.adjust.method = "BH"
)

```

```{r vf_count_analysis}

vf_per_genome <- total_genome_annotations %>%
  filter(!is.na(vf)) %>%
  group_by(genome) %>%
  summarise(vf_count = n(), .groups = "drop")

vf_summary <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) %>%
  group_by(host_type_3_2) %>%
  summarise(
    mean_kegg = mean(vf_count, na.rm = TRUE),
    median_kegg = median(vf_count, na.rm = TRUE),
    n_genomes = n(),
    .groups = "drop"
  )

vf_test <- vf_per_genome %>%
  left_join(total_genome_metadata, by = c("genome" = "ID")) 

ggplot(vf_test, aes(x = host_type_3_2, y = vf_count)) +
  geom_boxplot() +
  geom_jitter(aes(color = completeness), width = 0.2, alpha = 0.6) +
  scale_color_viridis_c()


kruskal.test(vf_count ~ host_type_3_2, data = vf_test)

pairwise.wilcox.test(
  vf_test$vf_count,
  vf_test$host_type_3_2,
  p.adjust.method = "BH"
)
```


```{r amr different per group}
amr_presence <- total_genome_annotations %>%
  filter(resistance_type == "AMR", !is.na(resistance_target)) %>%
  distinct(genome, resistance_target) %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

amr_sets <- amr_presence %>%
  group_by(host_type_3_2) %>%
  summarise(amr_list = list(unique(resistance_target)))

amr_freq <- amr_presence %>%
  group_by(host_type_3_2, resistance_target) %>%
  summarise(n = n(), .groups = "drop")

chisq.test(table(amr_presence$resistance_target, amr_presence$host_type_3_2))

human_adj_amr     <- amr_sets$amr_list[[which(amr_sets$host_type_3_2 == "human_adj")]]
domestic_amr  <- amr_sets$amr_list[[which(amr_sets$host_type_3_2 == "domestic")]]
wild_amr      <- amr_sets$amr_list[[which(amr_sets$host_type_3_2 == "wild")]]

Reduce(intersect, list(human_adj_amr, domestic_amr, wild_amr))

human_adj_only <- setdiff(human_adj_amr, union(domestic_amr, wild_amr))
domestic_only <- setdiff(domestic_amr, union(human_adj_amr, wild_amr))
wild_only <- setdiff(wild_amr, union(human_adj_amr, domestic_amr))

human_adj_only
domestic_only
wild_only

amr_binary <- amr_presence %>%
  distinct(host_type_3_2, resistance_target) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = host_type_3_2,
    values_from = present,
    values_fill = 0
  )

UpSetR::upset(
  as.data.frame(amr_binary[, -1]),  # remove resistance_target column
  sets = c("human_adj", "domestic", "wild"),
  order.by = "freq"
)



```

```{r vf different per group}
vf_presence <- total_genome_annotations %>%
  filter( !is.na(vf)) %>%
  distinct(genome, vf) %>%
  left_join(total_genome_metadata, by = c("genome" = "ID"))

vf_annotated <- vf_presence %>%
  left_join(vf_data, by = "vf")

vf_sets <- vf_presence %>%
  group_by(host_type_3_2) %>%
  summarise(vf_list = list(unique(vf)))

vf_freq <- vf_presence %>%
  group_by(host_type_3_2, vf) %>%
  summarise(n = n(), .groups = "drop")

chisq.test(table(vf_presence$vf, vf_presence$host_type_3_2))

human_adj_vf     <- vf_sets$vf_list[[which(vf_sets$host_type_3_2 == "human_adj")]]
domestic_vf  <- vf_sets$vf_list[[which(vf_sets$host_type_3_2 == "domestic")]]
wild_vf      <- vf_sets$vf_list[[which(vf_sets$host_type_3_2 == "wild")]]

Reduce(intersect, list(human_adj_vf, domestic_vf, wild_vf))

human_adj_only <- setdiff(human_adj_vf, union(domestic_vf, wild_vf))
domestic_only <- setdiff(domestic_vf, union(human_adj_vf, wild_vf))
wild_only <- setdiff(wild_vf, union(human_adj_vf, domestic_vf))


vf_wild_only_annot <- 
  tibble(vf = wild_only) %>% 
  left_join(vf_data, by = "vf")


vf_binary <- vf_presence %>%
  distinct(host_type_3_2, vf) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = host_type_3_2,
    values_from = present,
    values_fill = 0
  )

UpSetR::upset(
  as.data.frame(vf_binary[, -1]),  
  sets = c("human_adj", "domestic", "wild"),
  order.by = "freq"
)

vf_type_dist <- vf_annotated %>%
  group_by(host_type_3_2, vf_type) %>%
  summarise(n = n_distinct(vf), .groups = "drop") %>%
  group_by(host_type_3_2) %>%
  mutate(prop = n / sum(n))

ggplot(vf_type_dist, aes(x = host_type_3_2, y = prop, fill = vf_type)) +
  geom_bar(stat = "identity") +
  labs(y = "Proportion of VF types")

vf_domestic_only <- vf_data %>%
  filter(vf %in% domestic_only) %>%
  count(vf_type) %>%
  mutate(prop = n / sum(n))

ggplot(vf_domestic_only, aes(x = reorder(vf_type, -prop), y = prop, fill = vf_type)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "VF types unique to domestic hosts",
       y = "Proportion")



vf_unique_df <- dplyr::bind_rows(
  vf_data %>% filter(vf %in% domestic_only) %>% mutate(group = "domestic"),
  vf_data %>% filter(vf %in% wild_only) %>% mutate(group = "wild"),
  vf_data %>% filter(vf %in% human_adj_only) %>% mutate(group = "human_adj")
) %>%
  count(group, vf_type) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n))

ggplot(vf_unique_df, aes(x = group, y = prop, fill = vf_type)) +
  geom_bar(stat = "identity") +
  labs(y = "Proportion of VF types (unique only)")
```


```{r functional analysis all together}
# Build matrices
kegg_mat <- make_annotation_matrix(total_genome_annotations, "kegg")
amr_mat  <- make_annotation_matrix(total_genome_annotations, "resistance_target")
vf_mat <- make_annotation_matrix(total_genome_annotations, "vf")
cazy_mat <- make_annotation_matrix(total_genome_annotations, "cazy")
defense_mat <- make_annotation_matrix(total_genome_annotations, "defense")
antidefense_mat <- make_annotation_matrix(total_genome_annotations, "antidefense")

# Fisher + volcano
kegg_fisher <- run_fisher_tests(kegg_mat, total_genome_metadata, "kegg")
amr_fisher  <- run_fisher_tests(amr_mat, total_genome_metadata, "amr")
vf_fisher  <- run_fisher_tests(vf_mat, total_genome_metadata, "vf")
cazy_fisher  <- run_fisher_tests(cazy_mat, total_genome_metadata, "cazy")
defense_fisher  <- run_fisher_tests(defense_mat, total_genome_metadata, "defense")
antidefense_fisher  <- run_fisher_tests(antidefense_mat, total_genome_metadata, "antidefense")

plot_volcano(kegg_fisher, "kegg")
plot_volcano(amr_fisher, "amr")
plot_volcano(vf_fisher, "vf")
plot_volcano(cazy_fisher, "cazy")
plot_volcano(defense_fisher, "defense")
plot_volcano(antidefense_fisher, "antidefense")


# PERMANOVA
kegg <- pa_permanova(kegg_mat, total_genome_metadata)
kegg$betadisper
kegg$permanova

amr <- pa_permanova(amr_mat, total_genome_metadata)
amr$betadisper
amr$permanova

vf <- pa_permanova(vf_mat, total_genome_metadata)
vf$betadisper
vf$permanova

cazy <- pa_permanova(cazy_mat, total_genome_metadata)
cazy$betadisper
cazy$permanova

defense <- pa_permanova(defense_mat, total_genome_metadata)
defense$betadisper
defense$permanova

antidefense <- pa_permanova(antidefense_mat, total_genome_metadata)
antidefense$betadisper
antidefense$permanova


# PCOAs

#### ---------------------
#### KEGG
#### ---------------------
pcoa_res <- ape::pcoa(kegg$dist, correction = "cailliez")

ids_pa   <- rownames(kegg$pa_nz)
ids_pcoa <- rownames(pcoa_res$vectors)
ids_meta <- rownames(total_genome_metadata)

# common_ids <- Reduce(intersect, list(ids_pa, ids_pcoa, ids_meta))
# if (length(common_ids) == 0L) stop("[KEGG] No common IDs across pa_nz, PCoA vectors, and metadata.")
# 
# .report_drops("KEGG", ids_pa, ids_pcoa, ids_meta, common_ids)
# 
# # enforce order by PCoA vectors
# common_ids <- intersect(rownames(pcoa_res$vectors), common_ids)
# 
# pcoa_vec     <- pcoa_res$vectors[common_ids, , drop = FALSE]
# meta_aligned <- total_genome_metadata[common_ids, , drop = FALSE]
# 
# pcoa_df <- data.frame(
#   ID        = common_ids,
#   PC1       = pcoa_vec[, 1],
#   PC2       = pcoa_vec[, 2],
#   host_type_3 = meta_aligned$host_type_3,
#   host_order= meta_aligned$host_order,
#   row.names = NULL, check.names = FALSE
# )
# 
# var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
# 
# pcoa_kegg_pa <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_type_3_colors, name = "Host type") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of KEGG presence/absence across MAGs (colored by host type)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# pcoa_kegg_pa
# 
# ggplot(pcoa_df, aes(PC1, PC2, color = host_order)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_order_colors, name = "Host order") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of KEGG presence/absence across MAGs (colored by host order)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# 
# 
# #### ---------------------
# #### AMR
# #### ---------------------
# pcoa_res <- ape::pcoa(amr$dist, correction = "cailliez")
# 
# ids_pa   <- rownames(amr$pa_nz)
# ids_pcoa <- rownames(pcoa_res$vectors)
# ids_meta <- rownames(total_genome_metadata)
# 
# common_ids <- Reduce(intersect, list(ids_pa, ids_pcoa, ids_meta))
# if (length(common_ids) == 0L) stop("[AMR] No common IDs across pa_nz, PCoA vectors, and metadata.")
# 
# .report_drops("AMR", ids_pa, ids_pcoa, ids_meta, common_ids)
# 
# common_ids <- intersect(rownames(pcoa_res$vectors), common_ids)
# 
# pcoa_vec     <- pcoa_res$vectors[common_ids, , drop = FALSE]
# meta_aligned <- total_genome_metadata[common_ids, , drop = FALSE]
# 
# pcoa_df <- data.frame(
#   ID        = common_ids,
#   PC1       = pcoa_vec[, 1],
#   PC2       = pcoa_vec[, 2],
#   host_type_3 = meta_aligned$host_type_3,
#   host_order= meta_aligned$host_order,
#   row.names = NULL, check.names = FALSE
# )
# 
# var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
# 
# pcoa_amr_pa <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_type_3_colors, name = "Host type") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of AMR presence/absence across MAGs (colored by host type)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# pcoa_amr_pa
# 
# ggplot(pcoa_df, aes(PC1, PC2, color = host_order)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_order_colors, name = "Host order") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of AMR presence/absence across MAGs (colored by host order)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# 
# 
# #### ---------------------
# #### VF
# #### ---------------------
# pcoa_res <- ape::pcoa(vf$dist, correction = "cailliez")
# 
# ids_pa   <- rownames(vf$pa_nz)
# ids_pcoa <- rownames(pcoa_res$vectors)
# ids_meta <- rownames(total_genome_metadata)
# 
# common_ids <- Reduce(intersect, list(ids_pa, ids_pcoa, ids_meta))
# if (length(common_ids) == 0L) stop("[VF] No common IDs across pa_nz, PCoA vectors, and metadata.")
# 
# .report_drops("VF", ids_pa, ids_pcoa, ids_meta, common_ids)
# 
# common_ids <- intersect(rownames(pcoa_res$vectors), common_ids)
# 
# pcoa_vec     <- pcoa_res$vectors[common_ids, , drop = FALSE]
# meta_aligned <- total_genome_metadata[common_ids, , drop = FALSE]
# 
# pcoa_df <- data.frame(
#   ID        = common_ids,
#   PC1       = pcoa_vec[, 1],
#   PC2       = pcoa_vec[, 2],
#   host_type_3 = meta_aligned$host_type_3,
#   host_order= meta_aligned$host_order,
#   row.names = NULL, check.names = FALSE
# )
# 
# var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
# 
# pcoa_vf_pa <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_type_3_colors, name = "Host type") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of VF presence/absence across MAGs (colored by host type)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# pcoa_vf_pa
# 
# ggplot(pcoa_df, aes(PC1, PC2, color = host_order)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_order_colors, name = "Host order") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of VF presence/absence across MAGs (colored by host order)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# 
# 
# #### ---------------------
# #### CAZY
# #### ---------------------
# pcoa_res <- ape::pcoa(cazy$dist, correction = "cailliez")
# 
# ids_pa   <- rownames(cazy$pa_nz)
# ids_pcoa <- rownames(pcoa_res$vectors)
# ids_meta <- rownames(total_genome_metadata)
# 
# common_ids <- Reduce(intersect, list(ids_pa, ids_pcoa, ids_meta))
# if (length(common_ids) == 0L) stop("[CAZY] No common IDs across pa_nz, PCoA vectors, and metadata.")
# 
# .report_drops("CAZY", ids_pa, ids_pcoa, ids_meta, common_ids)
# 
# common_ids <- intersect(rownames(pcoa_res$vectors), common_ids)
# 
# pcoa_vec     <- pcoa_res$vectors[common_ids, , drop = FALSE]
# meta_aligned <- total_genome_metadata[common_ids, , drop = FALSE]
# 
# pcoa_df <- data.frame(
#   ID        = common_ids,
#   PC1       = pcoa_vec[, 1],
#   PC2       = pcoa_vec[, 2],
#   host_type_3 = meta_aligned$host_type_3,
#   host_order= meta_aligned$host_order,
#   row.names = NULL, check.names = FALSE
# )
# 
# var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
# 
# pcoa_cazy_pa <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_type_3_colors, name = "Host type") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of CAZy presence/absence across MAGs (colored by host type)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# pcoa_cazy_pa
# 
# ggplot(pcoa_df, aes(PC1, PC2, color = host_order)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_order_colors, name = "Host order") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of CAZy presence/absence across MAGs (colored by host order)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )
# 
# 
# #### ---------------------
# #### Defense
# #### ---------------------
# pcoa_res <- ape::pcoa(defense$dist, correction = "cailliez")
# 
# ids_pa   <- rownames(defense$pa_nz)
# ids_pcoa <- rownames(pcoa_res$vectors)
# ids_meta <- rownames(total_genome_metadata)
# 
# common_ids <- Reduce(intersect, list(ids_pa, ids_pcoa, ids_meta))
# if (length(common_ids) == 0L) stop("[Defense] No common IDs across pa_nz, PCoA vectors, and metadata.")
# 
# .report_drops("Defense", ids_pa, ids_pcoa, ids_meta, common_ids)
# 
# common_ids <- intersect(rownames(pcoa_res$vectors), common_ids)
# 
# pcoa_vec     <- pcoa_res$vectors[common_ids, , drop = FALSE]
# meta_aligned <- total_genome_metadata[common_ids, , drop = FALSE]
# 
# pcoa_df <- data.frame(
#   ID        = common_ids,
#   PC1       = pcoa_vec[, 1],
#   PC2       = pcoa_vec[, 2],
#   host_type_3 = meta_aligned$host_type_3,
#   host_order= meta_aligned$host_order,
#   row.names = NULL, check.names = FALSE
# )
# 
# var_exp <- round(100 * pcoa_res$values$Rel_corr_eig[1:2], 1)
# 
# pcoa_defense_pa <- ggplot(pcoa_df, aes(PC1, PC2, color = host_type_3)) +
#   geom_point(size = 2) +
#   scale_color_manual(values = host_type_3_colors, name = "Host type") +
#   theme_minimal() +
#   labs(
#     title = "PCoA of Defense presence/absence across MAGs (colored by host type)",
#     x = paste0("PC1 (", var_exp[1], "%)"),
#     y = paste0("PC2 (", var_exp[2], "%)")
#   )

```




### Phylogenetic tree of the species
Filter
```{r phylogenetic tree of the species}

top10_by_species <- total_genome_metadata %>%
  dplyr::filter(completeness > 99, contamination < 1) %>%
  dplyr::group_by(bacterial_species) %>%
  dplyr::arrange(dplyr::desc(completeness), contamination, .by_group = TRUE) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::ungroup() 

#write(top10_by_species, "top10_MAG_list.txt")
```




